### 2018-03-05 abstract.tex
compiles

A critically important, but surprisingly neglected in the software testing literature, aspect of system reliability is system predictability.  Many software systems are implemented using mechanisms (unsafe languages, mulitple threads, complex caching, stochastic algorithms) that can introduce behavioral non-determinism.  Users of software systems, including especially other software systems using library calls in a single-threaded context, however, expect that systems will behave deterministically from the perspective of having predictable results from the same sequential series of operations.  In this paper we show that lightweight techniques, requiring (almost) no effort on the part of a developer, can extend an existing test generation system to allow checking of determinism properties on tested code.  We introduce two notions of determinism to check (visible value trace determinism and failure determinism), inspired by real faults, and tailor these notions to the practical, automatic, checking of Python library code.

### 2018-03-05 implementation.tex
compiles

\section{Implementation}

We implemented our approach in the TSTL \cite{NFM15} system.

### 2018-03-05 relatedwork.tex
small additions

\section{Related Work}

Shi et. al \cite{DetermImp} examined what might be seen as a related, though in a sense the opposite, problem:  they detect code that assumes a deterministic implementation of a non-deterministic specification.  E.g., they detect instances when  iteration through items of a set is explicitly not guaranteed to provide any particular order of items, but code depends on the order produced by a given implementation.

### 2018-03-05 abstract.tex
better abstract

@@
 A critically important, but surprisingly neglected in the software testing literature, aspect of system reliability is system predictability.
-Many software systems are implemented using mechanisms (unsafe languages, mulitple threads, complex caching, stochastic algorithms) that can introduce behavioral non-determinism.
-Users of software systems, including especially other software systems using library calls in a single-threaded context, however, expect that systems will behave deterministically from the perspective of having predictable results from the same sequential series of operations.
-In this paper we show that lightweight techniques, requiring (almost) no effort on the part of a developer, can extend an existing test generation system to allow checking of determinism properties on tested code.
-We introduce two notions of determinism to check (visible value trace determinism and failure determinism), inspired by real faults, and tailor these notions to the practical, automatic, checking of Python library code.
+Many software systems are implemented using mechanisms (unsafe languages, mulitple threads, complex caching, stochastic algorithms, environmental dependencies) that can introduce behavioral nondeterminism.
+Users of software systems, especially other software systems using library calls in a single-threaded context, often expect that systems will behave deterministically in the sense that they have predictable results from the same sequential series of operations.
+Equally importantly, even when it does not impact correctness or user experience, nondeterminism must be understood and managed for effective automated testing.
+Nondeterministic behavior that is not either controlled or accounted for can result in flaky tests as well as causing problems for test reduction, differential testing, and automated regression test generation.
+We show that lightweight techniques, requiring (almost) no effort on the part of a developer, can be used to extend an existing test generation system to allow checking for problematic nondeterminism.
+We introduce three specific lightweight nondeterinism properties (visible value trace and final-value nondeterminism and failure nondeterminism), inspired by real faults, and tailor these notions to the practical, automatic, checking of Python library code.

### 2018-03-05 experiments.tex
better abstract

\section{Experimental Evaluation}

\subsection {Case Study: Redis-Py}

\subsection {Case Study: pyfakefs}

### 2018-03-06 relatedwork.tex
more related work

@@
 E.g., they detect instances when  iteration through items of a set is explicitly not guaranteed to provide any particular order of items, but code depends on the order produced by a given implementation.
+
+Other efforts \cite{ParallelDeterministic} have aimed to avoid nondeterminism in parallel implementations, by design, indicating the importance of avoiding nondeterministic behavior, even at the expense of adopting novel programming models, where the goal is not simply (as in, say, Rust) to avoid classic concurrency errors, but to enforce genuinely deterministic behavior.

### 2018-03-24 introduction.tex
more text

\section{Introduction}

For ourselves, we might prefer to think (and act as if) we had free
will; however, we generally prefer our software systems to be as
constrained in their actions as possible:  in other words, we wish
them to be largely deterministic, from our perspective.

WHY DETERMINISM:  Bugs, flaky tests, pain debugging.

\subsection{What is Determinism?}

A system is deterministic if, given its complete state at a point in
time, it is possible, in principle, to predict its future behavior
perfectly.  We say ``in principle'' because in the real world,
prediction may be possible but hopelessly impractical.  We write
complex software systems in many cases because we cannot predict their
behavior (if we could perform the calculations in advance, ourselves,
we would just do so).  As a consequence, in software, rather than
defining determinism in terms of prediction, we usually therefore
simply say that a system is deterministic if, given the same state and
inputs, it always produces the same outputs\footnote{Of course, this
  shift of definition is often adopted in arguments for physical or
  human behavioral determinism, due to the even greater difficulties
  of prediction in those arenas.}.  The change of
definition, we will see, is central to our approach, and comes with a
number of limitations.

Technically, many ``nondeterministic'' systems are not
nondeterministic in a strong sense, at all.  Given the complete state
of the system (which includes the entire state of the underlying
hardware, the operating system, storage devices, etc.), ignoring
quantum effects, and treating outside interventions such as network
traffic, human activity at an input device, etc. properly as inputs,
most software \emph{is} completely predictable.  What we actually
mean, usually, is that, given a certain limited abstraction of state
and of inputs, observable behavior is repeatable.  This abstraction,
for most software, is not expected to include many elements outside of
the software system itself.

Unexpected nondeterminism usually arises when there is an element of
state or input that \emph{is} critical to the produced behavior, but
the programmer has not anticipated.  E.g., it is beleived that the
behavior of a thread scheduler will not matter, but a race condition
in the code means that it does matter, after all.

\subsection{The High Cost of Unexpected Nondeterminism}

Unexpected nondeterminism is, unfortunately, usually only discovered
in a context that makes it very hard to debug.   The most common such
contexts are occasional rare failures of a system in deployment, and
flaky tests.

### 2018-03-24 abstract.tex
abstract fix

@@
 Users of software systems, especially other software systems using library calls in a single-threaded context, often expect that systems will behave deterministically in the sense that they have predictable results from the same sequential series of operations.
-Equally importantly, even when it does not impact correctness or user experience, nondeterminism must be understood and managed for effective automated testing.
+Equally importantly, even when it does not impact correctness or user experience, nondeterminism must be understood and managed for effective (especially automated) testing.
 Nondeterministic behavior that is not either controlled or accounted for can result in flaky tests as well as causing problems for test reduction, differential testing, and automated regression test generation.
 We show that lightweight techniques, requiring (almost) no effort on the part of a developer, can be used to extend an existing test generation system to allow checking for problematic nondeterminism.
-We introduce three specific lightweight nondeterinism properties (visible value trace and final-value nondeterminism and failure nondeterminism), inspired by real faults, and tailor these notions to the practical, automatic, checking of Python library code.
+We introduce a set of lightweight nondeterinism properties, inspired by real faults, and tailor these notions to the practical, automatic, checking of Python library code.

### 2018-03-24 concepts.tex
reorg

\section{Practical Lightweight Nondeterminism
  Detection and Debugging}

\subsection{Horizontal Determinism}

\subsubsection{Visible Value Determinism}

\subsubsection{Final State Determinism}

\subsubsection{Process-Level Determinism}

\subsection{Vertical Determinism}

Apple bugs.

\subsubsection{Failure Nondeterminism}

\subsection{Nondeterminism and Delta-Debugging}

### 2018-03-26 experiments.tex
3.3 hashing
edits
code example

@@
 
-\subsection {Case Study: pyfakefs}
+\subsection {Case Study: pyfakefs and OS X}
+
+\subsection{Case Study: Datarray Inference Algorithms}
+
+\url{https://github.com/BIDS/datarray}
+
+\begin{figure*}
+{\scriptsize
+\begin{code}
+@import inference\_algs
+@import datarray
+\vspace{0.1in}
+<@
+def flatten\_and\_sort(v):
+    return (sorted(map(flatten\_and\_sort,v),key=repr) if type(v) in [list,tuple] else
+                (flatten\_and\_sort(list(v.items())) if type(v) == dict else v))
+
+def psplit(P):
+    return ([P,1.0-P])
+@>
+\vspace{0.1in}
+pool: <P> 3
+pool: <cpts> 3
+pool: <evidence> 3 OPAQUE
+pool: <ename> 3
+pool: <event> 3
+\vspace{0.1in}
+<P> := 0.01 * <[0..100]>
+\vspace{0.1in}
+<ename> := "E" + str(<[1..5]>)
+\vspace{0.1in}
+\{Exception\} <event> := [datarray.DataArray(psplit(<P>), axes = [<ename>])]
+\{Exception\} <ename,1>!=<ename,2> -> <event> := [datarray.DataArray([[psplit(<P>)],psplit(<P>)],[<ename>,<ename>])]
+\vspace{0.1in}
+<cpts> := []
+~<cpts>.append(<event>[0])
+\vspace{0.1in}
+<evidence> := \{\}
+~<evidence>.update([(<ename>,0)])
+\vspace{0.1in}
+\{Exception\} print(flatten\_and\_sort(inference\_algs.calc\_marginals\_simple(<cpts>,<evidence>)))
+\{Exception\} print(flatten\_and\_sort(inference\_algs.calc\_marginals\_sumproduct(<cpts>,<evidence>)))
+\{Exception\} print(flatten\_and\_sort(inference\_algs.calc\_marginals\_jtree(<cpts>,<evidence>)))
+\end{code}
+}
+\caption {Complete TSTL harness for finding the hash-order bug in the datarray
+  inference algorithms.}
+\label{hashbug}
+\end{figure*}

### 2018-03-31 concepts.tex
determinism

@@
 \subsection{Nondeterminism and Delta-Debugging}
+
+In \emph{monotonic} cases, removing a part of a test cannot
+increase the probability of a predicate holding.  This is fairly
+common.  In these cases, when use a reduction algorithm to reduce $t$
+to $r$, $P(p(r) \leq P(p(t))$.  In \emph{non-monotonic} cases,
+however, there is no such upper bound.  Removing a step in $R$
+
+The probabilistic predicate of interest for determinism is always of
+the form:  {\bf the test will exhibit at least two different behaviors in $S$
+runs, with probability $P$}.  This predicate may be monotonic or
+non-monotonic, depending on the causes of the nondeterminism detected.

### 2018-04-01 concepts.tex
more on ddmin

@@
 
+Two purposes:  one is simply the usual purpose of test reduction: shrinking the test to make debugging
+easier; however, the other purpose is \emph{changing the probability of failure.}
+
 In \emph{monotonic} cases, removing a part of a test cannot
 increase the probability of a predicate holding.  This is fairly
-common.  In these cases, when use a reduction algorithm to reduce $t$
+common.  In these cases, when we use a reduction algorithm to reduce $t$
 to $r$, $P(p(r) \leq P(p(t))$.  In \emph{non-monotonic} cases,
-however, there is no such upper bound.  Removing a step in $R$
+however, there is no such upper bound.  Removing a step in $t$ may
+increase the probability that $t$ behaves nondeterministically.
 
@@
 non-monotonic, depending on the causes of the nondeterminism detected.
+
+The key changes required to standard delta-debugging implementations
+are simply removal of some ``sanity checks'' in the code:  first, many
+implementations (including the Python code provided by Zeller) assert
+that the predicate holds on the original test.  In probabilistic
+settings, this may not be true, and we may even be trying to find a
+subset where a predicate that is very far from holding on the original
+holds (a non-monotonic case where we aim to increase the probability
+of nondeterminism).

### 2018-04-04 concepts.tex
bib

--- 2018-03-31 concepts.tex
@@
 \subsection{Horizontal Determinism}
+
+\subsubsection{Determinism and Reflexive Differential Testing}
+
+Horizontal determinism can be best understood by thinking of
+nondeterminism detection as an unusual kind of \emph{differential
+  testing} \cite{Differential,ICSEDiff}.  In differential testing, a
+system is compared against a reference in order to ensure that
+it behaves equivalently, at some level of abstraction, to another
+implementation of the same functionality.  Differential testing is
+extremely powerful, in that any divergence of behavior, if divergence
+is correctly defined, indicates a functional correctness fault in (at
+least) one of the systems under test.  Being able to detect functional
+correctness errors without the cost of constructing a formal
+specification is extremely useful in automated test generation.  Differential testing is widely
+used for systems software components such as compilers
+\cite{Differential,csmith} and POSIX file systems \cite{CFV08,AMAI}, where multiple
+implementations are common.  The major limitation of differential
+testing, of course, is that multiple implementations of a system are
+almost as rare as good correcteness specifications.
+
+For the special case of detecting nondeterminism, however, \emph{a
+  system can serve as its own reference implementation.}
 
@@
 
+Delta-debugging \cite{DD}\footnote{We use delta-debugging to stand for
+  all test reduction algorithms, even those \cite{CReduce,tstl} that do not use the exact
+  binary-search that distinguishes delta-debugging \emph{per se}; the
+  differences are immaterial for our purposes.}  is a widely used method for reducing the
+size of failing tests, making them easier to understand and debug.
+Delta-debugging in the context of detecting nondeterminism has two
+purposes.  One is simply the usual goal of reducing the size of a
+test.  Identifying the cause of nondeterminism may be very easy in a
+test consisting of ten library function calls (it is one of these ten
+calls), but very difficult in a test consisting of a hundred library
+function calls.  This is no different than the common use of
+delta-debugging.  However, in nondeterminism detection,
+delta-debugging can also be used to \emph{change the probability of
+  nondeterministic behavior}.
+
 In \emph{monotonic} cases, removing a part of a test cannot
 increase the probability of a predicate holding.  This is fairly
-common.  In these cases, when use a reduction algorithm to reduce $t$
+common.  In these cases, when we use a reduction algorithm to reduce $t$
 to $r$, $P(p(r) \leq P(p(t))$.  In \emph{non-monotonic} cases,
-however, there is no such upper bound.  Removing a step in $R$
+however, there is no such upper bound.  Removing a step in $t$ may
+increase the probability that $t$ behaves nondeterministically.
 
@@
 non-monotonic, depending on the causes of the nondeterminism detected.
+
+The only absolutely necessary changes required to use standard
+delta-debugging implementations in nondeterminism detection
+are simply removal of some ``sanity checks'' in the code:  many
+implementations (including the Python code provided by Zeller) assert
+that the predicate holds on the original test.  In probabilistic
+settings, this may not be true, and we may even be trying to find a
+subset where a predicate that is very far from holding on the original test
+holds (in a non-monotonic case where we aim to increase the probability
+of nondeterminism).
+
+\subsubsection{``Publication Bias'' in Delta-Debugging}
+
+However, simply using delta-debugging off the shelf with a predicate
+like $P(fail) > 0.3 \wedge P(fail) < 1.0$, to force a test to be flaky, and force it to fail
+sufficiently often to be used in debugging, will often produce
+surprising and unfortunate results.  In many cases,
+delta-debugging will indeed reduce the large test to a small subset.
+And, in a technical sense, delta-debugging will work:  it will never
+convert a nondeterministic test to a completely deterministic test,
+because the reduced test that delta-debugging returns is always one
+where the predicate of interest has been seen to evaluate to true.
+However, if you run the resulting test, it will, in many cases, have a
+$P(fail)$ that is much, much smaller than
+0.3, perhaps as low as 0.01.  Why?
+
+The problem is analogous to the problem of publication bias in
+scientific fields \cite{ahmed2012assessment}.  A predicate like
+$P(fail) > 0.3 \wedge P(fail) < 1.0$ cannot be evaluated by
+determining the true probability of failure; rather, the test must be
+run some concrete number of times, and the number of failures counted.
+However, even if the number of samples $N$ is large, there is some
+probability (based on the sample size) of a result that diverges
+significantly from the actual probability of failure.  If the
+predicate were run only once, and the number of samples reasonably
+large, this would not matter.  However, by their nature,
+delta-debugging and other test reduction algorithms, explore a search
+space that often contains hundreds or even thousands or millions of
+candidate tests.  The predicate is evaluated on each of these, and so
+even with large $N$, it is extremely likely that some evaluation will
+produce a very poor estimate of $P(fail)$.  If such an evaluation
+causes the predicate to appear to hold for a test, delta-debugging is
+``stuck'' with the error, because (to our knowledge) no test reduction
+algorithms allow backtracking.  After such a mistake, finding further
+reductions will become harder, but may still be possible due to the
+same source of errors:  the number of experiments run is far larger
+than the probability of an incorrect result.
+
+It is the combination of
+a one-way bias on faulty evaluations (the consequences of a false
+positive for the predicate are much greater than for a false negative)
+and the huge number of experiments relative to error rate, that
+produces bad results, akin to the magnification of effect sizes
+in science due to publication bias.  The bias in delta-debugging tends
+towards producing a reduced test where probabilities are much smaller
+than demanded by a predicate, due to the pressure on delta-debugging
+to produce shorter tests.  Shorter tests have smaller probabilities,
+on average, due to two factors:  first, timing-induced nondeterminism
+has a much smaller temporal space to operate in, and second, if some
+test operation introduces a small probability of nondeterminism, and
+only many repetitions of that operation make the probability large,
+small tests obviously on average have fewer instances of the
+problematic operation.
+
+In scientific literature, the most frequently proposed solution is the
+use of replications:  repeated runs of ``successful'' experiments to
+minimize the probability that a result is a fluke due to publication
+bias.  One way to produce this effect would be to allow
+delta-debugging to backtrack if the probabilties observed in predicate
+evaluations suddely exhibit a strong discontinuity, a kind of ``paper
+retraction'' based on near-replications.  However, this requires
+modifying the delta-debugging implementations, which is difficult and
+sometimes not really feasible (e.g., few users of AFL-fuzz \cite{aflfuzz} will wish
+to change its C code for test reduction).  Ideally, the solution
+should be implementable simply by modifying the predictate that is
+evaluated.
+
+A costly but effective solution is to make $N$ large in comparison to
+the number of expected predicate evaluations performed during
+delta-debugging.  If the error rate is low enough, the problem
+disappears.  However, given the large number of evaluations performed,
+this will tend to make delta-debugging extremely slow.  We propose
+using a dynamic sampling approach, where $N$ is small, but if the
+predicate evaluates to true, $M$ repeated true evaluations
+(``replications'') are
+required before the predicate returns ``true'' to the delta-debugging
+algorithm.  Tuning $M$, any degree of confidence can be achieved, with
+the basic tradeoff being between finding a test with the desired
+probabilities, and the speed and effectiveness of delta-debugging.
+Because increasing $M$ makes false negatives more likely, larger $M$
+will usually result in less-than-optimal reduction of the original test.

### 2018-04-04 implementation.tex
3.3 hashing
edits
reorg
bib

@@
 We implemented our approach in the TSTL \cite{NFM15} system.
+
+Because TSTL supports differential testing \cite{tstlsttt}, horizontal
+nondeterminism detection can technically be implemented simply by
+declaring a system to be its own reference, using TSTL's notation for
+differential testing.  However, such an approach requires some effort
+on the part of the user to control which values are checked for
+equivalence, and does not easily support injecting timing differences,
+or using a new process, in checking the behavior.

### 2018-04-04 introduction.tex
bib

@@
 
-WHY DETERMINISM:  Bugs, flaky tests, pain debugging.
+Determinism is particularly important for testing and debugging, where
+being able to exactly reproduce system behavior is essential to
+productivity.  If executing ``the same test'' can produce
+significantly different behavior each time it is run, the consequences
+can be unfortunate.  Developers using a test exhibiting nondeterminism
+to debug a system
+face a challenge in reasoning about causality, in that an event
+observed may not even take place the next time the test is run.
+Automated test generation systems using test coverage results to drive
+the search for interesting inputs may be mislead when a run executes
+code only rarely.  And, most significantly, regression testing
+effectiveness can be significantly reduced if tests fail only
+intermittently and unpredictably, as a result of environmental
+factors rather than bugs in changed code.  Such behavior is
+unfortunately all too common:  Gao et
+al. \cite{Gao:2015:MSU:2818754.2818764} observed coverage differences
+of up to 184 lines of code for the same test, and false positive rates
+as high as 96\%.  Nondeterminism is sometimes problematic for
+developers, who will usually want to  assume that library code they
+use is deterministic; it is often disastrous for large-scale or highly
+automated testing.
 
@@
 
-A system is deterministic if, given its complete state at a point in
+A system is deterministic if, given the system's complete state at a point in
 time, it is possible, in principle, to predict its future behavior
@@
 
-Unexpected nondeterminism usually arises when there is an element of
+Because the
+programmer's abstraction of the system ignores  such a large number of
+details, very few tests are in fact ``deterministic'' in the sense
+that they eliminate all changes in behavior between executions.  Running the same test
+twice almost always results in differences, given a low enough level of
+abstraction, since the system load,
+cache contents, branch predictor history, etc. are almost never
+controlled for; however, this kind of
+nondeterminism is usually of no interest, unless the test involves
+extremely precise timing constraints.  Rather, what matters is when
+some kind of nondeterminism unexpectedly impacts computed values at a
+higher level of abstraction; in general, if the operating system
+itself is not buggy, low-level nondeterminism, by design, is invisible
+except in fine-grained performance testing or real-time systems.
+Unexpected nondeterminism usually arises when there is an element of higher-level
 state or input that \emph{is} critical to the produced behavior, but
-the programmer has not anticipated.  E.g., it is beleived that the
+the programmer has not anticipated.  E.g.,  when it is believed that the
 behavior of a thread scheduler will not matter, but a race condition
-in the code means that it does matter, after all.
+in the code means that it does matter, or when the order of items in
+an iterator on a hash table is important, and the hash values used are
+salted and thus not predictable.
 
@@
 contexts are occasional rare failures of a system in deployment, and
-flaky tests.
+regression tests that do not behave reliably (known as flaky tests).
+Furthermore, unexpected nondeterminism makes it difficult to use
+automated test generation to produce effective regression tests for a
+system.  While developers may know how to produce reliably
+deterministic unit tests, even in the presence of underlying
+nondeterminism, automated test generation, without a large investment
+in human time, usually does not have sufficient information to avoid
+producing the occasional such test.  Moreover, where a human might
+make an assertion about the final value produced during a unit test,
+an automatically generated regression test is usually most easily produced by
+simply asserting equality with all values produced during a reference
+run, since the final step of a test
+is likely to be somewhat arbitrary, generated in an effort to increase
+code coverage, not establish some functional correctness property.
+This may be one reason that automated testing is seldom used to
+produce general regression tests.  While failure-inducing tests
+produced automatically are often added to regression suites, the full
+set of passing tests is, to our knowledge, only infrequently preserved
+as part of a typical regression suite, perhaps because such tests are
+likely to be \emph{flaky} without considerable human effort.
+
+\subsubsection{Nondeterminism and Flaky Tests}
+
+In order to help ensure that they are reliable and secure, complex
+modern software systems usually include a large set of
+\emph{regression tests}.  A regression test suite is a (usually large)
+set of tests that can be run against a software system every time it
+is modified, to ensure that the modification has not broken the system
+in some way.  \emph{Flaky tests} \cite{miccoflaky} are regression
+tests that fail in an intermittent, unreliable fashion, and thus
+degrade the utility of regression testing.  The essence of a flaky
+test is that, for the same snapshot of test code and code under test
+(CUT), it sometimes fails and sometimes passes: the pass/fail result (disposition) of the
+test is not a deterministic property of the test code, code under
+test, and environment running the test.  This produces three serious
+problems: first, a flaky test often wastes developer time and delays
+software changes by forcing the investigation of code-under-test that
+is not actually incorrect.  Second, failures in flaky tests (for that
+reason) are often ignored, and therefore serious software faults may
+be missed.  Finally, to mitigate these problems, flaky tests are often
+run multiple times, wasting valuable computing resources and also
+delaying acceptance of code changes.  A canary in a coal mine is of
+little use if canaries frequently become ill for reasons unrelated to
+the presence of toxic gases.  Mining may stop for no good reason, or
+miners may learn to ignore the canary, leading to tragedy; a third,
+more ``practical'' option is that miners may carry so many redundant
+canaries into the coal mine that canary-care becomes a serious burden
+on mining.
+
+Flaky tests are, for us, simply a special case of general test
+nondeterminism, where the nondeterminism of test values is sufficient
+to cause the pass/fail result of the test to vary.  The definitions
+and techniques proposed in this paper all apply to flaky tests, as a
+special case of various kinds of nondeterminism described below.
+
+\subsection{Contributions:  Detecting and Debugging Nondeterminism}
+
+This paper proposes (1) a number of formal definitions of types of
+nondeterminism and (2) a number of techniques, based on these definitions, for detecting and debugging
+nondeterminism in property-based testing, particularly for library
+code.  The methods also work for testing at the system level, but are
+perhaps most effective for identifying the low-level sources of
+nondeterminism in libraries.  This enables a number of useful results:
+ the automated production of
+nondeterminism-free regression tests even for libraries with some
+sources of nondeterminism, the documentation of library
+nondeterminism, and, most importantly, detection and effective debugging of
+unexpected nondeterminism that is, itself, a bug.  We first
+describe the types of nondeterminism we aim to detect, and the
+extension of a widely used debugging tool to better handle
+nondeterminism as a test property, then describe an implementation of
+these ideas in a Python testing tool that is used on real-world
+projects.  Finally, we use a set
+of case studies of real-world Python libraries to demonstrate the
+utility of our ideas and understand the cost of adding nondeterminism
+detection to existing automated testing.

### 2018-04-04 relatedwork.tex
bib

@@
 \section{Related Work}
+
+Gao et al.
+generally considered the question of how to make tests repeatable \cite{Gao:2015:MSU:2818754.2818764}, in the context of system-wide user interaction tests.
+Their work focused on systemic factors, such as execution platform and persistent configuration and database files, in contrast to our focus on identifying nondeterminism at the library level.
+However, what they refer to as ``test harness factors'' includes delays, which can be of importance to nondeterminism at the library level when asynchronous behavior is involved.
 
@@
 E.g., they detect instances when  iteration through items of a set is explicitly not guaranteed to provide any particular order of items, but code depends on the order produced by a given implementation.
+This procedure, applied to Python code in a pre-3.3 environment, would have flagged many instances of the nondeterminism that arose on the introduction of salted hashing.
+Determinism is also sometimes used as a property in domain-specific testing efforts, e.g.
+in testing shader compilers \cite{shader}.
+
+The problem of test nondeterminism is closely related to the (extremely important in industrial settings) issue of flaky tests \cite{miccoflaky, luo2014empirical,palomba2017does,listfieldtestanalysis}.
+How to handle flaky tests in practice (when they cannot be avoided) is a major issue in Google-scale continuous testing \cite{memon2017taming}, and, as Memon et al.
+describe, the problem of flaky tests influences the general design of Google test automation.
+Previous work on flaky tests has either focused on test inter-dependence as a cause of flaky behavior \cite{LamZE2015}, or provided large-scale empirical examination of tests from one large open source project (Apache) \cite{luo2014empirical,palomba2017does}.
+Palomba and Zaidman \cite{palomba2017does} investigated the relationship between test code smells and flaky tests.
+The present paper, rather than focusing on flaky tests per se, investigates new tools for handling nondeterminism in property-based test generation.
+From our point of view, flaky behavior is simply a special case of nondeterminism, where the nondeterminism is sufficient to cause the test to have different pass/fail results at times.
 

### 2018-04-04 concepts.tex
more on replication

@@
 required before the predicate returns ``true'' to the delta-debugging
-algorithm.  Tuning $M$, any degree of confidence can be achieved, with
+algorithm.  A set of $M$ repeated false positives with
+$\frac{N}{M}$ samples each is much less likely than a false positive
+with $N$ samples; so long as we accept the resulting bias in favor of
+false negatives, we can therefore produce a reduced test with a
+desired $P(fail)$ much more cheaply:  the use of replications not only
+means we only pay the full sampling price on rare occasions, but a
+desired accuracy for $P$ can be obtained with a much smaller value $M
+\times N$ than a non-dynamically-sampled $N$.  For example, if we the predicate
+is $P(fail) \geq 0.5$, and the true probability for a candidate test
+is 0.25, using $N=8$ will give a false positive rate of over 10\%.
+Using 4 replications on just 2 samples ($M=4;N=2$) yields a false
+positive rate of about 3.7\%, yet requires only $X$ as many
+executions of the test.
+
+Tuning $M$, any degree of confidence can be achieved, with
 the basic tradeoff being between finding a test with the desired
@@
 Because increasing $M$ makes false negatives more likely, larger $M$
-will usually result in less-than-optimal reduction of the original test.
+will usually result in less-than-optimal reduction of the original
+test.

### 2018-04-04 concepts.tex
fix prob stuff

@@
 Using 4 replications on just 2 samples ($M=4;N=2$) yields a false
-positive rate of about 3.7\%, yet requires only $X$ as many
-executions of the test.
+positive rate of about 3.7\%, yet requires almost 60\% fewer
+executions of the test\footnote{The probability calculations for such comparisons
+  are relatively simple, but in real testing are
+  usually not very useful, since the true
+  probabilitiy distributions of test behaviors vary widely and
+  dynamically during the
+  delta-debugging process, making exact calculation unwieldy, even if
+  the changing probabilities were known; exact results would be based on fictional
+  probabilities, so gathering experimental data during a trial
+  delta-debugging run and tuning $N$ and $M$ to yield desired
+  results is  more effective.}.
 

### 2018-04-05 concepts.tex
small edit

@@
 predicate evaluates to true, $M$ repeated true evaluations
-(``replications'') are
+(``replications'', where the first true evaluation is counted as 1 replication) are
 required before the predicate returns ``true'' to the delta-debugging

### 2018-04-05 concepts.tex
figure for types of determinism

@@
   Detection and Debugging}
+
+\begin{figure}
+\includegraphics[width=\columnwidth]{types}
+\caption{Types of determinism}
+\label{fig:types}
+\end{figure}
+
+We define two basic types of determinism, shown in Figure
+\ref{fig:types}:  horizontal determinism and vertical determinism.  In
+horizontal determinism, which is what we usually think of when we
+think about deterministic behavior, a software system reliably
+produces the same behavior given the same steps of a test: the
+behavior of multiple executions that are ``the same'' in terms of
+inputs/actions can be aligned and checked for some type of equality.  In
+vertical determinism, rather than behavior across executions, we are
+interested in behavior within an execution, where repeating the same
+step of a test twice should result in the same behavior.  Obviously,
+vertical determinism is both rarer and more specialized than
+horizontal determinism.  However, in some instances, as we discuss
+below, vertical determinism can be extremely important, both for
+correctness and for ease-of-use of software interfaces.
 

### 2018-04-13 concepts.tex
closer to submittable

@@
 vertical determinism is both rarer and more specialized than
-horizontal determinism.  However, in some instances, as we discuss
-below, vertical determinism can be extremely important, both for
-correctness and for ease-of-use of software interfaces.
+horizontal determinism.  However, in those instances, vertical
+determinism is arguably more important than horizontal determinism, in
+that violations usually indicate a potentially serious fault.
 
@@
 testing, of course, is that multiple implementations of a system are
-almost as rare as good correcteness specifications.
+almost as rare as good correctness specifications.
 
 For the special case of detecting nondeterminism, however, \emph{a
-  system can serve as its own reference implementation.}
+  system can serve as its own reference implementation.}    The
+question, then, becomes one of deciding at what granularity the
+reference equivalence will be checked:  as discussed above,
+processor-instruction and memory-layout determinism is seldom
+necessary or even desired (it would greatly limit optimizations of
+code execution).  We propose two approaches to ``aligning'' an
+execution with itself.
 
@@
 
+Visible value determinism uses the human-accessible outputs (displayed
+or stored to a file), and \emph{values returned by functions or methods
+called as a library by other code}, as the criteria for determining if two executions
+are equivalent.  The idea is simple:  determinism is motivated by the
+desire to create consistent behavior for an observer, whether that
+observer is a human user, another software system, or a regression
+test.  The values output by a software element are the only values of
+interest to the observer.  In practice, of course, some values (time
+stamps, pointer addresses, etc.) are not expected to be deterministic
+by an observer; we call these values \emph{opaque} in that they are
+not interpretable as ``showing'' the internal workings of the code
+being tested for determinism.  Rather, they mask an abstraction,
+usually one managed by the operating system (system
+time, memory management).   Any mechanism for visible value
+determinism needs to support automatic and manual designation of some
+values as opaque.
+
 \subsubsection{Final State Determinism}
 
-\subsubsection{Process-Level Determinism}
+Visible value determinism has significant limitations.  While it
+provides the finest granularity for detecting nondeterminism, which is
+important for debugging purposes, it is also expensive, requiring
+checks on (and possibly storage of) a potentially very large number of
+values.  Additionally, in some cases so many values produced by a
+system are opaque that the annotation burden is inordinate.  In these
+cases, it is better to only compare final states of a computation
+performed by the system being checked for determinism.  The final
+state may have opaque components, but it is easy to define an
+abstraction that reduces the state to the actual value of interest.
+
+%\subsubsection{Core Sources of Difference:  Timing and Process}
 
@@
 
-Apple bugs.
+Vertical determinism is a property that expresses that some operations
+of a software system should, within the same trace, always behave the
+same way.  Usually, for interesting cases, this is dependent on some
+state of the system, though some operations should be completely
+state-independent.  E.g, the hash of a given bytestring returned by a
+security library should never change.  This is one aspect of
+\emph{pure} functions.  For nondeterminism checking, the interesting
+cases are non-pure: a function depends on system state, but should
+always depend on it in the same way, and should not, itself, change
+system state in a way that would change its behavior on a subsequent
+run.
 
-\subsubsection{Failure Nondeterminism}
+Many \emph{idempotent} operations fall into this category.
+Consider adding an element to an AVL tree implementing a set, not a
+multiset.  Assume the method call returns the parent node of the
+element, whether it was added or was already in the tree.  Calling
+this mehtod any number of times in sequence should always return the
+same value, though the first call may have modified the contents of
+the AVL tree.
+
+\subsubsection{Failure Determinism}
+
+A specialized case of vertical determinism is failure determinism.
+Failure determinism is the following restriction on an API:
+
+{\bf If a call fails, and indicates this to the caller, it should not modify system
+  state in any way.  Changes made before the failure should be rolled
+  back.}
+
+In other words, failure determinism is a property stating that an API
+is \emph{transactional} with respect to failures (it may not be so
+with respect to non-failing calls).
+
+From a user's perspective, some behaviors of the macOS High Sierra
+root exploit (CVE-2017-13872 \cite{applebug0}) exhibited failure
+nondeterminism.  Attempting to login with the root account with an
+empty password appeared to fail, then, on a repeated try, succeeded
+\cite{applebug1,applebug2}.
 

### 2018-04-13 experiments.tex
pyfakfefs

@@
 
-\subsection {Case Study: pyfakefs and OS X}
+\subsection {Case Study: pyfakefs}
+
+The {\tt pyfakefs} \cite{pyfakefs} module implements a fake file
+system that mocks the Python file system modules, to allow Python
+tests both to run faster by using an in-memory file system and to make
+or depend on file system changes that would
+not be safe or easily performed using real persistent storage.  Originally
+developed in 2006 at Google by Mike Bland, {\tt pyfakefs} is now used
+in over 2,000 Python tests, inside and outside Google.
+
+We introduced a subtle bug into {\tt pyfakefs}, where the {\tt remove}
+call checks that its target is not a directory, and returns the
+correct error, but still carries out the remove operation.  Using {\tt
+  os.remove} to delete directories does not break any file system
+invariants, but violates the Python {\tt os} specification (and,
+indirectly, the usual POSIX implementation behavior where {\tt unlink}
+does not work for directories).  Detecting this bug using the TSTL
+{\tt pyfakefs} harness is normally impossible without using another
+file system as a reference.  However, the fault is detected
+immediately, without using a reference, when we compile the harness
+with the {\tt --checkFailureDeterminism} flag.
 

### 2018-04-14 experiments.tex
edits

@@
 
-\subsection {Case Study: pyfakefs}
+\subsection {Case Study: {\tt pyfakefs}}
 
@@
 developed in 2006 at Google by Mike Bland, {\tt pyfakefs} is now used
-in over 2,000 Python tests, inside and outside Google.
+in over 2,000 Python tests, inside and outside Google \cite{pyfakefs}.
+
+The TSTL harness for {\tt pyfakefs} has been used to detect (and
+correct) over 50 faults.  However, the testing completely relies on
+the existence of a reference POSIX implementation.
 
@@
 {\tt pyfakefs} harness is normally impossible without using another
-file system as a reference.  However, the fault is detected
-immediately, without using a reference, when we compile the harness
+file system as a reference.  However, the fault was detected
+immediately, even without using a reference, when we compile the harness
 with the {\tt --checkFailureDeterminism} flag.
+Moreover, the overhead for adding a check for failure determinism was
+less than 8\%.
 

### 2018-04-14 experiments.tex
latest

@@
 tests both to run faster by using an in-memory file system and to make
-or depend on file system changes that would
-not be safe or easily performed using real persistent storage.  Originally
-developed in 2006 at Google by Mike Bland, {\tt pyfakefs} is now used
-in over 2,000 Python tests, inside and outside Google \cite{pyfakefs}.
+file system changes that would not be safe or easily performed using
+real persistent storage.  Originally developed in 2006 at Google by
+Mike Bland, {\tt pyfakefs} is now used in over 2,000 Python tests,
+inside and outside Google \cite{pyfakefs}.
 
@@
 correct) over 50 faults.  However, the testing completely relies on
-the existence of a reference POSIX implementation.
+the existence of a reference file system implementation.
 
@@
 file system as a reference.  However, the fault was detected
-immediately, even without using a reference, when we compile the harness
+immediately, even without using a reference, when we compiled the harness
 with the {\tt --checkFailureDeterminism} flag.
@@
 
-\subsection{Case Study: Datarray Inference Algorithms}
+\subsection{Case Study: {\tt datarray} Inference Algorithms}
 
-\url{https://github.com/BIDS/datarray}
+The {\tt datarray} module \cite{datarray} is a prototype
+implementation for numpy arrays with named axes to improve data
+management.  As part of its code, it provides a set of algorithms for
+inference in
+Bayesian belief networks \cite{russell2016artificial}.  An earlier
+version of these algorithms produced nondeterministic (and in some
+cases incorrect) results due to dependence on the order of values in
+an iterator over a Python dictionary.  This is a simple, clean example
+of a problem that appeared in numerous Python programs when Python
+version 3.3 introduced automatic random salting of hashes per-process,
+in order to mitigate hash-based denial of service attacks
+\cite{denial}.  Until version 3.6 this not only resulted in changes in
+exact hash values, but in the order of iteration on
+dictionaries\footnote{In version 3.6, the default dictionary
+  implementation became an ordered dictionary with consistent
+  iteration, though actual hashes remained nondeterministic.}.  While
+relatively few programs relied on exact hash values, many implicitly
+relied on them in that they only functioned correctly with a
+predictable order for dictionary iteration across processess.
+
+Figure \ref{hashbug} shows TSTL code for generating inputs to the {\tt
+  datarray} algorithms.  Running this harness using TSTL's {\tt
+  --checkProcessDeterminism} flag requires less than 10 seconds on
+average to produce a test exhibiting process-level nondeterminism.
 

### 2018-04-14 experiments.tex
fixup

@@
   --checkProcessDeterminism} flag requires less than 10 seconds on
-average to produce a test exhibiting process-level nondeterminism.
+average to produce a test exhibiting process-level nondeterminism in
+the {\tt calc\_marginals\_sumproduct} function (the only broken algorithm).
 

### 2018-04-15 experiments.tex
More text
latest

@@
 immediately, even without using a reference, when we compiled the harness
-with the {\tt --checkFailureDeterminism} flag.
+with the {\tt --checkFailureDeterminism} flag.  Because vertical
+reduction does not require running complete tests multiple times, and
+does not affect the delta-debugging algorithm's performance, reducing
+the failing test to 3 steps required less than a second.
 Moreover, the overhead for adding a check for failure determinism was
@@
 average to produce a test exhibiting process-level nondeterminism in
-the {\tt calc\_marginals\_sumproduct} function (the only broken algorithm).
+the {\tt calc\_marginals\_sumproduct} function (the only broken
+algorithm).  Reducing this test to a minimal test of only 6 steps
+requires another 92 seconds, and shows the problem very clearly.
 

### 2018-04-15 experiments.tex
ok

@@
 the failing test to 3 steps required less than a second.
-Moreover, the overhead for adding a check for failure determinism was
-less than 8\%.
+Moreover, the overhead for the check for failure determinism in a
+version of the code without the {\tt remove} error was
+less than 8\%.  Detecting the fault using a reference file system also
+required 17\% more time, and took about twice as long to reduce to a
+slightly longer failing test, which did not have {\tt remove} as its
+final operation (since further file system interaction is required to
+expose the bad state).
 

### 2018-04-15 experiments.tex
small edits

--- 2018-04-15 experiments.tex
@@
 the failing test to 3 steps required less than a second.
-Moreover, the overhead for adding a check for failure determinism was
-less than 8\%.
+Moreover, the overhead for the check for failure determinism in a
+version of the code without the {\tt remove} error was
+less than 8\%.  Detecting the fault using a reference file system 
+required 17\% more runtime before detection, and took over twice as
+long to reduce the failure to a
+slightly longer failing test, which did not have {\tt remove} as its
+final operation (since further file system interaction is required to
+expose the bad state).
 
@@
   datarray} algorithms.  Running this harness using TSTL's {\tt
-  --checkProcessDeterminism} flag requires less than 10 seconds on
+  --checkProcessDeterminism} flag required less than 10 seconds on
 average to produce a test exhibiting process-level nondeterminism in
 the {\tt calc\_marginals\_sumproduct} function (the only broken
-algorithm).  Reducing this test to a minimal test of only 6 steps
-requires another 92 seconds, and shows the problem very clearly.
+algorithm).  Reducing this 60 step test to a minimal test of only 6 steps,
+showing an extremely simple input producing the issue 
+required another 92 seconds.
 

### 2018-04-16 experiments.tex
fix

@@
 implementation for numpy arrays with named axes to improve data
-management.  As part of its code, it provides a set of algorithms for
+management, developed by the Berkeley Institute for Data Science.  As part of its code, it provides a set of algorithms for
 inference in

### 2018-04-16 introduction.tex
start on redis

@@
 and techniques proposed in this paper all apply to flaky tests, as a
-special case of various kinds of nondeterminism described below.
+special case of various kinds of \emph{horizontal} nondeterminism described below.
 

### 2018-04-16 concepts.tex
in progress

@@
 element, whether it was added or was already in the tree.  Calling
-this mehtod any number of times in sequence should always return the
+this method any number of times in sequence should always return the
 same value, though the first call may have modified the contents of
 the AVL tree.
+
+One application of vertical determinism, then, is to identify
+idempotent operations in an automated test generation harness, and
+simply automatically retry all such operations, checking that the
+result is unchanged.  The overhead of vertical nondeterminism detection will
+generally be much lower than that for horizontal nondeterminism
+(horizontal checks must re-execute an entire test; vertical checks
+only re-execute proportional to the number of idempotent operations
+performed).  However, identifying idempotent operations is a
+fairly serious specification burden.  Are there instances where a tool
+can automatically identify a limited kind of idempotent behavior?
 
@@
 
+Many library APIs are largely failure deterministic.  For instance, if
+we exclude actual I/O errors, most POSIX file system calls either
+succeed or do nothing (interfaces that may only partially succeed,
+such as {\tt read} tend to explicitly return the degree of success
+rather than signaling an error on partial success).  In fact, the
+design of the POSIX {\tt write} call is carefully crafted to largely maintain
+failure determinism.  If the failure mode is one that can be
+anticipated, such as insufficient space to write all bytes requested
+(but some bytes can be written), the call returns the actual bytes
+written, and does not set a failure code.
+
+In languages with a clear mechanism for expressing failure of a call
+(e.g., exceptions in Python and Java, or Option/Result types in Rust,
+Haskell, and ML), failure determinism can be automatically checked.
+Moreover, the expected overhead should be even lower than the general
+overhead for vertical determinism, in that we can expect most failing
+operations to be fast in an automated test generation setting (that the
+parameters to a call are illegal is the most common cause for
+failure).
+
 \subsection{Nondeterminism and Delta-Debugging}
@@
 function calls.  This is no different than the common use of
-delta-debugging.  However, in nondeterminism detection,
-delta-debugging can also be used to \emph{change the probability of
-  nondeterministic behavior}.
+delta-debugging.  However, in horizontal nondeterminism detection,
+delta-debugging also tends to \emph{change the probability of
+  nondeterministic behavior}; this can be both harmful and
+beneficial.  The behavior is part of a more general (and, to our
+knowledge, not previously investigate) issue:  using delta-debugging
+to minimize a test with respect to a predicate that only holds with a
+certain probability (the predicate itself is nondeterministic).
 
-In \emph{monotonic} cases, removing a part of a test cannot
-increase the probability of a predicate holding.  This is fairly
+In \emph{monotonic} probabilistic delta-debugging, removing a part of a test cannot
+increase the probability of the predicate holding.  This is fairly
 common.  In these cases, when we use a reduction algorithm to reduce $t$

### 2018-04-16 experiments.tex
start on redis
in progress

@@
 
-\subsection {Case Study: Redis-Py}
+\subsection {Case Study: {\tt redis-py}}
+
+The {\tt redis-py} \cite{redispy} module implements a widely-used Python interface
+to the popular Redis in-memory database, cache, and
+message broker \cite{redis}.
+
+Using TSTL's harness for {\tt redis-py}, both {\tt redis-py} and Redis
+itself can be tested; unfortunately, generating standalone
+high-coverage regression tests for {\tt redis-py} and Redis using TSTL
+was difficult, as numerous Redis commands introduce nondeterministic
+behavior.  Some of these are obvious (e.g., {\tt randomkey, srandmember}) by
+name.  Given an understanding of Redis semantics, that the various
+commands producing data with a limited lifetime (e.g., {\tt expire,
+  pexpire}) also introduce timing-based nondeterminism is also clear.
+However, that a command such as {\tt restore} takes an expiration
+argument is not obvious on inspection, nor is the behavior of {\tt
+  spop} which pops a random value from a set.  Moreover, it is
+difficult to guess whetehr the {\tt pipe} mechanism, which allows a
+large sequence of commands to be queued up in a pipe and executed all
+at once, introduced the potential for nondeterminism (does it execute
+sequentially before other command are handled, or is it in parallel
+with further commands).  Deep experience with Redis would make these
+issues clear, but tests using a library are often written by those not
+intimately familliar with the semantics of every used library (otherwise they would not have
+been as likely to introduce flaky tests in the first place).  This is
+even more true in the case of automated test generation, where the
+test engineer is often chosen for expertise in test generation tools,
+not in the domain of the software under test (SUT), and is seldom the
+original developer of the code.
+
+Using the original TSTL harness, approxoimately \emph{over 20\%} of all {\tt redis-py}
+regression tests (of length 200) generated were flaky.  Using the
+nondeterminism checker to reduce flaky tests to a minimal set allows
+an engineer to leave an overnight run that will, in 12 hours, produce
+a set of minimal flaky tests exposing the sources of
+nondeterminism/flakiness.  In this case, the probability of flakiness
+is not important, just the possibility, so we let the delta-debugging
+create a minimal test with some non-zero (but possibly very small)
+observed probability of nondeterminism.  The overhead for adding
+nondeterminism checking is
+
+After removing all identified sources of nondeterminism (the {\tt random-} functions, all
+expiration-related functionality and {\tt restore} calls with non-zero
+lifetimes, and {\tt spop}), but leaving all other suspect funcitonality in
+place (e.g. pipes), no flakiness was observed in a sample of over 1,000 length
+200 tests.  Moreover, because the removals were limited to acutally
+observed sources of flakiness, the code coverage loss was minimal.
+Average branch coverage, for regression tests of length 200 is reduced
+by less than 5 branches.  Obviously, it is
+impossible to test the removed calls, but the overall coverage loss is
+both minimal and known, and can be made up for using specially-crafted
+tests (for instance, wrapping the problematic calls in a way that does
+not check the values, or adding a delay after an expiration to allow
+the data to expire).  As a price to pay for the ability to produce
+fast-executing high-coverage full regression tests (not just tests for
+crashes and unexpected exceptions), this is a very acceptable price.
+Moreover, turning off sources of known, expected nondeterminism makes
+it possible to aggressively test {\tt redis-py} and Redis for
+unexpected nondeterminism arising from actual bugs.  The overhead for such
+determinism checks, with a 0 second delay between operations, is only
+20\%, much lower than the expected cost of running each test twice.
+This is because choosing the actions in a test (and determining which
+actions are enabled at each step) consumes a large part of the test
+generator's time; running the test again and checking equality is
+relatively inexpensive.
+
+\subsection{Case Study: {\tt datarray} Inference Algorithms}
+
+The {\tt datarray} module \cite{datarray} is a prototype
+implementation for numpy arrays with named axes to improve data
+management, developed by the Berkeley Institute for Data Science.  As part of its code, it provides a set of algorithms for
+inference in
+Bayesian belief networks \cite{russell2016artificial}.  An earlier
+version of these algorithms produced nondeterministic (and in some
+cases incorrect) results due to dependence on the order of values in
+an iterator over a Python dictionary.  This is a simple, clean example
+of a problem that appeared in numerous Python programs when Python
+version 3.3 introduced automatic random salting of hashes per-process,
+in order to mitigate hash-based denial of service attacks
+\cite{denial}.  Until version 3.6 this not only resulted in changes in
+exact hash values, but in the order of iteration on
+dictionaries\footnote{In version 3.6, the default dictionary
+  implementation became an ordered dictionary with consistent
+  iteration, though actual hashes remained nondeterministic.}.  While
+relatively few programs relied on exact hash values, many implicitly
+relied on them in that they only functioned correctly with a
+predictable order for dictionary iteration across processess.
+
+Figure \ref{hashbug} shows TSTL code for generating inputs to the {\tt
+  datarray} algorithms.  Running this harness using TSTL's {\tt
+  --checkProcessDeterminism} flag required less than 10 seconds on
+average to produce a test exhibiting process-level nondeterminism in
+the {\tt calc\_marginals\_sumproduct} function (the only broken
+algorithm).  Reducing this 60 step test to a minimal test of only 6 steps,
+showing an extremely simple input producing the issue, 
+required another 92 seconds.
+
+Removing the nondeterministic call, we can see that the cost of
+checking for process nondeterminism, with a zero delay on operations,
+is high, a 93\% slowdown.  This is due to the very high cost of
+using a new subprocess to re-run the test, and comparing values using
+text I/O on output pipes, rather than on values alone.
 
@@
 expose the bad state).
-
-\subsection{Case Study: {\tt datarray} Inference Algorithms}
-
-The {\tt datarray} module \cite{datarray} is a prototype
-implementation for numpy arrays with named axes to improve data
-management, developed by the Berkeley Institute for Data Science.  As part of its code, it provides a set of algorithms for
-inference in
-Bayesian belief networks \cite{russell2016artificial}.  An earlier
-version of these algorithms produced nondeterministic (and in some
-cases incorrect) results due to dependence on the order of values in
-an iterator over a Python dictionary.  This is a simple, clean example
-of a problem that appeared in numerous Python programs when Python
-version 3.3 introduced automatic random salting of hashes per-process,
-in order to mitigate hash-based denial of service attacks
-\cite{denial}.  Until version 3.6 this not only resulted in changes in
-exact hash values, but in the order of iteration on
-dictionaries\footnote{In version 3.6, the default dictionary
-  implementation became an ordered dictionary with consistent
-  iteration, though actual hashes remained nondeterministic.}.  While
-relatively few programs relied on exact hash values, many implicitly
-relied on them in that they only functioned correctly with a
-predictable order for dictionary iteration across processess.
-
-Figure \ref{hashbug} shows TSTL code for generating inputs to the {\tt
-  datarray} algorithms.  Running this harness using TSTL's {\tt
-  --checkProcessDeterminism} flag required less than 10 seconds on
-average to produce a test exhibiting process-level nondeterminism in
-the {\tt calc\_marginals\_sumproduct} function (the only broken
-algorithm).  Reducing this 60 step test to a minimal test of only 6 steps,
-showing an extremely simple input producing the issue 
-required another 92 seconds.
 
@@
 \end{figure*}
+
+\subsection {Vertical Determinism Case Study: {\tt pyfakefs}}
+
+The {\tt pyfakefs} \cite{pyfakefs} module implements a fake file
+system that mocks the Python file system modules, to allow Python
+tests both to run faster by using an in-memory file system and to make
+file system changes that would not be safe or easily performed using
+real persistent storage.  Originally developed in 2006 at Google by
+Mike Bland, {\tt pyfakefs} is now used in over 2,000 Python tests,
+inside and outside Google \cite{pyfakefs}.
+
+The TSTL harness for {\tt pyfakefs} has been used to detect (and
+correct) over 50 faults.  However, the testing completely relies on
+the existence of a reference file system implementation.
+
+We introduced a subtle bug into {\tt pyfakefs}, where the {\tt remove}
+call checks that its target is not a directory, and returns the
+correct error, but still carries out the remove operation.  Using {\tt
+  os.remove} to delete directories does not break any file system
+invariants, but violates the Python {\tt os} specification (and,
+indirectly, the usual POSIX implementation behavior where {\tt unlink}
+does not work for directories).  Detecting this bug using the TSTL
+{\tt pyfakefs} harness is normally impossible without using another
+file system as a reference.  However, the fault was detected
+immediately, even without using a reference, when we compiled the harness
+with the {\tt --checkFailureDeterminism} flag.  Because vertical
+reduction does not require running complete tests multiple times, and
+does not affect the delta-debugging algorithm's performance, reducing
+the failing test to 3 steps required less than a second.
+Moreover, the overhead for the check for failure determinism in a
+version of the code without the {\tt remove} error was
+less than 8\%.  Detecting the fault using a reference file system 
+required 17\% more runtime before detection, and took over twice as
+long to reduce the failure to a
+slightly longer failing test, which did not have {\tt remove} as its
+final operation (since further file system interaction is required to
+expose the bad state).

### 2018-04-16 experiments.tex
fix duplication of pyfakefs
fix cov

@@
 observed sources of flakiness, the code coverage loss was minimal.
-Average branch coverage, for regression tests of length 200 is reduced
-by less than 5 branches.  Obviously, it is
+Mean branch coverage, for regression tests of length 200 was reduced
+by less than 5 branches (a less than 1\% decrease).  Obviously, it is
 impossible to test the removed calls, but the overall coverage loss is
@@
 crashes and unexpected exceptions), this is a very acceptable price.
-Moreover, turning off sources of known, expected nondeterminism makes
+
+Moreover, turning off sources of expected nondeterminism makes
 it possible to aggressively test {\tt redis-py} and Redis for
@@
 text I/O on output pipes, rather than on values alone.
-
-\subsection {Case Study: {\tt pyfakefs}}
-
-The {\tt pyfakefs} \cite{pyfakefs} module implements a fake file
-system that mocks the Python file system modules, to allow Python
-tests both to run faster by using an in-memory file system and to make
-file system changes that would not be safe or easily performed using
-real persistent storage.  Originally developed in 2006 at Google by
-Mike Bland, {\tt pyfakefs} is now used in over 2,000 Python tests,
-inside and outside Google \cite{pyfakefs}.
-
-The TSTL harness for {\tt pyfakefs} has been used to detect (and
-correct) over 50 faults.  However, the testing completely relies on
-the existence of a reference file system implementation.
-
-We introduced a subtle bug into {\tt pyfakefs}, where the {\tt remove}
-call checks that its target is not a directory, and returns the
-correct error, but still carries out the remove operation.  Using {\tt
-  os.remove} to delete directories does not break any file system
-invariants, but violates the Python {\tt os} specification (and,
-indirectly, the usual POSIX implementation behavior where {\tt unlink}
-does not work for directories).  Detecting this bug using the TSTL
-{\tt pyfakefs} harness is normally impossible without using another
-file system as a reference.  However, the fault was detected
-immediately, even without using a reference, when we compiled the harness
-with the {\tt --checkFailureDeterminism} flag.  Because vertical
-reduction does not require running complete tests multiple times, and
-does not affect the delta-debugging algorithm's performance, reducing
-the failing test to 3 steps required less than a second.
-Moreover, the overhead for the check for failure determinism in a
-version of the code without the {\tt remove} error was
-less than 8\%.  Detecting the fault using a reference file system 
-required 17\% more runtime before detection, and took over twice as
-long to reduce the failure to a
-slightly longer failing test, which did not have {\tt remove} as its
-final operation (since further file system interaction is required to
-expose the bad state).
 

### 2018-04-16 experiments.tex
nodnet

@@
 cases incorrect) results due to dependence on the order of values in
-an iterator over a Python dictionary.  This is a simple, clean example
-of a problem that appeared in numerous Python programs when Python
-version 3.3 introduced automatic random salting of hashes per-process,
-in order to mitigate hash-based denial of service attacks
-\cite{denial}.  Until version 3.6 this not only resulted in changes in
-exact hash values, but in the order of iteration on
-dictionaries\footnote{In version 3.6, the default dictionary
-  implementation became an ordered dictionary with consistent
-  iteration, though actual hashes remained nondeterministic.}.  While
-relatively few programs relied on exact hash values, many implicitly
-relied on them in that they only functioned correctly with a
-predictable order for dictionary iteration across processess.
+an iterator over a Python dictionary (see Section \ref{sec:pnondet}).
 

### 2018-04-16 implementation.tex
closer to submittable
ok
nodnet

@@
 \section{Implementation}
+
+\subsection{Sources of Nondeterminism in Practice}
+
+\subsubsection{Process-Based Nondeterminism}
+
+\label{sec:pnondet}
+
+Some sources of nondeterminism require executing a test in a new
+process environment, because the source is inherently tied to the
+process environment.
+
+Address Space Layout Randomization (ASLR) \cite{ASLR}  is probably the most
+important source of nondeterminism that arises (only) from change in
+process.  ASLR scrambles the layout of
+
+Different CPU.
+
+This is a simple, clean example
+of a problem that appeared in numerous Python programs when Python
+version 3.3 introduced automatic random salting of hashes (see Section
+\ref{sec:pnondet} per-process,
+in order to mitigate hash-based denial of service attacks
+\cite{denial}.  Until version 3.6 this not only resulted in changes in
+exact hash values, but in the order of iteration on
+dictionaries\footnote{In version 3.6, the default dictionary
+  implementation became an ordered dictionary with consistent
+  iteration, though actual hashes remained nondeterministic.}.  While
+relatively few programs relied on exact hash values, many implicitly
+relied on them in that they only functioned correctly with a
+predictable order for dictionary iteration across processess.
+
+\subsection{Implementing Nondeterminism Detection}
 
@@
 or using a new process, in checking the behavior.
+
+Because TSTL has an interface to afl \cite{aflfuzz}, we can even use
+afl's sophisticated heuristics to perform very thorough, week-long
+checks for nondeterminism, using strategies built to find subtle
+security vulnerabilities in C programs.

### 2018-04-16 implementation.tex
ok

@@
 important source of nondeterminism that arises (only) from change in
-process.  ASLR scrambles the layout of
+process.  ASLR scrambles the layout of memory of the process in which
+an executable runs in order to make it
+harder to exploit memory-safety vulnerabilities.  As a side effect, it
+means that a test that, in some circumstances, causes a crash, may at
+other times not fail at all.
 
@@
 
-This is a simple, clean example
-of a problem that appeared in numerous Python programs when Python
+Another example of process-based nondeterminism was discovered by numerous Python programs when Python
 version 3.3 introduced automatic random salting of hashes (see Section
@@
 relied on them in that they only functioned correctly with a
-predictable order for dictionary iteration across processess.
+predictable order for dictionary iteration across processess.  Testing
+Python code for nondeterminism based on this hash seed requires
+running in a new process:  Python does not allow changing the salt, since changing hash values on-the-fly would
+break all existing dictionaries and other structures relying on hashing.
 

### 2018-04-16 introduction.tex
ok

@@
 describe the types of nondeterminism we aim to detect, and the
-extension of a widely used debugging tool to better handle
+extension of a widely used debugging algorithm to better handle
 nondeterminism as a test property, then describe an implementation of

### 2018-04-16 experiments.tex
code

@@
 create a minimal test with some non-zero (but possibly very small)
-observed probability of nondeterminism.  The overhead for adding
-nondeterminism checking is
+observed probability of nondeterminism.
 
@@
 unexpected nondeterminism arising from actual bugs.  The overhead for such
-determinism checks, with a 0 second delay between operations, is only
+determinism checks, with no delay between operations, is only
 20\%, much lower than the expected cost of running each test twice.
@@
 Removing the nondeterministic call, we can see that the cost of
-checking for process nondeterminism, with a zero delay on operations,
+checking for process nondeterminism, with no delay between operations,
 is high, a 93\% slowdown.  This is due to the very high cost of

### 2018-04-17 implementation.tex
more on implementation

@@
 \subsection{Sources of Nondeterminism in Practice}
+
+In order to implement a practical nondeterminism-detection tool, it is
+important to consider the sources of nondeterminism.  Most of these
+sources can, in principle, be detected simply by re-running a test
+twice and comparing results, either at all visible values or at the
+final state.  For example, if some library call uses a random number
+generator (perhaps in a stochastic algorithm), the value should be
+different on a second run.  Timing dependencies may change on a second
+execution, especially if a delay is added between steps of the test (a
+more principled version of the unfortunate tradition of testing
+concurrent code by adding ad-hoc {\tt sleep} statements).  However,
+simply running a test twice (possibly with a delay between steps) does
+not suffice in all cases.
 
@@
 
-Some sources of nondeterminism require executing a test in a new
+Some sources of nondeterminism unfortunately require executing a test in a new
 process environment, because the source is inherently tied to the
-process environment.
+process in which code runs; simply re-executing a test in the same
+process will not reveal the problem.
 
@@
 process.  ASLR scrambles the layout of memory of the process in which
-an executable runs in order to make it
-harder to exploit memory-safety vulnerabilities.  As a side effect, it
+an executable runs in order to make it harder to exploit memory-safety
+vulnerabilities in code.  As a side effect, it
 means that a test that, in some circumstances, causes a crash, may at
 other times not fail at all.
-
-Different CPU.
 
@@
 running in a new process:  Python does not allow changing the salt, since changing hash values on-the-fly would
-break all existing dictionaries and other structures relying on hashing.
+break all existing dictionaries and other structures relying on
+hashing.
+
+Other, less common, effects tied to process also exist.  A few tests
+may somehow depend on their actual PID (Process ID) or at least the
+parent process' ID.  Process change increases the likelyhood that on a multicore
+system a test will execut on a different CPU than in the first
+execution, which can have many subtle effects, mostly related to
+changed timing.
 
@@
 
-We implemented our approach in the TSTL \cite{NFM15} system.
+\begin{table*}
+\centering
+{%\scriptsize
 
+\caption{TSTL Method Calls for Nondeterminism Detection}
+\label{tab:methods}
+\begin{tabular}{l|l}
+%\hline
+Name & Purpose \\
+\hline
+\hline
+{\tt nondeterministic} & returns True if test exhibits final state nondeterminism in $k$
+  tries with delay $d$\\
+\hline
+{\tt stepNondeterministic} & as above, except checks all visible values
+                       (compares state at every step)\\
+\hline
+{\tt processNondeterministic} & as above, except runs test provided in
+                          subprocess and compares non-opaque output
+                          values \\
+\hline
+{\tt forceP} & wraps a predicate for test reduction (e.g. {\tt
+         nondeterministic}) with desired probability of holding,
+         sample size,\\
+&  and replications required \\
+\hline
+\hline
+
+%\hline
+\end{tabular}
+}
+\end{table*}
+
+\begin{table*}
+\centering
+{%\scriptsize
+\caption{TSTL Command Line Options for
+  Nondeterminism Detection}
+\label{tab:options}
+\begin{tabular}{l|l|l}
+%\hline
+Name &  Tool(s) & Purpose \\
+\hline
+\hline
+{\tt --checkDeterminism} & {\tt rt}, {\tt reduce} & Checks tests for deterministic
+                                  behavior (visible value in rt, final
+  value in reduce)\\
+\hline
+{\tt --checkProcessDeterminism} & {\tt rt}, {\tt reduce} & Checks determinism using more 
+                                         expensive process-changed 
+                                         execution \\
+\hline
+{\tt --determinismTries INT} & {\tt rt}, {\tt reduce} & Number of times to repeat test 
+                                  checking for nondeterministic
+                                  behavior \\
+\hline 
+{\tt --determinismDelay FLOAT} & {\tt rt}, {\tt reduce} & Delay between steps when 
+                                  re-executing to check for
+                                              determinism \\
+\hline 
+{\tt --checkFailureDeterminism} & tstl compiler & Produce a harness
+                                                  that checks
+                                                  determinism on all
+                                                  failing actions \\
+\hline 
+{\tt --probability P} & {\tt reduce} & force property with respect to which 
+                               test is being reduce to hold with 
+                                 probability $P$ \\
+\hline 
+{\tt --samples S} & {\tt reduce} & use $S$ samples when checking predicate
+                             probability\\
+\hline 
+{\tt --replications R} & {\tt reduce} & use $R$ replications when checking predicate probability\\
+\hline
+\hline
+
+%\hline
+\end{tabular}
+}
+\end{table*}
+
+We implemented our approach in the TSTL \cite{NFM15} system, a
+language and tool for property-based testing of Python code.
 Because TSTL supports differential testing \cite{tstlsttt}, horizontal
@@
 
+We therefore instead made nondeterminism detection a first-class
+property in TSTL, using TSTL's existing notation for marking some
+types of values as \emph{opaque} (not usefully compared for equality),
+used in automatic abstraction-based testing.  Tables \ref{tab:methods}
+and \ref{tab:options} show, respectively, the primary additions to the
+TSTL Python API and the TSTL command line tools.  The TSTL test
+generation tool is called {\tt rt}, since the default mode is a fast
+pure Random Tester, and {\tt reduce} is TSTL's tool for
+delta-debugging and test
+normalization (more aggressive reduction that also tends to ease
+debugging and test triage by making tests failing due to the same
+fault more similar \cite{onetest}).  With these additions,
+developers of TSTL-based testing tools, or developers using TSTL tools
+to test code can easily add nondeterminism to the set of properties
+checked.  First-class vertical nondeterminism checks are currently
+limited to failures:  {\tt --checkFailureDeterminism} causes the TSTL
+compiler to emit a harness where every action that causes an expected
+exception to be raised is repeated to ensure the same exception is
+still raised.  Users can also express that any action should exhibit
+vertical determinism by wrapping the action in a function that checks
+for failure and calls the code again.
+
+Unlike the other additions, explicitly designed for nondeterminism
+detection, the {\tt forceP} wrapper and {\tt --probability}, {\tt
+  --samples}, and {\tt --replications} reducer options are general
+tools for use with nondeterministic predicates in delta-debugging.  Of
+course, nondeterminism and flakyness are the most common probabilistic
+test properties we are aware of, but in theory these tools can be used
+for other probabilistic debugging problems (e.g., if the property of
+interest is that a value is above a certain threshold in 90\% of test
+executions, these TSTL additions can be used to find a small test where this property
+fails to hold).
+
 Because TSTL has an interface to afl \cite{aflfuzz}, we can even use

### 2018-04-18 concepts.tex
ok

@@
 to minimize a test with respect to a predicate that only holds with a
-certain probability (the predicate itself is nondeterministic).
+certain probability (the predicate itself is nondeterministic).  Note
+that while most descriptions of delta-debugging discuss reducing a
+test with respect to \emph{test failure} (hence the term
+``delta-\emph{debugging}'') we adopt the wider view of delta-debugging
+as reducing the length of a test while maintaining that an arbitrary
+predicate remains true of the shorter versions of the test.  The
+traditional predicate is ``the test fails'' but other predicates,
+especially those related to code-coverage, can also be useful
+\cite{icst2014,stvrcausereduce,NonAdeq}.  Our contribution here is to consider
+what happens when the predicate to be evaluated is not deterministic.
+``The test behaves flakily'' is a simple, important, example of such a predicate.
 
@@
 \subsubsection{``Publication Bias'' in Delta-Debugging}
+
+\begin{figure*}
+\centering 
+\begin{subfigure}{0.67\columnwidth}
+\centering
+\includegraphics[width=\columnwidth]{lengthddmin}
+\caption{Standard delta-debugging}
+\label{fig:p1}
+\end{subfigure}
+\begin{subfigure}{0.67\columnwidth}
+\centering
+\includegraphics[width=\columnwidth]{lengthddminforcep}
+\caption{P=0.5, 100 samples}
+\label{fig:p2}
+\end{subfigure}
+\begin{subfigure}{0.67\columnwidth}
+\centering
+\includegraphics[width=\columnwidth]{lengthddminforceprep}
+\caption{10 samples, 10 replications}
+\label{fig:p3}
+\end{subfigure}
+\caption{Delta-debugging of the same test, with different ways of
+  handling nondeterministic in the reduction predicate.}
+\end{figure*}
 
@@
 on average, due to two factors:  first, timing-induced nondeterminism
-has a much smaller temporal space to operate in, and second, if some
-test operation introduces a small probability of nondeterminism, and
-only many repetitions of that operation make the probability large,
-small tests obviously on average have fewer instances of the
-problematic operation.
+has a much smaller temporal space to operate in (the test may be over
+before a ``time-bomb'' set by an operation goes off, for example), and second, if some
+test operations introduce a small probability of nondeterminism, and
+only many repetitions of those operations make the probability large,
+small tests obviously, on average, have fewer instances of the
+problematic operations.
 
@@
 positive rate of about 3.7\%, yet requires almost 60\% fewer
-executions of the test\footnote{The probability calculations for such comparisons
+executions of the test.  The probability calculations for such comparisons
   are relatively simple, but in real testing are
@@
   delta-debugging run and tuning $N$ and $M$ to yield desired
-  results is  more effective.}.
+  results is  more effective.
 
@@
 test.
+
+To make the basic concepts, more clear, Figures
+\ref{fig:p1}-\ref{fig:p3} graphically show the interplay of
+delta-debugging and nondeterministic predicates for a simple example.
+In the example, tests consist of a sequence of operations that behave
+nondeterministically with (independent) probabilities of 0.01, 0.05,
+and 0.10, respectively, storing the value of the operation one of five array locations.
+There is also an operation that clears out an array location, making
+it possible to store another (possibly nondeterministic) value in that
+array slot.  An example test sequence would be:
+
+\begin{code}
+array[2] = op01();
+array[3] = op05();
+clear(array[2]);
+array[2] = op10();
+\end{code}
+
+The probability that this test behaves nondeterministically is
+0.84645, the result of the simple calculation $(1-0.01) \times
+(1-0.05) \times (1-0.10)$.  Figure \ref{fig:p1} shows one run of
+delta-debugging on a test of length 500.  The initial test is
+esentially guaranteed to behave nondeterministically; delta-debugging
+only checks that it \emph{can} behave nondeterministically, and so, in
+the course of reducing the test length to a test with only two
+operations, the delta-debugging algorithm also reduces the probability
+of nondeterminism to only about 20\%.  If we use a predicate that
+``forces'' the test to behave nondeterministically at least half the
+time, by sampling the predicate value 100 times and only returning
+true when at least 50 of the evaluations report true, we see the
+behavior in Figure \ref{fig:p2}:  the final test is slightly longer,
+but still falls well short of our target of exhibiting nondeterminism
+50\% of the time.  Finally, Figure \ref{fig:p3} shows what happens if
+we use the same target of 50\% nondeterminism, but use only 10
+samples, with 10 replications ($N$ = 10, $M$ = 10):  the test is not
+much longer than in Figure \ref{fig:p2}, but the probability of
+nondeterminism is above our target value, close to 60\%.

### 2018-04-18 conclusions.tex
ok

\section{Conclusions and Future Work}

Unexpected nondeterminism of software systems frustrates users,
whether they be humans or (more importantly) other software systems.
Nondeterminism is even more pernicious in software testing,
frustrating debugging efforts (``Heisenbugs'' are widely loathed \cite{Heisenbug}), and
leading to the costly problem of flaky tests
\cite{miccoflaky,listfieldtestanalysis}.

This paper proposes a formulation of types of nondeterminism, and a
practical approach to using automated test generation to detect
nondeterminism, especially in the library code that underlies most
systems. In addition to traditional \emph{horizontal} nondeterminism
(the phenonmenon behind Heisenbugs and flaky tests), we also discuss
the related concept of \emph{vertical} nondeterminism, which is more
frequently simply a software bug.  We implemented our approach in the
TSTL automated test generation system for Python, and demonstated the
simplicity and basic utility of the approach on three real-world examples.

As future work, we would like to make the automatic detection of
values that should not be visible (e.g., count towards nondeterminism)
possible: if a value (e.g., a timestamp) is always different in every
run, it is likely of no interest to developers.  This would make
testing libraries mixing deterministic behavior and nondeterministic
behavior, e.g. cryptographic libraries offering high-quality random
number generators, easier for users, who would only be shown
``surprising'' nondeterminism.  Similarly, in order to extend the
utility of vertical nondeterminism detection, we would like to
automatically identify usually-idempotent operations, and check them
for deviation from the expected behavior.  We would also like to lower the cost
of checking process nondeterminism, perhaps by using methods taken in
afl \cite{aflfuzz} to avoid high startup costs for test executions.

### 2018-04-18 implementation.tex
ok

@@
 means that a test that, in some circumstances, causes a crash, may at
-other times not fail at all.
+other times not fail at all.  Even if no inputs produce a crash,
+however, memory errors may produce variation in values that are
+overwritten or arise from uninitialized memory, and thus be detectable
+as nondeterminism.
 
-Another example of process-based nondeterminism was discovered by numerous Python programs when Python
-version 3.3 introduced automatic random salting of hashes (see Section
-\ref{sec:pnondet} per-process,
+Another example of process-based nondeterminism was discovered by numerous Python developers when Python
+version 3.3 introduced automatic random salting of hashes on a
+per-process basis (a new salt chosen for each Python process),
 in order to mitigate hash-based denial of service attacks
@@
   iteration, though actual hashes remained nondeterministic.}.  While
-relatively few programs relied on exact hash values, many implicitly
+relatively few programs rely on exact hash values, many implicitly
 relied on them in that they only functioned correctly with a
-predictable order for dictionary iteration across processess.  Testing
+predictable order for dictionary iteration.  Testing
 Python code for nondeterminism based on this hash seed requires
@@
 parent process' ID.  Process change increases the likelyhood that on a multicore
-system a test will execut on a different CPU than in the first
+system a test will execute on a different CPU than in the first
 execution, which can have many subtle effects, mostly related to
@@
                           values \\
+\hline
+{\tt P} & returns probability that a predicate holds for a test (with
+          optional number of sample to use in estimate) \\
 \hline
@@
 
-We implemented our approach in the TSTL \cite{NFM15} system, a
-language and tool for property-based testing of Python code.
+We implemented our approach in the TSTL \cite{NFM15,ISSTA15,tstlsttt} system, a
+language and tool for property-based testing of Python code. TSTL has
+been used to detect (and usually fix) errors in a number of widely
+used Python libraries, the Python implementation itself, the Solidity
+compiler and a Solidity static analysis tool, and Mac OS \cite{tstl}.
 Because TSTL supports differential testing \cite{tstlsttt}, horizontal
@@
 to test code can easily add nondeterminism to the set of properties
-checked.  First-class vertical nondeterminism checks are currently
+checked.
+
+First-class vertical nondeterminism checks are currently
 limited to failures:  {\tt --checkFailureDeterminism} causes the TSTL

### 2018-04-18 introduction.tex
ok

@@
 face a challenge in reasoning about causality, in that an event
-observed may not even take place the next time the test is run.
+observed may not even take place the next time the test is run; this
+pernicious phenomenon has been dubbed the dreaded ``Heisenbug'' \cite{Heisenbug}.
 Automated test generation systems using test coverage results to drive
@@
 developers, who will usually want to  assume that library code they
-use is deterministic; it is often disastrous for large-scale or highly
+use is deterministic; it is frequently vexing in debugging efforts; it is often disastrous for large-scale highly
 automated testing.

### 2018-04-18 experiments.tex
fix up

@@
 correct) over 50 faults.  However, the testing completely relies on
-the existence of a reference file system implementation.
+the existence of a reference file system implementation.  One purpose
+of failure nondeterminism is to make it somewhat easier to perform
+property-based testing of
+complex APIs like this even without a reference implementation, or in cases
+where (as is common, e.g. in file system development at NASA
+\cite{ICSEDiff,CFV08}) the implementations do not use the same error
+codes, making checking of behavior of failing calls more problematic.
 

### 2018-04-18 implementation.tex
fix up

@@
 We implemented our approach in the TSTL \cite{NFM15,ISSTA15,tstlsttt} system, a
-language and tool for property-based testing of Python code. TSTL has
+language and tool for property-based testing \cite{Hypothesis,ClaessenH00} of Python code. TSTL has
 been used to detect (and usually fix) errors in a number of widely

### 2018-04-18 experiments.tex
more details on inference algs

@@
 cases incorrect) results due to dependence on the order of values in
-an iterator over a Python dictionary (see Section \ref{sec:pnondet}).
+an iterator over a Python dictionary, on Python versions above 3.2, until 3.6 (see Section \ref{sec:pnondet}).
 
 Figure \ref{hashbug} shows TSTL code for generating inputs to the {\tt
-  datarray} algorithms.  Running this harness using TSTL's {\tt
+  datarray} algorithms.
+The {\tt flatten_and_sort} function is needed because we care about actual differences in probability values, not simply the order of list, tuple, or dictionary items.
+Running this harness using TSTL's {\tt
   --checkProcessDeterminism} flag required less than 10 seconds on

### 2018-04-18 experiments.tex
fix backslash

@@
   datarray} algorithms.
-The {\tt flatten_and_sort} function is needed because we care about actual differences in probability values, not simply the order of list, tuple, or dictionary items.
+The {\tt flatten\_and\_sort} function is needed because we care about actual differences in probability values, not simply the order of list, tuple, or dictionary items.
 Running this harness using TSTL's {\tt

### 2018-04-18 concepts.tex
new text, acks

@@
 beneficial.  The behavior is part of a more general (and, to our
-knowledge, not previously investigate) issue:  using delta-debugging
+knowledge, not previously investigated) issue:  using delta-debugging
 to minimize a test with respect to a predicate that only holds with a
@@
 much longer than in Figure \ref{fig:p2}, but the probability of
-nondeterminism is above our target value, close to 60\%.
+nondeterminism is above our target value, close to 60\%.  Note that
+this example is purely monotonic; the pattern of probability changes
+in non-monotonic settings can be even more difficult to predict and
+control, but the principle of forcing a probability by sampling and
+replication still holds.

### 2018-04-18 conclusions.tex
new text, acks

@@
 afl \cite{aflfuzz} to avoid high startup costs for test executions.
+
+More generally, we are interested in using test decomposition
+\cite{Composition} to more easily isolate, understand, and avoid
+nondeterminism is tests, and to mitigate the problem of flaky tests.
+Reliable detection of nondeterminism/flakiness is the first step to
+active solutions to eliminating nondeterministic/flaky test behavior.
+
+{\scriptsize {\bf Acknowledgements:}  The authors would like to thank John Micco,
+Jeff Listfield, and Celal Ziftci, all at Google, for discussions of
+the problem of flaky tests, Andreas Zeller and David R. MacIver for discussion of the
+problem of delta-debugging of nondeterministic properties, and Chris
+Colohan for discussion of sources of process-based nondeterminism.}

### 2018-04-18 implementation.tex
new text, acks

@@
 types of values as \emph{opaque} (not usefully compared for equality),
-used in automatic abstraction-based testing.  Tables \ref{tab:methods}
+used in automatic abstraction-based testing.  In TSTL, a test is a
+sequence of \emph{actions}; parameter values to method and function
+calls of the Software Under Test (SUT) are stored in finite-sized variable \emph{pools}.  A
+TSTL test is (essentially) a sequence of assignments to pool values,
+and method/function calls using pool values (including method calls on
+pool objects).  This formalism is a widely used and efficient way to
+represent unit tests \cite{Pachecho,AndrewsTR}.  In TSTL, \emph{visible
+value determinism} is based on comparing the values of \emph{all pool
+variables} after each test action (pool values other than the one
+assigned to must also be compared, since a common source of
+nondeterminism is when a call results in a change to a value passed as
+a parameter, or even due to a shared reference held by two values).
+TSTL omits comparison only of pools marked as {\tt OPAQUE} in the
+TSTL language test harness (and values that do not support equality
+checking).  \emph{Final State Determinism} simply performs the same
+comparison, but only on the final values of all pool variables after a test has
+finished executing.
+
+ Tables \ref{tab:methods}
 and \ref{tab:options} show, respectively, the primary additions to the

### 2018-04-18 concepts.tex
new draft, text

@@
 much longer than in Figure \ref{fig:p2}, but the probability of
-nondeterminism is above our target value, close to 60\%.  Note that
+nondeterminism is above our target value, close to 60\% (and, as a bonus, fewer
+test executions are required on average for each check of the predicate).  Note that
 this example is purely monotonic; the pattern of probability changes

### 2018-04-18 implementation.tex
new draft, text

@@
 to test code can easily add nondeterminism to the set of properties
-checked.
+checked.  It is even possible to write a TSTL harness that
+automatically checks itself for nondeterminism, even if the random
+tester is not run with {\tt --checkDeterminism}, simply by adding:
+
+{\scriptsize
+\begin{code}
+property: not self.nondeterministic(self.test())
+\end{code}
+}
+
+\noindent to the harness; in fact, by removing the {\tt not}, a test harness can
+even specify that behavior should not be deterministic, a property of
+possible use in some security-related libraries.
 

### 2018-04-18 introduction.tex
drafty

@@
 being able to exactly reproduce system behavior is essential to
-productivity.  If executing ``the same test'' can produce
+productivity \cite{Firesmith}.  If executing ``the same test'' can produce
 significantly different behavior each time it is run, the consequences

### 2018-04-19 concepts.tex
aflfuzz

@@
 modifying the delta-debugging implementations, which is difficult and
-sometimes not really feasible (e.g., few users of AFL-fuzz \cite{aflfuzz} will wish
+sometimes not really feasible (e.g., few users of afl-fuzz \cite{aflfuzz} will wish
 to change its C code for test reduction).  Ideally, the solution

### 2018-04-19 implementation.tex
drafty
other use of afl

@@
 pool objects).  This formalism is a widely used and efficient way to
-represent unit tests \cite{Pachecho,AndrewsTR}.  In TSTL, \emph{visible
+represent unit tests \cite{Pacheco,AndrewsTR}.  In TSTL, \emph{visible
 value determinism} is based on comparing the values of \emph{all pool
@@
 Unlike the other additions, explicitly designed for nondeterminism
-detection, the {\tt forceP} wrapper and {\tt --probability}, {\tt
+detection, the {\tt P} query, {\tt forceP} wrapper, and {\tt --probability}, {\tt
   --samples}, and {\tt --replications} reducer options are general
@@
 executions, these TSTL additions can be used to find a small test where this property
-fails to hold).
+fails to hold).  In addition to the basic probabilistic reduction
+issues discussed above, actual implementation of test reduction for
+nondeterminism introduces some subtle issues.  By default, for
+example, TSTL automatically removed non-enabled actions from tests
+during reduction, because these are never actually executed.  However,
+in timing based nondeterminism, such steps may be essential to
+causing the nondeterminism to appear (we had to turn off this feature
+in reducing the tests for the {\tt redis-py} example discussed below).
 
-Because TSTL has an interface to afl \cite{aflfuzz}, we can even use
+Because TSTL has a first-class interface to afl \cite{aflfuzz}, we can even use
 afl's sophisticated heuristics to perform very thorough, week-long
 checks for nondeterminism, using strategies built to find subtle
-security vulnerabilities in C programs.
+security vulnerabilities in C programs.  Equally importantly, since
+there is substantial overhead in nondeterminism detection, afl can be
+used to predict whether a program has potential nondeterminism; if the
+afl \emph{stability} statistic (see the afl documentation for details
+\cite{aflfuzz}) is lower than 100\%, it may indicate a problem.
+Because of the idiosyncracies of afl instrumentation and process
+behavior, this is not always a reliable guide, but it is a very low
+cost indicator produced as a by-product of fuzzing.

### 2018-04-19 experiments.tex
wording

@@
 high-coverage regression tests for {\tt redis-py} and Redis using TSTL
-was difficult, as numerous Redis commands introduce nondeterministic
-behavior.  Some of these are obvious (e.g., {\tt randomkey, srandmember}) by
-name.  Given an understanding of Redis semantics, that the various
+has proven difficult, as numerous Redis commands introduce nondeterministic
+behavior.   Using {\tt tstl\_afl\_fuzz}, we can see that the {\tt redis-py} has a 
+stability of only 56.26\%, a clear indicator of nondeterminism. Some
+of the problematic commands are obvious simply by inspection (e.g.,
+{\tt randomkey, srandmember}).  And, given an understanding of Redis
+semantics, it is also clear that the various
 commands producing data with a limited lifetime (e.g., {\tt expire,
-  pexpire}) also introduce timing-based nondeterminism is also clear.
+  pexpire}) also introduce timing-based nondeterminism.
+
 However, that a command such as {\tt restore} takes an expiration
-argument is not obvious on inspection, nor is the behavior of {\tt
+argument is not obvious to a non-expert, nor is the behavior of {\tt
   spop} which pops a random value from a set.  Moreover, it is

### 2018-04-19 experiments.tex
code

@@
 \subsection {Case Study: {\tt redis-py}}
+
+\begin{figure*}
+\centering 
+\begin{subfigure}{0.67\columnwidth}
+\centering
+\includegraphics[width=\columnwidth]{redisddmin}
+\caption{Standard delta-debugging}
+\label{fig:r1}
+\end{subfigure}
+\begin{subfigure}{0.67\columnwidth}
+\centering
+\includegraphics[width=\columnwidth]{redisddmin}
+\caption{P=0.5, 100 samples}
+\label{fig:r2}
+\end{subfigure}
+\begin{subfigure}{0.67\columnwidth}
+\centering
+\includegraphics[width=\columnwidth]{redisddmin}
+\caption{P=0.5, 10 samples, 10 replications}
+\label{fig:r3}
+\end{subfigure}
+\caption{Delta-debugging of the same test, with different ways of
+  handling nondeterministic in the reduction predicate.}
+\end{figure*}
 
@@
 
-Using the original TSTL harness, approxoimately \emph{over 20\%} of all {\tt redis-py}
+Using the original TSTL harness, approximately \emph{over 20\%} of all {\tt redis-py}
 regression tests (of length 200) generated were flaky.  Using the
 nondeterminism checker to reduce flaky tests to a minimal set allows
-an engineer to leave an overnight run that will, in 12 hours, produce
+us to simply set up an overnight run that will, in 12 hours, produce
 a set of minimal flaky tests exposing the sources of
@@
 
+Figures \ref{fig:r1}-\ref{fig:r3} show delta-debugging of a typical
+{\tt redis-py} nondeterministic test, originally with 287 steps.  The initial test behave
+nondeterministically about half the time.  Figure \ref{fig:r1} shows
+that this is a non-monotonic reduction problem, where removing steps
+can either increase or decrease the probability of nondeterminism.  At
+first, unmodified delta-debugging actually improves the probability of
+failure, but eventually it produces a test of length 16 that only behaves
+nondeterministically 24\% of the time.  The reduction takes only 38
+seconds.  For the purpose of identifying commands leading to
+nondeterminism, this is acceptable.  However, if we were actually
+debugging a complex nondeterminism bug in Redis itself, we might want
+a more reliably nondeterministic test.
+
 After removing all identified sources of nondeterminism (the {\tt random-} functions, all
@@
 lifetimes, and {\tt spop}), but leaving all other suspect funcitonality in
-place (e.g. pipes), no flakiness was observed in a sample of over 1,000 length
+place (e.g. pipes), no flakiness was observed in a sample of over 2,000 length
 200 tests.  Moreover, because the removals were limited to acutally

### 2018-04-19 experiments.tex
now too long, needs trim

@@
 \section{Experimental Evaluation}
+
+In order to evaluate our approach and implementation, we applied
+nondeterminism detection to three real world examples.  The primary
+points we wanted to explore were (1) whether lightweight, automatic,
+nondeterminism detection was able to reliably detect actual
+nondeterminism (2) whether the overhead of nondeterminism
+detection for real systems was acceptable and (3) the performance of
+delta-debugging for real nondeterministic predicates.
 

### 2018-04-19 experiments.tex
fixups

@@
 <P> := 0.01 * <[0..100]>
-\vspace{0.1in}
 <ename> := "E" + str(<[1..5]>)
-\vspace{0.1in}
 \{Exception\} <event> := [datarray.DataArray(psplit(<P>), axes = [<ename>])]
 \{Exception\} <ename,1>!=<ename,2> -> <event> := [datarray.DataArray([[psplit(<P>)],psplit(<P>)],[<ename>,<ename>])]
-\vspace{0.1in}
 <cpts> := []
 ~<cpts>.append(<event>[0])
-\vspace{0.1in}
 <evidence> := \{\}
@@
 \cite{ICSEDiff,CFV08}) the implementations do not use the same error
-codes, making checking of behavior of failing calls more problematic.
+codes.
 

### 2018-04-19 implementation.tex
fixups

@@
 \centering
-{%\scriptsize
+{\scriptsize
 
@@
 \hline
-{\tt forceP} & wraps a predicate for test reduction (e.g. {\tt
-         nondeterministic}) with desired probability of holding,
-         sample size,\\
-&  and replications required \\
+{\tt forceP} & wraps a predicate for test reduction, checking that it
+               has desired probability of holding, using
+         sample size and replications parameters \\
 \hline
@@
 \centering
-{%\scriptsize
+{\scriptsize
 \caption{TSTL Command Line Options for

### 2018-04-19 implementation.tex
fix layout

@@
 causing the nondeterminism to appear (we had to turn off this feature
-in reducing the tests for the {\tt redis-py} example discussed below).
+for the {\tt redis-py} example discussed below).
 

### 2018-04-19 experiments.tex
afl stability
change wording

@@
 showing an extremely simple input producing the issue, 
-required another 92 seconds.
+required another 92 seconds.  Interestingly, the way that {\tt
+  python-afl} is used in TSTL means that afl stability is 100\%,
+because the {\tt PYTHONHASHSEED} has already been chosen before the
+fork in afl executions.
 
@@
 using a new subprocess to re-run the test, and comparing values using
-text I/O on output pipes, rather than on values alone.
+text I/O on output pipes, rather than values in a Python dictionary.
 

### 2018-04-20 abstract.tex
ok

@@
 A critically important, but surprisingly neglected in the software testing literature, aspect of system reliability is system predictability.
-Many software systems are implemented using mechanisms (unsafe languages, mulitple threads, complex caching, stochastic algorithms, environmental dependencies) that can introduce behavioral nondeterminism.
+Many software systems are implemented using mechanisms (unsafe languages, multiple threads, complex caching, stochastic algorithms, environmental dependencies) that can introduce behavioral nondeterminism.
 Users of software systems, especially other software systems using library calls in a single-threaded context, often expect that systems will behave deterministically in the sense that they have predictable results from the same sequential series of operations.
@@
 We show that lightweight techniques, requiring (almost) no effort on the part of a developer, can be used to extend an existing test generation system to allow checking for problematic nondeterminism.
-We introduce a set of lightweight nondeterinism properties, inspired by real faults, and tailor these notions to the practical, automatic, checking of Python library code.
+We introduce a set of lightweight nondeterminism properties, inspired by real faults, and tailor these notions to the practical, automatic, checking of Python library code.

### 2018-04-20 concepts.tex
code
fixups
ok

@@
 
-From a user's perspective, some behaviors of the macOS High Sierra
+From a user's perspective, some behaviors of the Mac OS High Sierra
 root exploit (CVE-2017-13872 \cite{applebug0}) exhibited failure
@@
 \subsubsection{``Publication Bias'' in Delta-Debugging}
+\label{sec:pubbias}
 
@@
 \includegraphics[width=\columnwidth]{lengthddminforceprep}
-\caption{10 samples, 10 replications}
+\caption{P=0.5, 10 samples, 10 replications}
 \label{fig:p3}
 \end{subfigure}
-\caption{Delta-debugging of the same test, with different ways of
-  handling nondeterministic in the reduction predicate.}
+\caption{Delta-debugging of the same {\tt redis-py} test}
 \end{figure*}
@@
 bias.  One way to produce this effect would be to allow
-delta-debugging to backtrack if the probabilties observed in predicate
-evaluations suddely exhibit a strong discontinuity, a kind of ``paper
+delta-debugging to backtrack if the probabilities observed in predicate
+evaluations suddenly exhibit a strong discontinuity, a kind of ``paper
 retraction'' based on near-replications.  However, this requires
@@
 to change its C code for test reduction).  Ideally, the solution
-should be implementable simply by modifying the predictate that is
+should be implementable simply by modifying the predicate that is
 evaluated.
@@
   usually not very useful, since the true
-  probabilitiy distributions of test behaviors vary widely and
+  probability distributions of test behaviors vary widely and
   dynamically during the
@@
 
+{\scriptsize
 \begin{code}
@@
 \end{code}
+}
 
@@
 delta-debugging on a test of length 500.  The initial test is
-esentially guaranteed to behave nondeterministically; delta-debugging
+essentially guaranteed to behave nondeterministically; delta-debugging
 only checks that it \emph{can} behave nondeterministically, and so, in

### 2018-04-20 conclusions.tex
ok

--- 2018-04-18 conclusions.tex
@@
 systems. In addition to traditional \emph{horizontal} nondeterminism
-(the phenonmenon behind Heisenbugs and flaky tests), we also discuss
+(the phenomenon behind Heisenbugs and flaky tests), we also discuss
 the related concept of \emph{vertical} nondeterminism, which is more
 frequently simply a software bug.  We implemented our approach in the
-TSTL automated test generation system for Python, and demonstated the
+TSTL automated test generation system for Python, and demonstrated the
 simplicity and basic utility of the approach on three real-world examples.
@@
 afl \cite{aflfuzz} to avoid high startup costs for test executions.
+
+More generally, we are interested in using test decomposition
+\cite{Composition} to more easily isolate, understand, and avoid
+nondeterminism is tests, and to mitigate the problem of flaky tests.
+Reliable detection of nondeterminism is a first step towards
+evaluating such pro-active flaky test mitigations.
+
+{\scriptsize {\bf Acknowledgments:}  The authors would like to thank John Micco,
+Jeff Listfield, and Celal Ziftci at Google, for discussion of
+flaky tests, Andreas Zeller and David R. MacIver for discussion of the
+problem of probabilistic delta-debugging, and Chris
+Colohan for discussion of sources of process-based nondeterminism.}

### 2018-04-20 experiments.tex
graphs
ok

@@
 nondeterminism detection to three real world examples.  The primary
-points we wanted to explore were (1) whether lightweight, automatic,
-nondeterminism detection was able to reliably detect actual
-nondeterminism (2) whether the overhead of nondeterminism
-detection for real systems was acceptable and (3) the performance of
+points we wanted to explore were (1) whether our approach was able to reliably detect actual
+nondeterminism, (2) whether the overhead of nondeterminism
+detection for real systems was acceptable, and (3) the performance of
 delta-debugging for real nondeterministic predicates.
@@
 \includegraphics[width=\columnwidth]{redisddmin}
-\caption{Standard delta-debugging}
+\caption{Standard delta-debugging (38s)}
 \label{fig:r1}
@@
 \centering
-\includegraphics[width=\columnwidth]{redisddmin}
-\caption{P=0.5, 100 samples}
+\includegraphics[width=\columnwidth]{redisforcep}
+\caption{P=0.5, 100 samples (2131s)}
 \label{fig:r2}
@@
 \centering
-\includegraphics[width=\columnwidth]{redisddmin}
-\caption{P=0.5, 10 samples, 10 replications}
+\includegraphics[width=\columnwidth]{redisforceprep}
+\caption{P=0.5, 10 samples, 10 replications (617s)}
 \label{fig:r3}
 \end{subfigure}
-\caption{Delta-debugging of the same test, with different ways of
-  handling nondeterministic in the reduction predicate.}
+\caption{Delta-debugging of the same {\tt redis-py} nondeterministic test}
 \end{figure*}
@@
 Using TSTL's harness for {\tt redis-py}, both {\tt redis-py} and Redis
-itself can be tested; unfortunately, generating standalone
+itself can be tested; unfortunately, generating stand-alone
 high-coverage regression tests for {\tt redis-py} and Redis using TSTL
@@
   spop} which pops a random value from a set.  Moreover, it is
-difficult to guess whetehr the {\tt pipe} mechanism, which allows a
+difficult to guess whether the {\tt pipe} mechanism, which allows a
 large sequence of commands to be queued up in a pipe and executed all
@@
 issues clear, but tests using a library are often written by those not
-intimately familliar with the semantics of every used library (otherwise they would not have
+intimately familiar with the semantics of every used library (otherwise they would not have
 been as likely to introduce flaky tests in the first place).  This is
@@
 debugging a complex nondeterminism bug in Redis itself, we might want
-a more reliably nondeterministic test.
+a more reliably nondeterministic test.  Using the same parameters as
+in Section \ref{sec:pubbias}, we see the same pattern.  Simply making
+a predicate that ``forces'' the probability to remain high, with a
+large number of samples, does not work (Figure \ref{fig:r3}) and
+requires over 2000 seconds to produce a test with an even worse
+probability of nondeterminism.  Using 10 samples and 10 replications,
+on the other hand, actually improves the probability of
+nondeterministic behavior, and
+(in this case) even produces a smaller test (only 14 steps) in just
+over 10 minutes.
 
@@
 expiration-related functionality and {\tt restore} calls with non-zero
-lifetimes, and {\tt spop}), but leaving all other suspect funcitonality in
+lifetimes, and {\tt spop}), but leaving all other suspect functionality in
 place (e.g. pipes), no flakiness was observed in a sample of over 2,000 length
-200 tests.  Moreover, because the removals were limited to acutally
+200 tests.  Moreover, because the removals were limited to actually
 observed sources of flakiness, the code coverage loss was minimal.
@@
 fast-executing high-coverage full regression tests (not just tests for
-crashes and unexpected exceptions), this is a very acceptable price.
+crashes and unexpected exceptions), this seems acceptable.
 
@@
 checking for process nondeterminism, with no delay between operations,
-is high, a 93\% slowdown.  This is due to the very high cost of
-using a new subprocess to re-run the test, and comparing values using
-text I/O on output pipes, rather than values in a Python dictionary.
+is high, a 93\% slowdown.  This is due to the high cost of subprocess
+creation and communication.
 
@@
 complex APIs like this even without a reference implementation, or in cases
-where (as is common, e.g. in file system development at NASA
-\cite{ICSEDiff,CFV08}) the implementations do not use the same error
-codes.
+where the implementations do not use the same error
+codes (as is common, e.g. in NASA flight software
+\cite{ICSEDiff,CFV08}).
 
@@
 slightly longer failing test, which did not have {\tt remove} as its
-final operation (since further file system interaction is required to
-expose the bad state).
+final operation (since further operations are required to 
+expose the problem).

### 2018-04-20 implementation.tex
ok

@@
 may somehow depend on their actual PID (Process ID) or at least the
-parent process' ID.  Process change increases the likelyhood that on a multicore
+parent process' ID.  Process change increases the likelihood that on a multi-core
 system a test will execute on a different CPU than in the first
@@
 tools for use with nondeterministic predicates in delta-debugging.  Of
-course, nondeterminism and flakyness are the most common probabilistic
+course, nondeterminism and flakiness are the most common probabilistic
 test properties we are aware of, but in theory these tools can be used
@@
 \cite{aflfuzz}) is lower than 100\%, it may indicate a problem.
-Because of the idiosyncracies of afl instrumentation and process
+Because of the idiosyncrasies of afl instrumentation and process
 behavior, this is not always a reliable guide, but it is a very low

### 2018-04-20 concepts.tex
ok

@@
 overhead for vertical determinism, in that we can expect most failing
-operations to be fast in an automated test generation setting (that the
-parameters to a call are illegal is the most common cause for
-failure).
+operations to be fast (since in test generation, failure is usually
+due to invalid parameters).
 
@@
 replication still holds.
+
+Finally, we note that this problem is not limited to debugging
+nondeterminism/flakiness itself.  In some settings, we have observed
+delta-debugging taking a test that had such a high probability of
+failure we were unaware it was ``flaky'' (large runs showed there was
+indeed a small chance of the test passing) and transforming it into a
+test that only failed once in every 10 to 20 executions.
+Delta-debugging, without taking potential nondeterminism into account,
+can act as a flakiness magnifier, with unfortunate consequences for
+debugging.  Multiple evaluations, with replications, are our proposed solution.

### 2018-04-20 experiments.tex
ok

@@
 has proven difficult, as numerous Redis commands introduce nondeterministic
-behavior.   Using {\tt tstl\_afl\_fuzz}, we can see that the {\tt redis-py} has a 
+behavior:  thus the resulting tests are very often \emph{flaky}.   Using {\tt tstl\_afl\_fuzz}, we can see that the {\tt redis-py} has a 
 stability of only 56.26\%, a clear indicator of nondeterminism. Some

### 2018-04-20 implementation.tex
ok

@@
 exact hash values, but in the order of iteration on
-dictionaries\footnote{In version 3.6, the default dictionary
+dictionaries\footnote{In 3.6, the default dictionary
   implementation became an ordered dictionary with consistent
@@
 {\tt nondeterministic} & returns True if test exhibits final state nondeterminism in $k$
-  tries with delay $d$\\
+  tries with delay $d$, optionally only over pools in $\mathcal{P}$\\
 \hline
@@
 declaring a system to be its own reference, using TSTL's notation for
-differential testing.  However, such an approach requires some effort
-on the part of the user to control which values are checked for
-equivalence, and does not easily support injecting timing differences,
-or using a new process, in checking the behavior.
+differential testing.  However, such an approach requires considerable effort
+on the part of the user to express which values are checked for
+equivalence, and does not (without a great deal of effort) support injecting timing differences,
+or re-executing in a new process, in checking for nondeterminism.
 
@@
 checking).  \emph{Final State Determinism} simply performs the same
-comparison, but only on the final values of all pool variables after a test has
+comparison, but only on the final values of all pool variables (or a
+set of designated pools) after a test has
 finished executing.

### 2018-04-20 introduction.tex
ok

@@
 simply say that a system is deterministic if, given the same state and
-inputs, it always produces the same outputs\footnote{Of course, this
-  shift of definition is often adopted in arguments for physical or
-  human behavioral determinism, due to the even greater difficulties
-  of prediction in those arenas.}.  The change of
-definition, we will see, is central to our approach, and comes with a
-number of limitations.
+inputs, it always produces the same outputs.
 
@@
 an iterator on a hash table is important, and the hash values used are
-salted and thus not predictable.
+randomly salted.
 
@@
 run multiple times, wasting valuable computing resources and also
-delaying acceptance of code changes.  A canary in a coal mine is of
+delaying acceptance of code changes.  An analogy may make the extent
+of the problem in large-scale test automation clear:  a canary in a coal mine is of
 little use if canaries frequently become ill for reasons unrelated to
@@
 This paper proposes (1) a number of formal definitions of types of
-nondeterminism and (2) a number of techniques, based on these definitions, for detecting and debugging
-nondeterminism in property-based testing, particularly for library
-code.  The methods also work for testing at the system level, but are
-perhaps most effective for identifying the low-level sources of
-nondeterminism in libraries.  This enables a number of useful results:
+nondeterminism and (2) an implementation, based on these definitions, for detecting and debugging
+nondeterminism in property-based testing.  This enables a number of useful results:
  the automated production of
 nondeterminism-free regression tests even for libraries with some
-sources of nondeterminism, the documentation of library
-nondeterminism, and, most importantly, detection and effective debugging of
-unexpected nondeterminism that is, itself, a bug.  We first
-describe the types of nondeterminism we aim to detect, and the
-extension of a widely used debugging algorithm to better handle
-nondeterminism as a test property, then describe an implementation of
-these ideas in a Python testing tool that is used on real-world
+sources of nondeterminism, the documentation of
+nondeterminism, and, most importantly, detection and effective debugging for
+unexpected nondeterminism that is, itself, a bug.  We propose
+modifications to the widely used delta-debugging algorithm in order to better handle
+nondeterminism as a test property, and describe an implementation of
+our approach in a Python testing tool that is used on real-world
 projects.  Finally, we use a set
-of case studies of real-world Python libraries to demonstrate the
-utility of our ideas and understand the cost of adding nondeterminism
-detection to existing automated testing.
+of case studies of real, widely-used Python projects to demonstrate the
+utility of our ideas.

### 2018-04-20 concepts.tex
wording

@@
 delta-debugging taking a test that had such a high probability of
-failure we were unaware it was ``flaky'' (large runs showed there was
+failure we were unaware it was ``flaky'' (only hundreds of repeated executions showed there was
 indeed a small chance of the test passing) and transforming it into a

### 2018-04-26 concepts.tex
alex edit

@@
 common.  In these cases, when we use a reduction algorithm to reduce $t$
-to $r$, $P(p(r) \leq P(p(t))$.  In \emph{non-monotonic} cases,
+to $r$, $P(pred(r)) \leq P(pred(t))$.  In \emph{non-monotonic} cases,
 however, there is no such upper bound.  Removing a step in $t$ may
@@
 
-The probabilistic predicate of interest for determinism is always of
+The probabilistic predicate of interest for determinism is basically of
 the form:  {\bf the test will exhibit at least two different behaviors in $S$

### 2018-04-29 conclusions.tex
AFL capitalization change

@@
 of checking process nondeterminism, perhaps by using methods taken in
-afl \cite{aflfuzz} to avoid high startup costs for test executions.
+AFL \cite{aflfuzz} to avoid high startup costs for test executions.
 

### 2018-04-29 experiments.tex
AFL capitalization change

@@
 required another 92 seconds.  Interestingly, the way that {\tt
-  python-afl} is used in TSTL means that afl stability is 100\%,
+  python-afl} is used in TSTL means that AFL stability is 100\%,
 because the {\tt PYTHONHASHSEED} has already been chosen before the
-fork in afl executions.
+fork in AFL executions.
 

### 2018-04-29 implementation.tex
AFL capitalization change

@@
 
-Because TSTL has a first-class interface to afl \cite{aflfuzz}, we can even use
-afl's sophisticated heuristics to perform very thorough, week-long
+Because TSTL has a first-class interface to AFL \cite{aflfuzz}, we can even use
+AFL's sophisticated heuristics to perform very thorough, week-long
 checks for nondeterminism, using strategies built to find subtle
 security vulnerabilities in C programs.  Equally importantly, since
-there is substantial overhead in nondeterminism detection, afl can be
+there is substantial overhead in nondeterminism detection, AFL can be
 used to predict whether a program has potential nondeterminism; if the
-afl \emph{stability} statistic (see the afl documentation for details
+AFL \emph{stability} statistic (see the AFL documentation for details
 \cite{aflfuzz}) is lower than 100\%, it may indicate a problem.
-Because of the idiosyncrasies of afl instrumentation and process
+Because of the idiosyncrasies of AFL instrumentation and process
 behavior, this is not always a reliable guide, but it is a very low

### 2018-05-01 experiments.tex
fix example

@@
 <ename> := "E" + str(<[1..5]>)
-\{Exception\} <event> := [datarray.DataArray(psplit(<P>), axes = [<ename>])]
+\{Exception\} <event> := datarray.DataArray(psplit(<P>), axes = [<ename>])
 \{Exception\} <ename,1>!=<ename,2> -> <event> := [datarray.DataArray([[psplit(<P>)],psplit(<P>)],[<ename>,<ename>])]
 <cpts> := []
-~<cpts>.append(<event>[0])
+~<cpts>.append(<event>)
 <evidence> := \{\}

### 2018-07-18 experiments.tex
mutation experiment start

@@
 expose the problem).
+
+\subsubsection{Mutation analysis}
+
+We ran 2,350 mutants of pyfakefs.  Using the failure determinism check
+in 60 second test runs let us kill 98 mutants that were not killed
+during 60 seconds of testing without a reference or the failure
+determinism check, using the same seed.  Of these, 71 could not even be killed by 120
+seconds of testing using complete differential testing with Mac OS X
+as a reference.  In contrast, the kill rate for the entire set of
+mutants not killed without a reference, using 120 seconds of
+differential testing was...

### 2018-07-18 experiments.tex
more on mutants

@@
 seconds of testing using complete differential testing with Mac OS X
-as a reference.  In contrast, the kill rate for the entire set of
-mutants not killed without a reference, using 120 seconds of
+as a reference, a kill rate of only 27.6\%.  In contrast, the kill rate for the entire set of 1478
+mutants not killed when not using a reference, using 120 seconds of
 differential testing was...

### 2018-07-22 experiments.tex
ok

--- 2018-05-01 experiments.tex
@@
 expose the problem).
+
+\subsubsection{Mutation analysis}
+
+We ran 2,350 mutants of pyfakefs.  Using the failure determinism check
+in 60 second test runs let us kill 98 mutants that were not killed
+during 60 seconds of testing without a reference or the failure
+determinism check, using the exact same random seed (and thus same
+sequence of test operations, only changing the specification).  Of
+these additional mutants, 71 could not even be killed by 120
+seconds of testing using differential testing with the Mac OS X file system
+as a reference.  That is, more than 2/3 of the mutants killed by the
+failure determinism check could not be killed even with a much
+stronger specification that lacked a check for failure determinism,
+and was given more time to execute.  To put this in perspective, the
+mutation kill rate for the test harness with 60 seconds and no
+differential comparison was 37.1\%; for the full differential harness
+with 120 seconds it was 60\%.
+
+There is a price to be paid:  for mutants that could be killed both
+with and without the failure determinism check, the mean time to kill
+the mutant was 0.006 seconds larger when the failure determinism
+check was included, and this change was significant by Wilcoxon test
+at $p < 1 \times 10^{-15}$.
+
+In order to further investigate the additional specification power
+provided by failure determinism checks, we inspected the mutants
+killed by the failure determinism check but not by full differential
+testing.  The largest category of mutants not killed (22 of the 71 mutants) was what we refer
+to as ``exception swallowing'' mutants, which transform a Python
+statement into the same statement, but wrapped in a {\tt try} block
+with a catch that ignores any raised exceptions.  It is easy to
+understand that such mutants introduce faults in the handling of
+errors, and thus would tend to cause failurel nondeterminism.

### 2018-07-23 conclusions.tex
ok

@@
 Nondeterminism is even more pernicious in software testing,
-frustrating debugging efforts (``Heisenbugs'' are widely loathed \cite{Heisenbug}), and
+frustrating debugging efforts (Heisenbugs \cite{Heisenbug} and
+Mandelbugs \cite{GrottkeBugs,FaultTriggers} are widely loathed), and
 leading to the costly problem of flaky tests
@@
 systems. In addition to traditional \emph{horizontal} nondeterminism
-(the phenomenon behind Heisenbugs and flaky tests), we also discuss
+(the phenomenon behind Mandelbugs and flaky tests), we also discuss
 the related concept of \emph{vertical} nondeterminism, which is more

### 2018-07-23 experiments.tex
ok

--- 2018-07-18 experiments.tex
@@
 The TSTL harness for {\tt pyfakefs} has been used to detect (and
-correct) over 50 faults.  However, the testing completely relies on
+correct) over 80 faults.  However, the testing largely relies on
 the existence of a reference file system implementation.  One purpose
@@
 \cite{ICSEDiff,CFV08}).
+
+\subsubsection{Manually Inserted Fault}
 
@@
 less than 8\%.  Detecting the fault using a reference file system 
-required 17\% more runtime before detection, and took over twice as
+required 17\% more testing time before detection, and took over twice as
 long to reduce the failure to a
@@
 final operation (since further operations are required to 
-expose the problem).
+expose the bad file system state the operation introduces).
 
-\subsubsection{Mutation analysis}
+\subsubsection{Mutation Analysis}
 
-We ran 2,350 mutants of pyfakefs.  Using the failure determinism check
-in 60 second test runs let us kill 98 mutants that were not killed
-during 60 seconds of testing without a reference or the failure
-determinism check, using the same seed.  Of these, 71 could not even be killed by 120
-seconds of testing using complete differential testing with Mac OS X
-as a reference.  In contrast, the kill rate for the entire set of
-mutants not killed without a reference, using 120 seconds of
-differential testing was...
+We wanted to determine whether there were also simple faults that
+could be detected by failure nondeterminism, but \emph{not} detected
+by differential testing, and quantify the additional specification
+strength provided by failure determinism checks.  We therefore used
+{\tt universalmutator} \cite{RegExpMut} to produce 2,350 mutants of
+the {\tt pyfakefs} core file system code.   We restricted the generation
+to only mutate code covered during a 60 second run of the test
+harness, with differential testing turned off.  In this
+\emph{non-differential} mode, the harness will only detect a fault when it
+causes an unexpected exception to be raised, or results in a timeout
+due to, e.g., and infinite loop.
+
+We first analyzed the mutants using 60 seconds of non-differential
+testing, then with the full differential test plus
+an additional 60 seconds of testing (to account for the observed 17\%
+extra time to detect in the manually constructed example).   For both analyses, we tested
+the surviving mutants using the non-differential harness for 60
+seconds, but with failure nondeterminism checking.
+We used the same random seed for
+all runs, so that the only differences would be specification-based.
+
+Non-differential testing, without failure determinism checks, killed
+872 mutants, a 37.1\% kill rate.  Adding a failure determinism check
+allowed the testing to kill an additional 98 mutants.
+The differential harness, with an additional 60 seconds of testing
+time, improved the kill rate to 49.4\%.  Failure determinism added an
+additional 71 mutants to the total killed, even for this complete
+harness with a larger test budget.  The kill rates are generally low
+because {\tt universalmutator}, by design, includes many operations
+that can produce equivalent mutants, but can also produce hard-to-kill
+non-equivalent mutants not produced by other mutation tools, e.g. code
+that throws away exceptions raised by a function call, or code
+reversing a list.
+
+In order to further investigate the additional specification power
+provided by failure nondeterminism detection, we inspected the mutants
+killed by the failure determinism check but not by differential
+testing.  The largest category of mutants not killed (22 of the 71 mutants) was what we refer
+to as ``exception swallowing'' mutants, which transform a Python
+statement into the same statement, but wrapped in a {\tt try} block
+with a catch that ignores any raised exceptions, e.g., {\tt foo()} becomes:
+
+\begin{code}
+try: foo()
+except: pass
+\end{code}
+
+\noindent It is easy to
+see that such mutants may introduce faults in the handling of
+errors, and thus would tend to cause failure nondeterminism.  However,
+these are a minority of the mutations introducing failure
+nondeterminism.  Other mutation operators resulting in subtle flaws
+not (at least easily) detectable by differential testing compared to a
+correct file system include:  arithmetic operation changes, statement
+deletions, logical operator modifications, constant replacements
+(including replacement of a string with the empty string), and
+introducing a break into a loop.
+
+We also checked the cost of introducing failure determinism checking
+by analyzing mutants that could be killed by both methods.  Even
+though in some cases the failure determinism check allows a mutant to
+be detected sooner, the mean time to kill mutants was 0.006 seconds
+larger with failure determinism checking, and this change was significant by Wilcoxon test
+$(p < 1 \times 10^{-15})$.

### 2018-07-23 introduction.tex
ok

@@
 observed may not even take place the next time the test is run; this
-pernicious phenomenon has been dubbed the dreaded ``Heisenbug'' \cite{Heisenbug}.
+pernicious phenomenon has been sometimes referred to as the dreaded
+``Heisenbug'' \cite{Heisenbug}, though the more accurate usage is
+``Mandelbug'' \cite{GrottkeBugs,FaultTriggers}, since Heisenbugs, properly speaking, are bugs that disappear or
+alter behavior under instrumentation (especially in debugging).
 Automated test generation systems using test coverage results to drive

### 2018-07-23 relatedwork.tex
ok

@@
 \section{Related Work}
+
+At a very broad level, the topic of uncertainty (and thus nondeterminism) in software engineering has been addressed by Garlan \cite{GarlanUncertain}, Elbaum and Rosenblum \cite{Unknowns,lu2015roundtable}, and Ziv and Richardson \cite{UncertaintyPrinciple}.
+There is a general agreement that as systems become more complex, more distributed, and more \emph{statistical}, these problems will only grow \cite{lu2015roundtable}.
 
@@
 From our point of view, flaky behavior is simply a special case of nondeterminism, where the nondeterminism is sufficient to cause the test to have different pass/fail results at times.
+Our approach is, in a sense (especially the delta-debugging modifications) in line with Harman and O'Hearn's proposal to simply accept that ``All Tests Are Flaky'' \cite{StartupstoScaleups}, and work with probabilistically failing tests.
 
 Other efforts \cite{ParallelDeterministic} have aimed to avoid nondeterminism in parallel implementations, by design, indicating the importance of avoiding nondeterministic behavior, even at the expense of adopting novel programming models, where the goal is not simply (as in, say, Rust) to avoid classic concurrency errors, but to enforce genuinely deterministic behavior.
+
+Finally, Cotroneo et al.
+\cite{CompBugs,FaultTriggers}, and Grottke and Trivedi \cite{GrottkeBugs} have followed on early work on understanding bugs and how they manifest, including transient \cite{Transient} software faults.
+This work informs our attempt to identify sources of nondeterminism, and should provide other, context and project-specific, sources that could be introduced into our general framework.

### 2018-07-24 experiments.tex
final mutation kill numbers

@@
 872 mutants, a 37.1\% kill rate.  Adding a failure determinism check
-allowed the testing to kill an additional 98 mutants.
+allowed the testing to kill an additional 98 mutants, an 11\% improvement.
 The differential harness, with an additional 60 seconds of testing
-time, improved the kill rate to 49.4\%.  Failure determinism added an
-additional 71 mutants to the total killed, even for this complete
+time, killed 1148 mutants, improving the kill rate to 48.8\%.  Failure determinism added an
+additional 71 mutants to the total killed, a 6\% improvement even for this complete
 harness with a larger test budget.  The kill rates are generally low

### 2018-07-24 concepts.tex
afl info

@@
 large, this would not matter.  However, by their nature,
-delta-debugging and other test reduction algorithms, explore a search
+delta-debugging and other test reduction algorithms explore a search
 space that often contains hundreds or even thousands or millions of

### 2018-07-24 experiments.tex
afl info

@@
 
-Using the original TSTL harness, approximately \emph{over 20\%} of all {\tt redis-py}
+Using the original TSTL harness, just over 20\% of all {\tt redis-py}
 regression tests (of length 200) generated were flaky.  Using the
@@
 relatively inexpensive.
+
+Finally, AFL stability was 94.32\% with the harness allowing
+nondeterminism, but rose to 98.32\% using the harness with
+nondeterministic operations removed.
 

### 2018-07-24 introduction.tex
afl info

@@
 This paper proposes (1) a number of formal definitions of types of
-nondeterminism and (2) an implementation, based on these definitions, for detecting and debugging
-nondeterminism in property-based testing.  This enables a number of useful results:
+nondeterminism (\emph{horizontal} and \emph{vertical}) and (2) an implementation, based on these definitions, for detecting and debugging
+nondeterminism in property-based testing.  The implementation is based
+on an approach where (3) horizontal determinism is considered as a
+kind of \emph{reflexive differential testing} and (4) vertical
+determinism is specialized to the common case of \emph{failure determinsim}.
+We further propose modifications to the widely used delta-debugging algorithm in order to better handle
+nondeterminism as a test property
+
+This enables a number of useful results:
  the automated production of
@@
 nondeterminism, and, most importantly, detection and effective debugging for
-unexpected nondeterminism that is, itself, a bug.  We propose
-modifications to the widely used delta-debugging algorithm in order to better handle
-nondeterminism as a test property, and describe an implementation of
-our approach in a Python testing tool that is used on real-world
-projects.  Finally, we use a set
-of case studies of real, widely-used Python projects to demonstrate the
-utility of our ideas.
+unexpected nondeterminism that is, itself, a bug.  We use a set
+of case studies of real-world, widely-used, Python libraries to demonstrate the
+utility of our ideas.  Failure determinism directly
+produces a more than 5\% improvement in mutation score for an already highly
+effective automated test generation tool for a file system, with no
+additional specification or test design burden.

### 2018-11-03 experiments.tex
getting ready for TR

@@
 final operation (since further operations are required to 
-expose the bad file system state the operation introduces).
+expose the bad file system state the operation introduces).  For this
+hypothetical subtle bug, failure determinism checks not only make it
+possible to detect the fault without a reference implementation, it
+improves on detection speed and ease of debugging even compared to a
+full-blown reference implementation (the MacOS file system, operating
+on a RAM disk) and strong, hand-tuned, differential testing.
 
@@
 causes an unexpected exception to be raised, or results in a timeout
-due to, e.g., and infinite loop.
+due to, e.g., an infinite loop.
 
@@
 testing, then with the full differential test plus
-an additional 60 seconds of testing (to account for the observed 17\%
-extra time to detect in the manually constructed example).   For both analyses, we tested
+an additional 60 seconds of testing (to make sure we accounted for the observed 17\%
+extra time to detect in the manually constructed example: we want to
+maximize the chance to detect a bug for the methods we compare
+against).   For both comparisons, we then tested
 the surviving mutants using the non-differential harness for 60
-seconds, but with failure nondeterminism checking.
+seconds, but with failure nondeterminism checking turned on.
 We used the same random seed for
@@
 time, killed 1148 mutants, improving the kill rate to 48.8\%.  Failure determinism added an
-additional 71 mutants to the total killed, a 6\% improvement even for this complete
-harness with a larger test budget.  The kill rates are generally low
+additional 71 mutants to the total killed, a 6\% improvement even for
+this strong differential harness with a larger test budget.  The kill rates are generally low
 because {\tt universalmutator}, by design, includes many operations
@@
 provided by failure nondeterminism detection, we inspected the mutants
-killed by the failure determinism check but not by differential
-testing.  The largest category of mutants not killed (22 of the 71 mutants) was what we refer
-to as ``exception swallowing'' mutants, which transform a Python
-statement into the same statement, but wrapped in a {\tt try} block
-with a catch that ignores any raised exceptions, e.g., {\tt foo()} becomes:
+killed using the failure determinism check but not killed by the
+strong differential testing.  The largest category of mutants not
+killed (22 of the 71 mutants) was what we refer to as ``exception
+swallowing'' mutants, which transform a Python statement into the same
+statement, but wrapped in a {\tt try} block with a catch that ignores
+any raised exceptions, e.g., {\tt foo()} becomes:
 
@@
 errors, and thus would tend to cause failure nondeterminism.  However,
-these are a minority of the mutations introducing failure
+these are a minority of the mutations introducing hard-to-detect but
+non-equivalent failure
 nondeterminism.  Other mutation operators resulting in subtle flaws
@@
 (including replacement of a string with the empty string), and
-introducing a break into a loop.
+introducing a break into a loop.  The variety of mutant types suggests
+that no more specifically tailored strategy such as checking for
+exception propagation, will work as well as introducing a notion of
+failure nondeterminism.
 
@@
 larger with failure determinism checking, and this change was significant by Wilcoxon test
-$(p < 1 \times 10^{-15})$.
+$(p < 1 \times 10^{-15})$.  While significant, this cost is probably
+too small to be of any real practical importance.

### 2018-11-03 implementation.tex
getting ready for TR

@@
 
+\subsection{Implementing Horizontal Nondeterminism Detection}
+
 We implemented our approach in the TSTL \cite{NFM15,ISSTA15,tstlsttt} system, a
@@
 
-We therefore instead made nondeterminism detection a first-class
+We therefore instead made horizontal nondeterminism detection a first-class
 property in TSTL, using TSTL's existing notation for marking some
@@
 nondeterminism is when a call results in a change to a value passed as
-a parameter, or even due to a shared reference held by two values).
+a parameter, changes a shared reference held by two values).
 TSTL omits comparison only of pools marked as {\tt OPAQUE} in the
@@
 possible use in some security-related libraries.
+
+\subsection{Implementing Horizontal Nondeterminism Detection}
 

### 2018-11-03 experiments.tex
maybe ready for TR

@@
 
-The {\tt datarray} module \cite{datarray} is a prototype
-implementation for numpy arrays with named axes to improve data
-management, developed by the Berkeley Institute for Data Science.  As part of its code, it provides a set of algorithms for
-inference in
-Bayesian belief networks \cite{russell2016artificial}.  An earlier
-version of these algorithms produced nondeterministic (and in some
-cases incorrect) results due to dependence on the order of values in
-an iterator over a Python dictionary, on Python versions above 3.2, until 3.6 (see Section \ref{sec:pnondet}).
-
-Figure \ref{hashbug} shows TSTL code for generating inputs to the {\tt
-  datarray} algorithms.
-The {\tt flatten\_and\_sort} function is needed because we care about actual differences in probability values, not simply the order of list, tuple, or dictionary items.
-Running this harness using TSTL's {\tt
-  --checkProcessDeterminism} flag required less than 10 seconds on
-average to produce a test exhibiting process-level nondeterminism in
-the {\tt calc\_marginals\_sumproduct} function (the only broken
-algorithm).  Reducing this 60 step test to a minimal test of only 6 steps,
-showing an extremely simple input producing the issue, 
-required another 92 seconds.  Interestingly, the way that {\tt
-  python-afl} is used in TSTL means that AFL stability is 100\%,
-because the {\tt PYTHONHASHSEED} has already been chosen before the
-fork in AFL executions.
-
-Removing the nondeterministic call, we can see that the cost of
-checking for process nondeterminism, with no delay between operations,
-is high, a 93\% slowdown.  This is due to the high cost of subprocess
-creation and communication.
-
 \begin{figure*}
@@
 \end{figure*}
+
+The {\tt datarray} module \cite{datarray} is a prototype
+implementation for numpy arrays with named axes to improve data
+management, developed by the Berkeley Institute for Data Science.  As part of its code, it provides a set of algorithms for
+inference in
+Bayesian belief networks \cite{russell2016artificial}.  An earlier
+version of these algorithms produced nondeterministic (and in some
+cases incorrect) results due to dependence on the order of values in
+an iterator over a Python dictionary, on Python versions above 3.2, until 3.6 (see Section \ref{sec:pnondet}).
+
+Figure \ref{hashbug} shows TSTL code for generating inputs to the {\tt
+  datarray} algorithms.
+The {\tt flatten\_and\_sort} function is needed because we care about actual differences in probability values, not simply the order of list, tuple, or dictionary items.
+Running this harness using TSTL's {\tt
+  --checkProcessDeterminism} flag required less than 10 seconds on
+average to produce a test exhibiting process-level nondeterminism in
+the {\tt calc\_marginals\_sumproduct} function (the only broken
+algorithm).  Reducing this 60 step test to a minimal test of only 6 steps,
+showing an extremely simple input producing the issue, 
+required another 92 seconds.  Interestingly, the way that {\tt
+  python-afl} is used in TSTL means that AFL stability is 100\%,
+because the {\tt PYTHONHASHSEED} has already been chosen before the
+fork in AFL executions.
+
+Removing the nondeterministic call, we can see that the cost of
+checking for process nondeterminism, with no delay between operations,
+is high, a 93\% slowdown.  This is due to the high cost of subprocess
+creation and communication.
 
@@
 We first analyzed the mutants using 60 seconds of non-differential
-testing, then with the full differential test plus
-an additional 60 seconds of testing (to make sure we accounted for the observed 17\%
+testing \emph{without failure nondeterminism checks}, then with the
+full differential harness (also without failure nondeterminism checks)
+for 120 seconds; the additional 60 seconds is to make sure we accounted for the observed 17\%
 extra time to detect in the manually constructed example: we want to
-maximize the chance to detect a bug for the methods we compare
-against).   For both comparisons, we then tested
-the surviving mutants using the non-differential harness for 60
-seconds, but with failure nondeterminism checking turned on.
+maximize the chance to detect a bug using the strong version of the harness.   For both comparisons, we then tested
+all surviving mutants using the non-differential harness for 60
+seconds, this time with failure nondeterminism checking turned on.
 We used the same random seed for
@@
 The differential harness, with an additional 60 seconds of testing
-time, killed 1148 mutants, improving the kill rate to 48.8\%.  Failure determinism added an
+time, killed 1148 mutants, improving the kill rate to 48.8\%.  Failure
+nondeterminism checking added an
 additional 71 mutants to the total killed, a 6\% improvement even for
-this strong differential harness with a larger test budget.  The kill rates are generally low
+this strong differential harness with a larger test budget.  Using
+failure nondeterminism was the only way to push the mutation score
+above 50\%.  The kill rates are generally low
 because {\tt universalmutator}, by design, includes many operations
@@
 non-equivalent mutants not produced by other mutation tools, e.g. code
-that throws away exceptions raised by a function call, or code
+that throws away exceptions raised by a function call (see below), or code
 reversing a list.

### 2018-11-03 introduction.tex
maybe ready for TR

@@
 developers, who will usually want to  assume that library code they
-use is deterministic; it is frequently vexing in debugging efforts; it is often disastrous for large-scale highly
+use behaves in a reliably predictable fashion; nondeterminism is
+frequently vexing in debugging efforts; most importantly, perhaps, nondeterminism is \emph{often disastrous} for large-scale highly
 automated testing.

### 2019-02-07 implementation.tex
maybe ready for TR
start on reviewer requests

@@
 changed timing.
-
-\subsection{Implementing Nondeterminism Detection}
 
@@
 
-\subsection{Implementing Horizontal Nondeterminism Detection}
+\subsection{A Brief Introduction to TSTL}
 
-We implemented our approach in the TSTL \cite{NFM15,ISSTA15,tstlsttt} system, a
-language and tool for property-based testing \cite{Hypothesis,ClaessenH00} of Python code. TSTL has
+We implemented our approach in the TSTL \cite{NFM15,ISSTA15,tstlsttt}
+system, an open-source
+language and tool for property-based testing
+\cite{Hypothesis,ClaessenH00} of Python code.   TSTL, including full
+support for our nondeterminism detection methods, is available at
+\url{https://github.com/agroce/tstl}, and the {\tt examples} directory
+contains the source code needed to duplicate all of our experiments;
+TSTL can also be installed using Python's {\tt pip} tool, by typing
+{\tt pip install tstl}; you will need to clone the github repository
+to have access to the case studies used in the paper, however.
+
+TSTL has
 been used to detect (and usually fix) errors in a number of widely
@@
 compiler and a Solidity static analysis tool, and Mac OS \cite{tstl}.
+
+\subsection{Implementing Horizontal Nondeterminism Detection}
+
 Because TSTL supports differential testing \cite{tstlsttt}, horizontal
@@
 
-\subsection{Implementing Horizontal Nondeterminism Detection}
+\subsection{Implementing Vertical Nondeterminism Detection}
 
@@
 for failure and calls the code again.
+
+\subsection{Tools for Nondeterministic Delta-Debugging}
 
@@
 
+\subsection{Nondeterminism and AFL Stability}
+
 Because TSTL has a first-class interface to AFL \cite{aflfuzz}, we can even use
@@
 cost indicator produced as a by-product of fuzzing.
+
+\section{A Simple Example}

### 2019-02-09 concepts.tex
start on reviewer requests
formalism

@@
 system state in a way that would change its behavior on a subsequent
-run.
+call to that function.
 
@@
 operations to be fast (since in test generation, failure is usually
-due to invalid parameters).
+due to invalid parameters).  Checking equivalence of the full
+observable state, however, is still expensive, and requires defining
+the observable components of a state.  A more lightweight, but (as we
+will show) still quite powerful way of checking for failure nondeteterminism is to
+repeat the failing operation, and check if it still fails.  If not,
+this is a sign that the operation is not deterministic with respect to
+failure.  Note that this more restrictive notion of failure
+nondeterminism captures such issues as the Apple login bug.
+
+\subsection{Formal Definitions}
+
+We can formally distinguish the types of nondeterminism by using a
+variant of a labeled transition system (LTS) as our model:
+
+\begin{itemize}
+\item $S$ is a set of states.
+\item $V$ is a set of \emph{observable} states, where $|V| \leq |S|$.
+\item $v: S \rightarrow V$ is a total function that, given a state, maps it
+  into the set of observable states.  Every state has an observable
+  component, which may be the complete state, or only an aspect of the
+  full state; we even allow $V$ to contain a
+  single element, in which case no two states in $S$ can be
+  distinguished by an observer.
+\item $I \subseteq S$ is a set of initial states.
+\item $A$ is a set of \emph{actions}.
+\item $T \subseteq S \times A \times S$ is a transition relation.
+\end{itemize}
+
+We assume that the underlying behavior of a system may be
+deterministic, or nondeterministic, in the strict sense, by allowing
+for a transition relation that \emph{may} be a function of
+$S \times A$.  A \emph{trace} $t$ is a finite sequence $t = s_0
+\xrightarrow[]{\alpha_0} s_1 \xrightarrow[]{\alpha_1} \ldots
+\xrightarrow[]{\alpha_{n-1}} s_n$ where $s_0 \in I$ and $\forall n
+. (s_n, \alpha_n, s_{n+1}) \in T$.
+
+A pair of traces $(t_1, t_2)$ where $t_1 = s^1_0
+\xrightarrow[]{\alpha^1_0} s^1_1 \xrightarrow[]{\alpha^1_1} \ldots
+\xrightarrow[]{\alpha^1_{n-1}} s^1_n$ and $t_2 = s^2_0
+\xrightarrow[]{\alpha^2_0} s_1 \xrightarrow[]{\alpha^2_1} \ldots
+\xrightarrow[]{\alpha^2_{n-1}} s^2_n$ are said to show \emph{visible value
+  determinism} if $\exists i > 0$ such that: $\forall j < i . \alpha^1_j = \alpha^2_j \wedge
+v(s^1_j) = v(s^2_j)$ but $v(s^1_i) \neq v(s^2_i)$.  \emph{Final state
+  nondeterminism} is defined in the same way, except with the
+restriction that $i = n$.  A pair of traces may exhibit visible value
+nondeterminism but not final state nondeterminism, but any pair of
+traces exhibiting final state nondeterminism also demonstrate visible
+value nondeterminism.
+
+Vertical nondeterminism, in contrast, is demonstrated by a single trace
+$t$, not a pair of traces.  To define vertical nondeterminism with
+respect to idempotency, we can
+define $\mathcal{I} \subseteq A$, the set of \emph{idempotent}
+actions.  A trace $t s_0
+\xrightarrow[]{\alpha_0} s_1 \xrightarrow[]{\alpha_1} \ldots
+\xrightarrow[]{\alpha_{n-1}} s_n$ shows \emph{vertical nondeterminism
+  with respect to idempotent operations} if $\exists i < n$ such that
+$\alpha_i \in \mathcal{I} \wedge \alpha_i = \alpha_{i+1}$ and $v(s_i)
+\neq v(s_{i+2})$.
+
+To define failure nondeterminism, we extend the
+transition relation to include a notion of \emph{failure}: $T \subseteq S
+\times A \times S \times {\tt bool}$, where the boolean indicates
+whether the action \emph{failed}.  A trace $t = s_0
+\xrightarrow[]{\alpha_0} (s_1, F_0) \xrightarrow[]{\alpha_1} \ldots
+\xrightarrow[]{\alpha_{n-1}} (s_n, F_n)$ shows \emph{failure
+  nondeterminism} if $\exists i$ such that $F_{i+1}$ (that is,
+$\alpha_i$ \emph{fails}) and either (1) $v(s_{i}) \neq v(s_{i+1})$ or (2)
+$\alpha_i = \alpha_{i+1} \wedge \neg F_{i+2}$.  The second possibility
+captures our more lightweight approach to checking for failure nondeterminism.
 

### 2019-02-10 concepts.tex
fix n to i
fix
more text, fix formalisms

@@
 \section{Practical Lightweight Nondeterminism
-  Detection and Debugging}
+  Detection}
 
@@
 \xrightarrow[]{\alpha_0} s_1 \xrightarrow[]{\alpha_1} \ldots
-\xrightarrow[]{\alpha_{n-1}} s_n$ where $s_0 \in I$ and $\forall n
-. (s_n, \alpha_n, s_{n+1}) \in T$.
+\xrightarrow[]{\alpha_{n-1}} s_n$ where $s_0 \in I$ and $\forall i < n
+. (s_i, \alpha_i, s_{i+1}) \in T$.
 
@@
 $\alpha_i = \alpha_{i+1} \wedge \neg F_{i+2}$.  The second possibility
-captures our more lightweight approach to checking for failure nondeterminism.
+captures our more lightweight approach to checking for failure
+nondeterminism.  Note that unlike all of the other ways to demonstrate
+nondeterminism, this second type of failure nondeterminism does not
+require the use of $v$, or even the notion of observable components of
+a state, or (for that matter) any comparison of states.
 
-\subsection{Nondeterminism and Delta-Debugging}
+A further point to emphasize is that while horizontal determinism
+requires two traces, it does not require a test generation algorithm
+to produce pairs of traces, which would require substantial
+modification of most such tools (e.g., random testers or fuzzers, and
+even model checkers).  In general, testing tools do not produce
+\emph{traces}, they produce \emph{action sequences}.  The two traces
+used to show nondeterminism, by definition, have the same action
+sequences, which means that a testing tool only needs to search for an
+action sequence which, when executed twice, produces a $t_1$ and $t_2$
+meeting the requirements for showing nondeterminism ---
+non-equivalence of $v(s)$ for some $s$.
+
+\section{Nondeterminism and Delta-Debugging}
 
@@
 
-\subsubsection{``Publication Bias'' in Delta-Debugging}
+\subsection{``Publication Bias'' in Delta-Debugging}
 \label{sec:pubbias}
@@
 problematic operations.
+
+\subsection{Using Replications to Mitigate ``Publication Bias''}
 

### 2019-02-10 implementation.tex
more text, fix formalisms

@@
 more principled version of the unfortunate tradition of testing
-concurrent code by adding ad-hoc {\tt sleep} statements).  However,
-simply running a test twice (possibly with a delay between steps) does
+concurrent code by adding ad-hoc {\tt sleep} statements).   Most
+nondeterministic behavior in tests is probably attributable to a such
+effects that can vary with a repeated execution, in the same
+(approximate) enviroment, of the same test.  The two broad classes of
+nondeterminism that can usually be detected (with some probability) by
+simply executing a test twice in succession are:
+
+\begin{itemize}
+\item {\bf External Environment:} This class of nondeterminism arises
+  when a test's behavior depends on factors outside the program under
+  test, which are subject to uncontrolled variance.  Calling {\tt
+    random()} in Python is an obvious example; the call is almost
+  guaranteed to return a different value each time it is called in a
+  test, unless we explicitly re-seed the random number generator at
+  the beginning of each test.  Similarly, calls to {\tt time}
+  functions  will usually return different values on different runs,
+  and even if only elapsed time, rather than absolute, is used to
+  influence behavior, subtle variances in timing are to be expected
+  due to system load, I/O, and other external factors.  Finally, in
+  distributed system testing, the environment is often remote servers
+  (or clients) that may be unavailable, slower than expected, or
+  subject to faulty behavior for reasons not under the control of the test.
+\item {\bf Concurrency:} Perhaps the second most common cause for
+  nondeterminism in tests is when an algorithm is implemented using
+  concurrency, in the (incorrect) beleif that the result of the
+  algorithm remains deterministic.  E.g., consider a multi-threaded
+  implementation of a search algorithm, that returns the location of
+  an item in an unsorted list.  If the list may contain duplicates,
+  and a developer has not accounted for this possibility, the
+  algorithm may return different locations of an item for the same list,
+  depending on the scheduling of threads.  Arguably, this can be
+  subsumed under the first class, with the behavior of the
+  processor/scheduler as the ``enviromental'' factor; relying on
+  scheduling is, in a sense, equivalent to calling {\tt random} or
+  {\tt time}.
+\end{itemize}
+
+However, simply running a test twice in the same process (possibly with a delay between steps) does
 not suffice in all cases.

### 2019-02-20 experiments.tex
info on new experiment

@@
 nondeterministic operations removed.
+
+\subsubsection{Visible Value vs. Final State Nondeterminism}
+
+We also used {\tt redis-py} to investigate the tradeoff between
+overhead and ability to detect nondeterminism when using the two
+proposed approaches to horizontal determinism.  We produced 300 tests
+of length 100, using the known-nondeterministic version of the {\tt
+  redis-py} harness.  We then used {\tt nondeterministic} and {\tt
+  stepNondeterministic} calls, both with a delay of 0.005 seconds and
+a single additonal execution of the test, to
+check these tests for nondeterministic behavior.  Final state
+nondeterminism actually detected one instance of nondeterminism that
+visible value nondeterminism failed to detect; in general, final state
+nondeterminism is less able to detect nondeterminism, but in this
+instance the difference is less important than the nondeterminism of
+whether a particular test will exhibit nondeterministic behavior in a
+single run.  This difference is obviously not statistically
+significant; for the nondeterminism in {\tt redis-py}, with these
+parameters, the two approaches cannot be statistically distinguished
+as to effectiveness in detecting nondeterminism.
+
+Final state nondeterminism was also faster; visible value
+nondeterminism took, on average, 0.48\% longer to check for
+nondeterminism on the 300 tests.  The difference was significant, with
+$p$-value $< 1.19 \times 10^{-23}$ by a paired Wilcoxon test.
 

### 2019-02-20 experiments.tex
more on experiments

@@
 nondeterminism actually detected one instance of nondeterminism that
-visible value nondeterminism failed to detect; in general, final state
+visible value nondeterminism failed to detect (21 detections vs. 20); in general, final state
 nondeterminism is less able to detect nondeterminism, but in this
@@
 nondeterminism took, on average, 0.48\% longer to check for
-nondeterminism on the 300 tests.  The difference was significant, with
+nondeterminism on the 300 tests, 1.199 mean seconds vs. 1.205 mean seconds.  The difference was significant, with
 $p$-value $< 1.19 \times 10^{-23}$ by a paired Wilcoxon test.

### 2019-02-21 concepts.tex
fix idemopotent
discuss deflaker and elaborate on nondeterminism vs. flaky tests

@@
 
+\subsection{Nondeterminism vs. Flakiness}
+
+A key concept in this paper is that flaky tests (tests that sometimes
+pass and sometimes fail, recall) exist due to underlying
+nondeterministic behavior.  Our interest is in identifying such
+nondeterministic behavior, not in addressing flakiness \emph{per se}.
+Of course, the concepts are closely related.  We make ``did not behave
+nondeterministically'' an additional \emph{property} that a test can
+fail.  Due to the nature of (horizontal) nondeterminism, such failure
+is almost always going to be inherently flaky --- sometimes a test
+will behave nondeterministically, and sometimes it will not.  But the
+idea is a separate one from flakiness; in order to show horizontal
+nondeterminism, we need only see a change in values in a test, not in
+the disposition (failed/passed) of the test.  It may be that, without
+the additional property of determinism, the change cannot actually
+cause failure, and thus flakiness.  The real relationship between our
+nondeterminism and flakiness is that tests (automatically generated or
+manually constructued) that rely on code that can be shown to exhibit
+horizontal nondeterminism contain a potential source of flakiness.  If
+the code behaves nondeterministically, and the values that are
+observed to change can flow to influence the outcome of the test,
+flakiness may well result.
+
+Vertical nondeterminism, in fact, is potentially \emph{completely} unrelated
+to flakiness, in that some bugs (improper handling of hardware errors
+in a file system, for example) produce \emph{reliably}
+``nondeterministic'' (from a software library users' point of view)
+behavior.  Vertical nondeterminism may rely on ``real'' nondeterminism
+that could produce a flaky test, but it does not have to do so.
+
 \subsection{Formal Definitions}
@@
   with respect to idempotent operations} if $\exists i < n$ such that
-$\alpha_i \in \mathcal{I} \wedge \alpha_i = \alpha_{i+1}$ and $v(s_i)
+$\alpha_i \in \mathcal{I} \wedge \alpha_i = \alpha_{i+1}$ and $v(s_{i+1})
 \neq v(s_{i+2})$.

### 2019-02-21 experiments.tex
discuss deflaker and elaborate on nondeterminism vs. flaky tests

@@
 $p$-value $< 1.19 \times 10^{-23}$ by a paired Wilcoxon test.
+
+These results, of course, depend on the parameters of the check for
+nondeterminism.  If we increase the delay to 0.01 seconds and the
+number of replays of a test used to check for nondeterminism, we
+almost double the number of nondeterministic tests discovered.  Both
+visible value and final state approaches find 40 nondeterministic
+tests, though the overlap of tests thus detected is not quite perfect
+--- each method detects two tests the other does not.  The difference
+in overhead, interestingly, is very similar:  0.48\%; however, under
+these parameters, visible value nondeterminism is actually cheaper on average
+than final state nondeterminism, due to early termination of the
+additional replays, when nondeterminism is detected (the difference is
+again significant by Wilcoxon test, with $p < 1.44 \times 10^{-10}$.
+
+In practice, visible value nondeterminism and final state
+nondeterminism will often be very similar in both ability to detect
+nondeterministic behavior and overhead.  However, under unusual
+conditions, the behavior of the approaches can be very different.  The
+additional overhead of visible value nondeterminism is usually low
+because re-executing a test, not comparing state values for equality, is
+the primary cost in nondeterminism detection; however, if there are a
+very large number of state components, or some state components have
+an expensive equality check (e.g., recursive structures with cycle
+detection or depth limits), the overhead can grow, proportional to the
+length of tests under consideration.  Similarly, final state
+nondeterminism in a setting such as {\tt redis-py}, where divergences
+in behavior tend to propagate to other state components, or at least
+persist until termination of a test, detects nondeterminism quite
+effectively.  However, in a setting where changes in behavior can
+easily by overwritten by future test behavior, and do not causally
+influence future computation, final state nondeterminism may be almost useless.
 

### 2019-02-21 introduction.tex
discuss deflaker and elaborate on nondeterminism vs. flaky tests

@@
 We further propose modifications to the widely used delta-debugging algorithm in order to better handle
-nondeterminism as a test property
+nondeterminism as a test property.
+
+The underlying idea in this paper is that most approaches to
+nondeterminism/flaky test detection rely either on 1) identifying
+potential sources of nondeterminism such as test inter-dependence
+\cite{LamZE2015} or certain code smells \cite{palomba2017does}  or 2)
+used a heuristic approach, such as that taken by DeFlaker
+\cite{bell2018d}.  In the real world, people usually detect flaky
+tests by the most obvious heuristic of all:  they observe a test both
+fail and pass for the same code base.  In a sense, even DeFlaker
+relies on this method, but avoids having to observe a flaky test actually
+passing, by inferring that if a test was previously passing, and
+executes no changed code, its failure after a code change \emph{must}
+be due to flakiness/nondeterminism.  Our approach to the problem is
+essentially different, and largely orthogonal; we concentrate on the detection and avoidance of
+nondeterminism/flakiness using \emph{automated test generation},
+and make it possible for a developer or test engineer to detect when
+important values generated in automated testing differ without a
+change in the test itself or the code under test.  While some of these
+changes may be harmless, and unable to propagate to cause test
+failures, they are potential sources of flaky behavior.  To our
+knowledge, this is the first dynamic approach to the problem that
+allows detection of (potential) flakiness, based on real divergence in
+behavior, without having to observe a test actually fail due to nondeterminism.
 
@@
 nondeterminism, and, most importantly, detection and effective debugging for
-unexpected nondeterminism that is, itself, a bug.  We use a set
+unexpected nondeterminism that is, itself, a bug.  Moreover, by
+identifying (using automated testing) library behaviors that are
+nondeterministic, our approach should make it much easier to flag suspect
+statements in human-written flaky unit tests.
+
+We use a set
 of case studies of real-world, widely-used, Python libraries to demonstrate the

### 2019-02-21 relatedwork.tex
discuss deflaker and elaborate on nondeterminism vs. flaky tests

@@
 Palomba and Zaidman \cite{palomba2017does} investigated the relationship between test code smells and flaky tests.
-The present paper, rather than focusing on flaky tests per se, investigates new tools for handling nondeterminism in property-based test generation.
-From our point of view, flaky behavior is simply a special case of nondeterminism, where the nondeterminism is sufficient to cause the test to have different pass/fail results at times.
+
+Bell et al.
+proposed DeFlaker \cite{bell2018d}, which makes flaky tests much easier to detect by relying on the observation that if a test fails, and does not cover any changed code then (1) presumably it was a passing test in the past, or the tests would not be ``green'' before the latest code change so (2) the test is likely flaky.
+Comparing our approach with that of DeFlaker is difficult; DeFlaker applies only in the context of code \emph{changes} and is essentially a heuristic that identifies certain test failures as due to nondeterminism.
+Our approach is primarily focused on automatically providing a more extensive \emph{specification} in automated test generation, where a test ``failing'' is itself a sign of nondeterministic behavior, that could give rise to flaky behavior.
+
+The present paper, rather than focusing on flaky tests as such, therefore, investigates methods for handling nondeterminism in property-based test generation.
+Again, from our point of view, flaky behavior is simply a special case of nondeterminism, where the nondeterminism is sufficient to cause the test to have different pass/fail results at times.
 Our approach is, in a sense (especially the delta-debugging modifications) in line with Harman and O'Hearn's proposal to simply accept that ``All Tests Are Flaky'' \cite{StartupstoScaleups}, and work with probabilistically failing tests.

### 2019-02-22 experiments.tex
more

@@
 
-After removing all identified sources of nondeterminism (the {\tt random-} functions, all
-expiration-related functionality and {\tt restore} calls with non-zero
-lifetimes, and {\tt spop}), but leaving all other suspect functionality in
-place (e.g. pipes), no flakiness was observed in a sample of over 2,000 length
+\begin{figure*}
+\begin{code}
+{\bf REMOVED:}    
+<key> := <r>.randomkey() 
+<r>.expire(<key>,<int>)
+<r>.pexpire(<key>,<int>)
+\{redis.exceptions.ResponseError\} <r>.spop(<key>)
+\{redis.exceptions.ResponseError\} <r>.srandmember(<key>)
+\{redis.exceptions.ResponseError\} <r>.srandmember(<key>,<int>)
+<pipe>.expire(<key>,<int>)
+<pipe>.pexpire(<key>,<int>)
+\{redis.exceptions.ResponseError\} <pipe>.spop(<key>)
+\{redis.exceptions.ResponseError\} <pipe>.srandmember(<key>)
+\{redis.exceptions.ResponseError\} <pipe>.srandmember(<key>,<int>)
+\vspace{0.1in}
+{\bf MODIFIED:}
+\{redis.exceptions.ResponseError\} <r>.restore(<key>,<int>,<dump>)
+{\bf $\Rightarrow$}
+\{redis.exceptions.ResponseError\} <r>.restore(<key>,0,<dump>)
+\vspace{0.07in}
+\{redis.exceptions.ResponseError\} <pipe>.restore(<key>,<int>,<dump>)
+{\bf $\Rightarrow$}
+\{redis.exceptions.ResponseError\} <pipe>.restore(<key>,0,<dump>)
+\end{code}
+\caption{{\tt redis-py} calls removed or changed to avoid
+  nondeterminism.}
+\label{fig:flakysource}
+\end{figure*}
+
+Figure \ref{fig:flakysource} shows all calls removed or modified after
+identification of sources of nondeterminism.  After making these changes, no flakiness was observed in a sample of over 2,000 length
 200 tests.   Moreover, because the removals were limited to actually

### 2019-02-22 introduction.tex
more

--- 2018-11-03 introduction.tex
@@
 test is not a deterministic property of the test code, code under
-test, and environment running the test.  This produces three serious
+test, and environment in which the test is run\footnote{Again, in 
+  some sense the result \emph{is} clearly determined by the environment in
+  which the test is run, but from the point of view of a developer or
+  test engineer, the differences in environment are invisible and unknown.}.  This produces three serious
 problems: first, a flaky test often wastes developer time and delays
@@
 We further propose modifications to the widely used delta-debugging algorithm in order to better handle
-nondeterminism as a test property
+nondeterminism as a test property.
 
-This enables a number of useful results:
+The underlying idea in this paper is that most approaches to
+nondeterminism/flaky test detection and debugging rely either on 1) identifying
+potential sources of nondeterminism such as test inter-dependence
+\cite{LamZE2015} or certain code smells \cite{palomba2017does}  or 2)
+use a heuristic approach to mark failures as likely flaky, such as that taken by DeFlaker
+\cite{bell2018d}.  In the real world, people usually detect flaky
+tests by the most obvious heuristic of all:  they observe a test both
+fail and pass for the same code version.  In a sense, even DeFlaker
+relies on this method, but avoids having to observe a flaky test actually
+passing, by inferring that if a test was previously passing, and
+executes no changed code, its failure after a code change \emph{must}
+be due to flakiness/nondeterminism.  Our approach to the problem is
+fundamentally different, and essentially orthogonal; we concentrate on the detection and avoidance of
+nondeterminism/flakiness using \emph{automated test generation},
+and make it possible for a developer or test engineer to detect when
+important values generated in automated testing differ without a
+change in the test itself or the code under test.  While some of these
+changes may be harmless, and unable to propagate to cause test
+failures, they are potential sources of flaky behavior.  To our
+knowledge, this is the first approach to the problem that
+allows detection of (potential sources of) flakiness, based on real divergence in
+behavior, without having to observe a test actually fail due to nondeterminism.
+
+This approach enables a number of useful results:
  the automated production of
 nondeterminism-free regression tests even for libraries with some
-sources of nondeterminism, the documentation of
-nondeterminism, and, most importantly, detection and effective debugging for
-unexpected nondeterminism that is, itself, a bug.  We use a set
+sources of nondeterminism, the documentation of sources of
+nondeterminism that might produce problems for manually-constructed tests, and, most importantly, detection and effective debugging for
+unexpected nondeterminism that is, itself, a bug.  Additonally, by
+identifying (using automated testing) library behaviors that are
+nondeterministic, our approach should make it much easier to flag
+statements in existing human-written unit tests that may produce flaky
+behavior:  simply scan the
+code for calls to functions identified as sources of nondeterminism.
+
+We use a set
 of case studies of real-world, widely-used, Python libraries to demonstrate the

### 2019-02-22 concepts.tex
more tstl intro

@@
 behavior.  Vertical nondeterminism may rely on ``real'' nondeterminism
-that could produce a flaky test, but it does not have to do so.
+that could produce a flaky test, but it does not have to do so.  In
+other words, vertical nondeterminism is often not a problem for
+testing (flakiness), but simply a bug in the behavior of a software
+system, where an implicit specification on failed operations is
+consistently violated under certain conditions.  The behavior is
+``nondeterministic'' from a caller's point of view, however, in that
+there is an expectation that a call will always ``do the same thing,
+under the same circumstances'' which is the \emph{conceptual} common
+factor in all kinds of expected determinism.
 
 \subsection{Formal Definitions}
+\label{sec:formal}
 

### 2019-02-22 implementation.tex
more tstl intro

@@
 \url{https://github.com/agroce/tstl}, and the {\tt examples} directory
-contains the source code needed to duplicate all of our experiments;
+contains the source code needed to duplicate all of our
+experiments\footnote{In some cases, you will also need to install an
+  appropriate version of the tested software, by doing, e.g., {\tt pip
+    install pyfakefs}.};
 TSTL can also be installed using Python's {\tt pip} tool, by typing
@@
 compiler and a Solidity static analysis tool, and Mac OS \cite{tstl}.
+
+\begin{figure}
+\begin{code}
+pool: <int> 5
+pool: <l> 5
+pool: <s> 5
+\vspace{0.1in}
+<int> := <[1..20]>
+<l> := []
+<l> := list(<s>)
+<s> := set()
+<s> := set(<l>)
+\vspace{0.1in}
+<l>.append(<int>)
+\{ValueError\} <l>.remove(<int>)
+<s>.add(<int>)
+\{KeyError\} <s>.remove(<int>)
+\vspace{0.1in}
+property: len(<l>) >= len(set(<l,1>))
+property: len(<s>) == len(list(<s,1>))
+\end{code}
+\caption{Simple TSTL harness testing lists and sets}
+\label{fig:simple}
+\end{figure}
+
+In TSTL, a test consists of a sequence of
+\emph{actions}, where actions execute arbitrary Python code, but are
+expected to work by modifying the \emph{state} of a test, which is
+stored in a fixed set of \emph{pools} containing Python objects.  The
+TSTL actions defined by a test harness correspond to the set $A$ of
+actions in Section \ref{sec:formal}, and transitions between states of
+result from carrying out the Python code in an action.  Figure
+\ref{fig:simple} shows a very simple TSTL harness that tests Python's
+built-in {\tt set} and {\tt list} types.  The harness defines the
+(visible) state of a test as consisting of three named pools of Python
+objects, {\tt int}, {\tt l}, and {\tt s}.  These pools have no
+inherent associated type, though a type can be annotated as an
+additional constraint that TSTL will check holds during test
+execution.  Initially all the pools hold the special Python value {\tt
+  None} indicating they have not been initialized.  In terms of our formalism,
+$I = \{${\tt [<int> = [None, None, None, None, None]}, {\tt <l> =
+  [None, None, None, None, None]}, {\tt <s> = [None, None, None, None,
+  None]]}$\}$.  Actions with the {\tt :=} assignment operator
+initialize pools.  For example, the line of code {\tt l := []} defines a set of 5
+actions, each of which initializes a different member of the list pool
+{\tt l} to an empty list --- {\tt l0 = []} ... {\tt l4 = []}.  Actions
+without a {\tt :=} require that pools used in them already be
+initialized.  TSTL provides constructs like {\tt <[1..20]>} for
+choosing values in a range for use in a test, and allows a user to
+specify which exceptions an action may raise without causing the test
+to fail (as in {\tt \{KeyError\} <s>.remove(<int>)}).
+
+Compiling this harness using
+the TSTL compiler produces a Python module that allows test
+generation.  If we use the random testing tool provided by TSTL, {\tt
+  tstl\_rt}, it will randomly generate valid sequences of actions, and
+report no problems.  If we change the first property to read {\tt
+  property: len(<l>) == len(set(<l,1>))}, however, TSTL will report a
+failing test, shown in Figure \ref{fig:setlistfail}.
+
+\begin{figure}
+{\scriptsize
+\begin{code}
+int0 = 1                                                                 \# STEP 0
+ACTION: int0 = 1 
+int0 = None : <type 'NoneType'>
+=> int0 = 1 : <type 'int'>
+==================================================
+l0 = []                                                                  \# STEP 1
+ACTION: l0 = [] 
+l0 = None : <type 'NoneType'>
+=> l0 = [] : <type 'list'>
+==================================================
+l0.append(int0)                                                          \# STEP 2
+ACTION: l0.append(int0) 
+int0 = 1 : <type 'int'>
+l0 = [] : <type 'list'>
+=> l0 = [1] : <type 'list'>
+==================================================
+l0.append(int0)                                                          \# STEP 3
+ACTION: l0.append(int0) 
+int0 = 1 : <type 'int'>
+l0 = [1] : <type 'list'>
+=> l0 = [1, 1] : <type 'list'>
+==================================================
+ERROR: (<type 'exceptions.AssertionError'>,
+  AssertionError(), <traceback object at 0x1032b83f8>)
+TRACEBACK:
+  File "/Users/adg326/scratch/pyexample/sut.py",
+  line 13695, in check
+    assert len(self.p\_l[0]) == len(set(self.p\_l[0]))
+\end{code}
+}
+\caption{A failing test (showing that converting a list to a set may
+  not preserve the size of the object).}
+\label{fig:setlistfail}
+\end{figure}
 
@@
 types of values as \emph{opaque} (not usefully compared for equality),
-used in automatic abstraction-based testing.  In TSTL, a test is a
-sequence of \emph{actions}; parameter values to method and function
-calls of the Software Under Test (SUT) are stored in finite-sized variable \emph{pools}.  A
-TSTL test is (essentially) a sequence of assignments to pool values,
+used in automatic abstraction-based testing.  Recall that in TSTL, a
+test is (essentially) a sequence of assignments to pool values,
 and method/function calls using pool values (including method calls on
@@
 nondeterminism is when a call results in a change to a value passed as
-a parameter, changes a shared reference held by two values).
-TSTL omits comparison only of pools marked as {\tt OPAQUE} in the
+a parameter, changes a shared reference held by two values).  By
+default $V$ as defined in Section \ref{sec:formal} is just the set of all possible pool values, and $v$
+simply extracts the pool values from $S$.
+TSTL allows restriction of which state is visible by removing pools marked as {\tt OPAQUE} in the
 TSTL language test harness (and values that do not support equality
-checking).  \emph{Final State Determinism} simply performs the same
+checking) from $v(S)$.  \emph{Final State Determinism} simply performs the same
 comparison, but only on the final values of all pool variables (or a

### 2019-02-24 implementation.tex
simple example

@@
 \section{A Simple Example}
+
+\begin{figure}
+\begin{code}
+@import random
+\vspace{0.1in}
+<@
+storage = []
+\vspace{0.1in}
+def init():
+    global storage
+    storage = []
+\vspace{0.1in}
+def store(n):
+    global storage
+    if n in storage:
+       raise KeyError
+    storage.append(n)
+    if n < 0:
+       raise ValueError
+@>
+\vspace{0.1in}
+init: init()
+\vspace{0.1in}
+pool: <int> 5
+\vspace{0.1in}
+<int> := <[-10..10]>
+<int> := random.randint(-100,100)
+{KeyError,ValueError} store(<int>)
+\end{code}
+\caption{A trivial example containing (independent) vertical and
+  horizonal nondeterminisms}
+\label{fig:simpleexample}
+\end{figure}
+
+Figure \ref{fig:simpleexample} shows a self-contained example of a
+Python program and test harness with both vertical and horizontal nondeterminism.

### 2019-02-25 implementation.tex
example in progress

--- 2019-02-22 implementation.tex
@@
 \section{A Simple Example}
+
+\begin{figure}
+\begin{code}
+@import storage
+\vspace{0.1in}
+init: storage.init()
+\vspace{0.1in}
+pool: <int> 5
+\vspace{0.1in}
+<int> := <[-10..10]>
+<int> := storage.getNewVal()
+{KeyError,ValueError} storage.store(<int>)
+\vspace{0.1in}
+property: len(storage.contents()) == len(set(storage.contents()))
+\end{code}
+\caption{TSTL harness for Python library with (independent) vertical and
+  horizonal nondeterminisms}
+\label{fig:simpleexample}
+\end{figure}
+
+\begin{figure}
+\begin{code}
+import random
+
+storage = []
+
+def contents():
+    global storage
+    return storage
+
+def init():
+    global storage
+    storage = []
+
+def store(n):
+    global storage
+    if n in storage:
+       raise KeyError
+    storage.append(n)
+    if n < 0:
+       raise ValueError
+
+def getNewVal():
+    global storage
+    if len(storage) >= 1:
+        bot = min(storage)
+        top = max(storage)
+    else:
+        bot = -10
+        top = 10
+    v = random.randint(bot-1,top+1)
+    while v in storage:
+        v = random.randint(bot-1,top+1)
+    return v
+\end{code}
+\caption{Code for Python library with (independent) vertical and
+  horizonal nondeterminisms}
+\label{fig:simpleexample}
+\end{figure}
+
+Figure \ref{fig:simpleexample} shows a self-contained example of a
+Python program and test harness with both vertical and horizontal nondeterminism.

### 2019-02-25 implementation.tex
fix

@@
 <int> := storage.getNewVal()
-{KeyError,ValueError} storage.store(<int>)
+\{KeyError,ValueError\} storage.store(<int>)
 \vspace{0.1in}
@@
   horizonal nondeterminisms}
-\label{fig:simpleexample}
+\label{fig:simpleexamplecode}
 \end{figure}
 
-Figure \ref{fig:simpleexample} shows a self-contained example of a
-Python program and test harness with both vertical and horizontal nondeterminism.
+Figures \ref{fig:simpleexample} and \ref{fig:simpleexamplecode} show a self-contained example of a
+Python library and test harness with both vertical and horizontal nondeterminism.

### 2019-02-25 implementation.tex
full TSTL example

@@
 \begin{figure}
+{\scriptsize
 \begin{code}
-@import storage
+@from storage import *
 \vspace{0.1in}
-init: storage.init()
+init: init()
 \vspace{0.1in}
@@
 <int> := <[-10..10]>
-<int> := storage.getNewVal()
-\{KeyError,ValueError\} storage.store(<int>)
+<int> := getNewVal()
+\{KeyError,ValueError\} store(<int>)
 \vspace{0.1in}
-property: len(storage.contents()) == len(set(storage.contents()))
+property: len(contents()) == len(set(contents()))
 \end{code}
+}
 \caption{TSTL harness for Python library with (independent) vertical and
@@
 Figures \ref{fig:simpleexample} and \ref{fig:simpleexamplecode} show a self-contained example of a
-Python library and test harness with both vertical and horizontal nondeterminism.
+Python library and test harness with both vertical and horizontal
+nondeterminism.  The example is not meant to be realistic, but to
+better concretely explain the detection and debugging of
+nondeterminism.  A few additional TSTL features need to be described.
+First, lines beginning with an {\tt @} indicate raw Python code, here
+used to import the module to be tested.  Second, {\tt init:} indicates
+code that is to be executed at the beginning of each test, in
+order to restore the system under test to its initial state.
+
+\subsection{Horizontal Nondeterminism}
+
+If we compile and test this code without using any of the
+nondeterminism-detection functionality, TSTL will not report any
+failing tests.  The property (avoidance of duplicate entries in
+storage) holds:
+
+{\scriptsize
+\begin{code}
+ \$ tstl storage.tstl
+ ...
+ \$ tstl\_rt
+ Random testing using config=Config...
+ storage: TEST \#1396 STEP \#0 (0:00:30.000188) Mon Feb
+ 25 11:13:18 2019 [ 15 stmts 22 branches ] (no cov+ for
+ 1391 tests) 139500 TOTAL ACTIONS (4649.97/s) (test
+ 0.0/s) 2 cov+ tests
+\end{code}
+}
+
+TSTL reports information on the testing in a status line, here showing
+that it has performed 1,396 tests, with no failures, and two tests
+that increased code coverage.  TSTL can be run in a mode that stores
+minimized versions of the tests that obtained new coverage, for use
+as, e.g., regressions:
+
+{\scriptsize
+\begin{code}
+ \$ tstl storage.tstl
+ ...
+ \$ tstl\_rt --quickTests
+\end{code}
+}
+
+\begin{figure}
+{\bf quick.0.test:}
+\begin{code}
+self.p\_int[0] = getNewVal() 
+store(self.p\_int[0]) 
+store(self.p\_int[0]) 
+self.p\_int[0] = -10 
+store(self.p\_int[0]) 
+\end{code}
+\vspace{0.1in}
+{\bf quick.1.test:}
+\begin{code}
+self.p\_int[0] = -8
+store(self.p\_int[0]) 
+self.p\_int[0] = getNewVal() 
+\end{code}
+\caption{Coverage maximizing tests for Python library with (independent) vertical and
+  horizonal nondeterminisms}
+\label{fig:simpleexampletests}
+\end{figure}
+
+This will result in generation of two\footnote{In some rare cases,
+  TSTL may generate one test that achieves complete code coverage, but
+  this will only happen rarely here.} files, {\tt quick.0.test} and
+{\tt quick.1.test}.  Sometimes, TSTL will fail at this point with an internal error
+({\tt REDUCED TEST DOES NOT PRESERVE COVERAGE}), due to the attempt to reduce one of the coverage-obtaining
+tests losing coverage, which is a consequence of horizontal
+nondeterminism.  However, TSTL will often produce
+test files such as those shown in Figure
+\ref{fig:simpleexampletests}.  TSTL lets us convert these tests, which
+are in TSTL's internal format, into standalone Python regression tests
+that check the properties in the test harness and ensure that the
+behavior (values of pools) are the same as in the recorded run:
+
+{\scriptsize
+\begin{code}
+ \$ tstl\_standalone quick.0.test quick0.py --regression
+ ...
+ \$ tstl\_standalone quick.1.test quick1.py --regression
+\end{code}
+}
+
+However, if we execute these Python tests, they will be very flaky;
+sometimes they will pass, but {\tt quick0.py} will almost always fail, and
+{\tt quick1.py} will fail about two thirds of the time.  The reason,
+of course, is a trivial case of horizontal nondeterminism, due to the
+use of the {\tt random.randint} function in {\tt getNewValue}.  The
+regression test will record whatever value {\tt getNewValue} returned
+the when the TSTL test file was executed in order to produce the
+Python standalone test.  The value in future executions will seldom
+match this value.  In this case, removing the nondeterminism is easy
+(simply add a fixed random seed to the {\tt init:} line in the
+harness, e.g., {\tt init: init(); random.seed(0)}).  However, in
+real-world testing, the source of the nondeterminism, may not be
+obvious, and only some regression tests may (infrequently) fail.
+However, if we ask the random tester to check for nondeterminism, it
+quickly shows us what the problem is:
+
+{\scriptsize
+\begin{code}
+ \$ tstl\_rt --checkDeterminism
+ ...
+ int2 = getNewVal()              \# STEP 50
+ MISMATCH IN REPLAY VALUE:
+    int2 : 7 VS. -5
+ TEST WAS NOT DETERMINISTIC!  WRITING FAILURE AS
+ ndfail.test
+\end{code}
+}
+
+We can use TSTL's test reduction tool to produce a small test showing
+the problem:
+
+{\scriptsize
+\begin{code}
+ \$ tstl\_reduce ndfail.test hnd.test --checkDeterminism
+ ...
+ NEW LENGTH 1
+ int0 = getNewVal()              \# STEP 0
+
+ TEST WRITTEN TO hnd.test
+\end{code}
+}
+
+Nothing is required to produce nondeterminism other than calling {\tt
+  getNewVal}, and this test is more frequently nondeterministic than
+the original, since it calls {\tt getNewVal} without any stored data
+to restrict the range of random values.
+
+\subsection{Vertical Nondeterminism}
+
+The {\tt storage.py} program also contains a (again, unrealistically
+simple) example of \emph{vertical} nondeterminism.  The {\tt store}
+method is supposed to 1) allow only a single copy of a value to be in
+the storage and 2) to disallow negative values.  Unfortunately, the
+check for negatives is \emph{after} the value is stored, which means
+that a second attempt to store the same negative value will result in
+a {\tt KeyError}, not a {\tt ValueError}.  TSTL can detect this
+problem if we compile the harness with {\tt
+  --checkFailureDeterminism}, which produces a version of the testing
+interface that ensures failing actions (ones that raise an expected
+exception) always fail again if performed again without any
+intervening actions, and always fail with the same exception.  Note
+that the check for the same exception is a slight extension of our formal
+definition of failure determinism, recording the last observed
+exception as an additional visible state element.
+
+{\scriptsize
+\begin{code}
+ \$ tstl storage.tstl --checkFailureDeterminsim
+ ...
+ \$ tstl\_rt
+ ...
+ FINAL VERSION OF TEST, WITH LOGGED REPLAY:
+ int0 = -4                                   \# STEP 0
+ ACTION: int0 = -4 
+ int0 = None : <type 'NoneType'>
+ => int0 = -4 : <type 'int'>
+ ==================================================
+ store(int0)                                 \# STEP 1
+ ACTION: store(int0) 
+ int0 = -4 : <type 'int'>
+ RAISED EXPECTED EXCEPTION:
+   <type 'exceptions.ValueError'> 
+ RAISED EXCEPTION:
+   <type 'exceptions.KeyError'>  ON RETRY
+   ERROR: (<type 'exceptions.AssertionError'>,
+   AssertionError('Failure with different exception.',),
+ <traceback object at 0x105e9edd0>)
+\end{code}
+}
+
+With failure nondeterminism, we don't have to use TSTL's specialized reduction
+tool, since the failure does not require multiple replays of a test.
+The random tester itself can properly reduce the test case, since it
+is just a ``normal'' failure, exhibited by a single trace.

### 2019-02-28 implementation.tex
add new darko paper

@@
   (or clients) that may be unavailable, slower than expected, or
-  subject to faulty behavior for reasons not under the control of the test.
+  subject to faulty behavior for reasons not under the control of the
+  test.
 \item {\bf Concurrency:} Perhaps the second most common cause for
@@
 
-However, simply running a test twice in the same process (possibly with a delay between steps) does
+Lam et al. \cite{idflakies} note that flaky tests not due to order
+dependence are generally due to ``concurrency, timeouts,
+network/IO, etc.'' a description that is well accounted for by our two
+primary sources, and most of which can be detected, at least in
+principle, by multiple runs of a test within the same testing process.  However, simply running a test twice in the same process (possibly with a delay between steps) does
 not suffice in all cases.
@@
 changed timing.
+
+One common cause of flaky tests is \emph{test order dependence} \cite{idflakies}.  This
+is less of a concern in our setting, since many problems due to test
+order are easily exposed in random testing, since the
+order-dependence-inducing operations will seldom usually happen in the
+same order in randomly generated tests.  For example, consider the
+case where a test usually run ``early'' initializes an uninitialized
+value in a system, and a later test will only fail if it somehow is
+run before the initializing test.  In random testing, the order of
+initialization/access is likely to be random, and some run of a random
+tester is likely to expose the problem as a (deterministic, not flaky) failure.
+
+However, for cases where
+there is an order dependence that is hard to translate into an
+actual bug, using a fresh
+process may expose such dependencies --- the ``environment'' of the
+parent process will have the, e.g., flag set or value initialized, but
+the new process will not.
+
+\subsection{A Brief Introduction to TSTL}
 
@@
 \end{table*}
-
-\subsection{A Brief Introduction to TSTL}
 

### 2019-02-28 relatedwork.tex
add new darko paper

@@
 
+iDFlakies \cite{idflakies} is a framework and dataset for flaky tests, but again focuses on whole-tests, and detecting flakiness by actually observing it.
+Their work shows the order-dependence accounts for a little more than 50\% of flaky test causes, and they (like us) classify the rest as due to ``concurrency, timeouts,
+network/IO, etc.' --- the kinds of cause that our approach focuses on identifying.
+
 The present paper, rather than focusing on flaky tests as such, therefore, investigates methods for handling nondeterminism in property-based test generation.

### 2019-03-07 experiments.tex
value pool formalism, parallel sort example

@@
 In order to evaluate our approach and implementation, we applied
-nondeterminism detection to three real world examples.  The primary
+nondeterminism detection to real Python code, including some widely
+used systems ({\tt redis-py} and {\tt pyfakefs} are well-maintained,
+widely-used libraries).  The primary
 points we wanted to explore were (1) whether our approach was able to reliably detect actual
@@
 delta-debugging for real nondeterministic predicates.
+
+\subsection{Case Study: Parallel Sorting}
+
+Our first case study is a simple implementation of a parallel merge
+sort in Python
+\url{https://gist.github.com/stephenmcd/39ded69946155930c347}.  The
+merge sort and harness can be found in the TSTL GitHub repository
+\cite{tstl}, under {\tt examples/parallelsorts}.  When a sorting
+algorithm is implemented using concurrency, a key question to ask is
+whether the ordering of elements that are equal under the comparison
+operator is consistent across sorts of the same original sequence.
+This is related to the question of whether a sort is a stable sort
+(preserving the original relative order of equal-by-comparison elements),
+but imposes a less strict requirement on the sort; a concurrent sort
+might not be stable, but might still produce the same (non-stable)
+ordering from the same provided sequence, every time.  This is a
+particular instance of a common expectation noted above:  while a parallel
+implementation of an algorithm will likely have internal
+nondeterminism (due to scheduling of work), it is often expected to
+behave deterministically, from the point of view of a caller.
+
+This case study has two purposes:  first, to show that the
+nondeterminism detection features of TSTL can be used to, with
+acceptable overhead, show that a parallel sort implementation provides
+a consistent, deterministic ordering, and second to show that the
+mechanism can also detect the nondeterminism in a parallel sort that
+does exhibit nondeterministic behavior.  For the second purpose, we
+implemented a very simple swap-based parallel sorting algorithm that
+repeatedly has multiple threads scan through a sequence and swap
+out-of-order neighbors, until the sequence is sorted (a kind of
+parallel bubble-sort with a simpler termination criteria).
+
+\begin{figure}
+{\scriptsize
+\begin{code}
+@import mergesort
+@import swapsort
+\vspace{0.1in}
+pool: <val> 10
+pool: <data> 10
+pool: <sorted> 10
+\vspace{0.1in}
+<val> := <[-10..10]>
+\vspace{0.1in}
+<data> := []
+<data>.append((<val>,<val>))
+len(<data,2>) < len(<data,1>) -> <data>.extend(<data>)
+\vspace{0.1in}
+<sorted> := mergesort.merge\_sort\_parallel(<data>)
+<sorted> := swapsort.swap\_sort\_parallel(<data>)
+<sorted>
+\vspace{0.1in}
+property: swapsort.is\_sorted(<sorted>)
+\end{code}
+}
+\caption {Complete TSTL harness for checking two parallel sorting
+  algorithms for nondeterminism.}
+\label{fig:parallelcode}
+\end{figure}
+
+Figure \ref{fig:parallelcode} shows the complete TSTL harness for
+checking the parallel sorts.  This harness produces sequences of
+integer pairs, where only the first integer is used in the comparison
+operator of the sorting algorithms, and checks that the results of
+{\tt sort\_parallel} calls are in fact sorted, using the {\tt
+  swapsort}'s function for checking if a sequence is sorted.
 
@@
 to the popular Redis in-memory database, cache, and
-message broker \cite{redis}.
+message broker \cite{redis}.  The TSTL {\tt redis-py} harnesses can be found in
+the {\tt examples/redis} directory in the TSTL GitHub repo \cite{tstl}.
 
@@
 issues clear, but tests using a library are often written by those not
-intimately familiar with the semantics of every used library (otherwise they would not have
+intimately familiar with the semantics of every library call (otherwise they would not have
 been as likely to introduce flaky tests in the first place).  This is
@@
 cases incorrect) results due to dependence on the order of values in
-an iterator over a Python dictionary, on Python versions above 3.2, until 3.6 (see Section \ref{sec:pnondet}).
+an iterator over a Python dictionary, on Python versions above 3.2,
+until 3.6 (see Section \ref{sec:pnondet}).
 
 Figure \ref{hashbug} shows TSTL code for generating inputs to the {\tt
-  datarray} algorithms.
+  datarray} algorithms (it can be found in the {\tt
+  examples/datarray\_inference} directory in the TSTL repository \cite{tstl}).
 The {\tt flatten\_and\_sort} function is needed because we care about actual differences in probability values, not simply the order of list, tuple, or dictionary items.
@@
 The TSTL harness for {\tt pyfakefs} has been used to detect (and
-correct) over 80 faults.  However, the testing largely relies on
+correct) over 80 faults.  The harness can be found in the TSTL GitHub repository
+in the {\tt examples/pyfakefs} directory, and the bugs discovered can
+be viewed at \url{https://github.com/jmcgeheeiv/pyfakefs/issues?q=label%3ATSTL}.  However, the testing largely relies on
 the existence of a reference file system implementation.  One purpose
-of failure nondeterminism is to make it somewhat easier to perform
+of failure nondeterminism is to make it somewhat easier to perform effective
 property-based testing of
-complex APIs like this even without a reference implementation, or in cases
+complex APIs like this even without a complete reference implementation, or in cases
 where the implementations do not use the same error

### 2019-03-07 implementation.tex
value pool formalism, parallel sort example

@@
 the new process will not.
+
+\subsection{Pools and Visible State}
+
+While the formal definitions of nondeterminism proposed in Section
+\ref{sec:formal} offer a framework for checking for nondeterministic
+behavior, they leave unspecified the key notions of \emph{state},
+\emph{visible state}, and \emph{actions}.  For example, we could
+identify the state and visible state of a system as identical with the
+full set of memory locations accessed by that system, and actions with the
+individual processor instructions executed.  However, this
+definition is highly impractical for two reasons.  First, it is
+extremely inefficient to compare multiple executions for
+full-memory-state equivalence at every instruction step.  Second, and
+even more importantly, as noted in the introduction, very few programs
+will behave deterministically when considered at this low level,
+unless the full state of the operating system, processor, hardware
+devices, etc. are all taken as controlled inputs.
+
+What is a practical level of granularity for (visible) states and actions that
+can be efficiently checked, and matches (for the most part) developer
+and tester intuitions about what should behave deterministically?
+
+Andrews et al. formalized a value-pool-based model of unit tests that is widely used in
+actual testing tools, including Randoop \cite{Pacheco}, showing that
+all Java unit tests (and, more generally, unit tests in similar
+languages) can be represented in a \emph{canonical} form, where
+a test consists of three parts:
+
+\begin{enumerate}
+\item A declaration of a set of array variables, referred to as
+  \emph{value pools}, e.g., {\tt int [] intVP = new int[3];}.
+  \item A set of assignments of constant values to elements of
+    primitive type value pools, e.g. {\tt intVP[2] = 1;}.
+    \item A part in which all statements are assignments of calls of a
+      method to a value pool, with all arguments to the method also
+      taken from a value pool, e.g. {\tt intVP[2] = fooObjectVP[2].bar(intVP[0],
+        intVP[1]);} (these are referred to as \emph{array-canonical} statements).
+    \end{enumerate}
+
+Given such a form for unit tests, there is an obvious mapping to
+states and actions in our formalism for nondeterminism:  the state is
+the full state of the system, but visible state is restricted to
+\emph{the values stored in the value pools}.  Actions are the
+array-canonical statements that call code under test to modify the pools.  We can ignore constant assignments and
+declarations, since (assuming the language implementation itself is
+basically deterministic) these cannot introduce nondeterminism.
+Because assertions of correctness and \emph{what a test actually does}
+depend on the values in value pools, it seems likely that developers
+and testers expect determinism in unit tests, at the granularity of
+value pool assignments.
 
@@
 expected to work by modifying the \emph{state} of a test, which is
-stored in a fixed set of \emph{pools} containing Python objects.  The
+stored in a fixed set of pools containing Python objects.  TSTL's
+approach essentially follows the pool-based canonical form defined by
+Andrews et al., but relaxes many of its restrictions
+(e.g., a constant value can be used in a method call without first
+assigning it to a pool). The
 TSTL actions defined by a test harness correspond to the set $A$ of
-actions in Section \ref{sec:formal}, and transitions between states of
-result from carrying out the Python code in an action.  Figure
+actions in Section \ref{sec:formal}, and transitions between states
+result from executing out the Python code in an action; these actions
+are, roughly, the array-canonical statements of the value pool formalism.  Figure
 \ref{fig:simple} shows a very simple TSTL harness that tests Python's

### 2019-03-07 experiments.tex
parallel sorts now

@@
   swapsort}'s function for checking if a sequence is sorted.
+
+Running TSTL's random tester without checking for determinism, we can
+see that both sorting algorithms always produce sorted output.  If we run with {\tt --checkDeterminism --determinismDelay 0}, however, a
+test case showing the {\tt swapsort} behaving nondeterministically is
+almost instantly produced (it takes less than 3 seconds).  If we comment out the call to {\tt
+  swap\_sort\_parallel} and only check the parallel merge sort, however, TSTL
+finds no nondeterministic behavior.  Running tests of the merge sort
+with nondeterminism detection turned on results in a slowdown of approximately
+40\%.
+
+Reducing a lengthy test showing nondeterminism of the swapsort to a
+short test case using delta-debugging usually takes less than two
+minutes.  If the number of attempts to produce nondeterminism is
+increased in the reducer using the {\tt --determinismTries 10} option,
+the resulting test will be usually be very compact, e.g.:
+
+{\scriptsize
+\begin{code}
+data0 = []    
+val0 = 3      
+val1 = -9    
+data0.append((val0,val1)) 
+val2 = -1                         
+data0.append((val1,val2))
+data0.append((val1,val1))  
+sorted0 = swapsort.swap\_sort\_parallel(data0)      
+\end{code}
+}
+
+For this simple example, more sophisticated delta-debugging strategies
+are not needed; we only want to see the code behaving
+nondeterministically, and do not care about preserving a higher probability of nondeterminism.
 

### 2019-03-07 introduction.tex
value pool formalism, parallel sort example
parallel sorts now

@@
 This paper proposes (1) a number of formal definitions of types of
-nondeterminism (\emph{horizontal} and \emph{vertical}) and (2) an implementation, based on these definitions, for detecting and debugging
+nondeterminism (\emph{horizontal} and \emph{vertical}) (see Section \ref{sec:formal}) and (2) an implementation, based on these definitions, for detecting and debugging
 nondeterminism in property-based testing.  The implementation is based
 on an approach where (3) horizontal determinism is considered as a
-kind of \emph{reflexive differential testing} and (4) vertical
-determinism is specialized to the common case of \emph{failure determinsim}.
+kind of \emph{reflexive differential testing} (4) vertical
+determinism is specialized to the common case of \emph{failure
+  determinism}, and (5)  in both cases the formalism is made practical
+by using the \emph{value pool} model of unit tests to define a useful
+granularity for nondeterminism detection (see Section \ref{sec:pools}).
 We further propose modifications to the widely used delta-debugging algorithm in order to better handle
@@
 We use a set
-of case studies of real-world, widely-used, Python libraries to demonstrate the
+of case studies. including real-world, widely-used, Python libraries to demonstrate the
 utility of our ideas.  Failure determinism directly

### 2019-03-08 experiments.tex
threats
threats and typos

@@
 too small to be of any real practical importance.
+
+\section{Threats to Validity}
+
+There are several primary threats to validity:  first, our empirical
+results are limited to a small set of Python programs, ranging from
+relatively small and simple to large and complex libraries; the
+representative nature of these subjects is not clear.
+Furthermore, because
+no other tools implement the kind of approach taken here, based on
+automated generation of value pool based unit tests for a system
+(imperative property-based testing), to our knowledge, we were unable
+to perform a meaningful comparison with another nondeterminism
+detection tool.  Available tools, such as DeFlaker
+(\url{http://www.deflaker.org/}), perform a completely different task (DeFlaker
+is essentially a plugin for Maven regression tests, and cannot check
+nondeterminism at any smaller granularity than that of a whole tests's
+pass/fail result) and target the Java language.  However, the primary
+aim of our results is to show 1) that nondeterminism at the library
+level can exist in real Python programs, 2) that it can be detected
+by an implementation of the formalism proposed in this paper, and 3)
+that the overhead for such detection, and cost to delta-debug
+nondeterministic tests, is not prohibitive.  We believe that the case
+studies show that users testing Python libraries, and concerned about
+nondeterminism, would find the TSTL tools useful, and would be able to
+detect and debug nondeterministic behavior using our approach, a
+functionality that we believe to be unique in the automated test generation
+literature (and supported in no other tool).
+
+Another threat to validity is posed by implementation errors.
+TSTL and the mutation tool are tested for bugs by a set of Travis CI
+tests, and are (for academic research code) moderately well-used by
+external users, making this a lesser concern.  The nondeterminism
+detected by TSTL can be demonstrated with standalone tests TSTL
+generates that do not depend on any TSTL code.
+The source code for the tools and all case studies is
+available for inspection and replication in the TSTL repository.

### 2019-03-08 experiments.tex
improve threats

@@
 to perform a meaningful comparison with another nondeterminism
-detection tool.  Available tools, such as DeFlaker
+detection tool.  Available tools, such as DeFlaker \cite{bell2018d}
 (\url{http://www.deflaker.org/}), perform a completely different task (DeFlaker
@@
 nondeterminism at any smaller granularity than that of a whole tests's
-pass/fail result) and target the Java language.  However, the primary
+pass/fail result) and target the Java language.  iDFlakies
+\cite{idflakies}, similarly, ``does not further classify the causes of
+the flaky tests'' beyond identifying them as likely related to test
+order, or not.
+
+However, the primary
 aim of our results is to show 1) that nondeterminism at the library
@@
 functionality that we believe to be unique in the automated test generation
-literature (and supported in no other tool).
+literature (and supported in no other tool).  Comparison with a
+competing tool is not required to validate the basic idea, since
+precise empirical measurements of overheads, or improvements on an
+existing approach, are not proposed for evaluation.
 

### 2019-03-20 response.txt
ready to send to TR

Dear associate editor and reviewers,

We would like to thank the reviewers for their helpful comments.  We have addressed the comments, and submit a (much extended) revised version of the paper.

At the high-level, this revision provides the following major changes:

- We generally, throughout the paper, clarified the (obviously not clear in the first version) goal of our work, and how it relates to existing work on flaky-test detection.  In particular, we explain why comparison to all existing tools in the literature we are aware of is difficult or impossible due to fundamentally different purposes of the tools.

- We added a full formal definition of our concepts in II.D, and used this to clarify a large number of points, including the details and motivation of our implementation and key challenges

- We added sections IV.B and IV.C on the connection between TSTL's pool approach to test generation and the formalisms, and a short introduction to the TSTL language and tools

- We added Section V, a simple complete example of both horizontal and vertical determinism that should make the concepts and implementation of the approach much more clear to readers

- We added one additional case study, showing a parallel sort implementation that preserves output determinism and comparing to one that does not

- We added further detail and experimental results to the existing redis-py case study

- We added a section on threats to validity

- We provided more complete information on replicating the results: TSTL and the nondeterminism tools, and the case studies, were all open source and hosted on GitHub as of the first submission (TSTL itself has been on GitHub since 2015, the nondeterminsm tools were added to the public code and the pip release before the paper was submitted); however the paper did not make this clear at all.  How to install and use the nondeterminism detection tools should now be clear.

- We expanded related work to further clarify the position of this work, and compare to recent work/tools on flaky tests

These changes are mostly motivated by reviewer comments.  We also answer specific reviewer questions below.

Sincerely,
Alex Groce and Josie Holmes,
Northern Arizona University

=================================================================

Reviewer 1:

"How many times is the system re-executed, and how was this count decided on? When TSTL compares the values (intermediate program state) to detect non-determinism, how does it know to ignore irrelevant values?"

The default is to execute only one additional time, but this is fully under the control of the user, specified using the command line interface to tools, or using the API when programmed checking for nondeterminism is needed.  Adding OPAQUE annotations is probably easiest done in a simple feedback-driven manner:  initially run without any such annotations, produce (and delta-debug, if needed) tests showing nondeterminism, and mark pools as OPAQUE as needed.  Many cases, such as timestamps, will probably be obvious to users up front when developing a test harness.

"How much effort is it for a developer to find and mark these opaque values? How many false positives occur if this step is skipped or done incorrectly?"

The answer will greatly depend on the example.  In our experience, most harnesses do not have a huge number of pools, so that limits effort in general.  Many libraries will have no pools needing an OPAQUE label; even if there are logs, they may well not be stored in a pool for automated test generation.

"How is ASLR a representative source of non-determinism? Aren’t these memory bugs VERY very easily checked (especially checking for reads of uninitialized memory)?"

Valgrind etc. (or compiling with sanitizers if that's possible) often make this easy; ASLR is simply an example of process-level nondeterminism.  That said, in Python, such memory-nondeterminism often takes place at the level of C-implemented libraries, where recompiling may be difficult, and the overhead of running with Valgrind etc. may be prohibitive for automated test generation (Valgrind overhead is much larger than our overhead, though of course the detection is better!)  We believe the section on sources of nondeterminism is now somewhat more clear.

"How does TSTL know what the probabilities are of non-deterministic behavior, and moreover, where those sources of non-determinism are? Is it important that this be correctly estimated?"

TSTL knows nothing about nondeterminism other than what the formalism of differing pool values on re-execution tells it.  There is no need for estimating probabilties in detection -- either values are the same, or they are not.  There may be some need for users to know if nondeterminism is highly unlikely, and they need to re-run tests a large number of times, but this is outside the scope of TSTL, and really reflects the user's "tolerance" for possibly missing nondeterminism.  Delta-debugging automatically will retain the operations that produce (or are required for) nondeterminism, without any knowledge; it simply finds a subset of a test that retains the property of interest (exhibiting nondeterminism).

"How representative are the case study bugs of all non-determistic bugs?"

We think these capture major categories of nondeterminism, but in some sense each instance is idiosyncratic.  Performing a large-scale empirical study is difficult because only some libraries have a potential problem with nondeterminism, and a user must develop a TSTL test harness (to check for correctness, not nondeterminism) for each library.

The evaluation-related questions are not all answered, but we believe the paper now makes clear why we focused on the questions we did, and what the purpose of this work is.

=================================================================

Reviewer 2:

"As the authors recognize in the related work section, several papers have been studying automated solutions to deal with nondeterminism (e.g., all the methods able to locate flaky tests): can those previous techniques be employed instead of the one proposed in the paper? If so, how can they be adapted and how do they work when compared to the technique proposed in the paper?"

We believe the place of our work in the larger context of flaky-test work should now be clear, and how it differs fundamentally from, e.g. DeFlaker etc.

"For example, a simple yet widely used detection technique is to execute a test a number of times (e.g., ten times) and see whether the outcome from these executions is different. This aspect of flakiness may mean that checking the output values by means of differential testing might be not enough for detecting them all."

In a way, this is exactly what our implementation is doing, with the number of re-executions controlled by the user, except the process of fine-grained comparison (vs. pass/fail which may need many more runs) is automated.

As noted above, the tool is available on github at https://github.com/agroce/tstl; every example is found in the TSTL examples directory, and replicating our results should be a very simple matter of running the tool on these examples.

"For instance, is the technique a valid option for being adopted in a continuous integration pipeline? If so, why? If not, could you please specificate how and in which development moment a practitioners should use it?"

If you are using TSTL random testing in CI (which TSTL itself does, for example), then this is trivial to add to CI.  In general, the approach is simply another aspect of development, test execution, and debugging for applying automated test generation to Python libraries with TSTL.

"For instance, the authors use ‘universalmutator’ for mutation analysis: why this choice rather than more widely used tools like, e.g., Pitest?"

Pitest doesn't handle Python code.  To our knowledge, universalmutator is the best-supported mutation tool for Python available right now that supports scripted examination of the produced mutants.  The cosmic-ray project has become increasingly popular since we started this project, but seems to focus on running deterministic tests using a Python unit test framework, not interacting with TSTL or other command-line tools.

"Is the set of mutants generalizable? What is the percentage of equivalent mutants present in the case study (or at least an estimation - the authors cannot determine precisely this information)?"

The mutants will differ by mutation operator, but we expect the percent change in detected mutants to be similar.  The percentage of equivalent mutants is not known, though we used Trivial Compiler Equivalence on Python bytecode to throw out some of these.  The interest here is in additional mutants detected, so the absolute numbers of not percentages, should be independent of the equivalent mutant numbers.

"The same lack of details is true when considering the selection of the systems used in the case studies (e.g., why are they suitable for the paper?)."

They were chosen because redis-py is an example of a complex test harness that involves a system with limited nondeterminism that must be removed to produce regression tests, and pyfakefs is a good example of the kind of actual complex system where vertical nondeterminism is relevant. The other examples were more chosen to show particular kinds of nondeterminism.

=================================================================

Reviewer 3:

No specific questions here, but the issues about readability and references directed out general re-write and reorganization.

### 2019-05-09 experiments.tex
prep for JSS
jss submission

@@
 \centering 
-\begin{subfigure}{0.67\columnwidth}
+\begin{subfigure}{0.3\columnwidth}
 \centering
@@
 \end{subfigure}
-\begin{subfigure}{0.67\columnwidth}
+\begin{subfigure}{0.3\columnwidth}
 \centering
@@
 \begin{figure*}
+{\scriptsize
 \begin{code}
@@
 \end{code}
+}
 \caption{{\tt redis-py} calls removed or changed to avoid
@@
 
-\begin{figure*}
+\begin{figure}[t]
 {\scriptsize
@@
 \{Exception\} <event> := datarray.DataArray(psplit(<P>), axes = [<ename>])
-\{Exception\} <ename,1>!=<ename,2> -> <event> := [datarray.DataArray([[psplit(<P>)],psplit(<P>)],[<ename>,<ename>])]
+\{Exception\} <ename,1>!=<ename,2> -> <event> := [datarray.DataArray([[psplit(<P>)],
+   psplit(<P>)],[<ename>,<ename>])]
 <cpts> := []
@@
 \vspace{0.1in}
-\{Exception\} print(flatten\_and\_sort(inference\_algs.calc\_marginals\_simple(<cpts>,<evidence>)))
-\{Exception\} print(flatten\_and\_sort(inference\_algs.calc\_marginals\_sumproduct(<cpts>,<evidence>)))
-\{Exception\} print(flatten\_and\_sort(inference\_algs.calc\_marginals\_jtree(<cpts>,<evidence>)))
+\{Exception\} print(flatten\_and\_sort(inference\_algs.calc\_marginals\_simple(<cpts>,
+   <evidence>)))
+\{Exception\} print(flatten\_and\_sort(inference\_algs.calc\_marginals\_sumproduct(<cpts>,
+   <evidence>)))
+\{Exception\} print(flatten\_and\_sort(inference\_algs.calc\_marginals\_jtree(<cpts>,
+   <evidence>)))
 \end{code}
@@
 \label{hashbug}
-\end{figure*}
+\end{figure}
 

### 2019-08-15 concepts.tex
prep for JSS
start on revisions for JSS

@@
 \xrightarrow[]{\alpha^2_{n-1}} s^2_n$ are said to show \emph{visible value
-  determinism} if $\exists i > 0$ such that: $\forall j < i . \alpha^1_j = \alpha^2_j \wedge
+  nondeterminism} if $\exists i > 0$ such that: $\forall j < i . \alpha^1_j = \alpha^2_j \wedge
 v(s^1_j) = v(s^2_j)$ but $v(s^1_i) \neq v(s^2_i)$.  \emph{Final state
@@
 $\alpha_i \in \mathcal{I} \wedge \alpha_i = \alpha_{i+1}$ and $v(s_{i+1})
-\neq v(s_{i+2})$.
+\neq v(s_{i+2})$.  Note that vertical nondeterminism may exhibit only
+after a sequence of more than two successive applications of a (supposedly)
+idempotent operation; the definition only requires that after \emph{some}
+sequence of actions, possibly with $\alpha_{i-1} = \alpha_i$, the
+visible state changes.
 
@@
   nondeterminism} if $\exists i$ such that $F_{i+1}$ (that is,
-$\alpha_i$ \emph{fails}) and either (1) $v(s_{i}) \neq v(s_{i+1})$ or (2)
-$\alpha_i = \alpha_{i+1} \wedge \neg F_{i+2}$.  The second possibility
-captures our more lightweight approach to checking for failure
-nondeterminism.  Note that unlike all of the other ways to demonstrate
-nondeterminism, this second type of failure nondeterminism does not
+$\alpha_i$ \emph{fails}) and either (1) $v(s_{i}) \neq v(s_{i+1})$
+(the visible state changes) or (2)
+$\alpha_i = \alpha_{i+1} \wedge \neg F_{i+2}$ (the same action is
+repeated, but this time does not fail).  The second possibility
+offers a \emph{lightweight} approach to checking for failure
+nondeterminism.  Unlike all of the other ways to demonstrate
+nondeterminism we consider, this second type of failure nondeterminism does not
 require the use of $v$, or even the notion of observable components of
-a state, or (for that matter) any comparison of states.
+a state, or (for that matter) any comparison of states.  It only
+requires that we be able to say whether an operation failed.
 
@@
 size of failing tests, making them easier to understand and debug.
+The core idea of delta-debugging is to take a test that satisfies some
+property (usually the property is ``it fails in a particular way'')
+and produce a smaller test that satisfies the same property.
+Delta-debugging, as proposed by Zeller and Hildebrandt, relies on a
+modified binary search to produce this smaller test, though the
+following discussion relies only on the general structure of
+delta-debugging, and also applies to the greedy algorithms used to
+reduce failing tests in some work \cite{CReduce,onetest}.
+
 Delta-debugging in the context of detecting nondeterminism has two
@@
 \centering 
-\begin{subfigure}{0.67\columnwidth}
+\begin{subfigure}{0.30\columnwidth}
 \centering
@@
 \end{subfigure}
-\begin{subfigure}{0.67\columnwidth}
+\begin{subfigure}{0.30\columnwidth}
 \centering
@@
 \end{subfigure}
-\caption{Delta-debugging of the same {\tt redis-py} test}
+\caption{Delta-debugging of the same simple test}
 \end{figure*}

### 2019-08-26 concepts.tex
work on some issues

@@
 delta-debugging, and also applies to the greedy algorithms used to
-reduce failing tests in some work \cite{CReduce,onetest}.
+reduce failing tests in some work \cite{CReduce,onetest}.  At a high
+level delta-debugging and other test reduction/shrinking algorithms
+can be seen as following a general algorithmic structure.  Given a
+test case, $T_{\mathit{INIT}} : \mathit{test}$ and a predicate $\mathit{PRED}: \mathit{test} \rightarrow \mathit{bool}$
+
+\begin{enumerate}
+\item $T_{\mathit{CURR}} = T_{\mathit{INIT}}$
+\item Let $T_{\mathit{NEXT}} =$ a variation of $T_{\mathit{CURR}}$
+  that has not yet been tried.  If there are no further variations of
+  $T_{\mathit{CURR}}$, stop and return $T_{\mathit{CURR}}$ as the
+  reduced test.
+\item Add $T_{\mathit{NEXT}}$ to the set of variations that have been
+  tried.
+\item If $\mathit{PRED}(T_{\mathit{NEXT}})$, set $T_{\mathit{CURR}} =
+  T_{\mathit{NEXT}}$.
+\item Go to 2.
+\end{enumerate}
+
+The behavior of a particular algorithm for reducing a test can usually
+be specified simply be elaborating the construction and ordering of
+the variations considered.  For our purposes, the essential point is
+to note that any such algorithm will apply $\mathit{PRED}$ to a
+potentially very large number of tests.
 

### 2019-08-29 concepts.tex
changes to formalism/intro largely done

@@
 
-{\bf If a call fails, and indicates this to the caller, it should not modify system
-  state in any way.  Changes made before the failure should be rolled
-  back.}
+{\bf If a call \emph{fails}, and indicates this to the caller, it should not modify system
+  state in any way; changes should be ``rolled back.''}
 
@@
 (e.g., exceptions in Python and Java, or Option/Result types in Rust,
-Haskell, and ML), failure determinism can be automatically checked.
-Moreover, the expected overhead should be even lower than the general
+Haskell, and ML), failure determinism can be automatically checked,
+given a notion of general vertical determinism.
+Moreover, the expected overhead should be even lower than the usual
 overhead for vertical determinism, in that we can expect most failing
@@
 failure.  Note that this more restrictive notion of failure
-nondeterminism captures such issues as the Apple login bug.
+nondeterminism captures such issues as the Apple login bug.  The
+changes to visible or hidden state we are most concerned with are
+usually those related to the failure itself.
 
@@
 
+The concept of an action does not restrict this approach to reactive
+systems; it can be generalized to consider the only action that varies
+to be the selection of an input, $\alpha_0$, with the remainder of the
+actions being internal $\tau$ actions, in the usual fashion.
+
+\subsubsection{Horizontal Determinism}
+
 A pair of traces $(t_1, t_2)$ where $t_1 = s^1_0
@@
 
+A point to emphasize is that while horizontal determinism
+requires two traces, it does not require a test generation algorithm
+to \emph{produce} pairs of traces, which would require substantial
+modification of most such tools (e.g., random testers or fuzzers, and
+even model checkers).  In general, testing tools do not produce
+\emph{traces}, they produce \emph{action sequences} or \emph{inputs}.  The two traces
+used to show nondeterminism, by definition, have the same action
+sequences, which means that a testing tool only needs to search for an
+action sequence or input which, when executed twice, produces a $t_1$ and $t_2$
+meeting the requirements for showing nondeterminism ---
+non-equivalence of $v(s)$ for some $s$.
+
+\subsubsection{Vertical Determinism}
+
 Vertical nondeterminism, in contrast, is demonstrated by a single trace
@@
 $\alpha_i \in \mathcal{I} \wedge \alpha_i = \alpha_{i+1}$ and $v(s_{i+1})
-\neq v(s_{i+2})$.  Note that vertical nondeterminism may exhibit only
+\neq v(s_{i+2})$.
+
+Note that vertical nondeterminism may, in theory, exhibit only
 after a sequence of more than two successive applications of a (supposedly)
@@
 sequence of actions, possibly with $\alpha_{i-1} = \alpha_i$, the
-visible state changes.
+visible state changes.  In an implementation, we would usually restrict the
+search to a finite number of repetitions; for many faults, two
+``copies'' will suffice, since an operation is either truly idempotent or
+not idempotent.  Only the existence of \emph{non-visible} state changes that
+can eventually produce visible state changes \emph{after a succession of the
+same action} require more repetitions.
+
+\subsubsection{Failure Determinism}
 
@@
 transition relation to include a notion of \emph{failure}: $T \subseteq S
-\times A \times S \times {\tt bool}$, where the boolean indicates
-whether the action \emph{failed}.  A trace $t = s_0
+\times A \times S \times {\tt F: bool}$, where the boolean $F$ indicates
+whether the action $A$ \emph{failed}.
+
+A trace $t = s_0
 \xrightarrow[]{\alpha_0} (s_1, F_0) \xrightarrow[]{\alpha_1} \ldots
-\xrightarrow[]{\alpha_{n-1}} (s_n, F_n)$ shows \emph{failure
-  nondeterminism} if $\exists i$ such that $F_{i+1}$ (that is,
+\xrightarrow[]{\alpha_{n-1}} (s_n, F_{n-1})$ shows \emph{failure
+  nondeterminism} if $\exists i$ such that $F_{i}$ (that is,
 $\alpha_i$ \emph{fails}) and either (1) $v(s_{i}) \neq v(s_{i+1})$
 (the visible state changes) or (2)
-$\alpha_i = \alpha_{i+1} \wedge \neg F_{i+2}$ (the same action is
+$\alpha_i = \alpha_{i+1} \wedge \neg F_{i+1}$ (the same action is
 repeated, but this time does not fail).  The second possibility
@@
 nondeterminism we consider, this second type of failure nondeterminism does not
-require the use of $v$, or even the notion of observable components of
-a state, or (for that matter) any comparison of states.  It only
+require the use of $v$, the notion of observable components of
+a state, or any comparison of states.  It only
 requires that we be able to say whether an operation failed.
-
-A further point to emphasize is that while horizontal determinism
-requires two traces, it does not require a test generation algorithm
-to produce pairs of traces, which would require substantial
-modification of most such tools (e.g., random testers or fuzzers, and
-even model checkers).  In general, testing tools do not produce
-\emph{traces}, they produce \emph{action sequences}.  The two traces
-used to show nondeterminism, by definition, have the same action
-sequences, which means that a testing tool only needs to search for an
-action sequence which, when executed twice, produces a $t_1$ and $t_2$
-meeting the requirements for showing nondeterminism ---
-non-equivalence of $v(s)$ for some $s$.
 

### 2019-08-29 concepts.tex
fix through three maybe

@@
 can be seen as following a general algorithmic structure.  Given a
-test case, $T_{\mathit{INIT}} : \mathit{test}$ and a predicate $\mathit{PRED}: \mathit{test} \rightarrow \mathit{bool}$
+test case, $T_{\mathit{INIT}} : \mathit{test}$ and a predicate
+$\mathit{PRED}: \mathit{test} \rightarrow \mathit{bool}$, we
+\emph{reduce $T_{\mathit{INIT}}$ with respect to $\mathit{PRED}$} as follows:
 
@@
 knowledge, not previously investigated) issue:  using delta-debugging
-to minimize a test with respect to a predicate that only holds with a
+to minimize a test with respect to a predicate $\mathit{PRED}$ that only holds with a
 certain probability (the predicate itself is nondeterministic).  Note
@@
 predicate remains true of the shorter versions of the test.  The
-traditional predicate is ``the test fails'' but other predicates,
+most common $\mathit{PRED}$ is still ``the test fails'' but other predicates,
 especially those related to code-coverage, can also be useful
 \cite{icst2014,stvrcausereduce,NonAdeq}.  Our contribution here is to consider
-what happens when the predicate to be evaluated is not deterministic.
-``The test behaves flakily'' is a simple, important, example of such a predicate.
+what happens when $\mathit{PRED}$ not deterministic.
+``The test behaves flakily'' is a simple, important, example of such a $\mathit{PRED}$.
 
@@
 common.  In these cases, when we use a reduction algorithm to reduce $t$
-to $r$, $P(pred(r)) \leq P(pred(t))$.  In \emph{non-monotonic} cases,
+to $r$, $P(\mathit{PRED}(r)) \leq P(\mathit{PRED}(t))$.  In \emph{non-monotonic} cases,
 however, there is no such upper bound.  Removing a step in $t$ may
@@
 
-The probabilistic predicate of interest for determinism is basically of
+The $\mathit{PRED}$ of interest for horizontal determinism is basically of
 the form:  {\bf the test will exhibit at least two different behaviors in $S$
@@
 implementations (including the Python code provided by Zeller) assert
-that the predicate holds on the original test.  In probabilistic
+that $\mathit{PRED}$ holds on the original test.  In probabilistic
 settings, this may not be true, and we may even be trying to find a
@@
 
-However, simply using delta-debugging off the shelf with a predicate
-like $P(fail) > 0.3 \wedge P(fail) < 1.0$, to force a test to be flaky, and force it to fail
+However, simply using delta-debugging off the shelf with a $\mathit{PRED}$
+such as $P(fail) > 0.3 \wedge P(fail) < 1.0$, to force a test to be flaky, and force it to fail
 sufficiently often to be used in debugging, will often produce
@@
 convert a nondeterministic test to a completely deterministic test,
-because the reduced test that delta-debugging returns is always one
-where the predicate of interest has been seen to evaluate to true.
+because the reduced test $r$ that delta-debugging returns is always one
+such that $\mathit{PRED}(r)$ has evaluated to true at least once.
 However, if you run the resulting test, it will, in many cases, have a
@@
 The problem is analogous to the problem of publication bias in
-scientific fields \cite{ahmed2012assessment}.  A predicate like
+scientific fields \cite{ahmed2012assessment}.  We can think of each
+application of $\mathit{PRED}$ to a candidate variant
+$T_{\mathit{NEW}}$ as a scientific experiment.  A predicate like
 $P(fail) > 0.3 \wedge P(fail) < 1.0$ cannot be evaluated by
-determining the true probability of failure; rather, the test must be
+determining the true probability of failure, we have no more direct access to
+this value than a medical researcher has access to the true effect
+size and direction of a proposed treatment; rather, the test must be
 run some concrete number of times, and the number of failures counted.
@@
 candidate tests.  The predicate is evaluated on each of these, and so
-even with large $N$, it is extremely likely that some evaluation will
+even with large $N$, it is extremely likely that \emph{some} evaluation will
 produce a very poor estimate of $P(fail)$.  If such an evaluation
@@
 ``stuck'' with the error, because (to our knowledge) no test reduction
-algorithms allow backtracking.  After such a mistake, finding further
+algorithms allow backtracking (in normal, non-probabilistic settings,
+backtracking would not usually be useful).  After such a ``mistake,'' finding further
 reductions will become harder, but may still be possible due to the
-same source of errors:  the number of experiments run is far larger
-than the probability of an incorrect result.
+same source of errors:  \emph{the number of experiments run is far larger
+  than the probability of an incorrect result for a single
+  experiment}.  We define a \emph{false positive} in probabilistic
+delta debugging as a case where $\mathit{PRED}$ evaluates to true, but
+the actual probabilities required by $\mathit{PRED}$ do not hold, and
+a false negative to be the reverse, where ``bad luck'' makes it appear
+that the test does not satisfy $\mathit{PRED}$, but it does in fact do
+so.  A false negative's consequences are limited, since there will
+likely be another similar variation for which  $\mathit{PRED}$ will
+evaluated to true, but a false positive changes $T_{\mathit{CURR}}$.
+
+The analogy to publication bias is simple.  Assume that a paper in,
+e.g., a medical journal will only be published if it shows that a
+treatment is effective, with the statistical requirement that $p <
+0.05$ (that is, roughly, by some statistical test, the probability
+that the result is due to chance rather than reflecting reality, is
+less than 5\%).  Even if no researchers are dishonest, 5 of every 100 papers published
+in the journal will be report an effect with the wrong effect size or
+even effect direction!  Now consider a ``hot'' topic, a kind of
+treatment many research teams may investigate.  Major journals tend
+not to publish papers that say ``this new treatment, which is not
+established, may not work,'' (especially if the result is not ``the
+treatment certainly does not work'' but only ``the experiment did not
+establish $p < 0.05$''), but they  are very likely to publish a paper
+that says ``this exciting new treatment works!''  In addition to
+sometimes producing excitement about treatments that do not work, this
+bias tends to magnify the expected effect size of treatments, since
+the chance of getting a ``passing'' $p$-value is higher when an
+experiment finds a larger effect size, due to how most statistical
+tests work.
+The \emph{bias} in
+favor of positive results is shared by delta-debugging;
+delta-debugging's behavior is much more influenced by cases where
+$\mathit{PRED}$ holds than cases where it does not.  Looking at the
+literature, the unpublished papers are invisible, magnifying the
+effect on scientific opinion of the published experiments.
 
@@
 use of replications:  repeated runs of ``successful'' experiments to
-minimize the probability that a result is a fluke due to publication
+minimize the probability that a direction or effect size is a fluke due to publication
 bias.  One way to produce this effect would be to allow
@@
 the number of expected predicate evaluations performed during
-delta-debugging.  If the error rate is low enough, the problem
-disappears.  However, given the large number of evaluations performed,
+delta-debugging.  However, given the large number of evaluations performed,
 this will tend to make delta-debugging extremely slow.  We propose
@@
 required before the predicate returns ``true'' to the delta-debugging
-algorithm.  A set of $M$ repeated false positives with
+algorithm.  To evaluate the predicate $\mathit{PRED}$ given $N$,
+$M$, and desired probability threshold $p$, the algorithm is simple:
+
+\begin{enumerate}
+  \item  $i = 0$
+  \item Evaluate $\mathit{PRED}$ $N$ times.  If the number of cases
+  where $\mathit{PRED}$ was true is less than $p$, return false.
+  \item  $i = i + 1$
+  \item If $i < M$ go to step 1.
+  \item  Return true.
+\end{enumerate}
+
+A set of $M$ repeated false positives with
 $\frac{N}{M}$ samples each is much less likely than a false positive
@@
 desired accuracy for $P$ can be obtained with a much smaller value $M
-\times N$ than a non-dynamically-sampled $N$.  For example, if we the predicate
+\times N$ than a non-dynamically-sampled $N$.  For example, if $\mathit{PRED}$
 is $P(fail) \geq 0.5$, and the true probability for a candidate test

### 2019-08-30 concepts.tex
all probably ok up to experiments restructure

@@
 
+\section{Sources of Nondeterminism in Practice}
+\label{sec:sources}
+
+In order to implement a practical nondeterminism-detection tool, it is
+important to consider the real-world sources of nondeterminism.  Most of these
+sources can, in principle, be detected simply by re-running a test
+twice and comparing results, either at all visible values or at the
+final state.  For example, if some library call uses a random number
+generator (perhaps in a stochastic algorithm), the returned value should be
+different on a second run.  Timing dependencies may change on a second
+execution, especially if a delay is added between steps of the test (a
+more principled version of the unfortunate tradition of testing
+concurrent code by adding ad-hoc {\tt sleep} statements).   Most
+nondeterministic behavior in tests is probably attributable to such
+effects that can vary with a repeated execution, in the same
+(approximate) enviroment, of the same test.  The two broad classes of
+nondeterminism that can usually be detected (with some probability) by
+simply executing the same test twice in succession, without attempting
+to control any other factors, are:
+
+\begin{itemize}
+\item {\bf External Environment:} This class of nondeterminism arises
+  when a test's behavior depends on factors outside the program under
+  test, which are subject to uncontrolled variance.  Calling {\tt
+    random()} in Python is an obvious example; the call is almost
+  guaranteed to return a different value each time it is called in a
+  test, unless we explicitly re-seed the random number generator at
+  the beginning of each test.  Similarly, calls to {\tt time}
+  functions  will usually return different values on different runs,
+  and even if only elapsed time, rather than absolute, is used to
+  influence behavior, subtle variances in timing are to be expected
+  due to system load, I/O, and other external factors.  Finally, in
+  distributed system testing, the environment is often remote servers
+  (or clients) that may be unavailable, slower than expected, or
+  subject to faulty behavior for reasons not under the control of the
+  test.
+\item {\bf Concurrency:} Perhaps the second most common cause for
+  nondeterminism in tests is when an algorithm is implemented using
+  concurrency, in the (incorrect) belief that the result of the
+  algorithm remains deterministic.  E.g., consider a multi-threaded
+  implementation of a search algorithm, that returns the location of
+  an item in an unsorted list.  If the list may contain duplicates,
+  and a developer has not accounted for this possibility, the
+  algorithm may return different locations of an item for the same list,
+  depending on the scheduling of threads.  Arguably, this can be
+  subsumed under the first class, with the behavior of the
+  processor/scheduler as the ``enviromental'' factor; relying on
+  scheduling is, in a sense, equivalent to calling {\tt random} or
+  {\tt time}.
+\end{itemize}
+
+Lam et al. \cite{idflakies} note that flaky tests not due to order
+dependence are generally due to ``concurrency, timeouts,
+network/IO, etc.'' a description that is well accounted for by these two
+primary sources, and most of which can be detected, at least in
+principle, by multiple runs of a test within the same testing process.  However, simply running a test twice in the same process (possibly with a delay between steps) does
+not suffice in all cases.
+
+\subsection{Process-Based Nondeterminism}
+
+\label{sec:pnondet}
+
+Some sources of nondeterminism unfortunately require executing a test in a new
+process environment, because the source is inherently tied to the
+process in which code runs; simply re-executing a test in the same
+process will not reveal the problem.  In terms of our formalism, an
+action needs to be introduced that indicates the generation of a fresh
+process identity.
+
+Address Space Layout Randomization (ASLR) \cite{ASLR}  is probably the most
+important source of nondeterminism that arises (only) from change in
+process.  ASLR scrambles the layout of memory of the process in which
+an executable runs in order to make it harder to exploit memory-safety
+vulnerabilities in code.  As a side effect, it
+means that a test that, in some circumstances, causes a crash, may at
+other times not fail at all.  Even if no inputs produce a crash,
+however, memory errors may produce variation in values that are
+overwritten or arise from uninitialized memory, and thus be detectable
+as nondeterminism.
+
+Another example of process-based nondeterminism was discovered by numerous Python developers when Python
+version 3.3 introduced automatic random salting of hashes on a
+per-process basis (a new salt chosen for each Python process),
+in order to mitigate hash-based denial of service attacks
+\cite{denial}.  Until version 3.6 this not only resulted in changes in
+exact hash values, but in the order of iteration on
+dictionaries\footnote{In 3.6, the default dictionary
+  implementation became an ordered dictionary with consistent
+  iteration, though actual hashes remained nondeterministic.}.  While
+relatively few programs rely on exact hash values, many implicitly
+relied on them in that they only functioned correctly with a
+predictable order for dictionary iteration.  Testing
+Python code for nondeterminism based on this hash seed requires
+running in a new process:  Python does not allow changing the salt, since changing hash values on-the-fly would
+break all existing dictionaries and other structures relying on
+hashing.
+
+Other, less common, effects tied to process also exist.  A few tests
+may somehow depend on their actual PID (Process ID) or at least the
+parent process' ID.  Process change increases the likelihood that on a multi-core
+system a test will execute on a different CPU than in the first
+execution, which can have many subtle effects, mostly related to
+changed timing.
+
+One common cause of flaky tests is \emph{test order dependence} \cite{idflakies}.  This
+is less of a concern in our setting, since many problems due to test
+order are easily exposed in random testing, since the
+order-dependence-inducing operations will seldom usually happen in the
+same order in randomly generated tests.  For example, consider the
+case where a test usually run ``early'' initializes an uninitialized
+value in a system, and a later test will only fail if it somehow is
+run before the initializing test.  In random testing, the order of
+initialization/access is likely to be random, and some run of a random
+tester is likely to expose the problem as a (deterministic, not flaky) failure.
+
+However, for cases where
+there is an order dependence that is hard to translate into an
+actual bug, using a fresh
+process may expose such dependencies --- the ``environment'' of the
+parent process will have the, e.g., flag set or value initialized, but
+the new process will not.
+
 \section{Nondeterminism and Delta-Debugging}
@@
   \item  $i = 0$
-  \item Evaluate $\mathit{PRED}$ $N$ times.  If the number of cases
+  \item Evaluate $\mathit{PRED}$ $N$ times.  If the proportion of cases
   where $\mathit{PRED}$ was true is less than $p$, return false.
@@
 A set of $M$ repeated false positives with
-$\frac{N}{M}$ samples each is much less likely than a false positive
-with $N$ samples; so long as we accept the resulting bias in favor of
+$N = \frac{K}{M}$ samples each is much less likely than a false positive
+with $K$ samples; so long as we accept the resulting bias in favor of
 false negatives, we can therefore produce a reduced test with a

### 2019-08-30 implementation.tex
parallel sorts now
prep for JSS
paper
jss submission
all probably ok up to experiments restructure

@@
 
-\subsection{Sources of Nondeterminism in Practice}
-
-In order to implement a practical nondeterminism-detection tool, it is
-important to consider the sources of nondeterminism.  Most of these
-sources can, in principle, be detected simply by re-running a test
-twice and comparing results, either at all visible values or at the
-final state.  For example, if some library call uses a random number
-generator (perhaps in a stochastic algorithm), the value should be
-different on a second run.  Timing dependencies may change on a second
-execution, especially if a delay is added between steps of the test (a
-more principled version of the unfortunate tradition of testing
-concurrent code by adding ad-hoc {\tt sleep} statements).   Most
-nondeterministic behavior in tests is probably attributable to a such
-effects that can vary with a repeated execution, in the same
-(approximate) enviroment, of the same test.  The two broad classes of
-nondeterminism that can usually be detected (with some probability) by
-simply executing a test twice in succession are:
-
-\begin{itemize}
-\item {\bf External Environment:} This class of nondeterminism arises
-  when a test's behavior depends on factors outside the program under
-  test, which are subject to uncontrolled variance.  Calling {\tt
-    random()} in Python is an obvious example; the call is almost
-  guaranteed to return a different value each time it is called in a
-  test, unless we explicitly re-seed the random number generator at
-  the beginning of each test.  Similarly, calls to {\tt time}
-  functions  will usually return different values on different runs,
-  and even if only elapsed time, rather than absolute, is used to
-  influence behavior, subtle variances in timing are to be expected
-  due to system load, I/O, and other external factors.  Finally, in
-  distributed system testing, the environment is often remote servers
-  (or clients) that may be unavailable, slower than expected, or
-  subject to faulty behavior for reasons not under the control of the
-  test.
-\item {\bf Concurrency:} Perhaps the second most common cause for
-  nondeterminism in tests is when an algorithm is implemented using
-  concurrency, in the (incorrect) beleif that the result of the
-  algorithm remains deterministic.  E.g., consider a multi-threaded
-  implementation of a search algorithm, that returns the location of
-  an item in an unsorted list.  If the list may contain duplicates,
-  and a developer has not accounted for this possibility, the
-  algorithm may return different locations of an item for the same list,
-  depending on the scheduling of threads.  Arguably, this can be
-  subsumed under the first class, with the behavior of the
-  processor/scheduler as the ``enviromental'' factor; relying on
-  scheduling is, in a sense, equivalent to calling {\tt random} or
-  {\tt time}.
-\end{itemize}
-
-Lam et al. \cite{idflakies} note that flaky tests not due to order
-dependence are generally due to ``concurrency, timeouts,
-network/IO, etc.'' a description that is well accounted for by our two
-primary sources, and most of which can be detected, at least in
-principle, by multiple runs of a test within the same testing process.  However, simply running a test twice in the same process (possibly with a delay between steps) does
-not suffice in all cases.
-
-\subsubsection{Process-Based Nondeterminism}
-
-\label{sec:pnondet}
-
-Some sources of nondeterminism unfortunately require executing a test in a new
-process environment, because the source is inherently tied to the
-process in which code runs; simply re-executing a test in the same
-process will not reveal the problem.
-
-Address Space Layout Randomization (ASLR) \cite{ASLR}  is probably the most
-important source of nondeterminism that arises (only) from change in
-process.  ASLR scrambles the layout of memory of the process in which
-an executable runs in order to make it harder to exploit memory-safety
-vulnerabilities in code.  As a side effect, it
-means that a test that, in some circumstances, causes a crash, may at
-other times not fail at all.  Even if no inputs produce a crash,
-however, memory errors may produce variation in values that are
-overwritten or arise from uninitialized memory, and thus be detectable
-as nondeterminism.
-
-Another example of process-based nondeterminism was discovered by numerous Python developers when Python
-version 3.3 introduced automatic random salting of hashes on a
-per-process basis (a new salt chosen for each Python process),
-in order to mitigate hash-based denial of service attacks
-\cite{denial}.  Until version 3.6 this not only resulted in changes in
-exact hash values, but in the order of iteration on
-dictionaries\footnote{In 3.6, the default dictionary
-  implementation became an ordered dictionary with consistent
-  iteration, though actual hashes remained nondeterministic.}.  While
-relatively few programs rely on exact hash values, many implicitly
-relied on them in that they only functioned correctly with a
-predictable order for dictionary iteration.  Testing
-Python code for nondeterminism based on this hash seed requires
-running in a new process:  Python does not allow changing the salt, since changing hash values on-the-fly would
-break all existing dictionaries and other structures relying on
-hashing.
-
-Other, less common, effects tied to process also exist.  A few tests
-may somehow depend on their actual PID (Process ID) or at least the
-parent process' ID.  Process change increases the likelihood that on a multi-core
-system a test will execute on a different CPU than in the first
-execution, which can have many subtle effects, mostly related to
-changed timing.
-
-One common cause of flaky tests is \emph{test order dependence} \cite{idflakies}.  This
-is less of a concern in our setting, since many problems due to test
-order are easily exposed in random testing, since the
-order-dependence-inducing operations will seldom usually happen in the
-same order in randomly generated tests.  For example, consider the
-case where a test usually run ``early'' initializes an uninitialized
-value in a system, and a later test will only fail if it somehow is
-run before the initializing test.  In random testing, the order of
-initialization/access is likely to be random, and some run of a random
-tester is likely to expose the problem as a (deterministic, not flaky) failure.
-
-However, for cases where
-there is an order dependence that is hard to translate into an
-actual bug, using a fresh
-process may expose such dependencies --- the ``environment'' of the
-parent process will have the, e.g., flag set or value initialized, but
-the new process will not.
+The implementation of the above ideas is conceptually straightforward,
+and should be easy to apply in other automated test generation
+settings with a natural notion of actions.  We present out
+implementation by first sketching a mapping from the formalism to one
+particular popular way to represent automatically generated unit
+tests, in Section \ref{sec:pools}.  The particular implementations of
+horizontal (Section \ref{sec:horizimp} and vertical (Section
+\ref{sec:vertimp}) determinism checks, as well as probabilistic
+delta-debugging support (Section \ref{sec:ddimp}) are described after a general
+overview of the TSTL testing tool (Section \ref{sec:tstlintro}).
 
 \subsection{Pools and Visible State}
+\label{sec:pools}
 
@@
 \subsection{A Brief Introduction to TSTL}
-
-\begin{table*}
-\centering
-{\scriptsize
-
-\caption{TSTL Method Calls for Nondeterminism Detection}
-\label{tab:methods}
-\begin{tabular}{l|l}
-%\hline
-Name & Purpose \\
-\hline
-\hline
-{\tt nondeterministic} & returns True if test exhibits final state nondeterminism in $k$
-  tries with delay $d$, optionally only over pools in $\mathcal{P}$\\
-\hline
-{\tt stepNondeterministic} & as above, except checks all visible values
-                       (compares state at every step)\\
-\hline
-{\tt processNondeterministic} & as above, except runs test provided in
-                          subprocess and compares non-opaque output
-                          values \\
-\hline
-{\tt P} & returns probability that a predicate holds for a test (with
-          optional number of sample to use in estimate) \\
-\hline
-{\tt forceP} & wraps a predicate for test reduction, checking that it
-               has desired probability of holding, using
-         sample size and replications parameters \\
-\hline
-\hline
-
-%\hline
-\end{tabular}
-}
-\end{table*}
-
-\begin{table*}
-\centering
-{\scriptsize
-\caption{TSTL Command Line Options for
-  Nondeterminism Detection}
-\label{tab:options}
-\begin{tabular}{l|l|l}
-%\hline
-Name &  Tool(s) & Purpose \\
-\hline
-\hline
-{\tt --checkDeterminism} & {\tt rt}, {\tt reduce} & Checks tests for deterministic
-                                  behavior (visible value in rt, final
-  value in reduce)\\
-\hline
-{\tt --checkProcessDeterminism} & {\tt rt}, {\tt reduce} & Checks determinism using more 
-                                         expensive process-changed 
-                                         execution \\
-\hline
-{\tt --determinismTries INT} & {\tt rt}, {\tt reduce} & Number of times to repeat test 
-                                  checking for nondeterministic
-                                  behavior \\
-\hline 
-{\tt --determinismDelay FLOAT} & {\tt rt}, {\tt reduce} & Delay between steps when 
-                                  re-executing to check for
-                                              determinism \\
-\hline 
-{\tt --checkFailureDeterminism} & tstl compiler & Produce a harness
-                                                  that checks
-                                                  determinism on all
-                                                  failing actions \\
-\hline 
-{\tt --probability P} & {\tt reduce} & force property with respect to which 
-                               test is being reduce to hold with 
-                                 probability $P$ \\
-\hline 
-{\tt --samples S} & {\tt reduce} & use $S$ samples when checking predicate
-                             probability\\
-\hline 
-{\tt --replications R} & {\tt reduce} & use $R$ replications when checking predicate probability\\
-\hline
-\hline
-
-%\hline
-\end{tabular}
-}
-\end{table*}
+\label{sec:tstlintro}
 
@@
 \begin{figure}
+  {\scriptsize
 \begin{code}
@@
 \end{code}
+}
 \caption{Simple TSTL harness testing lists and sets}
@@
 approach essentially follows the pool-based canonical form defined by
-Andrews et al., but relaxes many of its restrictions
+Andrews et al. and described above, but relaxes many of its restrictions
 (e.g., a constant value can be used in a method call without first
@@
 \subsection{Implementing Horizontal Nondeterminism Detection}
+\label{sec:horizimp}
 
@@
 and method/function calls using pool values (including method calls on
-pool objects).  This formalism is a widely used and efficient way to
-represent unit tests \cite{Pacheco,AndrewsTR}.  In TSTL, \emph{visible
-value determinism} is based on comparing the values of \emph{all pool
+pool objects).  In TSTL, \emph{visible
+value determinism} is, as proposed above, based on comparing the values of \emph{all pool
 variables} after each test action (pool values other than the one
@@
 nondeterminism is when a call results in a change to a value passed as
-a parameter, changes a shared reference held by two values).  By
+a parameter, or changes a shared reference held by two values).  By
 default $V$ as defined in Section \ref{sec:formal} is just the set of all possible pool values, and $v$
@@
 
+\begin{table*}
+\centering
+{\scriptsize
+
+\caption{TSTL Method Calls for Nondeterminism Detection}
+\label{tab:methods}
+\begin{tabular}{l|p{8cm}}
+%\hline
+Name & Purpose \\
+\hline
+\hline
+{\tt nondeterministic} & returns True if test exhibits final state nondeterminism in $k$
+  tries with delay $d$, optionally only over pools in $\mathcal{P}$\\
+\hline
+{\tt stepNondeterministic} & as above, except checks all visible values
+                       (compares state at every step)\\
+\hline
+{\tt processNondeterministic} & as above, except runs test provided in
+                          subprocess and compares non-opaque output
+                          values \\
+\hline
+{\tt P} & returns probability that a predicate holds for a test (with
+          optional number of sample to use in estimate) \\
+\hline
+{\tt forceP} & wraps a predicate for test reduction, checking that it
+               has desired probability of holding, using
+         sample size and replications parameters \\
+\hline
+\hline
+
+%\hline
+\end{tabular}
+}
+\end{table*}
+
+\begin{table*}
+\centering
+{\scriptsize
+\caption{TSTL Command Line Options for
+  Nondeterminism Detection}
+\label{tab:options}
+{\scriptsize
+\begin{tabular}{l|l|p{6cm}}
+%\hline
+Name &  Tool(s) & Purpose \\
+\hline
+\hline
+{\tt --checkDeterminism} & {\tt rt}, {\tt reduce} & Checks tests for deterministic
+                                  behavior (visible value in rt, final
+  value in reduce)\\
+\hline
+{\tt --checkProcessDeterminism} & {\tt rt}, {\tt reduce} & Checks determinism using more 
+                                         expensive process-changed 
+                                         execution \\
+\hline
+{\tt --determinismTries INT} & {\tt rt}, {\tt reduce} & Number of times to repeat test 
+                                  checking for nondeterministic
+                                  behavior \\
+\hline 
+{\tt --determinismDelay FLOAT} & {\tt rt}, {\tt reduce} & Delay between steps when 
+                                  re-executing to check for
+                                              determinism \\
+\hline 
+{\tt --checkFailureDeterminism} & tstl compiler & Produce a harness
+                                                  that checks
+                                                  determinism on all
+                                                  failing actions \\
+\hline 
+{\tt --probability P} & {\tt reduce} & force property with respect to which 
+                               test is being reduce to hold with 
+                                 probability $P$ \\
+\hline 
+{\tt --samples S} & {\tt reduce} & use $S$ samples when checking predicate
+                             probability\\
+\hline 
+{\tt --replications R} & {\tt reduce} & use $R$ replications when checking predicate probability\\
+\hline
+\hline
+
+%\hline
+\end{tabular}
+}
+}
+\end{table*}
+
  Tables \ref{tab:methods}
 and \ref{tab:options} show, respectively, the primary additions to the
-TSTL Python API and the TSTL command line tools.  The TSTL test
+TSTL Python API and the TSTL command line tools; in addition to
+horizontal nondeterminism checking, these extensions also support failure
+nondeterminism and probabilistic delta-debugging.  The TSTL test
 generation tool is called {\tt rt}, since the default mode is a fast
@@
 
+Note that TSTL provides explicit support for exploring the real-world sources of
+nondeterminism described in Section \ref{sec:sources}, by supporting
+execution in a new process ({\tt processNondeterministic} API call and
+{\tt --checkProcessDeterminism} command line argument) and allowing
+control over timing changes in re-execution, by specifying a delay to
+introduce between actions (parameters to API calls, {\tt
+  --determinismDelay} command line arugment).  Changing the timing of
+steps is very effective when exact timing is somehow involved in
+results, and often effective in exposing concurrency-related
+nondeterminism, since the delay will frequently change the
+interleaving of threads.
+
+Because TSTL has a first-class interface to AFL \cite{aflfuzz}, we can even use
+AFL's sophisticated heuristics to perform very thorough, week-long
+checks for nondeterminism, using strategies built to find subtle
+security vulnerabilities in C programs.  Equally importantly, since
+there is substantial overhead in nondeterminism detection, AFL can be
+used to predict whether tested Python code is a good candidate for
+checking for horizontal nondeterminism; if the
+AFL \emph{stability} statistic (see the AFL documentation for details
+\cite{aflfuzz}) is lower than 100\%, it may indicate horizontal
+nondeterminism is present.
+Because of the idiosyncrasies of AFL instrumentation and process
+behavior, this is not always a reliable guide, but it is a very low
+cost indicator produced as a by-product of fuzzing for other problems.
+
 \subsection{Implementing Vertical Nondeterminism Detection}
+\label{sec:vertimp}
 
@@
 limited to failures:  {\tt --checkFailureDeterminism} causes the TSTL
-compiler to emit a harness where every action that causes an expected
+compiler to emit a harness where every action that causes an \emph{expected}
 exception to be raised is repeated to ensure the same exception is
-still raised.  Users can also express that any action should exhibit
-vertical determinism by wrapping the action in a function that checks
-for failure and calls the code again.
+still raised.  In terms of the formalism, if execution of action
+$\alpha_i$ raises an exception, then there are two possibilities:
+
+\begin{enumerate}
+\item If the TSTL harness does not explicitly allow the raised
+  exception, then the test fails.  There is no check for vertical
+  nondeterminism in this case, since the behavior of the system is
+  incorrect whether or not the behavior is idempotent.
+  \item If the harness does allow exception $e$ (e.g., as with {\tt
+      KeyError} and {\tt ValueError} in Figure \ref{fig:simple}), then
+    $F_i$ is true, and we repeat $\alpha_i$ in order to ensure that
+    $F_{i+1}$ is also true.
+    \end{enumerate}
+
+When a failure is indicated by some means other than raising an
+=exception, TSTL users must express this by explicitly writing
+a wrapper that raises an expected exception on failure.
 
 \subsection{Tools for Nondeterministic Delta-Debugging}
+\label{sec:ddimp}
 
@@
 for the {\tt redis-py} example discussed below).
-
-\subsection{Nondeterminism and AFL Stability}
-
-Because TSTL has a first-class interface to AFL \cite{aflfuzz}, we can even use
-AFL's sophisticated heuristics to perform very thorough, week-long
-checks for nondeterminism, using strategies built to find subtle
-security vulnerabilities in C programs.  Equally importantly, since
-there is substantial overhead in nondeterminism detection, AFL can be
-used to predict whether a program has potential nondeterminism; if the
-AFL \emph{stability} statistic (see the AFL documentation for details
-\cite{aflfuzz}) is lower than 100\%, it may indicate a problem.
-Because of the idiosyncrasies of AFL instrumentation and process
-behavior, this is not always a reliable guide, but it is a very low
-cost indicator produced as a by-product of fuzzing.
 
@@
 
-\begin{figure}
+\begin{figure}[t]
+  {\scriptsize
 \begin{code}
@@
   \end{code}
+  }
 \caption{Code for Python library with (independent) vertical and
@@
 \begin{figure}
+ {\scriptsize
 {\bf quick.0.test:}
@@
 \end{code}
+}
 \caption{Coverage maximizing tests for Python library with (independent) vertical and
@@
 
-The {\tt storage.py} program also contains a (again, unrealistically
+The {\tt storage.py} program also contains an (again, unrealistically
 simple) example of \emph{vertical} nondeterminism.  The {\tt store}

### 2019-09-04 experiments.tex
work on some issues
12 hour change

@@
 has proven difficult, as numerous Redis commands introduce nondeterministic
-behavior:  thus the resulting tests are very often \emph{flaky}.   Using {\tt tstl\_afl\_fuzz}, we can see that the {\tt redis-py} has a 
+behavior:  thus the resulting tests are very often \emph{flaky}.   Using {\tt tstl\_afl\_fuzz}, we can see that {\tt redis-py} has a 
 stability of only 56.26\%, a clear indicator of nondeterminism. Some
@@
 us to simply set up an overnight run that will, in 12 hours, produce
-a set of minimal flaky tests exposing the sources of
-nondeterminism/flakiness.  In this case, the probability of flakiness
+a set of minimal nondeterministic tests exposing the sources of
+nondeterminism/flakiness.  This process is unlikely to require 12
+hours, of course; our point is that it is completely automated,
+including generation of a set of minimal, distinct, tests showing
+different kinds of nondeterminism using test normalization~\cite{onetest}.  The actual amount of time given
+may be smaller, or, for that matter, larger (leaving an important
+library to self-document nondeterminism over a weekend is hardly a bad
+practice).  In this case, the probability of flakiness
 is not important, just the possibility, so we let the delta-debugging
 create a minimal test with some non-zero (but possibly very small)
-observed probability of nondeterminism.
+observed probability of nondeterminism.  Our 12 hour run identified
+all sources of nondeterminism successfully, according to a series of
+48 hour runs performed on the version of the harness we modified to
+remove nondeterminism.  We determined that the mean time to find all
+the sources of nondeterminism and produce minimal tests 3.2
+hours, which is not trivial, but also, for this purpose, not excessive.
 

### 2019-09-04 experiments.tex
edit exp

@@
 detection for real systems was acceptable, and (3) the performance of
-delta-debugging for real nondeterministic predicates.
+delta-debugging for real nondeterministic predicates.  We chose two of
+the subjects, {\tt redis-py} and {\tt pyfakefs}, based on the fact that they were previously-existing large
+examples of TSTL harnesses, and therefore serve to show how easily our
+approach can be integrated into an existing large-scale test
+generation effort.  The other two subjects were chosen, and harnesses
+written, specifically to demonstrate specific, important sources of
+nondeterminism in Python code: concurrency and hash salting.
 
-\subsection{Case Study: Parallel Sorting}
+Table \ref{tab:overviewexp} shows a summary of our experimental
+results.  In all cases, our approach was able to detect interesting
+nondeterminism.  The overhead on testing imposed by nondeterminism
+checks varied considerably, but in the worst case did not double the
+average cost of each action, and thus never halved the amount of
+testing performed.  Test reduction always reduced the size of
+generated tests by 90\% or more, thus showing that effective reduction
+has a large payoff.
 
-Our first case study is a simple implementation of a parallel merge
-sort in Python
-\url{https://gist.github.com/stephenmcd/39ded69946155930c347}.  The
+\begin{table}
+\centering
+\begin{tabular}{l|l|r|r|l}
+Subject & Detection & Overhead & Reduction & Reduction time \\
+\hline
+Parallel Sort & Yes (see below) & $\sim$40\% & & $\sim$2 minutes\\
+{\tt redis-py} & Yes (see below) & $\sim$20\% & $\sim$95\% & $\sim$10 minutes\\
+{\tt datarray} & Yes (see below) & $\sim$93\% & $\sim$90\% & $\sim$92
+  seconds\\
+{\tt pyfakefs} & Yes (see below) & $\sim$8\% & $\sim$99\% & $\sim$1 second  \\
+\end{tabular}
+\caption{Overview of experimental results}
+\label{tab:overviewexp}
+\end{table}
+
+\subsection{Parallel Sorting}
+
+Our first subject is a simple implementation of a parallel merge
+sort\footnote{\url{https://gist.github.com/stephenmcd/39ded69946155930c347}}.  The
 merge sort and harness can be found in the TSTL GitHub repository
@@
 
-This case study has two purposes:  first, to show that the
+This experimental subject serves two purposes:  first, to show that the
 nondeterminism detection features of TSTL can be used to, with
@@
 
-\subsection {Case Study: {\tt redis-py}}
+\subsection {The {\tt redis-py} Library}
 
@@
 
-\subsection{Case Study: {\tt datarray} Inference Algorithms}
+\subsection{Berkeley {\tt datarray} Inference Algorithms}
 
@@
 
-\subsection {Vertical Determinism Case Study: {\tt pyfakefs}}
+\subsection {Vertical Determinism: {\tt pyfakefs}}
 
@@
 
+\subsection{Summary of Results}
+
 \section{Threats to Validity}
@@
 generates that do not depend on any TSTL code.
-The source code for the tools and all case studies is
+The source code for the tools and all experimental subjects is
 available for inspection and replication in the TSTL repository.

### 2019-09-04 experiments.tex
result

@@
 testing performed.  Test reduction always reduced the size of
-generated tests by 90\% or more, thus showing that effective reduction
-has a large payoff.
+generated tests by 85\% or more, in less than 15 minutes, thus showing that effective reduction
+has a large payoff at an acceptable price.
 
@@
 \hline
-Parallel Sort & Yes (see below) & $\sim$40\% & & $\sim$2 minutes\\
+Parallel Sort & Yes (see below) & $\sim$40\% & $\sim$85\% & $\sim$2 minutes\\
 {\tt redis-py} & Yes (see below) & $\sim$20\% & $\sim$95\% & $\sim$10 minutes\\

### 2019-09-05 conclusions.tex
ready to ship new version, probably

@@
 
+Finally, it would be useful to implement versions of our formalisms in
+test generation tools for other languages with a suitable notion of
+action.  For example, the DeepState
+\cite{DeepState,DeepStateTutorial,deepstaterepo} tool for
+property-based fuzzing and symbolic-execution of C and C++ code defines
+actions that are choices between C++ lambdas in a test harness,
+produced by a nondeterminism operator, {\tt OneOf}.
+Adding annotations of visible state (perhaps simply in the form of
+blocks of address ranges in memory, a method often suitable for C and
+C++ code) would make it possible to search for unexpected nondeterminism in
+critical C and C++ libraries.
+
 {\scriptsize {\bf Acknowledgments:}  The authors would like to thank John Micco,

### 2019-09-05 experiments.tex
ready to ship new version, probably

@@
 nondeterminism detection to real Python code, including some widely
-used systems ({\tt redis-py} and {\tt pyfakefs} are well-maintained,
-widely-used libraries).  The primary
-points we wanted to explore were (1) whether our approach was able to reliably detect actual
-nondeterminism, (2) whether the overhead of nondeterminism
-detection for real systems was acceptable, and (3) the performance of
-delta-debugging for real nondeterministic predicates.  We chose two of
+used libraries (according to GitHub, {\tt pyfakefs} is used by at least
+338 other projects, and {\tt redis-py} is used by at least 50,000).  The primary
+points we wanted to explore were {\bf RQ1:} whether our approach was able to reliably detect actual
+nondeterminism, {\bf RQ2:} whether the overhead of nondeterminism
+detection for real systems was acceptable, and {\bf RQ3:} whether the performance of
+delta-debugging in this setting was acceptable, in terms of both time
+and results (amount of reduction in test size).  We chose two of
 the subjects, {\tt redis-py} and {\tt pyfakefs}, based on the fact that they were previously-existing large
@@
 approach can be integrated into an existing large-scale test
-generation effort.  The other two subjects were chosen, and harnesses
+generation effort.  The other two subjects were chosen, and new harnesses
 written, specifically to demonstrate specific, important sources of
@@
 results.  In all cases, our approach was able to detect interesting
-nondeterminism.  The overhead on testing imposed by nondeterminism
+nondeterminism (in fact, all nondeterminism of which we are aware).  The overhead on testing imposed by nondeterminism
 checks varied considerably, but in the worst case did not double the
-average cost of each action, and thus never halved the amount of
+average cost of each action (the expected overhead, since we
+essentially execute each test twice), and thus never halved the amount of
 testing performed.  Test reduction always reduced the size of
-generated tests by 85\% or more, in less than 15 minutes, thus showing that effective reduction
+generated tests by 85\% or more, in less than 15 minutes, thus showing that reduction
 has a large payoff at an acceptable price.
@@
 test case showing the {\tt swapsort} behaving nondeterministically is
-almost instantly produced (it takes less than 3 seconds).  If we comment out the call to {\tt
+almost instantly produced (it takes less than 3 seconds), thus
+answering yes to {\bf  RQ1}.  If we comment out the call to {\tt
   swap\_sort\_parallel} and only check the parallel merge sort, however, TSTL
@@
 with nondeterminism detection turned on results in a slowdown of approximately
-40\%.
+40\% ({\bf RQ2}.
 
@@
 increased in the reducer using the {\tt --determinismTries 10} option,
-the resulting test will be usually be very compact, e.g.:
+the resulting test will be usually be very compact, in this case
+reducing the original test from 56 to 8 steps.
 
@@
 
-The {\tt redis-py} \cite{redispy} module implements a widely-used Python interface
+The {\tt redis-py} \cite{redispy} module implements a (very) widely-used Python interface
 to the popular Redis in-memory database, cache, and
@@
 behavior:  thus the resulting tests are very often \emph{flaky}.   Using {\tt tstl\_afl\_fuzz}, we can see that {\tt redis-py} has a 
-stability of only 56.26\%, a clear indicator of nondeterminism. Some
+stability of only 56.26\%, a clear indicator of significant nondeterminism. Some
 of the problematic commands are obvious simply by inspection (e.g.,
@@
 commands producing data with a limited lifetime (e.g., {\tt expire,
-  pexpire}) also introduce timing-based nondeterminism.
+  pexpire}) introduce timing-based nondeterminism.
 
@@
 argument is not obvious to a non-expert, nor is the behavior of {\tt
-  spop} which pops a random value from a set.  Moreover, it is
+  spop}, which pops a random value from a set.  Moreover, it is
 difficult to guess whether the {\tt pipe} mechanism, which allows a
@@
 us to simply set up an overnight run that will, in 12 hours, produce
-a set of minimal nondeterministic tests exposing the sources of
-nondeterminism/flakiness.  This process is unlikely to require 12
-hours, of course; our point is that it is completely automated,
-including generation of a set of minimal, distinct, tests showing
-different kinds of nondeterminism using test normalization~\cite{onetest}.  The actual amount of time given
+a set of minimal nondeterministic tests exposing the root sources of
+nondeterminism/flakiness: we set the harness to repeatedly run, then
+produce a regression test.  If the test is flaky (exhibits
+nondeterminism in result, over a set of 3 runs), we use the TSTL reducer to shrink it to a
+normaliized \cite{onetest} test.  At the end of the testing period,
+all distinct tests exhibiting nondeterminism are reported to the
+user.  The process reduces the large number of different tests
+exhibiting flakiness to a small set of tests, each short, showing a
+different cause of nondeterminism.  The set produced is not perfect,
+there is some duplication of causes, but it is sufficient for
+identifying causes of nondeterminism.
+
+This process is unlikely to truly require 12
+hours, of course; our point is that it is completely automated and
+can be given time proportional to the need to find even
+low-probability sources of nondeterminism.  The actual amount of time given
 may be smaller, or, for that matter, larger (leaving an important
@@
 practice).  In this case, the probability of flakiness
-is not important, just the possibility, so we let the delta-debugging
+is not important, just the possibility, so we let the reducer
 create a minimal test with some non-zero (but possibly very small)
@@
 48 hour runs performed on the version of the harness we modified to
-remove nondeterminism.  We determined that the mean time to find all
+remove nondeterminism, which produced no flaky tests.  We determined that the mean time to find all
 the sources of nondeterminism and produce minimal tests 3.2
-hours, which is not trivial, but also, for this purpose, not excessive.
-
-Figures \ref{fig:r1}-\ref{fig:r3} show delta-debugging of a typical
-{\tt redis-py} nondeterministic test, originally with 287 steps.  The initial test behave
-nondeterministically about half the time.  Figure \ref{fig:r1} shows
-that this is a non-monotonic reduction problem, where removing steps
-can either increase or decrease the probability of nondeterminism.  At
-first, unmodified delta-debugging actually improves the probability of
-failure, but eventually it produces a test of length 16 that only behaves
-nondeterministically 24\% of the time.  The reduction takes only 38
-seconds.  For the purpose of identifying commands leading to
-nondeterminism, this is acceptable.  However, if we were actually
-debugging a complex nondeterminism bug in Redis itself, we might want
-a more reliably nondeterministic test.  Using the same parameters as
-in Section \ref{sec:pubbias}, we see the same pattern.  Simply making
-a predicate that ``forces'' the probability to remain high, with a
-large number of samples, does not work (Figure \ref{fig:r3}) and
-requires over 2000 seconds to produce a test with an even worse
-probability of nondeterminism.  Using 10 samples and 10 replications,
-on the other hand, actually improves the probability of
-nondeterministic behavior, and
-(in this case) even produces a smaller test (only 14 steps) in just
-over 10 minutes.
+hours, which is not trivial, but also, for this purpose, not
+excessive.  Our answer to {\bf RQ1} is thus a strong affirmative.
 
@@
 observed sources of flakiness, the code coverage loss was minimal.
-Mean branch coverage, for regression tests of length 200 was reduced
+Mean branch coverage, for regression tests of length 200, was reduced
 by less than 5 branches (a less than 1\% decrease).  Obviously, it is
@@
 fast-executing high-coverage full regression tests (not just tests for
-crashes and unexpected exceptions), this seems acceptable.
+crashes and unexpected exceptions), this seems acceptable.  AFL stability was 56.26\% with the harness allowing 
+nondeterminism, but rose to 98.32\% using the harness with 
+nondeterministic operations removed.
 
-Moreover, turning off sources of expected nondeterminism makes
+Removing sources of expected nondeterminism also makes
 it possible to aggressively test {\tt redis-py} and Redis for
@@
 determinism checks, with no delay between operations, is only
-20\%, much lower than the expected cost of running each test twice.
-This is because choosing the actions in a test (and determining which
+20\%, much lower than the expected cost of running each test twice,
+answering {\bf RQ2} also in the affirmative.  This is because choosing the actions in a test (and determining which
 actions are enabled at each step) consumes a large part of the test
-generator's time; running the test again and checking equality is
+generator's time with a complex library like {\tt redis-py}; running the test again and checking equality is
 relatively inexpensive.
 
-Finally, AFL stability was 94.32\% with the harness allowing
-nondeterminism, but rose to 98.32\% using the harness with
-nondeterministic operations removed.
+As to {\bf RQ3}, Figures \ref{fig:r1}-\ref{fig:r3} show delta-debugging of a typical
+{\tt redis-py} nondeterministic test, originally with 287 steps.  The initial test behaves
+nondeterministically about half the time.  Figure \ref{fig:r1} shows
+that this is a non-monotonic reduction problem, where removing steps
+can either increase or decrease the probability of nondeterminism.  At
+first, unmodified delta-debugging actually improves the probability of
+failure, but eventually it produces a test of length 16 that only behaves
+nondeterministically 24\% of the time.  The reduction takes only 38
+seconds.  For the purpose of identifying commands leading to
+nondeterminism, this is acceptable.  However, if we were actually
+debugging a complex nondeterminism bug in Redis itself, we might want
+a more reliably nondeterministic test.  Using the same parameters as
+in Section \ref{sec:pubbias}, we see the same pattern.  Simply making
+a predicate that ``forces'' the probability to remain high, with a
+large number of samples, does not work (Figure \ref{fig:r3}) and
+requires over 2000 seconds to produce a test with an even worse
+probability of nondeterminism.  Using 10 samples and 10 replications,
+on the other hand, actually improves the probability of
+nondeterministic behavior, and (in this case) even produces a smaller test (only 14 steps) in just
+over 10 minutes.
 
@@
 check these tests for nondeterministic behavior.  Final state
-nondeterminism actually detected one instance of nondeterminism that
-visible value nondeterminism failed to detect (21 detections vs. 20); in general, final state
+nondeterminism actually detected one instance (but not root cause) of nondeterminism that
+visible value nondeterminism failed to detect (21 detections
+vs. 20). In general, final state
 nondeterminism is less able to detect nondeterminism, but in this
@@
 additional replays, when nondeterminism is detected (the difference is
-again significant by Wilcoxon test, with $p < 1.44 \times 10^{-10}$.
+again significant by Wilcoxon test, with $p < 1.44 \times 10^{-10}$).
 
-In practice, visible value nondeterminism and final state
+In practice, we expect that visible value nondeterminism and final state
 nondeterminism will often be very similar in both ability to detect
@@
 the {\tt calc\_marginals\_sumproduct} function (the only broken
-algorithm).  Reducing this 60 step test to a minimal test of only 6 steps,
+algorithm), answering {\bf RQ1} again in the affirmative.  Reducing this 60 step test to a minimal test of only 6 steps,
 showing an extremely simple input producing the issue, 
-required another 92 seconds.  Interestingly, the way that {\tt
-  python-afl} is used in TSTL means that AFL stability is 100\%,
+required another 92 seconds ({\bf RQ3}).  Interestingly, the way that {\tt
+  python-afl} is used in TSTL means that AFL stability is 100\% even
+for the original, broken, code,
 because the {\tt PYTHONHASHSEED} has already been chosen before the
@@
 checking for process nondeterminism, with no delay between operations,
-is high, a 93\% slowdown.  This is due to the high cost of subprocess
-creation and communication.
+is high, a 93\% slowdown({\bf RQ2}.  This is due to the high cost of subprocess
+creation and communication. However, even with this essentially worst
+case behavior (where test runtime is very low compared to process overheads),
+we sacrifice less than half of the test executions for a given time budget.
 
@@
 {\tt pyfakefs} harness is normally impossible without using another
-file system as a reference.  However, the fault was detected
+file system as a reference.  However, the fault was detected  essentially
 immediately, even without using a reference, when we compiled the harness
-with the {\tt --checkFailureDeterminism} flag.  Because vertical
-reduction does not require running complete tests multiple times, and
-does not affect the delta-debugging algorithm's performance, reducing
-the failing test to 3 steps required less than a second.
-Moreover, the overhead for the check for failure determinism in a
+with the {\tt --checkFailureDeterminism} flag (an affirmative answer
+to {\bf RQ1}).  Moreover, the overhead for the check for failure determinism in a
 version of the code without the {\tt remove} error was
-less than 8\%.  Detecting the fault using a reference file system 
+less than 8\%  ({\bf RQ2}).  Detecting the fault using a reference file system 
 required 17\% more testing time before detection, and took over twice as
@@
 final operation (since further operations are required to 
-expose the bad file system state the operation introduces).  For this
-hypothetical subtle bug, failure determinism checks not only make it
-possible to detect the fault without a reference implementation, it
+expose the bad file system state the operation introduces).
+Because vertical
+reduction does not require running complete tests multiple times, and
+does not affect the delta-debugging algorithm's performance, reducing
+the failing test to 3 steps required less than a second ({\bf RQ3}).
+For this hypothetical subtle bug, failure determinism checks not only make it
+possible to detect the fault without a reference implementation, they
 improves on detection speed and ease of debugging even compared to a
@@
 by differential testing, and quantify the additional specification
-strength provided by failure determinism checks.  We therefore used
+strength provided by failure determinism checks, to further answer
+{\bf RQ1} in the special case of failure determinism.  We therefore used
 {\tt universalmutator} \cite{RegExpMut} to produce 2,350 mutants of
@@
 causes an unexpected exception to be raised, or results in a timeout
-due to, e.g., an infinite loop.
+due to, e.g., an infinite loop.  We could have applied mutation
+testing to {\tt redis-py}, but we are dubious of the value of such an
+analysis without a strong differential harness to compare to; the {\tt
+  redis-py} testing is limited in strength, and largely useful for
+producing high-coverage operation sequences from which to construct
+manual unit tests.
 
@@
 exception propagation, will work as well as introducing a notion of
-failure nondeterminism.
+failure nondeterminism.  This is a strong affirmative answer to {\bf
+  RQ1} at least for a large set of hypothetical bugs.
 
 We also checked the cost of introducing failure determinism checking
-by analyzing mutants that could be killed by both methods.  Even
+by analyzing mutants that could be killed by both methods ({\bf RQ2}).  Even
 though in some cases the failure determinism check allows a mutant to
@@
 larger with failure determinism checking, and this change was significant by Wilcoxon test
-$(p < 1 \times 10^{-15})$.  While significant, this cost is probably
+$(p < 1 \times 10^{-15})$.  While \emph{statistically} significant,
+this cost is almost certainly
 too small to be of any real practical importance.
@@
 \subsection{Summary of Results}
+
+\begin{itemize}
+\item {\bf RQ1: Is our approach able to reliably detect
+  nondeterminism?}  For all of our subjects, all nondeterminism we are
+aware of was detected.  The {\tt pyfakefs} experiments also show,
+using mutation testing, an additional ability to detect bugs over a
+strong differential test.
+\item {\bf RQ2: Is the overhead of nondeterminims detection
+    acceptable?} This depends on the criteria for acceptability, but
+  given the fault-detection and flaky-test avoidance capabilities
+  demonstrated, we think that so long as the testing is still able to
+  execute at least half as many tests, for the same time budget, the
+  answer is yes.  All subjects, even when using expensive
+  process-based checks, satisfied this constraint.  Failure
+  nondeterminism detection was particularly inexpensive, as expected.
+\item {\bf RQ3: Is the performance of delta-debugging/test reduction
+    acceptable?}  In all cases, reduction was relatively quick (a
+  little over 10 minutes in the worst case) and the amount of
+  reduction produced was large (85\% or better).
+  \end{itemize}
 

### 2019-09-09 experiments.tex
little more info
response, final touches

@@
 written, specifically to demonstrate specific, important sources of
-nondeterminism in Python code: concurrency and hash salting.
+nondeterminism in Python code: concurrency and hash salting.  In
+addition to our core research questions ({\bf RQ1-RQ3}) shared by all subjects, we
+report on particular aspects of using the TSTL nondeterminism tools
+specific to each subject.
 
@@
 \begin{tabular}{l|l|r|r|l}
-Subject & Detection & Overhead & Reduction & Reduction time \\
+& \multicolumn{1}{c}{\bf RQ1} & \multicolumn{1}{|c|}{\bf RQ2} &
+                                                              \multicolumn{2}{c}{\bf RQ3}\\
+  \hline
+Subject & Detection & Overhead & Reduction \%
+                                                        & Reduction time\\
 \hline

### 2019-09-09 implementation.tex
missing paren
response, final touches

@@
 and should be easy to apply in other automated test generation
-settings with a natural notion of actions.  We present out
+settings with a natural notion of actions.  We present our
 implementation by first sketching a mapping from the formalism to one
@@
 tests, in Section \ref{sec:pools}.  The particular implementations of
-horizontal (Section \ref{sec:horizimp} and vertical (Section
+horizontal (Section \ref{sec:horizimp}) and vertical (Section
 \ref{sec:vertimp}) determinism checks, as well as probabilistic
@@
 {\tt P} & returns probability that a predicate holds for a test (with
-          optional number of sample to use in estimate) \\
+          optional number of samples to use in estimate) \\
 \hline
@@
 for the {\tt redis-py} example discussed below).
+
+\subsection{Key Challenges}
+The key challenge in all of these implementations was choosing
+appropriate default values.  In particular, how many times to retry
+tests (horizontal determinism) or actions (vertical/failure determinism), what delays to use, and numbers of replications
+required in nondeterministic delta-debugging.  Because our confidence
+that these defaults are ideal is low, we made it easy to change the
+values used, as shown by the presence of command line options and
+optional function parameters in Tables \ref{tab:methods} and \ref{tab:options}.
 

### 2019-09-10 highlights.tex
stuff'

\documentclass{article}

\begin{document}

\noindent {\bf DOES IT ALWAYS DO THAT? PRACTICAL AUTOMATIC LIGHT- WEIGHT NONDETERMINISM AND FLAKY TEST DETECTION AND DEBUGGING FOR PYTHON LIBRARIES}

\vspace{0.1in}

\noindent {\bf Highlights:}

\vspace{0.1in}

\noindent This article presents a new formalism and characterization of types of nondeterminism in tests (and software libraries) that can lead to ``flaky tests,'' a pernicious problem in real-world large-scale testing.

\vspace{0.1in}

\noindent The article describes a formalism distinguishing horizontal (where the same test does different things when run multiple times, the traditional ``flaky test'' behavior) and vertical determinism (where an operation in the same test does different things, when this is incorrect).  The articles then describes how the formalisms presented can be used to implement effective detection of nondeterminism for Python code.

\vspace{0.1in}

\noindent The article shows that nondeterminism detection can be relatively easy to use, and inexpensive, and that the novel approach of checking for failure determinism can detect up to 6\% additional program mutants over a strong differential testing approach, for a real-world mock file system library.

\vspace{0.1in}

\noindent In addition to detection of nondeterminism, the paper presents the first examination of delta-debugging/test case reduction in a probabilistic setting, where a property that a test should satisfy only holds with a certain probability, or, more generally, the property describes a probability over a property of a test.  The key concept introduced is that the issues in this setting can be understood in terms of the scientific replication problem, and similar methods used to control false positives.  The above experimental results on real-world Python libraries show that the approach is effective and inexpensive.

\end{document}

### 2020-01-07 ISSTA/response.tex
issta?

\documentclass{letter}
%\def\today{September 10, 2019}
\usepackage[letterpaper,left=2.5cm, right=2.5cm, top=2.5cm, bottom=2.5cm]{geometry}

\begin{document}
\begin{letter}{}

  \address{Alex Groce\\School of Informatics, Computing \& Cyber Systems\\Northern Arizona University}

  \opening{Dear editor and reviewers:}

We would like to thank the anonymous reviewers for their help in
significantly improving the paper.  Based on the comments, we
corrected errors in the formalism, re-organized the paper by moving
the sources of determinism discussion to earlier in the paper and
providing more overview of implementation choices, including key
challenges, and re-formulated the experimental section in terms of
explicit research questions and answers to those questions, including
a summary table.

The section numbering is changed.  There is a new section 3, moved
from the implementation section, on sources of nondeterminism.  The
delta-debugging is now section 4, implementation is section 5, the
example is section 6, and section 7 is the experimental evaluation.
Section 8, 9, and 10 present threats to validity, related work, and
conclusions, respectively.

Below are our detailed responses to the reviewer comments.

{\bf R1:}

\begin{enumerate}
\item  Revised Formalism, fixed page 12 mistake, revised formal definition of failure nondeterminism.  We now discuss the issue fo multiple executions in vertical determinism.  In particular, our original formulation actually allowed for this, but this was not made clear.

\item We revised the section substantially, and added definitions and pseudo-code to make the concepts more precise and easy to follow.

\item Section 4 has been restructured and summarized more precisely.

\item Section 6 has been revised to make research questions explicit,
  our goals in selecting the subjects explicit, and presented as
  (limited) experiment rather than a series of case studies.
  \end{enumerate}

{\bf R2:}

\begin{enumerate}
\item We now much more thoroughly introduce delta debugging.  We explained our motivation in using mutation for vertical determinism only (it relates to an actual high-quality specification of a different kind to compare to; for redis-py, the "spec" is largely either failure or nondeterminism represented by flaky tests, so the comparison with mutants would have very little value and be difficult to interpret.  Comparison with a strong differential specification is easy to understand, and is certainly meaningful.

\item We discuss in threats and related work why we did not compare to, e.g., DeFlaker:  the settings and uses are almost completely different.  We were unable to devise a reasonable experiment that would compare the tools, since essentially they are orthogonal.

\item The 12 hours mentioned is rather misleading, we now elaborate what
this meant, in a way that should make it possible to understand the
result much more easily.  Basically, 12 hours \emph{sufficed} and any
amount of time can be spent that is needed to satisfy neede certainty.
This is basically independent of the flaky test suite size, though
reducing those might also be a (weaker) approach to the same goal.
\end{enumerate}

  \closing{Sincerely,\\Alex Groce (agroce@gmail.com)\\Associate
    Professor\\
    School of Informatics, Computing \& Cyber Systems\\Northern
    Arizona University \vspace{0.1in}\\Josie Holmes\\Affiliated Researcher\\
    School of Informatics, Computing \& Cyber Systems\\Northern Arizona University}

\end{letter}
\end{document}

### 2020-01-07 ISSTA/abstract.tex
issta?

A critically important, but surprisingly neglected, aspect of system reliability is system predictability.  Many software systems are implemented using mechanisms (unsafe languages, concurrency, caching, stochastic algorithms, environmental dependencies) that can introduce behavioral nondeterminism.  Users of software systems, especially other software using library calls in a single-threaded context, often expect that systems will behave deterministically in the sense that they have predictable results from the same series of operations.  Equally importantly, even when it does not impact correctness, nondeterminism must be understood and managed for effective (especially automated) testing.  Nondeterministic behavior that is not either controlled or accounted for can result in flaky tests as well as causing problems for test reduction, differential testing, and automated regression test generation.  We show that lightweight techniques, requiring little effort on the part of developers, can be used to extend an existing testing system to allow detection of nondeterminism.  We introduce a set of lightweight nondeterminism properties, inspired by real faults, and tailor these notions to the practical, automatic, checking of Python library code.  In particular, we demonstrate that our proposed failure nondeterminism can improve mutation score by 6\% for a strong, full differential test harness for a widely used mock file system.

### 2020-01-07 ISSTA/bio.tex
issta?

\documentclass{article}

\begin{document}

{\bf Alex Groce:} Alex Groce received his PhD in Computer Science from Carnegie Mellon University in 2005, and worked at NASA's Jet Propulsion Laboratory, as part of the Laboratory for Reliable Software, until 2009, when he joined the faculty at Oregon State University, and was promoted to Associate Professor in 2015.  In 2017, he moved to Northern Arizona University, where he is an associate professor of Computer Science.

{\bf Josie Holmes:}  Josie Holmes graduated from Northern Arizona University in 2004 with a BS in Geography, and received the MSGIS from Pennsylvania State University in 2016.  She is the author of multiple papers in software testing and analysis.

\end{document}

### 2020-01-07 ISSTA/cover.tex
issta?

\documentclass{letter}
%\def\today{January 30, 2016}
\usepackage[letterpaper,left=2.5cm, right=2.5cm, top=2.5cm, bottom=2.5cm]{geometry}

\begin{document}
\begin{letter}{}

  \address{Alex Groce\\School of Informatics, Computing \& Cyber Systems\\Northern Arizona University}

  \opening{Dear editor:}

  Attached you will find our submission, ``Does It Always Do That?  Practical Automatic Lightweight Nondeterminism
  and Flaky Test
  Detection and Debugging for Python Libraries.''  This manuscript is original and not in submission elsewhere.  Reviewers knowledgable in delta debugging or random testing systems might be most appropriate for the core content, which concerns the testing mechanics as much as the concepts involved in nondeterminism.

  \closing{Sincerely,\\Alex Groce (agroce@gmail.com)\\Associate
    Professor\\
    School of Informatics, Computing \& Cyber Systems\\Northern Arizona University}

\end{letter}
\end{document}

### 2020-01-07 ISSTA/cover2.tex
issta?

\documentclass{letter}
%\def\today{September 10, 2019}
\usepackage[letterpaper,left=2.5cm, right=2.5cm, top=2.5cm, bottom=2.5cm]{geometry}

\begin{document}
\begin{letter}{}

  \address{Alex Groce\\School of Informatics, Computing \& Cyber Systems\\Northern Arizona University}

  \opening{Dear editor:}

  Attached you will find a revision of our submission, ``Does It Always Do That?  Practical Automatic Lightweight Nondeterminism
  and Flaky Test
  Detection and Debugging for Python Libraries.''  We believe this
  revision addresses all reviewer concerns in the two reviews, and
  detail our response in the response to reviewers.

  \closing{Sincerely,\\Alex Groce (agroce@gmail.com)\\Associate
    Professor\\
    School of Informatics, Computing \& Cyber Systems\\Northern
    Arizona University \vspace{0.1in}\\Josie Holmes\\Affiliated Researcher\\
    School of Informatics, Computing \& Cyber Systems\\Northern Arizona University}

\end{letter}
\end{document}

### 2020-01-07 ISSTA/highlights.tex
issta?

\documentclass{article}

\begin{document}

\noindent {\bf DOES IT ALWAYS DO THAT? PRACTICAL AUTOMATIC LIGHT- WEIGHT NONDETERMINISM AND FLAKY TEST DETECTION AND DEBUGGING FOR PYTHON LIBRARIES}

\vspace{0.1in}

\noindent {\bf Highlights:}

\vspace{0.1in}

\noindent This article presents a new approach to ``flaky tests'' and nondeterminism.

\vspace{0.1in}

\noindent The article presents a formalism distinguishing horizontal and vertical determinism.

\vspace{0.1in}

\noindent The approach is implemented in an open source Python library testing system.

\vspace{0.1in}

\noindent Failure determinism can detect 6\% more program mutants than a strong specification.

\vspace{0.1in}

\noindent The article presents the first examination of test reduction for probabilistic properties.

\end{document}

### 2020-01-07 ISSTA/relatedwork.tex
issta?

--- 2019-02-28 relatedwork.tex
@@
 
-The present paper, rather than focusing on flaky tests as such, therefore, investigates methods for handling nondeterminism in property-based test generation.
+The present paper, rather than focusing on flaky tests as such, therefore, investigates methods for handling nondeterminism in property-based test generation \cite{ClaessenH00,Papadakis:2011:PIT:2034654.2034663}.
 Again, from our point of view, flaky behavior is simply a special case of nondeterminism, where the nondeterminism is sufficient to cause the test to have different pass/fail results at times.
@@
 
+Vertical/failure determinism is a less-studied concept, and to our knowledge the formulation here, as a kind of determinism in a different ``direction,'' has not previously appeared; the kinds of errors that are exposed, however, are not new.
+In particular, we believe that many faults discovered using failure determinism are related to the propagation of error conditions in code, which has been studied in, e.g., file systems, using static analysis \cite{FileProp}.
+
+TSTL's approach to test generation, and our instantiation of the formal definitions of nondeterminism are based on a formulation of unit tests using \emph{pools} of values \cite{AndrewsTR}, which provides a practical solution to the problem of defining the visible state to be compared when checking for nondeterminism.
+
 Finally, Cotroneo et al.

### 2020-01-10 ISSTA/abstract.tex
abs

@@
 We show that lightweight techniques, requiring little effort on the part of developers, can be used to extend an existing testing system to allow detection of nondeterminism.
-We introduce a set of lightweight nondeterminism properties, inspired by real faults, and tailor these notions to the practical, automatic, checking of Python library code.
+We introduce a set of lightweight nondeterminism properties, inspired by real faults, and tailor these notions to the practical, automatic, checking and debugging (through improvements to the delta-debugging algorithm that support probabilistic properties) of Python library code.
 In particular, we demonstrate that our proposed failure nondeterminism can improve mutation score by 6\% for a strong, full differential test harness for a widely used mock file system.

### 2020-01-10 ISSTA/implementation.tex
issta shrink start

--- 2019-09-09 implementation.tex
@@
 
+\begin{comment}
 The implementation of the above ideas is conceptually straightforward,
@@
 overview of the TSTL testing tool (Section \ref{sec:tstlintro}).
+\end{comment}
 
@@
       method to a value pool, with all arguments to the method also
-      taken from a value pool, e.g. {\tt intVP[2] = fooObjectVP[2].bar(intVP[0],
+      taken from a value pool, e.g. {\tt intVP[2] = fooVP[2].bar(intVP[0],
         intVP[1]);} (these are referred to as \emph{array-canonical} statements).
@@
 
-\subsection{A Brief Introduction to TSTL}
-\label{sec:tstlintro}
+%\subsection{A Brief Introduction to TSTL}
+%\label{sec:tstlintro}
 
@@
   appropriate version of the tested software, by doing, e.g., {\tt pip
-    install pyfakefs}.};
+    install pyfakefs}.}.
+\begin{comment}
 TSTL can also be installed using Python's {\tt pip} tool, by typing
@@
 \end{figure}
+\end{comment}
 
@@
 result from executing out the Python code in an action; these actions
-are, roughly, the array-canonical statements of the value pool formalism.  Figure
+are, roughly, the array-canonical statements of the value pool
+formalism.
+
+\begin{comment}
+  Figure
 \ref{fig:simple} shows a very simple TSTL harness that tests Python's
@@
 
+\end{comment}
+
 \subsection{Implementing Horizontal Nondeterminism Detection}
@@
 
+\begin{comment}
 \begin{table*}
@@
 
+\end{comment}
+
 Note that TSTL provides explicit support for exploring the real-world sources of
 nondeterminism described in Section \ref{sec:sources}, by supporting
-execution in a new process ({\tt processNondeterministic} API call and
-{\tt --checkProcessDeterminism} command line argument) and allowing
+execution in a new process and allowing
 control over timing changes in re-execution, by specifying a delay to
-introduce between actions (parameters to API calls, {\tt
-  --determinismDelay} command line arugment).  Changing the timing of
+introduce between actions.  Changing the timing of
 steps is very effective when exact timing is somehow involved in
@@
 
+\begin{comment}
 \subsection{Tools for Nondeterministic Delta-Debugging}
@@
 values used, as shown by the presence of command line options and
-optional function parameters in Tables \ref{tab:methods} and \ref{tab:options}.
+optional function parameters in Tables \ref{tab:methods} and
+\ref{tab:options}.
+\end{comment}
 

### 2020-01-10 ISSTA/concepts.tex
shrink

--- 2019-08-30 concepts.tex
@@
 is correctly defined, indicates a functional correctness fault in (at
-least) one of the systems under test.  Being able to detect functional
-correctness errors without the cost of constructing a formal
-specification is extremely useful in automated test generation.  Differential testing is widely
+least) one of the systems under test, and  is widely
 used for systems software components such as compilers
@@
 implementations are common.  The major limitation of differential
-testing, of course, is that multiple implementations of a system are
+testing is that multiple implementations of a system are
 almost as rare as good correctness specifications.
@@
   system can serve as its own reference implementation.}    The
-question, then, becomes one of deciding at what granularity the
-reference equivalence will be checked:  as discussed above,
+problem, then, becomes one of deciding at what granularity the
+reference equivalence will be checked:  e.g., 
 processor-instruction and memory-layout determinism is seldom
-necessary or even desired (it would greatly limit optimizations of
-code execution).  We propose two approaches to ``aligning'' an
+necessary or even desired.  We propose two approaches to ``aligning'' an
 execution with itself.
@@
 called as a library by other code}, as the criteria for determining if two executions
-are equivalent.  The idea is simple:  determinism is motivated by the
+are equivalent.  Determinism is motivated by the
 desire to create consistent behavior for an observer, whether that
 observer is a human user, another software system, or a regression
-test.  The values output by a software element are the only values of
-interest to the observer.  In practice, of course, some values (time
+test.  In practice, of course, some values (time
 stamps, pointer addresses, etc.) are not expected to be deterministic
@@
 time, memory management).   Any mechanism for visible value
-determinism needs to support automatic and manual designation of some
+determinism needs to support designation of some
 values as opaque.
@@
 Visible value determinism has significant limitations.  While it
-provides the finest granularity for detecting nondeterminism, which is
+provides the finest granularity, which is
 important for debugging purposes, it is also expensive, requiring
-checks on (and possibly storage of) a potentially very large number of
+checks on a potentially very large number of
 values.  Additionally, in some cases so many values produced by a
@@
 idempotent operations in an automated test generation harness, and
-simply automatically retry all such operations, checking that the
+automatically retry all such operations, checking that the
 result is unchanged.  The overhead of vertical nondeterminism detection will
@@
 A specialized case of vertical determinism is failure determinism.
-Failure determinism is the following restriction on an API:
-
-{\bf If a call \emph{fails}, and indicates this to the caller, it should not modify system
-  state in any way; changes should be ``rolled back.''}
-
-In other words, failure determinism is a property stating that an API
-is \emph{transactional} with respect to failures (it may not be so
-with respect to non-failing calls).
-
-From a user's perspective, some behaviors of the Mac OS High Sierra
+Failure determinism is the following restriction on an API: {\bf If a call \emph{fails}, and indicates this to the caller, it should not modify system
+  state in any way; changes should be ``rolled back.''}   In other words, failure determinism is a property stating that an API
+is \emph{transactional} with respect to failing calls.  From a user's perspective, some behaviors of the Mac OS High Sierra
 root exploit (CVE-2017-13872 \cite{applebug0}) exhibited failure
@@
 overhead for vertical determinism, in that we can expect most failing
-operations to be fast (since in test generation, failure is usually
+operations to be fast (since failure is usually
 due to invalid parameters).  Checking equivalence of the full
@@
 this is a sign that the operation is not deterministic with respect to
-failure.  Note that this more restrictive notion of failure
-nondeterminism captures such issues as the Apple login bug.  The
-changes to visible or hidden state we are most concerned with are
-usually those related to the failure itself.
+failure.  Note that this more restrictive notion still captures such issues as the Apple login bug.
 
@@
 fail.  Due to the nature of (horizontal) nondeterminism, such failure
-is almost always going to be inherently flaky --- sometimes a test
+is almost always going to be inherently flaky---sometimes a test
 will behave nondeterministically, and sometimes it will not.  But the
@@
 nondeterminism, we need only see a change in values in a test, not in
-the disposition (failed/passed) of the test.  It may be that, without
+the disposition (failed/passed) of the test.
+\begin{comment}
+  It may be that, without
 the additional property of determinism, the change cannot actually
@@
 flakiness may well result.
+\end{comment}
 
@@
 ``nondeterministic'' (from a software library users' point of view)
-behavior.  Vertical nondeterminism may rely on ``real'' nondeterminism
-that could produce a flaky test, but it does not have to do so.  In
-other words, vertical nondeterminism is often not a problem for
-testing (flakiness), but simply a bug in the behavior of a software
+behavior.   That is, vertical nondeterminism is often simply a bug in a software
 system, where an implicit specification on failed operations is
@@
 under the same circumstances'' which is the \emph{conceptual} common
-factor in all kinds of expected determinism.
+factor in ``determinism.''
 
@@
 We can formally distinguish the types of nondeterminism by using a
-variant of a labeled transition system (LTS) as our model:
+variant of a labeled transition system (LTS) as our model, where (1) $S$
+is a set of states; (2) $V$ is a set of \emph{observable} states, where
+$|V| \leq |S|$; (3) $v: S \rightarrow V$ is a total function that, given a state, maps it
+  into the set of observable states, such that every state has an observable
+  component, which may be the complete state, or only an aspect of the
+  full state; (4)  $I \subseteq S$ is a set of initial states; (5) $A$
+  is a set of \emph{actions}; and (6) $T \subseteq S \times A \times S$ is a transition relation.
 
+  \begin{comment}
 \begin{itemize}
@@
 \end{itemize}
+\end{comment}
 
@@
 \xrightarrow[]{\alpha_{n-1}} s_n$ where $s_0 \in I$ and $\forall i < n
-. (s_i, \alpha_i, s_{i+1}) \in T$.
-
-The concept of an action does not restrict this approach to reactive
+. (s_i, \alpha_i, s_{i+1}) \in T$.  The concept of an action does not restrict this approach to reactive
 systems; it can be generalized to consider the only action that varies
@@
 modification of most such tools (e.g., random testers or fuzzers, and
-even model checkers).  In general, testing tools do not produce
+even model checkers), only to find an
+action sequence or input which, when executed twice, produces a $t_1$ and $t_2$
+meeting the requirements for showing nondeterminism.
+
+\begin{comment}In general, testing tools do not produce
 \emph{traces}, they produce \emph{action sequences} or \emph{inputs}.  The two traces
@@
 non-equivalence of $v(s)$ for some $s$.
+\end{comment}
 
@@
 
-Vertical nondeterminism, in contrast, is demonstrated by a single trace
-$t$, not a pair of traces.  To define vertical nondeterminism with
+To define vertical nondeterminism with
 respect to idempotency, we can
@@
 $\alpha_i \in \mathcal{I} \wedge \alpha_i = \alpha_{i+1}$ and $v(s_{i+1})
-\neq v(s_{i+2})$.
-
-Note that vertical nondeterminism may, in theory, exhibit only
+\neq v(s_{i+2})$.  Note that vertical nondeterminism may, in theory, exhibit only
 after a sequence of more than two successive applications of a (supposedly)
@@
 ``copies'' will suffice, since an operation is either truly idempotent or
-not idempotent.  Only the existence of \emph{non-visible} state changes that
-can eventually produce visible state changes \emph{after a succession of the
-same action} require more repetitions.
+not idempotent.
+%Only the existence of \emph{non-visible} state changes that
+%can eventually produce visible state changes \emph{after a succession of the
+%same action} require more repetitions.
 
@@
 \times A \times S \times {\tt F: bool}$, where the boolean $F$ indicates
-whether the action $A$ \emph{failed}.
-
-A trace $t = s_0
+whether the action $A$ \emph{failed}.   A trace $t = s_0
 \xrightarrow[]{\alpha_0} (s_1, F_0) \xrightarrow[]{\alpha_1} \ldots
@@
 a state, or any comparison of states.  It only
-requires that we be able to say whether an operation failed.
+requires that we know whether an operation failed.
 
-\section{Sources of Nondeterminism in Practice}
+\section{Sources of Nondeterminisms}
 \label{sec:sources}
@@
 primary sources, and most of which can be detected, at least in
-principle, by multiple runs of a test within the same testing process.  However, simply running a test twice in the same process (possibly with a delay between steps) does
-not suffice in all cases.
+principle, by multiple runs of a test within the same testing process.  %However, simply running a test twice in the same process (possibly with a delay between steps) does
+%not suffice in all cases.
 
@@
 process environment, because the source is inherently tied to the
-process in which code runs; simply re-executing a test in the same
-process will not reveal the problem.  In terms of our formalism, an
+process in which code runs.  In terms of our formalism, an
 action needs to be introduced that indicates the generation of a fresh
-process identity.
-
-Address Space Layout Randomization (ASLR) \cite{ASLR}  is probably the most
+process.  Address Space Layout Randomization (ASLR) \cite{ASLR}  is probably the most
 important source of nondeterminism that arises (only) from change in
@@
 other times not fail at all.  Even if no inputs produce a crash,
-however, memory errors may produce variation in values that are
-overwritten or arise from uninitialized memory, and thus be detectable
+however, memory errors may produce variation in values and thus be detectable
 as nondeterminism.
@@
 
-Other, less common, effects tied to process also exist.  A few tests
-may somehow depend on their actual PID (Process ID) or at least the
-parent process' ID.  Process change increases the likelihood that on a multi-core
-system a test will execute on a different CPU than in the first
-execution, which can have many subtle effects, mostly related to
-changed timing.
+%Other, less common, effects tied to process also exist.  A few tests
+%may somehow depend on their actual PID (Process ID) or at least the
+%parent process' ID.  Process change increases the likelihood that on a multi-core
+%system a test will execute on a different CPU than in the first
+%execution, which can have many subtle effects, mostly related to
+%changed timing.
 
@@
 order are easily exposed in random testing, since the
-order-dependence-inducing operations will seldom usually happen in the
+order-dependence-inducing operations will seldom happen in the
 same order in randomly generated tests.  For example, consider the
@@
 tester is likely to expose the problem as a (deterministic, not flaky) failure.
-
 However, for cases where
@@
 actual bug, using a fresh
-process may expose such dependencies --- the ``environment'' of the
+process may expose such dependencies---the ``environment'' of the
 parent process will have the, e.g., flag set or value initialized, but
@@
 \cite{icst2014,stvrcausereduce,NonAdeq}.  Our contribution here is to consider
-what happens when $\mathit{PRED}$ not deterministic.
+what happens when $\mathit{PRED}$ is not deterministic.
 ``The test behaves flakily'' is a simple, important, example of such a $\mathit{PRED}$.
@@
 increase the probability that $t$ behaves nondeterministically.
-
 The $\mathit{PRED}$ of interest for horizontal determinism is basically of
@@
 
-\subsection{Using Replications to Mitigate ``Publication Bias''}
+\subsection{Replication Mitigates  ``Publication Bias''}
 

### 2020-01-10 ISSTA/introduction.tex
shrink

--- 2019-03-07 introduction.tex
@@
 Automated test generation systems using test coverage results to drive
-the search for interesting inputs may be mislead when a run executes
+the search for interesting inputs may go awry when a run executes
 code only rarely.  And, most significantly, regression testing
@@
 we would just do so).  As a consequence, in software, rather than
-defining determinism in terms of prediction, we usually therefore
-simply say that a system is deterministic if, given the same state and
+defining determinism in terms of prediction, we usually 
+say that a system is deterministic if, given the same state and
 inputs, it always produces the same outputs.
@@
 of the system (which includes the entire state of the underlying
-hardware, the operating system, storage devices, etc.), ignoring
-quantum effects, and treating outside interventions such as network
-traffic, human activity at an input device, etc. properly as inputs,
+hardware, the operating system, storage devices, etc.),
 most software \emph{is} completely predictable.  What we actually
-mean, usually, is that, given a certain limited abstraction of state
-and of inputs, observable behavior is repeatable.  This abstraction,
-for most software, is not expected to include many elements outside of
-the software system itself.
+mean is that, given a certain limited abstraction of state
+and inputs, observable behavior is repeatable.  %This abstraction,
+%for most software, is not expected to include many elements outside of
+%the software system itself.
 
@@
 controlled for; however, this kind of
-nondeterminism is usually of no interest, unless the test involves
-extremely precise timing constraints.  Rather, what matters is when
+nondeterminism is usually of no interest.  Rather, what matters is when
 some kind of nondeterminism unexpectedly impacts computed values at a
-higher level of abstraction; in general, if the operating system
+higher level of abstraction, e.g. when it is believed that the
+behavior of a thread scheduler will not matter, but a race condition
+in the code means that it does matter, or when the order of items in
+an iterator on a hash table is important, and the hash values used are
+randomly salted.
+\begin{comment} in general, if the operating system
 itself is not buggy, low-level nondeterminism, by design, is invisible
@@
 randomly salted.
+\end{comment}
 
-\subsection{The High Cost of Unexpected Nondeterminism}
+%\subsection{The High Cost of Unexpected Nondeterminism}
 
@@
 system.  While developers may know how to produce reliably
-deterministic unit tests, even in the presence of underlying
-nondeterminism, automated test generation, without a large investment
-in human time, usually does not have sufficient information to avoid
+deterministic unit tests, automated test generation usually does not have sufficient information to avoid
 producing the occasional such test.  Moreover, where a human might
@@
 an automatically generated regression test is usually most easily produced by
-simply asserting equality with all values produced during a reference
+asserting equality with all values produced during a reference
 run, since the final step of a test
-is likely to be somewhat arbitrary, generated in an effort to increase
-code coverage, not establish some functional correctness property.
+is arbitrary, generated in an effort to increase
+code coverage, not establish functional correctness.
 This may be one reason that automated testing is seldom used to
@@
 and techniques proposed in this paper all apply to flaky tests, as a
-special case of various kinds of \emph{horizontal} nondeterminism described below.
+special case of various kinds of \emph{horizontal} nondeterminism
+described below.  Our focus, however, unlike all other tools and
+approaches we are aware of, is on detecting the \emph{sources} of
+flaky-ness \emph{before} they propagate to cause actual test failure.
 
-\subsection{Contributions:  Detecting and Debugging Nondeterminism}
+\subsection{Contributions}
 
@@
 
-The underlying idea in this paper is that most approaches to
+Previous practical approaches to
 nondeterminism/flaky test detection and debugging rely either on 1) identifying
@@
 sources of nondeterminism, the documentation of sources of
-nondeterminism that might produce problems for manually-constructed tests, and, most importantly, detection and effective debugging for
+nondeterminism that might produce problems for manually-constructed
+tests, and, most importantly, detection and \emph{effective debugging}
+(using a novel modification of delta-debugging designed for
+nondeterminism) for
 unexpected nondeterminism that is, itself, a bug.  Additonally, by
-identifying (using automated testing) library behaviors that are
+identifying library behaviors that are
 nondeterministic, our approach should make it much easier to flag
@@
 behavior:  simply scan the
-code for calls to functions identified as sources of nondeterminism.
-
-We use a set
+code for calls to such functions.  We use a set
 of case studies. including real-world, widely-used, Python libraries to demonstrate the

### 2020-01-10 ISSTA/concepts.tex
shrinking

@@
 
+\begin{comment}
 \subsection{Nondeterminism vs. Flakiness}
@@
 the disposition (failed/passed) of the test.
-\begin{comment}
+%\begin{comment}
   It may be that, without
@@
 flakiness may well result.
-\end{comment}
+%\end{comment}
 
@@
 factor in ``determinism.''
+\end{comment}
 
@@
 
-\section{Sources of Nondeterminisms}
+\section{Sources of Nondeterminism}
 \label{sec:sources}
@@
 (approximate) enviroment, of the same test.  The two broad classes of
-nondeterminism that can usually be detected (with some probability) by
-simply executing the same test twice in succession, without attempting
+nondeterminism that can usually be detected by
+executing the same test twice in succession, without attempting
 to control any other factors, are:
@@
   test, which are subject to uncontrolled variance.  Calling {\tt
-    random()} in Python is an obvious example; the call is almost
-  guaranteed to return a different value each time it is called in a
-  test, unless we explicitly re-seed the random number generator at
-  the beginning of each test.  Similarly, calls to {\tt time}
+    random()} in Python is an obvious example. Similarly, calls to {\tt time}
   functions  will usually return different values on different runs,
-  and even if only elapsed time, rather than absolute, is used to
-  influence behavior, subtle variances in timing are to be expected
-  due to system load, I/O, and other external factors.  Finally, in
-  distributed system testing, the environment is often remote servers
-  (or clients) that may be unavailable, slower than expected, or
-  subject to faulty behavior for reasons not under the control of the
-  test.
-\item {\bf Concurrency:} Perhaps the second most common cause for
+  and even if only elapsed time is used, due to system load, I/O, and
+  other external factors.
+  %Finally, in
+  %distributed system testing, the environment is often remote servers
+  %(or clients) that may be unavailable, slower than expected, or
+  %subject to faulty behavior for reasons not under the control of the
+  %test.
+\item {\bf Concurrency:} Ae second common cause for
   nondeterminism in tests is when an algorithm is implemented using
@@
   and a developer has not accounted for this possibility, the
-  algorithm may return different locations of an item for the same list,
-  depending on the scheduling of threads.  Arguably, this can be
-  subsumed under the first class, with the behavior of the
-  processor/scheduler as the ``enviromental'' factor; relying on
-  scheduling is, in a sense, equivalent to calling {\tt random} or
-  {\tt time}.
+  algorithm may return different locations of an item
+  depending on the scheduling of threads.
+  %Arguably, this can be
+  %subsumed under the first class, with the behavior of the
+  %processor/scheduler as the ``enviromental'' factor; relying on
+  %scheduling is, in a sense, equivalent to calling {\tt random} or
+  %{\tt time}.
 \end{itemize}

### 2020-01-10 ISSTA/conclusions.tex
shrinking

--- 2019-09-05 conclusions.tex
@@
 AFL \cite{aflfuzz} to avoid high startup costs for test executions.
-
 More generally, we are interested in using test decomposition
 \cite{Composition} to more easily isolate, understand, and avoid
-nondeterminism is tests, and to mitigate the problem of flaky tests.
+nondeterminism in tests, and to mitigate the problem of flaky tests.
 Reliable detection of nondeterminism is a first step towards
@@
 
+\begin{comment}
 Finally, it would be useful to implement versions of our formalisms in
@@
 Colohan for discussion of sources of process-based nondeterminism.}
+\end{comment}

### 2020-01-14 ISSTA/abstract.tex
shorter

--- 2020-01-07 ISSTA/abstract.tex
@@
 Nondeterministic behavior that is not either controlled or accounted for can result in flaky tests as well as causing problems for test reduction, differential testing, and automated regression test generation.
-We show that lightweight techniques, requiring little effort on the part of developers, can be used to extend an existing testing system to allow detection of nondeterminism.
-We introduce a set of lightweight nondeterminism properties, inspired by real faults, and tailor these notions to the practical, automatic, checking of Python library code.
-In particular, we demonstrate that our proposed failure nondeterminism can improve mutation score by 6\% for a strong, full differential test harness for a widely used mock file system.
+We show that lightweight techniques, requiring little effort on the part of developers, can be used to extend an existing testing system to allow detection and debugging of nondeterminism.
+We show that our methods can improve mutation score by 6\% for a strong, full differential test harness for a widely used mock file system, and make delta-debugging practical for probabilistic faults.

### 2020-01-14 ISSTA/abstract.tex
shorter

@@
 Many software systems are implemented using mechanisms (unsafe languages, concurrency, caching, stochastic algorithms, environmental dependencies) that can introduce behavioral nondeterminism.
-Users of software systems, especially other software using library calls in a single-threaded context, often expect that systems will behave deterministically in the sense that they have predictable results from the same series of operations.
-Equally importantly, even when it does not impact correctness, nondeterminism must be understood and managed for effective (especially automated) testing.
-Nondeterministic behavior that is not either controlled or accounted for can result in flaky tests as well as causing problems for test reduction, differential testing, and automated regression test generation.
+Users of software systems, especially other software systems using a library, often expect that systems will behave deterministically.
+Even when it does not impact correctness, nondeterminism must be understood and managed for effective automated testing.
+Nondeterministic behavior can result in flaky tests as well as causing problems for test reduction, differential testing, and automated regression test generation.
 We show that lightweight techniques, requiring little effort on the part of developers, can be used to extend an existing testing system to allow detection and debugging of nondeterminism.
-We show that our methods can improve mutation score by 6\% for a strong, full differential test harness for a widely used mock file system, and make delta-debugging practical for probabilistic faults.
+We show that our methods can improve mutation score by 6\% for a strong, full differential test harness for a widely used mock file system, and show how to make delta-debugging effective for probabilistic faults in general.

### 2020-01-14 ISSTA/introduction.tex
contribs

@@
 behavior, without having to observe a test actually fail due to nondeterminism.
+we additionally introduce the first variation of delta-debugging that properly handles probabilistic faults and reduction properties.
 
-This approach enables a number of useful results:
- the automated production of
+Our methods enable automated production of
 nondeterminism-free regression tests even for libraries with some
-sources of nondeterminism, the documentation of sources of
+sources of nondeterminism, documentation of sources of
 nondeterminism that might produce problems for manually-constructed
-tests, and, most importantly, detection and \emph{effective debugging}
-(using a novel modification of delta-debugging designed for
-nondeterminism) for
-unexpected nondeterminism that is, itself, a bug.  Additonally, by
-identifying library behaviors that are
-nondeterministic, our approach should make it much easier to flag
-statements in existing human-written unit tests that may produce flaky
-behavior:  simply scan the
-code for calls to such functions.  We use a set
-of case studies. including real-world, widely-used, Python libraries to demonstrate the
-utility of our ideas.  Failure determinism directly
+tests, and, most importantly, detection of 
+unexpected nondeterminism that is, itself, a bug.  We use a set
+of case studies. including real-world, widely-used, Python libraries, to demonstrate the
+utility of our ideas.  Our novel notion of failure determinism directly
 produces a more than 5\% improvement in mutation score for an already highly

### 2020-01-14 ISSTA/abstract.tex
shrink

@@
 Nondeterministic behavior can result in flaky tests as well as causing problems for test reduction, differential testing, and automated regression test generation.
-We show that lightweight techniques, requiring little effort on the part of developers, can be used to extend an existing testing system to allow detection and debugging of nondeterminism.
-We show that our methods can improve mutation score by 6\% for a strong, full differential test harness for a widely used mock file system, and show how to make delta-debugging effective for probabilistic faults in general.
+We show that lightweight techniques, requiring little effort on the part of developers, can extend an existing testing system to allow detection and debugging of nondeterminism.
+We show how to make delta-debugging effective for probabilistic faults in general, and that our methods can improve mutation score by 6\% for a strong, full differential test harness for a widely used mock file system.

### 2020-01-14 ISSTA/introduction.tex
shrink

@@
 
+\begin{comment}
 Previous practical approaches to
@@
 allows detection of (potential sources of) flakiness, based on real divergence in
-behavior, without having to observe a test actually fail due to nondeterminism.
-we additionally introduce the first variation of delta-debugging that properly handles probabilistic faults and reduction properties.
+behavior, without having to observe a test actually fail due to
+nondeterminism.  we additionally introduce the first variation of
+delta-debugging that properly handles probabilistic faults and
+reduction properties.
+\end{comment}
 
@@
 nondeterminism that might produce problems for manually-constructed
-tests, and, most importantly, detection of 
+tests, and, most importantly, detection and debugging of 
 unexpected nondeterminism that is, itself, a bug.  We use a set

### 2020-01-14 ISSTA/relatedwork.tex
shrink

@@
 Our approach is, in a sense (especially the delta-debugging modifications) in line with Harman and O'Hearn's proposal to simply accept that ``All Tests Are Flaky'' \cite{StartupstoScaleups}, and work with probabilistically failing tests.
+To our
+knowledge, this is the first approach to the problem that
+allows detection of (potential sources of) flakiness, based on real divergence in
+behavior, without having to observe a test actually fail due to
+nondeterminism.  we additionally introduce the first variation of
+delta-debugging that properly handles probabilistic faults and
+reduction properties.
 
@@
 
+\begin{comment}
 Finally, Cotroneo et al.
@@
 This work informs our attempt to identify sources of nondeterminism, and should provide other, context and project-specific, sources that could be introduced into our general framework.
+\end{comment}

### 2020-01-16 ISSTA/experiments.tex
issta shrink start
shrink
shrinking

--- 2019-09-09 experiments.tex
@@
   \centering
+  {\scriptsize
 \begin{tabular}{l|l|r|r|l}
@@
 \end{tabular}
+}
 \caption{Overview of experimental results}
@@
 @import swapsort
-\vspace{0.1in}
+%\vspace{0.1in}
 pool: <val> 10
@@
 pool: <sorted> 10
-\vspace{0.1in}
+%\vspace{0.1in}
 <val> := <[-10..10]>
-\vspace{0.1in}
+%\vspace{0.1in}
 <data> := []
@@
 len(<data,2>) < len(<data,1>) -> <data>.extend(<data>)
-\vspace{0.1in}
+%\vspace{0.1in}
 <sorted> := mergesort.merge\_sort\_parallel(<data>)
@@
 <sorted>
-\vspace{0.1in}
+%\vspace{0.1in}
 property: swapsort.is\_sorted(<sorted>)
@@
 
-\begin{figure*}
+\begin{figure}
 \centering 
@@
 \caption{Delta-debugging of the same {\tt redis-py} nondeterministic test}
-\end{figure*}
+\end{figure}
 
@@
 
+\begin{comment}
 \begin{figure*}
@@
 \end{figure*}
+\end{comment}
 
-Figure \ref{fig:flakysource} shows all calls removed or modified after
+We removed 11 calls from the {\tt redis} harness, and  modified 2
+calls, after
 identification of sources of nondeterminism.  After making these changes, no flakiness was observed in a sample of over 2,000 length
@@
 @import datarray
-\vspace{0.1in}
+%\vspace{0.1in}
 <@
@@
 @>
-\vspace{0.1in}
+%\vspace{0.1in}
 pool: <P> 3
@@
 pool: <event> 3
-\vspace{0.1in}
+%\vspace{0.1in}
 <P> := 0.01 * <[0..100]>
@@
 ~<evidence>.update([(<ename>,0)])
-\vspace{0.1in}
+%\vspace{0.1in}
 \{Exception\} print(flatten\_and\_sort(inference\_algs.calc\_marginals\_simple(<cpts>,
@@
 
+\begin{comment}
 \subsection{Summary of Results}
@@
 \end{itemize}
+\end{comment}
 
-\section{Threats to Validity}
+\subsection{Threats to Validity}
 
@@
 the flaky tests'' beyond identifying them as likely related to test
-order, or not.
-
-However, the primary
+order, or not.  However, the primary
 aim of our results is to show 1) that nondeterminism at the library
@@
 that the overhead for such detection, and cost to delta-debug
-nondeterministic tests, is not prohibitive.  We believe that the case
+nondeterministic tests, is not prohibitive.
+\begin{comment}
+We believe that the case
 studies show that users testing Python libraries, and concerned about
@@
 existing approach, are not proposed for evaluation.
-
+\end{comment}
 Another threat to validity is posed by implementation errors.
 TSTL and the mutation tool are tested for bugs by a set of Travis CI
-tests, and are (for academic research code) moderately well-used by
-external users, making this a lesser concern.  The nondeterminism
-detected by TSTL can be demonstrated with standalone tests TSTL
-generates that do not depend on any TSTL code.
+tests, and the nondeterminism
+detected by TSTL can be demonstrated with standalone tests that do not depend on any TSTL code.
 The source code for the tools and all experimental subjects is

### 2020-01-16 ISSTA/implementation.tex
shrink
shrinking

@@
     \item A part in which all statements are assignments of calls of a
-      method to a value pool, with all arguments to the method also
-      taken from a value pool, e.g. {\tt intVP[2] = fooVP[2].bar(intVP[0],
+      method to a value pool, with all arguments also
+      taken from a pool, e.g. {\tt intVP[2] = fooVP[2].bar(intVP[0],
         intVP[1]);} (these are referred to as \emph{array-canonical} statements).
@@
 
-\subsection{Implementing Horizontal Nondeterminism Detection}
+\subsection{Horizontal Nondeterminism Detection}
 \label{sec:horizimp}
@@
 
-\subsection{Implementing Vertical Nondeterminism Detection}
+\subsection{Vertical Nondeterminism Detection}
 \label{sec:vertimp}
@@
 
+\begin{comment}
 \section{A Simple Example}
@@
 init: init()
-\vspace{0.1in}
 pool: <int> 5
@@
 \begin{code}
-import random
-
-storage = []
-
-def contents():
+import random                                   def getNewVal():
+storage = []                                            global storage
+def contents():                                        if len(storage) >= 1:
     global storage
     return storage
-
 def init():
@@
     storage = []
-
 def store(n):
@@
 is just a ``normal'' failure, exhibited by a single trace.
+
+\end{comment}

### 2020-01-16 ISSTA/concepts.tex
shrink
big shrink

@@
 element, whether it was added or was already in the tree.  Calling
-this method any number of times in sequence should always return the
-same value, though the first call may have modified the contents of
-the AVL tree.
+this method any number of times in a row should always return the
+same value
 
-One application of vertical determinism, then, is to identify
+One approach, then, would be to identify
 idempotent operations in an automated test generation harness, and
@@
 only re-execute proportional to the number of idempotent operations
-performed).  However, identifying idempotent operations is a
-fairly serious specification burden.  Are there instances where a tool
+performed).  However, identifying idempotent operations is an additional specification burden.  Are there instances where a tool
 can automatically identify a limited kind of idempotent behavior?
@@
 empty password appeared to fail, then, on a repeated try, succeeded
-\cite{applebug1,applebug2}.
-
-Many library APIs are largely failure deterministic.  For instance, if
+\cite{applebug1,applebug2}.   Many library APIs are largely failure deterministic.  For instance, if
 we exclude actual I/O errors, most POSIX file system calls either
-succeed or do nothing (interfaces that may only partially succeed,
-such as {\tt read} tend to explicitly return the degree of success
-rather than signaling an error on partial success).  In fact, the
-design of the POSIX {\tt write} call is carefully crafted to largely maintain
-failure determinism.  If the failure mode is one that can be
-anticipated, such as insufficient space to write all bytes requested
-(but some bytes can be written), the call returns the actual bytes
-written, and does not set a failure code.
+succeed or do nothing; interfaces that may only partially succeed,
+such as {\tt read} and {\tt write} tend to explicitly return a degree of success, rather than signalling total failure, when appropriate.
 
@@
 (e.g., exceptions in Python and Java, or Option/Result types in Rust,
-Haskell, and ML), failure determinism can be automatically checked,
-given a notion of general vertical determinism.
-Moreover, the expected overhead should be even lower than the usual
-overhead for vertical determinism, in that we can expect most failing
+Haskell, and ML), failure determinism can easily be automatically checked.
+Moreover, the expected overhead should be even lower than for other vertical determinism, in that we can expect most failing
 operations to be fast (since failure is usually
@@
 the observable components of a state.  A more lightweight, but (as we
-will show) still quite powerful way of checking for failure nondeteterminism is to
-repeat the failing operation, and check if it still fails.  If not,
-this is a sign that the operation is not deterministic with respect to
-failure.  Note that this more restrictive notion still captures such issues as the Apple login bug.
+will show) still powerful way of checking for failure nondeteterminism is to
+\emph{repeat the failing operation, and check if it still fails}.  If not,
+this is a sign that the operation is not failure deterministic; this more restrictive notion still captures such issues as the Apple login bug.
 
@@
 %\end{comment}
-
-Vertical nondeterminism, in fact, is potentially \emph{completely} unrelated
+Vertical nondeterminism, in fact, is usually \emph{completely} unrelated
 to flakiness, in that some bugs (improper handling of hardware errors
@@
 ``nondeterministic'' (from a software library users' point of view)
-behavior.   That is, vertical nondeterminism is often simply a bug in a software
-system, where an implicit specification on failed operations is
-consistently violated under certain conditions.  The behavior is
-``nondeterministic'' from a caller's point of view, however, in that
+behavior. The behavior is only
+``nondeterministic'' from a caller's point of view in that
 there is an expectation that a call will always ``do the same thing,
@@
 
+\begin{comment}
 A point to emphasize is that while horizontal determinism
@@
 meeting the requirements for showing nondeterminism.
+\end{comment}
 
@@
 In order to implement a practical nondeterminism-detection tool, it is
-important to consider the real-world sources of nondeterminism.  Most of these
+important to consider the real-world sources of nondeterminism.  
+\begin{comment}
+Most of these
 sources can, in principle, be detected simply by re-running a test
@@
 more principled version of the unfortunate tradition of testing
-concurrent code by adding ad-hoc {\tt sleep} statements).   Most
-nondeterministic behavior in tests is probably attributable to such
+concurrent code by adding ad-hoc {\tt sleep} statements).
+\end{comment}
+Most nondeterministic behavior in tests is probably attributable to simple
 effects that can vary with a repeated execution, in the same
@@
   %test.
-\item {\bf Concurrency:} Ae second common cause for
+\item {\bf Concurrency:} A second common cause for
   nondeterminism in tests is when an algorithm is implemented using
   concurrency, in the (incorrect) belief that the result of the
-  algorithm remains deterministic.  E.g., consider a multi-threaded
+  algorithm remains deterministic.  E.g., a multi-threaded
   implementation of a search algorithm, that returns the location of
-  an item in an unsorted list.  If the list may contain duplicates,
-  and a developer has not accounted for this possibility, the
-  algorithm may return different locations of an item
-  depending on the scheduling of threads.
+  an item in an unsorted list; in the presence of duplicate items, thread scheduling may change the index returned.
+  %If the list may contain duplicates,
+  %and a developer has not accounted for this possibility, the
+  %algorithm may return different locations of an item
+  %depending on the scheduling of threads.
   %Arguably, this can be
@@
 network/IO, etc.'' a description that is well accounted for by these two
-primary sources, and most of which can be detected, at least in
-principle, by multiple runs of a test within the same testing process.  %However, simply running a test twice in the same process (possibly with a delay between steps) does
+primary sources, and most of which can be detected easily
+%, at least in principle, 
+by multiple runs of a test.
+%However, simply running a test twice in the same process (possibly with a delay between steps) does
 %not suffice in all cases.
@@
 
-Some sources of nondeterminism unfortunately require executing a test in a new
-process environment, because the source is inherently tied to the
+Some sources of nondeterminism unfortunately require executing a test in a \emph{new
+process environment}, because the source is inherently tied to the
 process in which code runs.  In terms of our formalism, an
@@
 other times not fail at all.  Even if no inputs produce a crash,
-however, memory errors may produce variation in values and thus be detectable
-as nondeterminism.
+however, memory errors may produce variation in values and thus be detectable.
+%as nondeterminism.
 
@@
 version 3.3 introduced automatic random salting of hashes on a
-per-process basis (a new salt chosen for each Python process),
+per-process basis,
 in order to mitigate hash-based denial of service attacks
@@
 exact hash values, but in the order of iteration on
-dictionaries\footnote{In 3.6, the default dictionary
-  implementation became an ordered dictionary with consistent
-  iteration, though actual hashes remained nondeterministic.}.  While
-relatively few programs rely on exact hash values, many implicitly
+dictionaries; in Python 3.6, the default dictionary implementation became an ordered dictionary with consistent iteration, though actual hashes remained nondeterministic.
+While very few programs rely on exact hash values, many
 relied on them in that they only functioned correctly with a
 predictable order for dictionary iteration.  Testing
-Python code for nondeterminism based on this hash seed requires
+Python code for nondeterminism based on hash seed requires
 running in a new process:  Python does not allow changing the salt, since changing hash values on-the-fly would
-break all existing dictionaries and other structures relying on
-hashing.
+break all existing dictionaries.
 
@@
 
-One common cause of flaky tests is \emph{test order dependence} \cite{idflakies}.  This
-is less of a concern in our setting, since many problems due to test
-order are easily exposed in random testing, since the
-order-dependence-inducing operations will seldom happen in the
-same order in randomly generated tests.  For example, consider the
-case where a test usually run ``early'' initializes an uninitialized
-value in a system, and a later test will only fail if it somehow is
-run before the initializing test.  In random testing, the order of
-initialization/access is likely to be random, and some run of a random
-tester is likely to expose the problem as a (deterministic, not flaky) failure.
-However, for cases where
-there is an order dependence that is hard to translate into an
-actual bug, using a fresh
-process may expose such dependencies---the ``environment'' of the
-parent process will have the, e.g., flag set or value initialized, but
-the new process will not.
+%One common cause of flaky tests is \emph{test order dependence} \cite{idflakies}.  This
+%is less of a concern in our setting, since many problems due to test
+%order are easily exposed in random testing, since the
+%order-dependence-inducing operations will seldom happen in the
+%same order in randomly generated tests.
+%For example, consider the
+%case where a test usually run ``early'' initializes an uninitialized
+%value in a system, and a later test will only fail if it somehow is
+%run before the initializing test.  In random testing, the order of
+%initialization/access is likely to be random, and some run of a random
+%tester is likely to expose the problem as a (deterministic, not flaky) failure.
+%However, for cases where
+%there is an order dependence that is hard to translate into an
+%actual bug, using a fresh
+%process may expose such dependencies---the ``environment'' of the
+%parent process will have the, e.g., a flag set or value initialized, but
+%the new process will not.
 
@@
 Delta-debugging, as proposed by Zeller and Hildebrandt, relies on a
-modified binary search to produce this smaller test, though the
+modified binary search, but the
 following discussion relies only on the general structure of
-delta-debugging, and also applies to the greedy algorithms used to
-reduce failing tests in some work \cite{CReduce,onetest}.  At a high
-level delta-debugging and other test reduction/shrinking algorithms
-can be seen as following a general algorithmic structure.  Given a
+delta-debugging \cite{CReduce,onetest}, which is as follows.  Given a
 test case, $T_{\mathit{INIT}} : \mathit{test}$ and a predicate
 $\mathit{PRED}: \mathit{test} \rightarrow \mathit{bool}$, we
-\emph{reduce $T_{\mathit{INIT}}$ with respect to $\mathit{PRED}$} as follows:
+reduce $T_{\mathit{INIT}}$ with respect to $\mathit{PRED}$
+as follows:
 
@@
 
+\begin{comment}
 The behavior of a particular algorithm for reducing a test can usually
@@
 potentially very large number of tests.
+\end{comment}
 
@@
 calls), but very difficult in a test consisting of a hundred library
-function calls.  This is no different than the common use of
+function calls.  This is the typical use of
 delta-debugging.  However, in horizontal nondeterminism detection,
@@
 as reducing the length of a test while maintaining that an arbitrary
-predicate remains true of the shorter versions of the test.  The
-most common $\mathit{PRED}$ is still ``the test fails'' but other predicates,
-especially those related to code-coverage, can also be useful
-\cite{icst2014,stvrcausereduce,NonAdeq}.  Our contribution here is to consider
+predicate still holds.   E.g., predicates, related to code-coverage can be very useful
+\cite{icst2014,stvrcausereduce,NonAdeq}.  Our contribution is to consider
 what happens when $\mathit{PRED}$ is not deterministic.
@@
 
-In \emph{monotonic} probabilistic delta-debugging, removing a part of a test cannot
-increase the probability of the predicate holding.  This is fairly
+In \emph{monotonic} probabilistic delta-debugging, removing part of a test \emph{cannot
+increase the probability of the predicate holding}.  This is fairly
 common.  In these cases, when we use a reduction algorithm to reduce $t$
@@
 however, there is no such upper bound.  Removing a step in $t$ may
-increase the probability that $t$ behaves nondeterministically.
+\emph{increase} the probability that $t$ behaves nondeterministically.
+\begin{comment}
 The $\mathit{PRED}$ of interest for horizontal determinism is basically of
@@
 non-monotonic, depending on the causes of the nondeterminism detected.
+\end{comment}
+In probabilistic
+settings, we may want to exploit non-monotonicity to find a
+subset of a test where a predicate that is very far from holding (in terms of probability, $P$) on the original test now
+holds.  For instance, a test may be very infrequently nondeterministic, and we want to find a subset where there is much greater probability of variance.
 
+\begin{comment}
 The only absolutely necessary changes required to use standard
@@
 of nondeterminism).
+\end{comment}
 
@@
 \caption{Delta-debugging of the same simple test}
-\end{figure*}
+\end{figure}
 
 However, simply using delta-debugging off the shelf with a $\mathit{PRED}$
-such as $P(fail) > 0.3 \wedge P(fail) < 1.0$, to force a test to be flaky, and force it to fail
+such as $P(fail) > 0.3 \wedge P(fail) < 0.7$, to force a test to be highly flaky, and force it to fail
 sufficiently often to be used in debugging, will often produce
@@
 $P(fail) > 0.3 \wedge P(fail) < 1.0$ cannot be evaluated by
-determining the true probability of failure, we have no more direct access to
+determining the true probability of failure; we have no more direct access to
 this value than a medical researcher has access to the true effect
-size and direction of a proposed treatment; rather, the test must be
+size and direction of a proposed medical treatment; rather, the test must be
 run some concrete number of times, and the number of failures counted.
-However, even if the number of samples $N$ is large, there is some
+Even if the number of samples $N$ is large, there is some
 probability (based on the sample size) of a result that diverges
@@
 delta-debugging and other test reduction algorithms explore a search
-space that often contains hundreds or even thousands or millions of
+space that often contains thousands or even millions of
 candidate tests.  The predicate is evaluated on each of these, and so
 even with large $N$, it is extremely likely that \emph{some} evaluation will
-produce a very poor estimate of $P(fail)$.  If such an evaluation
-causes the predicate to appear to hold for a test, delta-debugging is
-``stuck'' with the error, because (to our knowledge) no test reduction
-algorithms allow backtracking (in normal, non-probabilistic settings,
-backtracking would not usually be useful).  After such a ``mistake,'' finding further
+produce a very poor estimate of $P(fail)$.  At this point, delta-debugging is
+``stuck'' with the error, because test reduction
+algorithms usually do not allow backtracking.  After such a ``mistake,'' finding further
 reductions will become harder, but may still be possible due to the
@@
 delta debugging as a case where $\mathit{PRED}$ evaluates to true, but
-the actual probabilities required by $\mathit{PRED}$ do not hold, and
-a false negative to be the reverse, where ``bad luck'' makes it appear
-that the test does not satisfy $\mathit{PRED}$, but it does in fact do
-so.  A false negative's consequences are limited, since there will
-likely be another similar variation for which  $\mathit{PRED}$ will
-evaluated to true, but a false positive changes $T_{\mathit{CURR}}$.
+the actual probability constraints of $\mathit{PRED}$ are not satsified.  False negatives
+also exist, but are usually harmless, since it often just forces a different, equally
+desirable reduction path.
 
@@
 treatment is effective, with the statistical requirement that $p <
-0.05$ (that is, roughly, by some statistical test, the probability
-that the result is due to chance rather than reflecting reality, is
+0.05$ (that is, by some statistical test, the probability
+that the result is due to chance rather than reflecting reality is
 less than 5\%).  Even if no researchers are dishonest, 5 of every 100 papers published
@@
 
+\begin{comment}
 It is the combination of
@@
 problematic operations.
+\end{comment}
 
@@
 retraction'' based on near-replications.  However, this requires
-modifying the delta-debugging implementations, which is difficult and
+modifying delta-debugging implementations, which is difficult and
 sometimes not really feasible (e.g., few users of afl-fuzz \cite{aflfuzz} will wish
 to change its C code for test reduction).  Ideally, the solution
-should be implementable simply by modifying the predicate that is
-evaluated.
+should be implementable simply by modifying $\mathit{PRED}$.
 
-A costly but effective solution is to make $N$ large in comparison to
+A costly but somewhat effective solution is to make $N$ large in comparison to
 the number of expected predicate evaluations performed during
@@
 positive rate of about 3.7\%, yet requires almost 60\% fewer
-executions of the test.  The probability calculations for such comparisons
-  are relatively simple, but in real testing are
+executions of the test.  The probability calculations for such comparisons are
+relatively simple, but in real testing are
 usually not very useful, since the true
@@
 dynamically during the
-  delta-debugging process, making exact calculation unwieldy, even if
-  the changing probabilities were known; exact results would be based on fictional
-  probabilities, so gathering experimental data during a trial
+delta-debugging process; gathering experimental data during a trial
 delta-debugging run and tuning $N$ and $M$ to yield desired
-  results is  more effective.
-
-Tuning $M$, any degree of confidence can be achieved, with
+results is  more effective.  In fact, by tuning $M$, any degree of confidence can be achieved, with
 the basic tradeoff being between finding a test with the desired
@@
 nondeterministically with (independent) probabilities of 0.01, 0.05,
-and 0.10, respectively, storing the value of the operation one of five array locations.
+and 0.10, respectively, storing the value of the operation in one of five array locations.
 There is also an operation that clears out an array location, making
@@
 operations, the delta-debugging algorithm also reduces the probability
-of nondeterminism to only about 20\%.  If we use a predicate that
+of nondeterminism to about 20\%.  If we use a predicate that
 ``forces'' the test to behave nondeterministically at least half the
@@
 much longer than in Figure \ref{fig:p2}, but the probability of
-nondeterminism is above our target value, close to 60\% (and, as a bonus, fewer
+nondeterminism is above our target value, close to 60\% (and, as a bonus, many fewer
 test executions are required on average for each check of the predicate).  Note that
 this example is purely monotonic; the pattern of probability changes
-in non-monotonic settings can be even more difficult to predict and
+in non-monotonic settings can be more difficult to predict and
 control, but the principle of forcing a probability by sampling and
@@
 Finally, we note that this problem is not limited to debugging
-nondeterminism/flakiness itself.  In some settings, we have observed
+nondeterminism or known flakiness.  We have often observed
 delta-debugging taking a test that had such a high probability of
-failure we were unaware it was ``flaky'' (only hundreds of repeated executions showed there was
-indeed a small chance of the test passing) and transforming it into a
+failure that we were unaware it was ``flaky'' (only hundreds of repeated executions showed there was
+indeed a small chance the could pass) and transforming it into a
 test that only failed once in every 10 to 20 executions.
-Delta-debugging, without taking potential nondeterminism into account,
-can act as a flakiness magnifier, with unfortunate consequences for
-debugging.  Multiple evaluations, with replications, are our proposed solution.
+\emph{Delta-debugging can easily act as a flakiness multiplier.} Multiple evaluations using replications are our proposed solution.

### 2020-01-16 ISSTA/implementation.tex
small shrink

@@
 execution in a new process and allowing
-control over timing changes in re-execution, by specifying a delay to
-introduce between actions.  Changing the timing of
-steps is very effective when exact timing is somehow involved in
-results, and often effective in exposing concurrency-related
+control over timing changes.  Changing the timing of
+steps is also effective in exposing concurrency-related
 nondeterminism, since the delay will frequently change the
@@
 First-class vertical nondeterminism checks are currently
-limited to failures:  {\tt --checkFailureDeterminism} causes the TSTL
-compiler to emit a harness where every action that causes an \emph{expected}
-exception to be raised is repeated to ensure the same exception is
-still raised.  In terms of the formalism, if execution of action
-$\alpha_i$ raises an exception, then there are two possibilities:
+limited to failures.  
+%{\tt --checkFailureDeterminism} causes the TSTL
+%compiler to emit a harness where every action that causes an \emph{expected}
+%exception to be raised is repeated to ensure the same exception is
+still raised.
+%In terms of the formalism, if execution of action
+$\alpha_i$ raises an exception that does not, itself, indicate a test failure, then 
+    $F_i$ is true, and TSTL repeats $\alpha_i$ in order to ensure that
+    $F_{i+1}$ is also true.
+    %\end{enumerate}
 
-\begin{enumerate}
-\item If the TSTL harness does not explicitly allow the raised
-  exception, then the test fails.  There is no check for vertical
-  nondeterminism in this case, since the behavior of the system is
-  incorrect whether or not the behavior is idempotent.
-  \item If the harness does allow exception $e$ (e.g., as with {\tt
-      KeyError} and {\tt ValueError} in Figure \ref{fig:simple}), then
-    $F_i$ is true, and we repeat $\alpha_i$ in order to ensure that
-    $F_{i+1}$ is also true.
-    \end{enumerate}
-
-When a failure is indicated by some means other than raising an
-=exception, TSTL users must express this by explicitly writing
-a wrapper that raises an expected exception on failure.
+%When a failure is indicated by some means other than raising an
+%exception, TSTL users must express this by explicitly writing
+%a wrapper that raises an expected exception on failure.
 

### 2020-01-16 ISSTA/conclusions.tex
modest shrink

@@
 \cite{miccoflaky,listfieldtestanalysis}.
-
 This paper proposes a formulation of types of nondeterminism, and a
@@
 TSTL automated test generation system for Python, and demonstrated the
-simplicity and basic utility of the approach on three real-world examples.
+simplicity and utility of the approach on real-world examples.
 
@@
 possible: if a value (e.g., a timestamp) is always different in every
-run, it is likely of no interest to developers.  This would make
+run, it is likely opaque.  This would make
 testing libraries mixing deterministic behavior and nondeterministic
-behavior, e.g. cryptographic libraries offering high-quality random
-number generators, easier for users, who would only be shown
-``surprising'' nondeterminism.  Similarly, in order to extend the
-utility of vertical nondeterminism detection, we would like to
-automatically identify usually-idempotent operations, and check them
-for deviation from the expected behavior.  We would also like to lower the cost
-of checking process nondeterminism, perhaps by using methods taken in
-AFL \cite{aflfuzz} to avoid high startup costs for test executions.
-More generally, we are interested in using test decomposition
+behavior, e.g. cryptographic librarie, much easier. Similarly, in order to extend the
+utility of vertical nondeterminism, we would like to
+automatically identify usually-idempotent operations.
+%We would also like to lower the cost
+%of checking process nondeterminism, perhaps by using methods taken in
+%AFL \cite{aflfuzz} to avoid high startup costs for test executions.
+Finally, we are interested in using test decomposition
 \cite{Composition} to more easily isolate, understand, and avoid
-nondeterminism in tests, and to mitigate the problem of flaky tests.
+nondeterminism in tests, and to mitigate flaky tests.
 Reliable detection of nondeterminism is a first step towards
-evaluating such pro-active flaky test mitigations.
+evaluating such fine-grained proactive flaky test mitigations.
 

### 2020-01-16 ISSTA/relatedwork.tex
minor shrink

--- 2020-01-07 ISSTA/relatedwork.tex
@@
 Their work focused on systemic factors, such as execution platform and persistent configuration and database files, in contrast to our focus on identifying nondeterminism at the library level.
-However, what they refer to as ``test harness factors'' includes delays, which can be of importance to nondeterminism at the library level when asynchronous behavior is involved.
-
+However, what their ``test harness factors'' include delays, which can be of importance to nondeterminism at the library level.
 Shi et.
-al \cite{DetermImp} examined what might be seen as a related, though in a sense the opposite, problem:  they detect code that assumes a deterministic implementation of a non-deterministic specification.
-E.g., they detect instances when  iteration through items of a set is explicitly not guaranteed to provide any particular order of items, but code depends on the order produced by a given implementation.
-This procedure, applied to Python code in a pre-3.3 environment, would have flagged many instances of the nondeterminism that arose on the introduction of salted hashing.
-Determinism is also sometimes used as a property in domain-specific testing efforts, e.g.
-in testing shader compilers \cite{shader}.
+al \cite{DetermImp} examined what might be seen as a related, though in a sense, opposite, problem:  they detect code that assumes a deterministic implementation of a non-deterministic specification.
+E.g., they detect instances when  iteration through items of a set is explicitly not guaranteed to provide any particular order of items, but code depends on the order.
+Determinism is also sometimes used as a property in domain-specific testing, e.g.
+for shader compilers \cite{shader}.
 
@@
 Bell et al.
-proposed DeFlaker \cite{bell2018d}, which makes flaky tests much easier to detect by relying on the observation that if a test fails, and does not cover any changed code then (1) presumably it was a passing test in the past, or the tests would not be ``green'' before the latest code change so (2) the test is likely flaky.
-Comparing our approach with that of DeFlaker is difficult; DeFlaker applies only in the context of code \emph{changes} and is essentially a heuristic that identifies certain test failures as due to nondeterminism.
-Our approach is primarily focused on automatically providing a more extensive \emph{specification} in automated test generation, where a test ``failing'' is itself a sign of nondeterministic behavior, that could give rise to flaky behavior.
+proposed DeFlaker \cite{bell2018d}, which makes flaky tests much easier to detect by relying on the observation that if a test fails, and does not cover any changed code then (1) presumably it was a passing test in the past, or the tests would not be ``green'' before the change so (2) the test is likely flaky.
+Comparing our approach with DeFlaker is difficult; DeFlaker applies only in the context of code \emph{changes} and is essentially a heuristic that identifies failures as due to nondeterminism.
+Our approach is primarily focused on automatically providing a more extensive \emph{specification} in automated test generation, where a test ``failing'' is itself a sign of nondeterministic behavior.
+iDFlakies \cite{idflakies} is a framework and dataset for flaky tests, but again focuses on whole-tests, and detecting flakiness by actually observing it.
+Their work shows the order-dependence accounts for a little more than 50\% of flaky test causes, and they (like us) classify the rest as due to concurrency, timeouts,
+network/IO, etc.---the kinds of cause that our approach focuses on identifying.
+The present paper, rather than focusing on flaky tests as such, investigates methods for handling nondeterminism in property-based test generation \cite{ClaessenH00,Papadakis:2011:PIT:2034654.2034663}.
+From our point of view, flaky behavior is simply a special case of nondeterminism, where the nondeterminism is sufficient to cause the test to have different pass/fail results.
+Our approach is, in a sense (especially the delta-debugging modifications) in line with Harman and O'Hearn's proposal to simply accept that ``All Tests Are Flaky'' \cite{StartupstoScaleups}, and work with probabilistically failing tests.
+To our
+knowledge, this is the first approach to the problem that
+allows detection of potential sources of flakiness, based on real divergence in
+behavior, without having to observe a test actually fail.  We additionally introduce the first variation of
+delta-debugging that properly handles probabilistic reduction criteria.
 
-iDFlakies \cite{idflakies} is a framework and dataset for flaky tests, but again focuses on whole-tests, and detecting flakiness by actually observing it.
-Their work shows the order-dependence accounts for a little more than 50\% of flaky test causes, and they (like us) classify the rest as due to ``concurrency, timeouts,
-network/IO, etc.' --- the kinds of cause that our approach focuses on identifying.
-
-The present paper, rather than focusing on flaky tests as such, therefore, investigates methods for handling nondeterminism in property-based test generation \cite{ClaessenH00,Papadakis:2011:PIT:2034654.2034663}.
-Again, from our point of view, flaky behavior is simply a special case of nondeterminism, where the nondeterminism is sufficient to cause the test to have different pass/fail results at times.
-Our approach is, in a sense (especially the delta-debugging modifications) in line with Harman and O'Hearn's proposal to simply accept that ``All Tests Are Flaky'' \cite{StartupstoScaleups}, and work with probabilistically failing tests.
-
-Other efforts \cite{ParallelDeterministic} have aimed to avoid nondeterminism in parallel implementations, by design, indicating the importance of avoiding nondeterministic behavior, even at the expense of adopting novel programming models, where the goal is not simply (as in, say, Rust) to avoid classic concurrency errors, but to enforce genuinely deterministic behavior.
+Other efforts \cite{ParallelDeterministic} have aimed to avoid nondeterminism in parallel implementations, by design, indicating the importance of avoiding nondeterministic behavior, even at the expense of adopting novel programming models, where the goal is not simply (as in, say, Rust) to avoid classic concurrency errors, but to enforce true determinism.
 
 Vertical/failure determinism is a less-studied concept, and to our knowledge the formulation here, as a kind of determinism in a different ``direction,'' has not previously appeared; the kinds of errors that are exposed, however, are not new.
-In particular, we believe that many faults discovered using failure determinism are related to the propagation of error conditions in code, which has been studied in, e.g., file systems, using static analysis \cite{FileProp}.
+In particular, we believe that many such faults are related to the propagation of error conditions, which has been studied using static analysis \cite{FileProp}.
 

### 2020-01-16 ISSTA/abstract.tex
big pass

@@
 A critically important, but surprisingly neglected, aspect of system reliability is system predictability.
-Many software systems are implemented using mechanisms (unsafe languages, concurrency, caching, stochastic algorithms, environmental dependencies) that can introduce behavioral nondeterminism.
-Users of software systems, especially other software systems using a library, often expect that systems will behave deterministically.
-Even when it does not impact correctness, nondeterminism must be understood and managed for effective automated testing.
-Nondeterministic behavior can result in flaky tests as well as causing problems for test reduction, differential testing, and automated regression test generation.
+Many software systems are implemented using mechanisms (unsafe languages, concurrency, caching, stochastic algorithms, environmental dependencies) that can introduce unexpected and unwanted behavioral nondeterminism.
+Such nondeterministic behavior can result in software bugs and flaky tests as well as causing problems for test reduction, differential testing, and automated regression test generation.
 We show that lightweight techniques, requiring little effort on the part of developers, can extend an existing testing system to allow detection and debugging of nondeterminism.

### 2020-01-16 ISSTA/concepts.tex
big pass

@@
 \ref{fig:types}:  horizontal determinism and vertical determinism.  In
-horizontal determinism, which is what we usually think of when we
-think about deterministic behavior, a software system reliably
+horizontal determinism a software system reliably
 produces the same behavior given the same steps of a test: the
 behavior of multiple executions that are ``the same'' in terms of
-inputs/actions can be aligned and checked for some type of equality.  In
+inputs/actions can be aligned and checked for equality.  In
 vertical determinism, rather than behavior across executions, we are
 interested in behavior within an execution, where repeating the same
-step of a test twice should result in the same behavior.  Obviously,
-vertical determinism is both rarer and more specialized than
-horizontal determinism.  However, in those instances, vertical
-determinism is arguably more important than horizontal determinism, in
-that violations usually indicate a potentially serious fault.
+step of a test twice should result in the same behavior.  Vertical determinism is both rarer and more specialized than
+horizontal determinism.  However, vertical
+determinism violations usually indicate a potentially serious fault.
 
@@
 it behaves equivalently, at some level of abstraction, to another
-implementation of the same functionality.  Differential testing is
-extremely powerful, in that any divergence of behavior, if divergence
-is correctly defined, indicates a functional correctness fault in (at
+implementation.  Differential testing is
+extremely powerful, in that any (properly defined) divergence of
+behavior indicates a functional correctness fault in (at
 least) one of the systems under test, and  is widely
-used for systems software components such as compilers
-\cite{Differential,csmith} and POSIX file systems \cite{CFV08,AMAI}, where multiple
-implementations are common.  The major limitation of differential
+used for systems software such as compilers
+\cite{Differential,csmith} and file systems \cite{CFV08,AMAI}.  The major limitation of differential
 testing is that multiple implementations of a system are
 almost as rare as good correctness specifications.
-
 For the special case of detecting nondeterminism, however, \emph{a
@@
 We can formally distinguish the types of nondeterminism by using a
-variant of a labeled transition system (LTS) as our model, where (1) $S$
+variant of a labeled transition system (LTS)l, where (1) $S$
 is a set of states; (2) $V$ is a set of \emph{observable} states, where
@@
 
-We assume that the underlying behavior of a system may be
-deterministic, or nondeterministic, in the strict sense, by allowing
+We assume that the \emph{underlying} behavior of a system may be
+deterministic, or nondeterministic  by allowing
 for a transition relation that \emph{may} be a function of
@@
 to be the selection of an input, $\alpha_0$, with the remainder of the
-actions being internal $\tau$ actions, in the usual fashion.
+actions being internal $\tau$ actions.
 
@@
 restriction that $i = n$.  A pair of traces may exhibit visible value
-nondeterminism but not final state nondeterminism, but any pair of
-traces exhibiting final state nondeterminism also demonstrate visible
-value nondeterminism.
+nondeterminism but not final state nondeterminism.
 
@@
 search to a finite number of repetitions; for many faults, two
-``copies'' will suffice, since an operation is either truly idempotent or
-not idempotent.
+``copies'' will suffice.
 %Only the existence of \emph{non-visible} state changes that
@@
 effects that can vary with a repeated execution, in the same
-(approximate) enviroment, of the same test.  The two broad classes of
+(approximate) enviroment, of the same test.  There are two broad classes of
 nondeterminism that can usually be detected by
 executing the same test twice in succession, without attempting
-to control any other factors, are:
+to control any other factors.  {\bf External Environment} nondeterminism arises
+  when a test's behavior depends on factors outside the program under
+  test, which are subject to uncontrolled variance.  Calling {\tt
+    random()} in Python is an obvious example. Similarly, calls to {\tt time}
+  functions  will usually return different values on different runs,
+  and even if only elapsed time is used, due to system load, I/O, and
+  other external factors.  {\bf Concurrency} is a second common cause for
+  nondeterminism, when an algorithm is implemented using
+  concurrency, in the (incorrect) belief that the result of the
+  algorithm remains deterministic.  E.g., a multi-threaded
+  implementation of a search algorithm, that returns the location of
+  an item in an unsorted list; in the presence of duplicate items, thread scheduling may change the index returned.
 
+\begin{comment}
 \begin{itemize}
@@
 network/IO, etc.'' a description that is well accounted for by these two
-primary sources, and most of which can be detected easily
+primary sources.
+%, and most of which can be detected easily
 %, at least in principle, 
-by multiple runs of a test.
+%by multiple runs of a test.
 %However, simply running a test twice in the same process (possibly with a delay between steps) does
@@
 
-Delta-debugging \cite{DD}\footnote{We use delta-debugging to stand for
-  all test reduction algorithms, even those \cite{CReduce,tstl} that do not use the exact
-  binary-search that distinguishes delta-debugging \emph{per se}; the
-  differences are immaterial for our purposes.}  is a widely used method for reducing the
+Delta-debugging \cite{DD}  is a widely used method for reducing the
 size of failing tests, making them easier to understand and debug.
-The core idea of delta-debugging is to take a test that satisfies some
-property (usually the property is ``it fails in a particular way'')
-and produce a smaller test that satisfies the same property.
-Delta-debugging, as proposed by Zeller and Hildebrandt, relies on a
-modified binary search, but the
-following discussion relies only on the general structure of
-delta-debugging \cite{CReduce,onetest}, which is as follows.  Given a
+The core idea of delta-debugging is to take a test with some
+property (usually ``it fails'')
+and produce a smaller test that has the same property.
+Delta-debugging as originally proposed uses a
+modified binary search, but we rely only on the general structure: \cite{CReduce,onetest}:  Given a
 test case, $T_{\mathit{INIT}} : \mathit{test}$ and a predicate
@@
 \item Let $T_{\mathit{NEXT}} =$ a variation of $T_{\mathit{CURR}}$
-  that has not yet been tried.  If there are no further variations of
-  $T_{\mathit{CURR}}$, stop and return $T_{\mathit{CURR}}$ as the
-  reduced test.
+  that has not yet been tried.  If none exist, stop and return $T_{\mathit{CURR}}$.
 \item Add $T_{\mathit{NEXT}}$ to the set of variations that have been
@@
 
-Delta-debugging in the context of detecting nondeterminism has two
+Delta-debugging in the context of nondeterminism has two
 purposes.  One is simply the usual goal of reducing the size of a
 test.  Identifying the cause of nondeterminism may be very easy in a
-test consisting of ten library function calls (it is one of these ten
-calls), but very difficult in a test consisting of a hundred library
-function calls.  This is the typical use of
-delta-debugging.  However, in horizontal nondeterminism detection,
+test consisting of ten library function calls but very difficult with
+more than a hundred calls.  In horizontal nondeterminism detection, however,
 delta-debugging also tends to \emph{change the probability of
   nondeterministic behavior}; this can be both harmful and
-beneficial.  The behavior is part of a more general (and, to our
-knowledge, not previously investigated) issue:  using delta-debugging
-to minimize a test with respect to a predicate $\mathit{PRED}$ that only holds with a
-certain probability (the predicate itself is nondeterministic).  Note
+beneficial.  This behavior is part of a more general (and, to our
+knowledge, not previously investigated) issue: reducing a test with
+respect to an arbitrary predicate $\mathit{PRED}$ (including failure,
+but also code coverage, etc. \cite{icst2014,stvrcausereduce,NonAdeq}) that \emph{only holds with a
+certain probability} (the predicate itself is nondeterministic).
+\begin{comment}Note
 that while most descriptions of delta-debugging discuss reducing a
@@
 predicate still holds.   E.g., predicates, related to code-coverage can be very useful
-\cite{icst2014,stvrcausereduce,NonAdeq}.  Our contribution is to consider
-what happens when $\mathit{PRED}$ is not deterministic.
-``The test behaves flakily'' is a simple, important, example of such a $\mathit{PRED}$.
+\cite{icst2014,stvrcausereduce,NonAdeq}.
+\end{comment}
+%We want to consider
+%what happens when $\mathit{PRED}$ is not deterministic.
+``The test behaves flakily'' is an obvious relevant example of such a $\mathit{PRED}$.
 
-In \emph{monotonic} probabilistic delta-debugging, removing part of a test \emph{cannot
-increase the probability of the predicate holding}.  This is fairly
-common.  In these cases, when we use a reduction algorithm to reduce $t$
-to $r$, $P(\mathit{PRED}(r)) \leq P(\mathit{PRED}(t))$.  In \emph{non-monotonic} cases,
+For \emph{monotonic} $\mathit{PRED}$, removing part of a test \emph{cannot
+increase the probability of the predicate holding}: when we reduce $t$
+to $r$, $P(\mathit{PRED}(r)) \leq P(\mathit{PRED}(t))$.  This is fairly
+common.  In \emph{non-monotonic} cases,
 however, there is no such upper bound.  Removing a step in $t$ may
-\emph{increase} the probability that $t$ behaves nondeterministically.
+\emph{increase} the probability of $\mathit{PRED}$.
 \begin{comment}
@@
 In probabilistic
-settings, we may want to exploit non-monotonicity to find a
-subset of a test where a predicate that is very far from holding (in terms of probability, $P$) on the original test now
-holds.  For instance, a test may be very infrequently nondeterministic, and we want to find a subset where there is much greater probability of variance.
+settings, we can often exploit non-monotonicity.  For instance, a test may be
+``almost non-flaky'' and we want $P(\mathit{fail})$ to be closer to
+50\%, to aid debugging.
 
@@
 \includegraphics[width=\columnwidth]{lengthddmin}
-\caption{Standard delta-debugging}
+\caption{No modification}
 \label{fig:p1}
@@
 \includegraphics[width=\columnwidth]{lengthddminforcep}
-\caption{P=0.5, 100 samples}
+\caption{N=100}
 \label{fig:p2}
@@
 \includegraphics[width=\columnwidth]{lengthddminforceprep}
-\caption{P=0.5, 10 samples, 10 replications}
+\caption{M=10, N=10}
 \label{fig:p3}
 \end{subfigure}
-\caption{Delta-debugging of the same simple test}
+\caption{Reducing the same simple test; x-axis is time/changing
+  $T_\mathit{CURR}$; y-axis is true $P(\mathit{PRED})$.}
 \end{figure}
@@
 However, simply using delta-debugging off the shelf with a $\mathit{PRED}$
-such as $P(fail) > 0.3 \wedge P(fail) < 0.7$, to force a test to be highly flaky, and force it to fail
+such as $P(\mathit{fail}) > 0.3 \wedge P(\mathit{fail}) < 0.7$, to force a test to be highly flaky, and force it to fail
 sufficiently often to be used in debugging, will often produce
@@
 However, if you run the resulting test, it will, in many cases, have a
-$P(fail)$ that is much, much smaller than
+$P(\mathit{fail})$ that is much, much smaller than
 0.3, perhaps as low as 0.01.  Why?
@@
 $T_{\mathit{NEW}}$ as a scientific experiment.  A predicate like
-$P(fail) > 0.3 \wedge P(fail) < 1.0$ cannot be evaluated by
+$P(\mathit{fail}) > 0.3 \wedge P(\mathit{fail}) < 1.0$ cannot be evaluated by
 determining the true probability of failure; we have no more direct access to
@@
 predicate were run only once, and the number of samples reasonably
-large, this would not matter.  However, by their nature,
-delta-debugging and other test reduction algorithms explore a search
-space that often contains thousands or even millions of
-candidate tests.  The predicate is evaluated on each of these, and so
-even with large $N$, it is extremely likely that \emph{some} evaluation will
-produce a very poor estimate of $P(fail)$.  At this point, delta-debugging is
-``stuck'' with the error, because test reduction
+large, this would not matter.  However, test reduction algorithms explore a search
+space that often contains thousands or millions of
+tests.  The predicate is evaluated on each of these, and so
+even with large $N$, it is likely that \emph{some} evaluation will
+produce a poor estimate of $P(\mathit{fail})$.  At this point, reduction is
+``stuck'' with the error, because the
 algorithms usually do not allow backtracking.  After such a ``mistake,'' finding further
-reductions will become harder, but may still be possible due to the
-same source of errors:  \emph{the number of experiments run is far larger
-  than the probability of an incorrect result for a single
-  experiment}.  We define a \emph{false positive} in probabilistic
+reductions will become harder, but may still be possible due to
+another unlucky evaluation.  We define a \emph{false positive} in probabilistic
 delta debugging as a case where $\mathit{PRED}$ evaluates to true, but
-the actual probability constraints of $\mathit{PRED}$ are not satsified.  False negatives
-also exist, but are usually harmless, since it often just forces a different, equally
-desirable reduction path.
+the actual probability constraints of $\mathit{PRED}$ are not
+satsified (false negatives
+also exist, but are less of a problem).
 
@@
 treatment is effective, with the statistical requirement that $p <
-0.05$ (that is, by some statistical test, the probability
-that the result is due to chance rather than reflecting reality is
-less than 5\%).  Even if no researchers are dishonest, 5 of every 100 papers published
+0.05$.  Even if no researchers are dishonest, 5 of every 100 papers published
 in the journal will be report an effect with the wrong effect size or
-even effect direction!  Now consider a ``hot'' topic, a kind of
+even direction!  Now consider a ``hot'' topic, a kind of
 treatment many research teams may investigate.  Major journals tend
 not to publish papers that say ``this new treatment, which is not
-established, may not work,'' (especially if the result is not ``the
-treatment certainly does not work'' but only ``the experiment did not
-establish $p < 0.05$''), but they  are very likely to publish a paper
-that says ``this exciting new treatment works!''  In addition to
-sometimes producing excitement about treatments that do not work, this
-bias tends to magnify the expected effect size of treatments, since
-the chance of getting a ``passing'' $p$-value is higher when an
-experiment finds a larger effect size, due to how most statistical
-tests work.
+established, may not work,'' (especially if the result is not ``we
+accept the null hypothesis'' but only ``the experiment did not
+show $p < 0.05$''), but they  are very likely to publish a paper
+that says ``this exciting new treatment works!'' 
 The \emph{bias} in
-favor of positive results is shared by delta-debugging;
-delta-debugging's behavior is much more influenced by cases where
-$\mathit{PRED}$ holds than cases where it does not.  Looking at the
-literature, the unpublished papers are invisible, magnifying the
-effect on scientific opinion of the published experiments.
+favor of positive results is shared by test reduction, which;
+is much more influenced by cases where
+$\mathit{PRED}$ holds than cases where it does not.  The ``unpublished papers'' are invisible, magnifying the
+effect of the ``published'' experiments.
 
@@
 use of replications:  repeated runs of ``successful'' experiments to
-minimize the probability that a direction or effect size is a fluke due to publication
-bias.  One way to produce this effect would be to allow
-delta-debugging to backtrack if the probabilities observed in predicate
-evaluations suddenly exhibit a strong discontinuity, a kind of ``paper
-retraction'' based on near-replications.  However, this requires
-modifying delta-debugging implementations, which is difficult and
-sometimes not really feasible (e.g., few users of afl-fuzz \cite{aflfuzz} will wish
-to change its C code for test reduction).  Ideally, the solution
-should be implementable simply by modifying $\mathit{PRED}$.
+minimize the probability that a direction or effect size is a fluke.  One way to produce this effect would be to allow
+reducers to backtrack if the probabilities observed in predicate
+evaluations suddenly exhibit a strong discontinuity.  However, this requires
+modifying the implementations, which is difficult and
+sometimes not really feasible.  Ideally, the solution
+should be implementable simply by modifying $\mathit{PRED}$ itself.
 
-A costly but somewhat effective solution is to make $N$ large in comparison to
+A costly but plausible solution is to make $N$ large in comparison to
 the number of expected predicate evaluations performed during
 delta-debugging.  However, given the large number of evaluations performed,
-this will tend to make delta-debugging extremely slow.  We propose
+this will tend to make reduction extremely slow, since $N$ must be
+very large indeed.  We propose
 using a dynamic sampling approach, where $N$ is small, but if the
 predicate evaluates to true, $M$ repeated true evaluations
-(``replications'', where the first true evaluation is counted as 1 replication) are
-required before the predicate returns ``true'' to the delta-debugging
-algorithm.  To evaluate the predicate $\mathit{PRED}$ given $N$,
-$M$, and desired probability threshold $p$, the algorithm is simple:
+(``replications'') are
+required before the predicate is counted as holding.  To evaluate the predicate $\mathit{PRED}$ given $N$,
+$M$, and desired probability bound $p$, the algorithm is:
 
-\begin{enumerate}
-  \item  $i = 0$
-  \item Evaluate $\mathit{PRED}$ $N$ times.  If the proportion of cases
-  where $\mathit{PRED}$ was true is less than $p$, return false.
-  \item  $i = i + 1$
-  \item If $i < M$ go to step 1.
-  \item  Return true.
-\end{enumerate}
+\begin{code}
+for $i = 0 \ldots M-1$
+   $T =$ \# times $\mathit{PRED}$ is TRUE over $N$ evaluations.
+   if $\frac{T}{N} < p$, return FALSE.
+return TRUE
+\end{code}
 
@@
 false negatives, we can therefore produce a reduced test with a
-desired $P(fail)$ much more cheaply:  the use of replications not only
+desired $P(\mathit{fail})$ much more cheaply:  the use of replications not only
 means we only pay the full sampling price on rare occasions, but a
@@
 \times N$ than a non-dynamically-sampled $N$.  For example, if $\mathit{PRED}$
-is $P(fail) \geq 0.5$, and the true probability for a candidate test
+is $P(\mathit{fail}) \geq 0.5$, and the true probability for a candidate test
 is 0.25, using $N=8$ will give a false positive rate of over 10\%.
@@
 the basic tradeoff being between finding a test with the desired
-probabilities, and the speed and effectiveness of delta-debugging.
+probabilities, and the speed and effectiveness of reduction.
 Because increasing $M$ makes false negatives more likely, larger $M$

### 2020-01-16 ISSTA/experiments.tex
big pass

@@
 delta-debugging in this setting was acceptable, in terms of both time
-and results (amount of reduction in test size).  We chose two of
+and results (amount of reduction).  We chose two of
 the subjects, {\tt redis-py} and {\tt pyfakefs}, based on the fact that they were previously-existing large
-examples of TSTL harnesses, and therefore serve to show how easily our
-approach can be integrated into an existing large-scale test
+TSTL harnesses, and serve to show how easily our
+approach can be integrated into an existing test
 generation effort.  The other two subjects were chosen, and new harnesses
@@
 addition to our core research questions ({\bf RQ1-RQ3}) shared by all subjects, we
-report on particular aspects of using the TSTL nondeterminism tools
+report on aspects of using the TSTL nondeterminism tools
 specific to each subject.
@@
 tests, and the nondeterminism
-detected by TSTL can be demonstrated with standalone tests that do not depend on any TSTL code.
-The source code for the tools and all experimental subjects is
-available for inspection and replication in the TSTL repository.
+detected by TSTL can be demonstrated with standalone tests that do not
+depend on our code.
+The source code for our TSTL version and all experimental subjects will be
+made available for inspection and replication.

### 2020-01-16 ISSTA/implementation.tex
big pass

@@
 
-While the formal definitions of nondeterminism proposed in Section
-\ref{sec:formal} offer a framework for checking for nondeterministic
-behavior, they leave unspecified the key notions of \emph{state},
+While the formal definitions above offer a framework for checking nondeterminism, they leave unspecified the key notions of \emph{state},
 \emph{visible state}, and \emph{actions}.  For example, we could
 identify the state and visible state of a system as identical with the
-full set of memory locations accessed by that system, and actions with the
-individual processor instructions executed.  However, this
-definition is highly impractical for two reasons.  First, it is
+full set of memory locations accessed by that system, and actions with
+processor instructions.  However, this
+definition is highly impractical.  First, it is
 extremely inefficient to compare multiple executions for
@@
 even more importantly, as noted in the introduction, very few programs
-will behave deterministically when considered at this low level,
-unless the full state of the operating system, processor, hardware
-devices, etc. are all taken as controlled inputs.
-
-What is a practical level of granularity for (visible) states and actions that
+are deterministic in this strict sense.  What is a practical level of granularity for (visible) states and actions that
 can be efficiently checked, and matches (for the most part) developer
@@
 actual testing tools, including Randoop \cite{Pacheco}, showing that
-all Java unit tests (and, more generally, unit tests in similar
-languages) can be represented in a \emph{canonical} form, where
+unit tests can be represented in a \emph{canonical} form, where
 a test consists of three parts:
@@
       taken from a pool, e.g. {\tt intVP[2] = fooVP[2].bar(intVP[0],
-        intVP[1]);} (these are referred to as \emph{array-canonical} statements).
+        intVP[1]);} (referred to as \emph{array-canonical} statements).
     \end{enumerate}
@@
 
-We implemented our approach in the TSTL \cite{NFM15,ISSTA15,tstlsttt}
+We implemented our approach as a modification of the TSTL \cite{NFM15,ISSTA15,tstlsttt}
 system, an open-source
 language and tool for property-based testing
-\cite{Hypothesis,ClaessenH00} of Python code.   TSTL, including full
-support for our nondeterminism detection methods, is available at
-\url{https://github.com/agroce/tstl}, and the {\tt examples} directory
-contains the source code needed to duplicate all of our
-experiments\footnote{In some cases, you will also need to install an
-  appropriate version of the tested software, by doing, e.g., {\tt pip
-    install pyfakefs}.}.
+\cite{Hypothesis,ClaessenH00} of Python code.
+%The TSTL {\tt examples} directory
+%contains the SUTs needed to replicate our
+%experiments.
 \begin{comment}
@@
 
-\subsection{Horizontal Nondeterminism Detection}
-\label{sec:horizimp}
+\subsection{Nondeterminism Detection}
+%\label{sec:horizimp}
 
@@
 
-Note that TSTL provides explicit support for exploring the real-world sources of
-nondeterminism described in Section \ref{sec:sources}, by supporting
-execution in a new process and allowing
-control over timing changes.  Changing the timing of
+We provide support for the real-world sources of
+nondeterminism described in Section \ref{sec:sources}, via
+control over timing and/or execution in a new process.  Changing the timing of
 steps is also effective in exposing concurrency-related
-nondeterminism, since the delay will frequently change the
+nondeterminism, since it often changes the
 interleaving of threads.
 
-Because TSTL has a first-class interface to AFL \cite{aflfuzz}, we can even use
+Because TSTL has an interface to AFL \cite{aflfuzz}, we can use
 AFL's sophisticated heuristics to perform very thorough, week-long
-checks for nondeterminism, using strategies built to find subtle
-security vulnerabilities in C programs.  Equally importantly, since
+checks for nondeterminism.  Moreover, since
 there is substantial overhead in nondeterminism detection, AFL can be
@@
 \cite{aflfuzz}) is lower than 100\%, it may indicate horizontal
-nondeterminism is present.
+nondeterminism.
 Because of the idiosyncrasies of AFL instrumentation and process
@@
 
-\subsection{Vertical Nondeterminism Detection}
-\label{sec:vertimp}
+%\subsection{Vertical Nondeterminism Detection}
+%\label{sec:vertimp}
 
@@
 %exception to be raised is repeated to ensure the same exception is
-still raised.
-%In terms of the formalism, if execution of action
+%still raised.
+In terms of the formalism, if execution of action
 $\alpha_i$ raises an exception that does not, itself, indicate a test failure, then 

### 2020-01-16 ISSTA/introduction.tex
big pass

@@
 of up to 184 lines of code for the same test, and false positive rates
-as high as 96\%.  Nondeterminism is sometimes problematic for
-developers, who will usually want to  assume that library code they
-use behaves in a reliably predictable fashion; nondeterminism is
+as high as 96\%.  Nondeterminism is often problematic for
+developers, who want to  assume that library code behaves in a  predictable fashion; nondeterminism is
 frequently vexing in debugging efforts; most importantly, perhaps, nondeterminism is \emph{often disastrous} for large-scale highly
@@
 perfectly.  We say ``in principle'' because in the real world,
-prediction may be possible but hopelessly impractical.  We write
-complex software systems in many cases because we cannot predict their
-behavior (if we could perform the calculations in advance, ourselves,
-we would just do so).  As a consequence, in software, rather than
+prediction may be possible but  impractical.  We write
+complex software systems because we cannot predict their
+behavior (if we could perform the calculations in advance, 
+we would just do so).  As a consequence, rather than
 defining determinism in terms of prediction, we usually 
@@
 inputs, it always produces the same outputs.  
-
 Technically, many ``nondeterministic'' systems are not
-nondeterministic in a strong sense, at all.  Given the complete state
+nondeterministic in a strong sense.  Given the complete state
 of the system (which includes the entire state of the underlying
@@
 most software \emph{is} completely predictable.  What we actually
-mean is that, given a certain limited abstraction of state
-and inputs, observable behavior is repeatable.  %This abstraction,
+mean is that, \emph{given a certain limited abstraction of state
+and inputs, observable behavior is repeatable}.  %This abstraction,
 %for most software, is not expected to include many elements outside of
@@
 programmer's abstraction of the system ignores  such a large number of
-details, very few tests are in fact ``deterministic'' in the sense
-that they eliminate all changes in behavior between executions.  Running the same test
+details, very few tests are ``deterministic'' in the sense
+that they eliminate all changes between executions.  Running the same test
 twice almost always results in differences, given a low enough level of
@@
 cache contents, branch predictor history, etc. are almost never
-controlled for; however, this kind of
-nondeterminism is usually of no interest.  Rather, what matters is when
+controlled for; however, this is seldom of interest.  What matters is when
 some kind of nondeterminism unexpectedly impacts computed values at a
@@
 behavior of a thread scheduler will not matter, but a race condition
-in the code means that it does matter, or when the order of items in
-an iterator on a hash table is important, and the hash values used are
-randomly salted.
+in the code means that it does matter.
+
 \begin{comment} in general, if the operating system
@@
 This may be one reason that automated testing is seldom used to
-produce general regression tests.  While failure-inducing tests
-produced automatically are often added to regression suites, the full
-set of passing tests is, to our knowledge, only infrequently preserved
-as part of a typical regression suite, perhaps because such tests are
-likely to be \emph{flaky} without considerable human effort.
+produce general regression tests.  While bugs found by tools are added
+to regression suites, the full
+set of passing tests is seldom preserved, perhaps because such tests are
+likely to be \emph{flaky}.
+%without considerable human effort.
 
@@
 modern software systems usually include a large set of
-\emph{regression tests}.  A regression test suite is a (usually large)
-set of tests that can be run against a software system every time it
-is modified, to ensure that the modification has not broken the system
-in some way.  \emph{Flaky tests} \cite{miccoflaky} are regression
-tests that fail in an intermittent, unreliable fashion, and thus
-degrade the utility of regression testing.  The essence of a flaky
+\emph{regression tests}.  A regression test suite is a
+set of tests that can be run against softwarre every time it
+is modified, to ensure that the modification has not broken the system.  \emph{Flaky tests} \cite{miccoflaky} are regression
+tests that fail in an intermittent, unreliable fashion.  The essence of a flaky
 test is that, for the same snapshot of test code and code under test
@@
 test is not a deterministic property of the test code, code under
-test, and environment in which the test is run\footnote{Again, in 
-  some sense the result \emph{is} clearly determined by the environment in
-  which the test is run, but from the point of view of a developer or
-  test engineer, the differences in environment are invisible and unknown.}.  This produces three serious
+test, and testing environment.  This produces three serious
 problems: first, a flaky test often wastes developer time and delays
 software changes by forcing the investigation of code-under-test that
-is not actually incorrect.  Second, failures in flaky tests (for that
-reason) are often ignored, and therefore serious software faults may
+is not actually incorrect.  Second, failures in flaky tests are often ignored, and therefore serious software faults may
 be missed.  Finally, to mitigate these problems, flaky tests are often
-run multiple times, wasting valuable computing resources and also
-delaying acceptance of code changes.  An analogy may make the extent
-of the problem in large-scale test automation clear:  a canary in a coal mine is of
+run multiple times, wasting computing resources and
+delaying acceptance of code changes.  As an analogy, we note that  a canary in a coal mine is of
 little use if canaries frequently become ill for reasons unrelated to
@@
 more ``practical'' option is that miners may carry so many redundant
-canaries into the coal mine that canary-care becomes a serious burden
-on mining.
+canaries into the coal mine that canary-care itself becomes a serious burden.
 
@@
 special case of various kinds of \emph{horizontal} nondeterminism
-described below.  Our focus, however, unlike all other tools and
-approaches we are aware of, is on detecting the \emph{sources} of
-flaky-ness \emph{before} they propagate to cause actual test failure.
+described below.  Our focus, however, is on detecting sources of
+flaky-ness in libraries, before they propagate.
+%unlike all other tools and
+%approaches we are aware of, is on detecting the \emph{sources} of
+%flaky-ness \emph{before} they propagate to cause actual test failure.
 
@@
 This paper proposes (1) a number of formal definitions of types of
-nondeterminism (\emph{horizontal} and \emph{vertical}) (see Section \ref{sec:formal}) and (2) an implementation, based on these definitions, for detecting and debugging
+nondeterminism (\emph{horizontal} and \emph{vertical}) and (2) an implementation, based on these definitions, for detecting and debugging
 nondeterminism in property-based testing.  The implementation is based
@@
   determinism}, and (5)  in both cases the formalism is made practical
-by using the \emph{value pool} model of unit tests to define a useful
-granularity for nondeterminism detection (see Section \ref{sec:pools}).
-We further propose modifications to the widely used delta-debugging algorithm in order to better handle
+by using the \emph{value pool} model of unit tests.
+We also introduce necessary modifications to the widely used delta-debugging algorithm in order to better handle
 nondeterminism as a test property.

### 2020-01-16 ISSTA/relatedwork.tex
big pass

@@
 iDFlakies \cite{idflakies} is a framework and dataset for flaky tests, but again focuses on whole-tests, and detecting flakiness by actually observing it.
-Their work shows the order-dependence accounts for a little more than 50\% of flaky test causes, and they (like us) classify the rest as due to concurrency, timeouts,
-network/IO, etc.---the kinds of cause that our approach focuses on identifying.
+Their work shows the order-dependence accounts for a little more than 50\% of flaky test causes, and they (like us) classify the rest as due to ``concurrency, timeouts,
+network/IO, etc.''---the kinds of cause that our approach focuses on identifying.
 The present paper, rather than focusing on flaky tests as such, investigates methods for handling nondeterminism in property-based test generation \cite{ClaessenH00,Papadakis:2011:PIT:2034654.2034663}.

### 2020-01-16 ISSTA/experiments.tex
more

@@
 generated tests by 85\% or more, in less than 15 minutes, thus showing that reduction
-has a large payoff at an acceptable price.
+has a large payoff at an acceptable price.  Values here and discussed
+below are, unless otherwise noted, the mean of 30 runs; we omit variance due to little
+variance exhibited, but can provide full details in our replication package.
 
@@
 
-Our first subject is a simple implementation of a parallel merge
-sort\footnote{\url{https://gist.github.com/stephenmcd/39ded69946155930c347}}.  The
-merge sort and harness can be found in the TSTL GitHub repository
-\cite{tstl}, under {\tt examples/parallelsorts}.  When a sorting
+Our first subject is an implementation of a parallel merge
+sort\footnote{\url{https://gist.github.com/stephenmcd/39ded69946155930c347}}. When a sorting
 algorithm is implemented using concurrency, a key question to ask is
-whether the ordering of elements that are equal under the comparison
-operator is consistent across sorts of the same original sequence.
-This is related to the question of whether a sort is a stable sort
-(preserving the original relative order of equal-by-comparison elements),
-but imposes a less strict requirement on the sort; a concurrent sort
+whether the ordering of (equal under comparison) elements is consistent.
+This is related to the question of whether a sort is stable,
+but imposes a less strict requirement; a concurrent sort
 might not be stable, but might still produce the same (non-stable)
 ordering from the same provided sequence, every time.  This is a
-particular instance of a common expectation noted above:  while a parallel
+particular instance of a common expectation:  while a parallel
 implementation of an algorithm will likely have internal
-nondeterminism (due to scheduling of work), it is often expected to
+nondeterminism, it is often expected to
 behave deterministically, from the point of view of a caller.
@@
 out-of-order neighbors, until the sequence is sorted (a kind of
-parallel bubble-sort with a simpler termination criteria).
+parallel bubble-sort).
 
@@
 @import swapsort
-%\vspace{0.1in}
+
 pool: <val> 10
@@
 pool: <sorted> 10
-%\vspace{0.1in}
+
 <val> := <[-10..10]>
-%\vspace{0.1in}
+
 <data> := []
@@
 len(<data,2>) < len(<data,1>) -> <data>.extend(<data>)
-%\vspace{0.1in}
+
 <sorted> := mergesort.merge\_sort\_parallel(<data>)
@@
 <sorted>
-%\vspace{0.1in}
+
 property: swapsort.is\_sorted(<sorted>)
@@
 operator of the sorting algorithms, and checks that the results of
-{\tt sort\_parallel} calls are in fact sorted, using the {\tt
-  swapsort}'s function for checking if a sequence is sorted.
-
+{\tt sort\_parallel} calls are sorted..
 Running TSTL's random tester without checking for determinism, we can
-see that both sorting algorithms always produce sorted output.  If we run with {\tt --checkDeterminism --determinismDelay 0}, however, a
-test case showing the {\tt swapsort} behaving nondeterministically is
-almost instantly produced (it takes less than 3 seconds), thus
+see that both sorting algorithms always produce sorted output.  The
+horizontal determinism detector, however, always generates a 
+test case showing the {\tt swapsort} behaving nondeterministically in
+less than 3 seconds, thus
 answering yes to {\bf  RQ1}.  If we comment out the call to {\tt
@@
 with nondeterminism detection turned on results in a slowdown of approximately
-40\% ({\bf RQ2}.
+40\% ({\bf RQ2} with little variance.
 
-Reducing a lengthy test showing nondeterminism of the swapsort to a
-short test case using delta-debugging usually takes less than two
-minutes.  If the number of attempts to produce nondeterminism is
-increased in the reducer using the {\tt --determinismTries 10} option,
-the resulting test will be usually be very compact, in this case
+Reducing a lengthy test showing nondeterminism of the swapsort usually takes less than two
+minutes (out of 10 trials, only one took four minutes).  If the number of attempts to produce nondeterminism is
+increased in the reducer, the resulting test will be usually be very compact, e.g.
 reducing the original test from 56 to 8 steps.
@@
 
-For this simple example, more sophisticated delta-debugging strategies
-are not needed; we only want to see the code behaving
-nondeterministically, and do not care about preserving a higher probability of nondeterminism.
+%For this simple example, more sophisticated reduction strategies
+%are not needed; we only want to see the code behaving
+%nondeterministically, and do not care about preserving a higher probability of nondeterminism.
 
@@
 has proven difficult, as numerous Redis commands introduce nondeterministic
-behavior:  thus the resulting tests are very often \emph{flaky}.   Using {\tt tstl\_afl\_fuzz}, we can see that {\tt redis-py} has a 
+behavior:  thus the resulting tests are very often \emph{flaky}.
+Using AFL, we can see that {\tt redis-py} has a 
 stability of only 56.26\%, a clear indicator of significant nondeterminism. Some
@@
 exhibiting flakiness to a small set of tests, each short, showing a
-different cause of nondeterminism.  The set produced is not perfect,
-there is some duplication of causes, but it is sufficient for
+different cause of nondeterminism.  The set produced is not perfect:
+There is some duplication of causes, but it is sufficient for
 identifying causes of nondeterminism.
@@
 remove nondeterminism, which produced no flaky tests.  We determined that the mean time to find all
-the sources of nondeterminism and produce minimal tests 3.2
+the sources of nondeterminism and produce minimal tests is 3.2
 hours, which is not trivial, but also, for this purpose, not
@@
 
-We removed 11 calls from the {\tt redis} harness, and  modified 2
+We removed 11 calls from the {\tt redis-py} harness, and  modified 2
 calls, after
@@
 large number of samples, does not work (Figure \ref{fig:r3}) and
-requires over 2000 seconds to produce a test with an even worse
+requires over 2,000 seconds to produce a test with an even worse
 probability of nondeterminism.  Using 10 samples and 10 replications,
@@
 visible value and final state approaches find 40 nondeterministic
-tests, though the overlap of tests thus detected is not quite perfect
---- each method detects two tests the other does not.  The difference
+tests, though the overlap of tests thus detected is not quite perfect---each method detects two tests the other does not.  The difference
 in overhead, interestingly, is very similar:  0.48\%; however, under
@@
 @import datarray
-%\vspace{0.1in}
 <@
-def flatten\_and\_sort(v):
-    return (sorted(map(flatten\_and\_sort,v),key=repr) if type(v) in [list,tuple] else
-                (flatten\_and\_sort(list(v.items())) if type(v) == dict else v))
+def flat\_sort(v):
+    return (sorted(map(flat\_sort,v),key=repr) if type(v) in [list,tuple]
+                else (flat\_sort(list(v.items())) if type(v) == dict else v))
 
@@
 @>
-%\vspace{0.1in}
 pool: <P> 3
@@
 pool: <event> 3
-%\vspace{0.1in}
+
 <P> := 0.01 * <[0..100]>
 <ename> := "E" + str(<[1..5]>)
-\{Exception\} <event> := datarray.DataArray(psplit(<P>), axes = [<ename>])
-\{Exception\} <ename,1>!=<ename,2> -> <event> := [datarray.DataArray([[psplit(<P>)],
-   psplit(<P>)],[<ename>,<ename>])]
+\{Exception\} <event> := datarray.DataArray(psplit(<P>),axes=[<ename>])
+\{Exception\} <ename,1>!=<ename,2> -> <event> := [datarray.DataArray([[
+   psplit(<P>)], psplit(<P>)],[<ename>,<ename>])]
 <cpts> := []
@@
 ~<evidence>.update([(<ename>,0)])
-%\vspace{0.1in}
-\{Exception\} print(flatten\_and\_sort(inference\_algs.calc\_marginals\_simple(<cpts>,
+
+\{Exception\} print(flat\_sort(inference\_algs.calc\_marginals\_simple(<cpts>,
    <evidence>)))
-\{Exception\} print(flatten\_and\_sort(inference\_algs.calc\_marginals\_sumproduct(<cpts>,
+\{Exception\} print(flat\_sort(inference\_algs.calc\_marginals\_sumproduct(<cpts>,
    <evidence>)))
-\{Exception\} print(flatten\_and\_sort(inference\_algs.calc\_marginals\_jtree(<cpts>,
+\{Exception\} print(flat\_sort(inference\_algs.calc\_marginals\_jtree(<cpts>,
    <evidence>)))
@@
 Figure \ref{hashbug} shows TSTL code for generating inputs to the {\tt
-  datarray} algorithms (it can be found in the {\tt
-  examples/datarray\_inference} directory in the TSTL repository \cite{tstl}).
-The {\tt flatten\_and\_sort} function is needed because we care about actual differences in probability values, not simply the order of list, tuple, or dictionary items.
-Running this harness using TSTL's {\tt
-  --checkProcessDeterminism} flag required less than 10 seconds on
-average to produce a test exhibiting process-level nondeterminism in
+  datarray} algorithms.  The {\tt flat\_sort} function is
+needed because we care about actual differences in probability values,
+not simply the order of list, tuple, or dictionary items.  Running
+this harness consistently requires less than 10 seconds to produce a test exhibiting process-level nondeterminism in
 the {\tt calc\_marginals\_sumproduct} function (the only broken
@@
 showing an extremely simple input producing the issue, 
-required another 92 seconds ({\bf RQ3}).  Interestingly, the way that {\tt
+required another 92 seconds ({\bf RQ3}) on average.  Interestingly, the way that {\tt
   python-afl} is used in TSTL means that AFL stability is 100\% even
@@
 because the {\tt PYTHONHASHSEED} has already been chosen before the
-fork in AFL executions.
-
-Removing the nondeterministic call, we can see that the cost of
+fork in AFL.  Removing the nondeterministic call, we can see that the cost of
 checking for process nondeterminism, with no delay between operations,
-is high, a 93\% slowdown({\bf RQ2}.  This is due to the high cost of subprocess
+is high, a mean 93\% slowdown ({\bf RQ2}.  This is due to the high cost of subprocess
 creation and communication. However, even with this essentially worst
 case behavior (where test runtime is very low compared to process overheads),
-we sacrifice less than half of the test executions for a given time budget.
+we sacrifice less than half of the test executions.
 

### 2020-01-16 ISSTA/experiments.tex
ok

@@
 generated tests by 85\% or more, in less than 15 minutes, thus showing that reduction
-has a large payoff at an acceptable price.  Values here and discussed
-below are, unless otherwise noted, the mean of 30 runs; we omit variance due to little
-variance exhibited, but can provide full details in our replication package.
+has a large payoff at an acceptable price.  Values are, unless otherwise noted, the mean of 30 runs.
 
@@
 Our first subject is an implementation of a parallel merge
-sort\footnote{\url{https://gist.github.com/stephenmcd/39ded69946155930c347}}. When a sorting
+sort ({\url{https://gist.github.com/stephenmcd/39ded69946155930c347}). When a sorting
 algorithm is implemented using concurrency, a key question to ask is
@@
 might not be stable, but might still produce the same (non-stable)
-ordering from the same provided sequence, every time.  This is a
-particular instance of a common expectation:  while a parallel
+ordering from the same provided sequence, every time.  While a parallel
 implementation of an algorithm will likely have internal
@@
 
+\begin{comment}
 \begin{figure}
@@
 Figure \ref{fig:parallelcode} shows the complete TSTL harness for
-checking the parallel sorts.  This harness produces sequences of
+checking the parallel sorts.
+\end{comment}
+
+Our TSTL harness for parallel sorting produces sequences of
 integer pairs, where only the first integer is used in the comparison
 operator of the sorting algorithms, and checks that the results of
-{\tt sort\_parallel} calls are sorted..
+are sorted..
 Running TSTL's random tester without checking for determinism, we can
@@
 answering yes to {\bf  RQ1}.  If we comment out the call to {\tt
-  swap\_sort\_parallel} and only check the parallel merge sort, however, TSTL
-finds no nondeterministic behavior.  Running tests of the merge sort
-with nondeterminism detection turned on results in a slowdown of approximately
-40\% ({\bf RQ2} with little variance.
+  swap\_sort\_parallel} and only check the parallel merge sort, TSTL
+finds no nondeterminism.  This results in a slowdown of approximately
+40\% ({\bf RQ2}).
 
@@
 
+\begin{comment}
 {\scriptsize
@@
 }
-
+\end{comment}
 %For this simple example, more sophisticated reduction strategies
@@
 \includegraphics[width=\columnwidth]{redisddmin}
-\caption{Standard delta-debugging (38s)}
+\caption{No modification}
 \label{fig:r1}
@@
 \includegraphics[width=\columnwidth]{redisforcep}
-\caption{P=0.5, 100 samples (2131s)}
+\caption{N=100 (2131s)}
 \label{fig:r2}
@@
 \includegraphics[width=\columnwidth]{redisforceprep}
-\caption{P=0.5, 10 samples, 10 replications (617s)}
+\caption{M=10, N=10}
 \label{fig:r3}
 \end{subfigure}
-\caption{Delta-debugging of the same {\tt redis-py} nondeterministic test}
+\caption{Reducing a {\tt redis-py} nondeterministic test}
 \end{figure}
@@
 to the popular Redis in-memory database, cache, and
-message broker \cite{redis}.  The TSTL {\tt redis-py} harnesses can be found in
-the {\tt examples/redis} directory in the TSTL GitHub repo \cite{tstl}.
-
+message broker \cite{redis}.  
 Using TSTL's harness for {\tt redis-py}, both {\tt redis-py} and Redis
 itself can be tested; unfortunately, generating stand-alone
-high-coverage regression tests for {\tt redis-py} and Redis using TSTL
+high-coverage regression tests for them
 has proven difficult, as numerous Redis commands introduce nondeterministic
@@
 of the problematic commands are obvious simply by inspection (e.g.,
-{\tt randomkey, srandmember}).  And, given an understanding of Redis
-semantics, it is also clear that the various
+{\tt randomkey, srandmember}). It is also clear that 
 commands producing data with a limited lifetime (e.g., {\tt expire,
@@
 difficult to guess whether the {\tt pipe} mechanism, which allows a
-large sequence of commands to be queued up in a pipe and executed all
-at once, introduced the potential for nondeterminism (does it execute
-sequentially before other command are handled, or is it in parallel
-with further commands).  Deep experience with Redis would make these
+large sequence of commands to be queued up and executed
+at once, introduces the potential for nondeterminism. Deep experience with Redis would make these
 issues clear, but tests using a library are often written by those not
@@
 test engineer is often chosen for expertise in test generation tools,
-not in the domain of the software under test (SUT), and is seldom the
-original developer of the code.
+not in the domain of the SUT, and is seldom the
+original developer.
 
@@
 nondeterminism checker to reduce flaky tests to a minimal set allows
-us to simply set up an overnight run that will, in 12 hours, produce
+us to set up an overnight run that will, in 12 hours, produce
 a set of minimal nondeterministic tests exposing the root sources of
@@
 exhibiting flakiness to a small set of tests, each short, showing a
-different cause of nondeterminism.  The set produced is not perfect:
-There is some duplication of causes, but it is sufficient for
-identifying causes of nondeterminism.
+different cause of nondeterminism.  
+There is some duplication of causes, but the tests are sufficient for
+identifying all nondeterminism sources.
 
@@
 
+\begin{comment}
 \begin{figure}[t]
@@
 
+\begin{comment}
 Figure \ref{hashbug} shows TSTL code for generating inputs to the {\tt
@@
 needed because we care about actual differences in probability values,
-not simply the order of list, tuple, or dictionary items.  Running
-this harness consistently requires less than 10 seconds to produce a test exhibiting process-level nondeterminism in
+not simply the order of list, tuple, or dictionary items.
+\end{comment}
+Running our TSTL
+harness to test {\tt datarray} consistently requires less than 10 seconds to produce a test exhibiting process-level nondeterminism in
 the {\tt calc\_marginals\_sumproduct} function (the only broken

### 2020-01-16 ISSTA/implementation.tex
ok

@@
 control over timing and/or execution in a new process.  Changing the timing of
-steps is also effective in exposing concurrency-related
-nondeterminism, since it often changes the
-interleaving of threads.
+steps is also often effective in exposing concurrency-related
+nondeterminism.
 
@@
 checking for horizontal nondeterminism; if the
-AFL \emph{stability} statistic (see the AFL documentation for details
-\cite{aflfuzz}) is lower than 100\%, it may indicate horizontal
+AFL \emph{stability} statistic \cite{aflfuzz} is lower than 100\%, it may indicate horizontal
 nondeterminism.
@@
 behavior, this is not always a reliable guide, but it is a very low
-cost indicator produced as a by-product of fuzzing for other problems.
+cost indicator.
 

### 2020-01-16 ISSTA/concepts.tex
ok
almost there

@@
 
-\section{Sources of Nondeterminism}
+\subsection{Sources of Nondeterminism}
 \label{sec:sources}
@@
 
-\subsection{Process-Based Nondeterminism}
+\paragraph{Process-Based Nondeterminism}
 
@@
 
-\section{Nondeterminism and Delta-Debugging}
+\section{Probabilistic Test Reduction}
 
@@
 In probabilistic
-settings, we can often exploit non-monotonicity.  For instance, a test may be
+settings, we can often exploit non-monotonicity.  E.g., a test may be
 ``almost non-flaky'' and we want $P(\mathit{fail})$ to be closer to
@@
 
-\subsection{``Publication Bias'' in Delta-Debugging}
+\subsection{``Publication Bias'' in Reduction}
 \label{sec:pubbias}
@@
 is much more influenced by cases where
-$\mathit{PRED}$ holds than cases where it does not.  The ``unpublished papers'' are invisible, magnifying the
-effect of the ``published'' experiments.
+$\mathit{PRED}$ holds than cases where it does not.
+%The ``unpublished papers'' are invisible, magnifying the
+%effect of the ``published'' experiments.
 
@@
 \begin{code}
-array[2] = op01();
-array[3] = op05();
-clear(array[2]);
-array[2] = op10();
+array[2] = op01(); array[3] = op05(); clear(array[2]); array[2] = op10();
 \end{code}

### 2020-01-16 ISSTA/experiments.tex
almost there

@@
 written, specifically to demonstrate specific, important sources of
-nondeterminism in Python code: concurrency and hash salting.  In
+nondeterminism in Python code: concurrency and hash salting.
+\begin{comment}In
 addition to our core research questions ({\bf RQ1-RQ3}) shared by all subjects, we
@@
 specific to each subject.
-
+\end{comment}
 Table \ref{tab:overviewexp} shows a summary of our experimental
 results.  In all cases, our approach was able to detect interesting
-nondeterminism (in fact, all nondeterminism of which we are aware).  The overhead on testing imposed by nondeterminism
-checks varied considerably, but in the worst case did not double the
+nondeterminism (in fact, all nondeterminism of which we are aware).  The overhead imposed by nondeterminism
+checks varied,, but in the worst case did not double the
 average cost of each action (the expected overhead, since we
 essentially execute each test twice), and thus never halved the amount of
-testing performed.  Test reduction always reduced the size of
+testing.  Test reduction always reduced the size of
 generated tests by 85\% or more, in less than 15 minutes, thus showing that reduction
@@
 these parameters, visible value nondeterminism is actually cheaper on average
-than final state nondeterminism, due to early termination of the
-additional replays, when nondeterminism is detected (the difference is
-again significant by Wilcoxon test, with $p < 1.44 \times 10^{-10}$).
+than final state nondeterminism, due to early termination when nondeterminism is detected ($p < 1.44 \times 10^{-10}$).
 
@@
 inside and outside Google \cite{pyfakefs}.
+The TSTL harness for {\tt pyfakefs} has been used to detect (and
+correct) over 80 faults.
+%\url{https://github.com/jmcgeheeiv/pyfakefs/issues?q=label\%3ATSTL}.
 
-The TSTL harness for {\tt pyfakefs} has been used to detect (and
-correct) over 80 faults.  The harness can be found in the TSTL GitHub repository
-in the {\tt examples/pyfakefs} directory, and the bugs discovered can
-be viewed at \url{https://github.com/jmcgeheeiv/pyfakefs/issues?q=label%3ATSTL}.  However, the testing largely relies on
-the existence of a reference file system implementation.  One purpose
-of failure nondeterminism is to make it somewhat easier to perform effective
-property-based testing of
-complex APIs like this even without a complete reference implementation, or in cases
-where the implementations do not use the same error
-codes (as is common, e.g. in NASA flight software
-\cite{ICSEDiff,CFV08}).
-
-\subsubsection{Manually Inserted Fault}
+\paragraph{Manually Inserted Fault}
 
@@
 call checks that its target is not a directory, and returns the
-correct error, but still carries out the remove operation.  Using {\tt
-  os.remove} to delete directories does not break any file system
-invariants, but violates the Python {\tt os} specification (and,
-indirectly, the usual POSIX implementation behavior where {\tt unlink}
-does not work for directories).  Detecting this bug using the TSTL
+correct error, but still carries out the remove.  Using {\tt
+  os.remove} to delete directories violates the Python {\tt os}
+specification (and the POSIX standard).  Detecting this bug using the TSTL
 {\tt pyfakefs} harness is normally impossible without using another
 file system as a reference.  However, the fault was detected  essentially
-immediately, even without using a reference, when we compiled the harness
-with the {\tt --checkFailureDeterminism} flag (an affirmative answer
-to {\bf RQ1}).  Moreover, the overhead for the check for failure determinism in a
-version of the code without the {\tt remove} error was
+immediately with our failure determinism checks (an affirmative answer
+to {\bf RQ1}).  Moreover, the overhead for the check in a
+version of the code without the error was
 less than 8\%  ({\bf RQ2}).  Detecting the fault using a reference file system 
 required 17\% more testing time before detection, and took over twice as
-long to reduce the failure to a
+long to reduce the failure, to a
 slightly longer failing test, which did not have {\tt remove} as its
 final operation (since further operations are required to 
-expose the bad file system state the operation introduces).
+expose the bad state).
 Because vertical
@@
 the failing test to 3 steps required less than a second ({\bf RQ3}).
-For this hypothetical subtle bug, failure determinism checks not only make it
-possible to detect the fault without a reference implementation, they
-improves on detection speed and ease of debugging even compared to a
-full-blown reference implementation (the MacOS file system, operating
-on a RAM disk) and strong, hand-tuned, differential testing.
+For this hypothetical subtle bug, failure determinism even improves on strong, hand-tuned, differential testing.
 
-\subsubsection{Mutation Analysis}
+\paragraph{Mutation Analysis}
 
@@
 {\tt universalmutator} \cite{RegExpMut} to produce 2,350 mutants of
-the {\tt pyfakefs} core file system code.   We restricted the generation
-to only mutate code covered during a 60 second run of the test
-harness, with differential testing turned off.  In this
-\emph{non-differential} mode, the harness will only detect a fault when it
-causes an unexpected exception to be raised, or results in a timeout
-due to, e.g., an infinite loop.  We could have applied mutation
+the {\tt pyfakefs} core file system code.   %We restricted the generation
+%to only mutate code covered during a 60 second run of the test
+%harness.
+\begin{comment}
+  We could have applied mutation
 testing to {\tt redis-py}, but we are dubious of the value of such an
@@
 
-We first analyzed the mutants using 60 seconds of non-differential
+In each run we first analyzed the mutants using 60 seconds of non-differential
 testing \emph{without failure nondeterminism checks}, then with the
@@
 maximize the chance to detect a bug using the strong version of the harness.   For both comparisons, we then tested
-all surviving mutants using the non-differential harness for 60
-seconds, this time with failure nondeterminism checking turned on.
-We used the same random seed for
-all runs, so that the only differences would be specification-based.
+all surviving mutants using a non-differential harness for 60
+seconds, this time with failure nondeterminism checking.
+%We used the same random seed for
+%all runs, so that the only differences would be specification-based.
+\end{comment}
 
-Non-differential testing, without failure determinism checks, killed
-872 mutants, a 37.1\% kill rate.  Adding a failure determinism check
-allowed the testing to kill an additional 98 mutants, an 11\% improvement.
+Non-differential testing, without failure determinism checks,
+consistently (in all runs) killed
+872 mutants, a 37.1\% kill rate.
+%Adding a failure determinism check
+%allowed the testing to (again, consistently) kill an additional 98 mutants, an 11\% improvement.
 The differential harness, with an additional 60 seconds of testing
-time, killed 1148 mutants, improving the kill rate to 48.8\%.  Failure
-nondeterminism checking added an
-additional 71 mutants to the total killed, a 6\% improvement even for
-this strong differential harness with a larger test budget.  Using
-failure nondeterminism was the only way to push the mutation score
+time (to make sure we accounted for the observed 17\%
+extra time to detect in the manually constructed example: we want to
+maximize the chance to detect a bug using the strong version of the harness), consistently killed 1,148 mutants, improving the kill rate to 48.8\%.  Failure
+nondeterminism checking consistently added an
+additional 71 mutants to that total, \emph{a 6\% improvement even for
+this strong differential harness with a larger test budget}, nearly as
+large an improvement as the gain from using a full reference
+implementation for comparison.  Using
+failure nondeterminism was the \emph{only} way to push the mutation score
 above 50\%.  The kill rates are generally low
-because {\tt universalmutator}, by design, includes many operations
-that can produce equivalent mutants, but can also produce hard-to-kill
-non-equivalent mutants not produced by other mutation tools, e.g. code
-that throws away exceptions raised by a function call (see below), or code
-reversing a list.
+because {\tt universalmutator} includes many operations
+thatproduce hard-to-kill
+non-equivalent mutants not produced by other mutation tools.  This is a strong affirmative answer to {\bf
+  RQ1} at least for a large set of hypothetical bugs.
 
+\begin{comment}
 In order to further investigate the additional specification power
@@
 statement, but wrapped in a {\tt try} block with a catch that ignores
-any raised exceptions, e.g., {\tt foo()} becomes:
-
-\begin{code}
-try: foo()
-except: pass
-\end{code}
-
-\noindent It is easy to
+any raised exceptions.
+It is easy to
 see that such mutants may introduce faults in the handling of
-errors, and thus would tend to cause failure nondeterminism.  However,
-these are a minority of the mutations introducing hard-to-detect but
-non-equivalent failure
-nondeterminism.  Other mutation operators resulting in subtle flaws
+errors, and thus would tend to cause failure nondeterminism.  Other mutation operators resulting in subtle flaws
 not (at least easily) detectable by differential testing compared to a
-correct file system include:  arithmetic operation changes, statement
+correct file system included:  arithmetic operation changes, statement
 deletions, logical operator modifications, constant replacements
@@
 too small to be of any real practical importance.
+\end{comment}
 
@@
 representative nature of these subjects is not clear.
-Furthermore, because
-no other tools implement the kind of approach taken here, based on
-automated generation of value pool based unit tests for a system
-(imperative property-based testing), to our knowledge, we were unable
+Because
+no other tools implement the kind of approach taken here, we were unable
 to perform a meaningful comparison with another nondeterminism
-detection tool.  Available tools, such as DeFlaker \cite{bell2018d}
-(\url{http://www.deflaker.org/}), perform a completely different task (DeFlaker
-is essentially a plugin for Maven regression tests, and cannot check
-nondeterminism at any smaller granularity than that of a whole tests's
-pass/fail result) and target the Java language.  iDFlakies
+detection tool.  DeFlaker \cite{bell2018d}
+performs a completely different task, and cannot check
+nondeterminism at any smaller granularity than a whole test.  iDFlakies
 \cite{idflakies}, similarly, ``does not further classify the causes of
 the flaky tests'' beyond identifying them as likely related to test
-order, or not.  However, the primary
+order.  However, the primary
 aim of our results is to show 1) that nondeterminism at the library
 level can exist in real Python programs, 2) that it can be detected
-by an implementation of the formalism proposed in this paper, and 3)
-that the overhead for such detection, and cost to delta-debug
-nondeterministic tests, is not prohibitive.
+by our methods, and 3)
+that the overhead for such detection, and test reduction, is acceptable.
 \begin{comment}
@@
 tests, and the nondeterminism
-detected by TSTL can be demonstrated with standalone tests that do not
-depend on our code.
-The source code for our TSTL version and all experimental subjects will be
+detected by TSTL can be demonstrated with standalone tests.
+Our source code and experimental subjects will be
 made available for inspection and replication.

### 2020-01-16 ISSTA/implementation.tex
almost there

@@
 
-\subsection{Pools and Visible State}
+\paragraph{Pools and Visible State}
 \label{sec:pools}
@@
 and tester intuitions about what should behave deterministically?
-
 Andrews et al. formalized a value-pool-based model of unit tests that is widely used in
-actual testing tools, including Randoop \cite{Pacheco}, showing that
-unit tests can be represented in a \emph{canonical} form, where
-a test consists of three parts:
+tools, including Randoop \cite{Pacheco}, where
+unit tests are represented in a \emph{canonical} form based on:
 
@@
 TSTL actions defined by a test harness correspond to the set $A$ of
-actions in Section \ref{sec:formal}, and transitions between states
-result from executing out the Python code in an action; these actions
-are, roughly, the array-canonical statements of the value pool
-formalism.
+actions in Section \ref{sec:formal}, and transitions 
+result from executing action code, roughly, the array-canonical
+statements.
 

### 2020-01-16 ISSTA/relatedwork.tex
almost there

@@
 
-At a very broad level, the topic of uncertainty (and thus nondeterminism) in software engineering has been addressed by Garlan \cite{GarlanUncertain}, Elbaum and Rosenblum \cite{Unknowns,lu2015roundtable}, and Ziv and Richardson \cite{UncertaintyPrinciple}.
+At a broad level, the topic of uncertainty (and thus nondeterminism) in software engineering has been addressed by Garlan \cite{GarlanUncertain}, Elbaum and Rosenblum \cite{Unknowns,lu2015roundtable}, and Ziv and Richardson \cite{UncertaintyPrinciple}.
 There is a general agreement that as systems become more complex, more distributed, and more \emph{statistical}, these problems will only grow \cite{lu2015roundtable}.
-
 Gao et al.
-generally considered the question of how to make tests repeatable \cite{Gao:2015:MSU:2818754.2818764}, in the context of system-wide user interaction tests.
-Their work focused on systemic factors, such as execution platform and persistent configuration and database files, in contrast to our focus on identifying nondeterminism at the library level.
-However, what their ``test harness factors'' include delays, which can be of importance to nondeterminism at the library level.
+examined the question of how to make tests repeatable \cite{Gao:2015:MSU:2818754.2818764}, in the context of system-wide user interaction tests, focusing on systemic factors, such as execution platform and persistent configuration and database files, in contrast to our focus on  the library level.
+However, their ``test harness factors'' also included delays.
 Shi et.
-al \cite{DetermImp} examined what might be seen as a related, though in a sense, opposite, problem:  they detect code that assumes a deterministic implementation of a non-deterministic specification.
+al \cite{DetermImp} examined what might be seen as the opposite, problem:  they detect code that assumes a deterministic implementation of a non-deterministic specification.
 E.g., they detect instances when  iteration through items of a set is explicitly not guaranteed to provide any particular order of items, but code depends on the order.

### 2020-01-16 ISSTA/conclusions.tex
shrink conc

@@
 nondeterminism, especially in the library code that underlies most
-systems. In addition to traditional \emph{horizontal} nondeterminism
-(the phenomenon behind Mandelbugs and flaky tests), we also discuss
-the related concept of \emph{vertical} nondeterminism, which is more
-frequently simply a software bug.  We implemented our approach in the
+systems. In addition to traditional \emph{horizontal} nondeterminism, we introduce
+the concept of \emph{vertical} nondeterminism bugs.  
+\begin{comment}
+We implemented our approach in the
 TSTL automated test generation system for Python, and demonstrated the
 simplicity and utility of the approach on real-world examples.
+\end{comment}
 
-As future work, we would like to make the automatic detection of
-values that should not be visible (e.g., count towards nondeterminism)
-possible: if a value (e.g., a timestamp) is always different in every
-run, it is likely opaque.  This would make
-testing libraries mixing deterministic behavior and nondeterministic
-behavior, e.g. cryptographic librarie, much easier. Similarly, in order to extend the
-utility of vertical nondeterminism, we would like to
+As future work, we would like to support the automatic detection of opaque
+values; if a value (e.g., a timestamp) is different in \emph{every}
+trace, it is likely opaque.  This would make
+testing libraries mixing deterministic and nondeterministic
+behavior, e.g. cryptographic libraries, much easier. Similarly, for
+vertical nondeterminism, we would like to
 automatically identify usually-idempotent operations.
@@
 Finally, we are interested in using test decomposition
-\cite{Composition} to more easily isolate, understand, and avoid
-nondeterminism in tests, and to mitigate flaky tests.
+\cite{Composition} to more isolate, understand, and mitigate
+nondeterminism in unit tests, and to mitigate flaky tests.
 Reliable detection of nondeterminism is a first step towards
-evaluating such fine-grained proactive flaky test mitigations.
+evaluating such fine-grained proactive flaky test mitigation.
 

### 2020-01-20 ISSTA/concepts.tex
ship it

@@
 \end{comment}
-
 Lam et al. \cite{idflakies} note that flaky tests not due to order
@@
 vulnerabilities in code.  As a side effect, it
-means that a test that, in some circumstances, causes a crash, may at
-other times not fail at all.  Even if no inputs produce a crash,
-however, memory errors may produce variation in values and thus be detectable.
+means can easily produce flaky tests that sometimes crash.  The
+nondeterminism may usually be visible at the value level, even if it does not
+cause a crash, however.
 %as nondeterminism.

### 2020-01-20 ISSTA/conclusions.tex
ship it

@@
 This paper proposes a formulation of types of nondeterminism, and a
-practical approach to using automated test generation to detect
+practical approach to using automated test generation to detect and understand
 nondeterminism, especially in the library code that underlies most
 systems. In addition to traditional \emph{horizontal} nondeterminism, we introduce
-the concept of \emph{vertical} nondeterminism bugs.  
+the concept of \emph{vertical} nondeterminism bugs, and show their
+additional specification power compared to a
+strong differential testing harness.  
 \begin{comment}
@@
 nondeterminism in unit tests, and to mitigate flaky tests.
-Reliable detection of nondeterminism is a first step towards
-evaluating such fine-grained proactive flaky test mitigation.
+Reliable detection of nondeterminism is a critical tool in making such
+an approach efficient and reliable.
 

### 2020-03-31 ISSTA/concepts.tex
qrs version

@@
 We define two basic types of determinism, shown in Figure
-\ref{fig:types}:  horizontal determinism and vertical determinism.  In
+\ref{fig:types}:  \emph{horizontal} and \emph{vertical} determinism.  In
 horizontal determinism a software system reliably
@@
 this method any number of times in a row should always return the
-same value
+same value.  One approach, then, would be to identify
+idempotent operations 
+automatically retry all such operations, checking that the
+result is unchanged.  However, identifying idempotent operations is an additional specification burden.  Are there instances where a tool
+can automatically identify likely idempotency?
 
-One approach, then, would be to identify
-idempotent operations in an automated test generation harness, and
-automatically retry all such operations, checking that the
-result is unchanged.  The overhead of vertical nondeterminism detection will
-generally be much lower than that for horizontal nondeterminism
-(horizontal checks must re-execute an entire test; vertical checks
-only re-execute proportional to the number of idempotent operations
-performed).  However, identifying idempotent operations is an additional specification burden.  Are there instances where a tool
-can automatically identify a limited kind of idempotent behavior?
+%\subsubsection{Failure Determinism}
 
-\subsubsection{Failure Determinism}
-
-A specialized case of vertical determinism is failure determinism.
+We claim the answer is yes: in the specialized case of  failure determinism.
 Failure determinism is the following restriction on an API: {\bf If a call \emph{fails}, and indicates this to the caller, it should not modify system
@@
 
-\paragraph{Process-Based Nondeterminism}
+\subsubsection{Process-Based Nondeterminism}
 
@@
 respect to an arbitrary predicate $\mathit{PRED}$ (including failure,
-but also code coverage, etc. \cite{icst2014,stvrcausereduce,NonAdeq}) that \emph{only holds with a
+but also code coverage, etc. \cite{icst2014,stvrcausereduce}) that \emph{only holds with a
 certain probability} (the predicate itself is nondeterministic).
@@
 predicate still holds.   E.g., predicates, related to code-coverage can be very useful
-\cite{icst2014,stvrcausereduce,NonAdeq}.
+\cite{icst2014,stvrcausereduce}.
 \end{comment}
@@
 \begin{code}
-array[2] = op01(); array[3] = op05(); clear(array[2]); array[2] = op10();
+  array[2] = op01(); array[3] = op05();
+  clear(array[2]); array[2] = op10();
 \end{code}

### 2020-03-31 ISSTA/conclusions.tex
fix
qrs version

@@
 trace, it is likely opaque.  This would make
-testing libraries mixing deterministic and nondeterministic
-behavior, e.g. cryptographic libraries, much easier. Similarly, for
+testing libraries with intended nondeterminism much easier. Similarly, for
 vertical nondeterminism, we would like to
-automatically identify usually-idempotent operations.
+automatically identify idempotent operations.
 %We would also like to lower the cost
@@
 Finally, we are interested in using test decomposition
-\cite{Composition} to more isolate, understand, and mitigate
+\cite{Composition} to isolate and understand
 nondeterminism in unit tests, and to mitigate flaky tests.

### 2020-03-31 ISSTA/experiments.tex
FITS NOW
qrs version

@@
 as to effectiveness in detecting nondeterminism.
-
 Final state nondeterminism was also faster; visible value
@@
 case behavior (where test runtime is very low compared to process overheads),
-we sacrifice less than half of the test executions.
+we don't quite halve test executions.
 
@@
 
-\paragraph{Manually Inserted Fault}
+\subsubsection{Manually Inserted Fault}
 
@@
 the failing test to 3 steps required less than a second ({\bf RQ3}).
-For this hypothetical subtle bug, failure determinism even improves on strong, hand-tuned, differential testing.
+Here, failure determinism even improves on hand-tuned differential testing.
 
-\paragraph{Mutation Analysis}
+\subsubsection{Mutation Analysis}
 
@@
 non-equivalent mutants not produced by other mutation tools.  This is a strong affirmative answer to {\bf
-  RQ1} at least for a large set of hypothetical bugs.
+  RQ1}.
 

### 2020-03-31 ISSTA/implementation.tex
qrs version

@@
 
-\paragraph{Pools and Visible State}
+%\subsection{Pools and Visible State}
 \label{sec:pools}
@@
 and tester intuitions about what should behave deterministically?
-Andrews et al. formalized a value-pool-based model of unit tests that is widely used in
+Andrews et al. formalized a value-pool-based model that is widely used in
 tools, including Randoop \cite{Pacheco}, where
-unit tests are represented in a \emph{canonical} form based on:
+unit tests are \emph{canonically} given as:
 
@@
 
-We implemented our approach as a modification of the TSTL \cite{NFM15,ISSTA15,tstlsttt}
+We implemented our approach as a modification of the TSTL \cite{tstlsttt}
 system, an open-source

### 2020-03-31 ISSTA/introduction.tex
ok
qrs version

@@
 being able to exactly reproduce system behavior is essential to
-productivity \cite{Firesmith}.  If executing ``the same test'' can produce
+productivity.  If executing ``the same test'' can produce
 significantly different behavior each time it is run, the consequences
@@
 
-\subsection{What is Determinism?}
+%\subsection{What is Determinism?}
 
@@
 
-\subsubsection{Nondeterminism and Flaky Tests}
+%\subsection{Nondeterminism and Flaky Tests}
 
-In order to help ensure that they are reliable and secure, complex
+Complex
 modern software systems usually include a large set of
@@
 special case of various kinds of \emph{horizontal} nondeterminism
-described below.  Our focus, however, is on detecting sources of
-flaky-ness in libraries, before they propagate.
+described below.  Our focus is on detecting sources of
+flaky-ness in libraries, \emph{before} they propagate.
 %unlike all other tools and
@@
 
-\subsection{Contributions}
+%\subsection{Contributions}
 
-This paper proposes (1) a number of formal definitions of types of
+{\bf Contributions:} This paper proposes (1) a number of formal definitions of types of
 nondeterminism (\emph{horizontal} and \emph{vertical}) and (2) an implementation, based on these definitions, for detecting and debugging
@@
 tests, and, most importantly, detection and debugging of 
-unexpected nondeterminism that is, itself, a bug.  We use a set
+unexpected nondeterminism bugs.  We use a set
 of case studies. including real-world, widely-used, Python libraries, to demonstrate the

### 2020-03-31 ISSTA/relatedwork.tex
qrs version

@@
 network/IO, etc.''---the kinds of cause that our approach focuses on identifying.
-The present paper, rather than focusing on flaky tests as such, investigates methods for handling nondeterminism in property-based test generation \cite{ClaessenH00,Papadakis:2011:PIT:2034654.2034663}.
+The present paper, rather than focusing on flaky tests as such, investigates methods for handling nondeterminism in property-based test generation \cite{ClaessenH00}.
 From our point of view, flaky behavior is simply a special case of nondeterminism, where the nondeterminism is sufficient to cause the test to have different pass/fail results.

### 2020-03-31 ISSTA/implementation.tex
fix andrews ref

@@
 and tester intuitions about what should behave deterministically?
-Andrews et al. formalized a value-pool-based model that is widely used in
+Andrews et al. formalized a value-pool-based model \cite{AndrewsTR} that is widely used in
 tools, including Randoop \cite{Pacheco}, where

### 2020-04-01 ISSTA/implementation.tex
12 but tells more

@@
 are deterministic in this strict sense.  What is a practical level of granularity for (visible) states and actions that
-can be efficiently checked, and matches (for the most part) developer
+can be efficiently checked, and matches developer
 and tester intuitions about what should behave deterministically?
@@
 language and tool for property-based testing
-\cite{Hypothesis,ClaessenH00} of Python code.
+\cite{Hypothesis,ClaessenH00} of Python code.  TSTL has
+been used to detect (and usually fix) errors in a number of widely
+used Python libraries, the Python implementation itself, the Solidity
+compiler and a Solidity static analysis tool, and Mac OS.
+ TSTL takes a harness (a
+definition of \emph{what} to test, and what properties to check) and
+generates tests using random testing and various other approaches.
+TSTL is open source, and it and all examples used in this paper are
+available at:  \url{https://github.com/agroce/tstl}.  We implemented
+horizontal determinism checking (with both timing and process-based
+differentiation of executions), failure determinism checking, and a
+set of probabilistic delta-debugging strategies in TSTL, in order to
+examine their effectiveness, and to make them available to users
+performing Python library testing.
 %The TSTL {\tt examples} directory
@@
 approach essentially follows the pool-based canonical form defined by
-Andrews et al. and described above, but relaxes many of its restrictions
+Andrews et al. \cite{AndrewsTR} and described above, but relaxes many of its restrictions
 (e.g., a constant value can be used in a method call without first

### 2020-07-06 ISSTA/concepts.tex
12 but tells more
start QRS shrink

@@
-\section{Practical Lightweight Nondeterminism
+\section{Lightweight Nondeterminism
   Detection}
@@
 \begin{figure}
-\includegraphics[width=\columnwidth]{types}
+  \centering
+\includegraphics[width=0.6\columnwidth]{types}
 \caption{Types of determinism}
@@
   implementation of a search algorithm, that returns the location of
-  an item in an unsorted list; in the presence of duplicate items, thread scheduling may change the index returned.
+  an item in an unsorted list; in the presence of duplicate items,
+  thread scheduling may change the index returned.  Lam et al. \cite{idflakies} note that flaky tests not due to order
+dependence are generally due to ``concurrency, timeouts,
+network/IO, etc.'' a description that is well accounted for by these two
+primary sources.
 
@@
 \end{comment}
-Lam et al. \cite{idflakies} note that flaky tests not due to order
-dependence are generally due to ``concurrency, timeouts,
-network/IO, etc.'' a description that is well accounted for by these two
-primary sources.
+
 %, and most of which can be detected easily
@@
 an executable runs in order to make it harder to exploit memory-safety
-vulnerabilities in code.  As a side effect, it
-means can easily produce flaky tests that sometimes crash.  The
-nondeterminism may usually be visible at the value level, even if it does not
-cause a crash, however.
+vulnerabilities in code, and different orders may cause/avoid crashes.
 %as nondeterminism.
@@
 dictionaries; in Python 3.6, the default dictionary implementation became an ordered dictionary with consistent iteration, though actual hashes remained nondeterministic.
-While very few programs rely on exact hash values, many
-relied on them in that they only functioned correctly with a
+While few programs relied on exact hash values, many
+relied on a
 predictable order for dictionary iteration.  Testing
 Python code for nondeterminism based on hash seed requires
-running in a new process:  Python does not allow changing the salt, since changing hash values on-the-fly would
-break all existing dictionaries.
+running in a new process:  Python does not allow changing the salt.
 
@@
 $P(\mathit{fail})$ that is much, much smaller than
-0.3, perhaps as low as 0.01.  Why?
+0.3, perhaps as low as 0.01.  In fact, we have often observed
+delta-debugging taking a test that had such a high probability of
+failure that we were unaware it was ``flaky'' (only hundreds of repeated executions showed there was
+a small chance it could pass) and transforming it into a
+test that only failed once in every 10 to 20 executions.
+\emph{Delta-debugging can easily act as a flakiness multiplier.}  Why?
 
@@
 nondeterministically with (independent) probabilities of 0.01, 0.05,
-and 0.10, respectively, storing the value of the operation in one of five array locations.
-There is also an operation that clears out an array location, making
-it possible to store another (possibly nondeterministic) value in that
-array slot.  An example test sequence would be:
-
-{\scriptsize
-\begin{code}
-  array[2] = op01(); array[3] = op05();
-  clear(array[2]); array[2] = op10();
-\end{code}
-}
-
-The probability that this test behaves nondeterministically is
-0.84645, the result of the simple calculation $(1-0.01) \times
-(1-0.05) \times (1-0.10)$.  Figure \ref{fig:p1} shows one run of
-delta-debugging on a test of length 500.  The initial test is
-essentially guaranteed to behave nondeterministically; delta-debugging
-only checks that it \emph{can} behave nondeterministically, and so, in
+and 0.10, respectively.  Figure \ref{fig:p1} shows one run of
+delta-debugging on a test of length 500.  In
 the course of reducing the test length to a test with only two
 operations, the delta-debugging algorithm also reduces the probability
-of nondeterminism to about 20\%.  If we use a predicate that
+of nondeterminism from almost 100\% to about 20\%.  If we use a predicate that
 ``forces'' the test to behave nondeterministically at least half the
@@
 nondeterminism is above our target value, close to 60\% (and, as a bonus, many fewer
-test executions are required on average for each check of the predicate).  Note that
-this example is purely monotonic; the pattern of probability changes
-in non-monotonic settings can be more difficult to predict and
-control, but the principle of forcing a probability by sampling and
-replication still holds.
-
-Finally, we note that this problem is not limited to debugging
-nondeterminism or known flakiness.  We have often observed
-delta-debugging taking a test that had such a high probability of
-failure that we were unaware it was ``flaky'' (only hundreds of repeated executions showed there was
-indeed a small chance the could pass) and transforming it into a
-test that only failed once in every 10 to 20 executions.
-\emph{Delta-debugging can easily act as a flakiness multiplier.} Multiple evaluations using replications are our proposed solution.
+test executions are required on average for each check of the predicate).

### 2020-07-06 ISSTA/experiments.tex
start QRS shrink

@@
 algorithm is implemented using concurrency, a key question to ask is
-whether the ordering of (equal under comparison) elements is consistent.
+whether the ordering of (equal under comparison) elements is
+consistent.  TSTL can be used to show that a parallel sort implementation provides
+a consistent, deterministic ordering.
+\begin{comment}
 This is related to the question of whether a sort is stable,
@@
 mechanism can also detect the nondeterminism in a parallel sort that
-does exhibit nondeterministic behavior.  For the second purpose, we
+does exhibit nondeterministic behavior.
+\end{comment}
+For contrast, we also 
 implemented a very simple swap-based parallel sorting algorithm that
@@
 are sorted..
-Running TSTL's random tester without checking for determinism, we can
-see that both sorting algorithms always produce sorted output.  The
+Both sorting algorithms always produce sorted output.  The
 horizontal determinism detector, however, always generates a 
@@
 Reducing a lengthy test showing nondeterminism of the swapsort usually takes less than two
-minutes (out of 10 trials, only one took four minutes).  If the number of attempts to produce nondeterminism is
-increased in the reducer, the resulting test will be usually be very compact, e.g.
+minutes (out of 10 trials, only one took four minutes).  The resulting test will be usually be very compact, e.g.
 reducing the original test from 56 to 8 steps.
@@
   pexpire}) introduce timing-based nondeterminism.
-
 However, that a command such as {\tt restore} takes an expiration
-argument is not obvious to a non-expert, nor is the behavior of {\tt
-  spop}, which pops a random value from a set.  Moreover, it is
+argument is not obvious to a non-expert, and it is very
 difficult to guess whether the {\tt pipe} mechanism, which allows a
 large sequence of commands to be queued up and executed
-at once, introduces the potential for nondeterminism. Deep experience with Redis would make these
+at once, introduces nondeterminism.
+\begin{comment}
+  Deep experience with Redis would make these
 issues clear, but tests using a library are often written by those not
@@
 original developer.
+\end{comment}
 
@@
 
+\begin{comment}
 These results, of course, depend on the parameters of the check for
@@
 easily by overwritten by future test behavior, and do not causally
-influence future computation, final state nondeterminism may be almost useless.
+influence future computation, final state nondeterminism may be almost
+useless.
+\end{comment}
 
@@
 the flaky tests'' beyond identifying them as likely related to test
-order.  However, the primary
-aim of our results is to show 1) that nondeterminism at the library
-level can exist in real Python programs, 2) that it can be detected
-by our methods, and 3)
-that the overhead for such detection, and test reduction, is acceptable.
-\begin{comment}
-We believe that the case
-studies show that users testing Python libraries, and concerned about
-nondeterminism, would find the TSTL tools useful, and would be able to
-detect and debug nondeterministic behavior using our approach, a
-functionality that we believe to be unique in the automated test generation
-literature (and supported in no other tool).  Comparison with a
-competing tool is not required to validate the basic idea, since
-precise empirical measurements of overheads, or improvements on an
-existing approach, are not proposed for evaluation.
-\end{comment}
+order.  
 Another threat to validity is posed by implementation errors.
-TSTL and the mutation tool are tested for bugs by a set of Travis CI
-tests, and the nondeterminism
-detected by TSTL can be demonstrated with standalone tests.
-Our source code and experimental subjects will be
-made available for inspection and replication.
+%TSTL and the mutation tool are tested for bugs by a set of Travis CI
+%tests, and the nondeterminism
+%detected by TSTL can be demonstrated with standalone tests.

### 2020-07-06 ISSTA/implementation.tex
start QRS shrink

@@
 or re-executing in a new process, in checking for nondeterminism.
-
 We therefore instead made horizontal nondeterminism detection a first-class
 property in TSTL, using TSTL's existing notation for marking some
-types of values as \emph{opaque} (not usefully compared for equality),
-used in automatic abstraction-based testing.  Recall that in TSTL, a
-test is (essentially) a sequence of assignments to pool values,
-and method/function calls using pool values (including method calls on
-pool objects).  In TSTL, \emph{visible
-value determinism} is, as proposed above, based on comparing the values of \emph{all pool
-variables} after each test action (pool values other than the one
-assigned to must also be compared, since a common source of
-nondeterminism is when a call results in a change to a value passed as
-a parameter, or changes a shared reference held by two values).  By
-default $V$ as defined in Section \ref{sec:formal} is just the set of all possible pool values, and $v$
-simply extracts the pool values from $S$.
-TSTL allows restriction of which state is visible by removing pools marked as {\tt OPAQUE} in the
-TSTL language test harness (and values that do not support equality
-checking) from $v(S)$.  \emph{Final State Determinism} simply performs the same
+types of values as \emph{opaque} (not usefully compared for equality)  In TSTL, \emph{visible
+value determinism} is thus based on comparing the values of \emph{all
+non-opaque pool
+variables} after each test action. 
+\emph{Final State Determinism} simply performs the same
 comparison, but only on the final values of all pool variables (or a

### 2020-07-06 ISSTA/introduction.tex
start QRS shrink

@@
 time, it is possible, in principle, to predict its future behavior
-perfectly.  We say ``in principle'' because in the real world,
+perfectly.  In the real world,
 prediction may be possible but  impractical.  We write
 complex software systems because we cannot predict their
-behavior (if we could perform the calculations in advance, 
-we would just do so).  As a consequence, rather than
+behavior.  As a consequence, rather than
 defining determinism in terms of prediction, we usually 
@@
 cache contents, branch predictor history, etc. are almost never
-controlled for; however, this is seldom of interest.  What matters is when
+controlled for; however, what matters is when
 some kind of nondeterminism unexpectedly impacts computed values at a
-higher level of abstraction, e.g. when it is believed that the
-behavior of a thread scheduler will not matter, but a race condition
-in the code means that it does matter.
+higher level of abstraction.
 
@@
 code coverage, not establish functional correctness.
-This may be one reason that automated testing is seldom used to
-produce general regression tests.  While bugs found by tools are added
+While bugs found by tools are added
 to regression suites, the full
-set of passing tests is seldom preserved, perhaps because such tests are
-likely to be \emph{flaky}.
+set of passing tests is seldom preserved.
 %without considerable human effort.

### 2020-07-06 ISSTA/relatedwork.tex
start QRS shrink

@@
 
+\begin{comment}
 Other efforts \cite{ParallelDeterministic} have aimed to avoid nondeterminism in parallel implementations, by design, indicating the importance of avoiding nondeterministic behavior, even at the expense of adopting novel programming models, where the goal is not simply (as in, say, Rust) to avoid classic concurrency errors, but to enforce true determinism.
+\end{comment}
 
@@
 
-TSTL's approach to test generation, and our instantiation of the formal definitions of nondeterminism are based on a formulation of unit tests using \emph{pools} of values \cite{AndrewsTR}, which provides a practical solution to the problem of defining the visible state to be compared when checking for nondeterminism.
+%TSTL's approach to test generation, and our instantiation of the formal definitions of nondeterminism are based on a formulation of unit tests using \emph{pools} of values \cite{AndrewsTR}, which provides a practical solution to the problem of defining the visible state to be compared when checking for nondeterminism.
 

### 2020-07-06 ISSTA/concepts.tex
paper

@@
 the observable components of a state.  A more lightweight, but (as we
-will show) still powerful way of checking for failure nondeteterminism is to
-\emph{repeat the failing operation, and check if it still fails}.  If not,
-this is a sign that the operation is not failure deterministic; this more restrictive notion still captures such issues as the Apple login bug.
+will show) still powerful way of checking for failure nondeteterminism
+is to simply
+\emph{repeat the failing operation, and check if it still fails}.  This more restrictive notion still captures such issues as the Apple login bug.
 
@@
 nondeterminism we consider, this second type of failure nondeterminism does not
-require the use of $v$, the notion of observable components of
-a state, or any comparison of states.  It only
-requires that we know whether an operation failed.
+require the use of $v$.
 
@@
 an executable runs in order to make it harder to exploit memory-safety
-vulnerabilities in code, and different orders may cause/avoid crashes.
+vulnerabilities in code, but also makes tests more flaky
 %as nondeterminism.

### 2020-07-11 ISSTA/concepts.tex
shrinking

@@
 step of a test twice should result in the same behavior.  Vertical determinism is both rarer and more specialized than
-horizontal determinism.  However, vertical
-determinism violations usually indicate a potentially serious fault.
+horizontal determinism.
 
@@
 
-\subsubsection{Visible Value Determinism}
-
-Visible value determinism uses the human-accessible outputs (displayed
+\emph{Visible value determinism} uses the human-accessible outputs (displayed
 or stored to a file), and \emph{values returned by functions or methods
@@
 
-\subsubsection{Final State Determinism}
-
-Visible value determinism has significant limitations.  While it
-provides the finest granularity, which is
-important for debugging purposes, it is also expensive, requiring
-checks on a potentially very large number of
-values.  Additionally, in some cases so many values produced by a
-system are opaque that the annotation burden is inordinate.  In these
-cases, it is better to only compare final states of a computation
-performed by the system being checked for determinism.  The final
-state may have opaque components, but it is easy to define an
-abstraction that reduces the state to the actual value of interest.
+While visisble value determinism provides very fine granularity, which
+is important for debugging purposes, it is also expensive, requiring
+checks on a potentially very large number of values.  It is often
+sufficient to only compare final states of a computation performed by
+the system, which we refer to as \emph{final state determinism}.
 
@@
 automatically retry all such operations, checking that the
-result is unchanged.  However, identifying idempotent operations is an additional specification burden.  Are there instances where a tool
-can automatically identify likely idempotency?
+result is unchanged.  Identifying idempotent operations, however, imposes a specification burden.
 
@@
 
-We claim the answer is yes: in the specialized case of  failure determinism.
-Failure determinism is the following restriction on an API: {\bf If a call \emph{fails}, and indicates this to the caller, it should not modify system
-  state in any way; changes should be ``rolled back.''}   In other words, failure determinism is a property stating that an API
-is \emph{transactional} with respect to failing calls.  From a user's perspective, some behaviors of the Mac OS High Sierra
+However, for the specialized case of  failure determinism, no
+specification is required.
+Failure determinism is the following restriction on an API: {\bf If a call/action \emph{fails}, and indicates this to the caller, it should not modify system
+  state in any way; changes should be ``rolled back.''}   Some behaviors of the Mac OS High Sierra
 root exploit (CVE-2017-13872 \cite{applebug0}) exhibited failure
@@
 (e.g., exceptions in Python and Java, or Option/Result types in Rust,
-Haskell, and ML), failure determinism can easily be automatically checked.
-Moreover, the expected overhead should be even lower than for other vertical determinism, in that we can expect most failing
-operations to be fast (since failure is usually
-due to invalid parameters).  Checking equivalence of the full
+Haskell, and ML), failure determinism can be automatically checked.
+The overhead should be even lower than for other vertical determinism, in that we can expect most failing
+operations to be fast.  Checking equivalence of full
 observable state, however, is still expensive, and requires defining
@@
 We can formally distinguish the types of nondeterminism by using a
-variant of a labeled transition system (LTS)l, where (1) $S$
+variant of a labeled transition system (LTS), where (1) $S$
 is a set of states; (2) $V$ is a set of \emph{observable} states, where

### 2020-07-11 ISSTA/implementation.tex
shrinking

@@
 extremely inefficient to compare multiple executions for
-full-memory-state equivalence at every instruction step.  Second, and
-even more importantly, as noted in the introduction, very few programs
+full-memory-state equivalence at every instruction step.  Second, very few programs
 are deterministic in this strict sense.  What is a practical level of granularity for (visible) states and actions that
@@
 depend on the values in value pools, it seems likely that developers
-and testers expect determinism in unit tests, at the granularity of
-value pool assignments.
+and testers expect determinism \emph{at the granularity of
+value pool assignments}.
 
@@
 actions in Section \ref{sec:formal}, and transitions 
-result from executing action code, roughly, the array-canonical
+result from executing action code,---roughly, the array-canonical
 statements.
@@
 
-\subsection{Nondeterminism Detection}
+%\subsection{Nondeterminism Detection}
 %\label{sec:horizimp}
@@
 property in TSTL, using TSTL's existing notation for marking some
-types of values as \emph{opaque} (not usefully compared for equality)  In TSTL, \emph{visible
+types of values as \emph{opaque} (not usefully compared for equality).  In TSTL, \emph{visible
 value determinism} is thus based on comparing the values of \emph{all
@@
 
-\end{comment}
-
 We provide support for the real-world sources of
@@
 nondeterminism.
+\end{comment}
 

### 2020-07-11 ISSTA/introduction.tex
shrinking

@@
 being able to exactly reproduce system behavior is essential to
-productivity.  If executing ``the same test'' can produce
-significantly different behavior each time it is run, the consequences
-can be unfortunate.  Developers using a test exhibiting nondeterminism
+productivity.  Developers using a test exhibiting nondeterminism
 to debug a system
-face a challenge in reasoning about causality, in that an event
-observed may not even take place the next time the test is run; this
+face a serious challenge in reasoning about causality; this
 pernicious phenomenon has been sometimes referred to as the dreaded
@@
 ``Mandelbug'' \cite{GrottkeBugs,FaultTriggers}, since Heisenbugs, properly speaking, are bugs that disappear or
-alter behavior under instrumentation (especially in debugging).
-Automated test generation systems using test coverage results to drive
-the search for interesting inputs may go awry when a run executes
-code only rarely.  And, most significantly, regression testing
-effectiveness can be significantly reduced if tests fail only
+alter behavior under instrumentation.
+Even more significantly, regression testing
+effectiveness can be significantly reduced if tests cover code or,
+worse yet, fail only
 intermittently and unpredictably, as a result of environmental
@@
 developers, who want to  assume that library code behaves in a  predictable fashion; nondeterminism is
-frequently vexing in debugging efforts; most importantly, perhaps, nondeterminism is \emph{often disastrous} for large-scale highly
+frequently vexing in debugging efforts; nondeterminism is \emph{often disastrous} for large-scale highly
 automated testing.
@@
 defining determinism in terms of prediction, we usually 
-say that a system is deterministic if, given the same state and
-inputs, it always produces the same outputs.  
-Technically, many ``nondeterministic'' systems are not
-nondeterministic in a strong sense.  Given the complete state
-of the system (which includes the entire state of the underlying
-hardware, the operating system, storage devices, etc.),
-most software \emph{is} completely predictable.  What we actually
-mean is that, \emph{given a certain limited abstraction of state
+say that a system is deterministic if, \emph{given a certain limited abstraction of state
 and inputs, observable behavior is repeatable}.  %This abstraction,
@@
 Because the
-programmer's abstraction of the system ignores  such a large number of
+programmer's abstraction of the system ignores a large number of
 details, very few tests are ``deterministic'' in the sense
@@
 system.  While developers may know how to produce reliably
-deterministic unit tests, automated test generation usually does not have sufficient information to avoid
-producing the occasional such test.  Moreover, where a human might
-make an assertion about the final value produced during a unit test,
-an automatically generated regression test is usually most easily produced by
-asserting equality with all values produced during a reference
-run, since the final step of a test
-is arbitrary, generated in an effort to increase
-code coverage, not establish functional correctness.
-While bugs found by tools are added
-to regression suites, the full
-set of passing tests is seldom preserved.
+deterministic unit tests, automated test generation usually does not
+have sufficient information to do this.
 %without considerable human effort.

### 2020-07-11 ISSTA/relatedwork.tex
shrinking

@@
 Gao et al.
-examined the question of how to make tests repeatable \cite{Gao:2015:MSU:2818754.2818764}, in the context of system-wide user interaction tests, focusing on systemic factors, such as execution platform and persistent configuration and database files, in contrast to our focus on  the library level.
-However, their ``test harness factors'' also included delays.
+examined the question of how to make tests repeatable \cite{Gao:2015:MSU:2818754.2818764}, in the context of system-wide user interaction tests, focusing on systemic factors in contrast to our focus on  the library level.
 Shi et.
 al \cite{DetermImp} examined what might be seen as the opposite, problem:  they detect code that assumes a deterministic implementation of a non-deterministic specification.
-E.g., they detect instances when  iteration through items of a set is explicitly not guaranteed to provide any particular order of items, but code depends on the order.
+%E.g., they detect instances when  iteration through items of a set is explicitly not guaranteed to provide any particular order of items, but code depends on the order.
 Determinism is also sometimes used as a property in domain-specific testing, e.g.
@@
 
-The problem of test nondeterminism is closely related to the (extremely important in industrial settings) issue of flaky tests \cite{miccoflaky, luo2014empirical,palomba2017does,listfieldtestanalysis}.
+The problem of test nondeterminism is closely related to the issue of flaky tests \cite{miccoflaky, luo2014empirical,palomba2017does,listfieldtestanalysis}.
 How to handle flaky tests in practice (when they cannot be avoided) is a major issue in Google-scale continuous testing \cite{memon2017taming}, and, as Memon et al.
 describe, the problem of flaky tests influences the general design of Google test automation.
-Previous work on flaky tests has either focused on test inter-dependence as a cause of flaky behavior \cite{LamZE2015}, or provided large-scale empirical examination of tests from one large open source project (Apache) \cite{luo2014empirical,palomba2017does}.
-Palomba and Zaidman \cite{palomba2017does} investigated the relationship between test code smells and flaky tests.
+Previous work on flaky tests has either focused on test inter-dependence \cite{LamZE2015}, or large-scale empirical examination \cite{luo2014empirical,palomba2017does}.  
+Bell et al.
+proposed DeFlaker \cite{bell2018d}, which makes flaky tests much easier to detect by relying on the observation that if a test fails, and does not cover any changed code then the test is likely flaky.
+iDFlakies \cite{idflakies} is a framework and dataset for flaky tests, but again focuses on whole-tests, and detecting flakiness by actually observing it.
+Their work shows the order-dependence accounts for a little more than 50\% of flaky test causes, and they classify the rest as due to ``concurrency, timeouts,
+network/IO, etc.''---the causes that our approach focuses on identifying.
 
-Bell et al.
-proposed DeFlaker \cite{bell2018d}, which makes flaky tests much easier to detect by relying on the observation that if a test fails, and does not cover any changed code then (1) presumably it was a passing test in the past, or the tests would not be ``green'' before the change so (2) the test is likely flaky.
-Comparing our approach with DeFlaker is difficult; DeFlaker applies only in the context of code \emph{changes} and is essentially a heuristic that identifies failures as due to nondeterminism.
-Our approach is primarily focused on automatically providing a more extensive \emph{specification} in automated test generation, where a test ``failing'' is itself a sign of nondeterministic behavior.
-iDFlakies \cite{idflakies} is a framework and dataset for flaky tests, but again focuses on whole-tests, and detecting flakiness by actually observing it.
-Their work shows the order-dependence accounts for a little more than 50\% of flaky test causes, and they (like us) classify the rest as due to ``concurrency, timeouts,
-network/IO, etc.''---the kinds of cause that our approach focuses on identifying.
-The present paper, rather than focusing on flaky tests as such, investigates methods for handling nondeterminism in property-based test generation \cite{ClaessenH00}.
-From our point of view, flaky behavior is simply a special case of nondeterminism, where the nondeterminism is sufficient to cause the test to have different pass/fail results.
-Our approach is, in a sense (especially the delta-debugging modifications) in line with Harman and O'Hearn's proposal to simply accept that ``All Tests Are Flaky'' \cite{StartupstoScaleups}, and work with probabilistically failing tests.
-To our
-knowledge, this is the first approach to the problem that
-allows detection of potential sources of flakiness, based on real divergence in
-behavior, without having to observe a test actually fail.  We additionally introduce the first variation of
-delta-debugging that properly handles probabilistic reduction criteria.
-
-\begin{comment}
-Other efforts \cite{ParallelDeterministic} have aimed to avoid nondeterminism in parallel implementations, by design, indicating the importance of avoiding nondeterministic behavior, even at the expense of adopting novel programming models, where the goal is not simply (as in, say, Rust) to avoid classic concurrency errors, but to enforce true determinism.
-\end{comment}
-
-Vertical/failure determinism is a less-studied concept, and to our knowledge the formulation here, as a kind of determinism in a different ``direction,'' has not previously appeared; the kinds of errors that are exposed, however, are not new.
+We introduce the first variation of
+delta-debugging that properly handles probabilistic reduction criteria, in line with Harman and O'Hearn's proposal to simply accept that ``All Tests Are Flaky'' \cite{StartupstoScaleups}, and work with probabilistically failing tests..
+Vertical/failure determinism is also a less-studied concept, and to our knowledge our formulation is novel.
+The kinds of errors that are exposed, however, are not new.
 In particular, we believe that many such faults are related to the propagation of error conditions, which has been studied using static analysis \cite{FileProp}.

### 2020-07-13 ISSTA/concepts.tex
shrinking agian

@@
 interested in behavior within an execution, where repeating the same
-step of a test twice should result in the same behavior.  Vertical determinism is both rarer and more specialized than
-horizontal determinism.
+step of a test twice should result in the same behavior.
 
@@
 automatically retry all such operations, checking that the
-result is unchanged.  Identifying idempotent operations, however, imposes a specification burden.
+result is unchanged.  Identifying idempotent operations does impose a specification burden.
 
@@
   when a test's behavior depends on factors outside the program under
-  test, which are subject to uncontrolled variance.  Calling {\tt
-    random()} in Python is an obvious example. Similarly, calls to {\tt time}
-  functions  will usually return different values on different runs,
-  and even if only elapsed time is used, due to system load, I/O, and
-  other external factors.  {\bf Concurrency} is a second common cause for
-  nondeterminism, when an algorithm is implemented using
-  concurrency, in the (incorrect) belief that the result of the
-  algorithm remains deterministic.  E.g., a multi-threaded
+  test, which are subject to uncontrolled variance.  Calls to {\tt
+    time} or {\tt random} are obvious examples.  {\bf Concurrency} is a second common cause.  E.g., a multi-threaded
   implementation of a search algorithm, that returns the location of
   an item in an unsorted list; in the presence of duplicate items,
-  thread scheduling may change the index returned.  Lam et al. \cite{idflakies} note that flaky tests not due to order
-dependence are generally due to ``concurrency, timeouts,
-network/IO, etc.'' a description that is well accounted for by these two
-primary sources.
-
+  thread scheduling may change the index returned.  
 \begin{comment}
@@
 
-Another example of process-based nondeterminism was discovered by numerous Python developers when Python
+Another example of process-based nondeterminism arose when Python
 version 3.3 introduced automatic random salting of hashes on a
@@
 exact hash values, but in the order of iteration on
-dictionaries; in Python 3.6, the default dictionary implementation became an ordered dictionary with consistent iteration, though actual hashes remained nondeterministic.
-While few programs relied on exact hash values, many
+dictionaries.  Many Python programs
 relied on a
-predictable order for dictionary iteration.  Testing
-Python code for nondeterminism based on hash seed requires
-running in a new process:  Python does not allow changing the salt.
-
+predictable ordering.  
 %Other, less common, effects tied to process also exist.  A few tests
@@
 false negatives, we can therefore produce a reduced test with a
-desired $P(\mathit{fail})$ much more cheaply:  the use of replications not only
-means we only pay the full sampling price on rare occasions, but a
+desired $P(\mathit{fail})$ much more cheaply, and a
 desired accuracy for $P$ can be obtained with a much smaller value $M
-\times N$ than a non-dynamically-sampled $N$.  For example, if $\mathit{PRED}$
+\times N$ than a non-dynamically-sampled $N$.
+
+\begin{comment}For example, if $\mathit{PRED}$
 is $P(\mathit{fail}) \geq 0.5$, and the true probability for a candidate test
@@
 test.
+\end{comment}
 
@@
 nondeterminism is above our target value, close to 60\% (and, as a bonus, many fewer
-test executions are required on average for each check of the predicate).
+test executions were performed).

### 2020-07-13 ISSTA/experiments.tex
paper
shrinking agian

@@
 minutes (out of 10 trials, only one took four minutes).  The resulting test will be usually be very compact, e.g.
-reducing the original test from 56 to 8 steps.
+reducing the test from 56 to 8 steps.
 
@@
 Using the original TSTL harness, just over 20\% of all {\tt redis-py}
-regression tests (of length 200) generated were flaky.  Using the
+regression tests (of length 200) generated were flaky.  We determined
+that the mean time to find all (known to us, over many experimental runs)
+sources of nondeterminism and produce minimal tests is 3.2 
+hours.  Our answer to {\bf RQ1} is thus a strong affirmative.
+
+\begin{comment}
+  Using the
 nondeterminism checker to reduce flaky tests to a minimal set allows
@@
 48 hour runs performed on the version of the harness we modified to
-remove nondeterminism, which produced no flaky tests.  We determined that the mean time to find all
-the sources of nondeterminism and produce minimal tests is 3.2
-hours, which is not trivial, but also, for this purpose, not
-excessive.  Our answer to {\bf RQ1} is thus a strong affirmative.
+remove nondeterminism, which produced no flaky tests.
+\end{comment}
 
@@
 identification of sources of nondeterminism.  After making these changes, no flakiness was observed in a sample of over 2,000 length
-200 tests.   Moreover, because the removals were limited to actually
-observed sources of flakiness, the code coverage loss was minimal.
-Mean branch coverage, for regression tests of length 200, was reduced
-by less than 5 branches (a less than 1\% decrease).  Obviously, it is
-impossible to test the removed calls, but the overall coverage loss is
-both minimal and known, and can be made up for using specially-crafted
-tests (for instance, wrapping the problematic calls in a way that does
-not check the values, or adding a delay after an expiration to allow
-the data to expire).  As a price to pay for the ability to produce
-fast-executing high-coverage full regression tests (not just tests for
-crashes and unexpected exceptions), this seems acceptable.  AFL stability was 56.26\% with the harness allowing 
+200 tests.   Resulting code coverage loss was minimal.
+Mean branch coverage  was reduced
+by less than 5 branches (a less than 1\% decrease).  AFL stability was 56.26\% with the harness allowing 
 nondeterminism, but rose to 98.32\% using the harness with 
@@
 
-Removing sources of expected nondeterminism also makes
-it possible to aggressively test {\tt redis-py} and Redis for
-unexpected nondeterminism arising from actual bugs.  The overhead for such
+The overhead for
 determinism checks, with no delay between operations, is only
 20\%, much lower than the expected cost of running each test twice,
-answering {\bf RQ2} also in the affirmative.  This is because choosing the actions in a test (and determining which
-actions are enabled at each step) consumes a large part of the test
-generator's time with a complex library like {\tt redis-py}; running the test again and checking equality is
-relatively inexpensive.
+answering {\bf RQ2} also in the affirmative.
 
@@
 {\tt redis-py} nondeterministic test, originally with 287 steps.  The initial test behaves
-nondeterministically about half the time.  Figure \ref{fig:r1} shows
-that this is a non-monotonic reduction problem, where removing steps
-can either increase or decrease the probability of nondeterminism.  At
-first, unmodified delta-debugging actually improves the probability of
-failure, but eventually it produces a test of length 16 that only behaves
+nondeterministically about half the time.  Unmodified delta-debugging
+eventually produces a test of length 16 that only behaves
 nondeterministically 24\% of the time.  The reduction takes only 38
@@
 
+\begin{comment}
 \subsubsection{Visible Value vs. Final State Nondeterminism}
@@
 nondeterminism on the 300 tests, 1.199 mean seconds vs. 1.205 mean seconds.  The difference was significant, with
-$p$-value $< 1.19 \times 10^{-23}$ by a paired Wilcoxon test.
+$p < 1.19 \times 10^{-23}$.
 
-\begin{comment}
+%\begin{comment}
 These results, of course, depend on the parameters of the check for
@@
 expose the bad state).
-Because vertical
-reduction does not require running complete tests multiple times, and
-does not affect the delta-debugging algorithm's performance, reducing
-the failing test to 3 steps required less than a second ({\bf RQ3}).
-Here, failure determinism even improves on hand-tuned differential testing.
+Reducing
+the failing test to just 3 steps required less than a second ({\bf RQ3}).
 
@@
 large an improvement as the gain from using a full reference
-implementation for comparison.  Using
+implementation.  Using
 failure nondeterminism was the \emph{only} way to push the mutation score
@@
 
-There are several primary threats to validity:  first, our empirical
+Our empirical
 results are limited to a small set of Python programs, ranging from
@@
 to perform a meaningful comparison with another nondeterminism
-detection tool.  DeFlaker \cite{bell2018d}
-performs a completely different task, and cannot check
-nondeterminism at any smaller granularity than a whole test.  iDFlakies
-\cite{idflakies}, similarly, ``does not further classify the causes of
-the flaky tests'' beyond identifying them as likely related to test
-order.  
-Another threat to validity is posed by implementation errors.
+detection tool.  
 %TSTL and the mutation tool are tested for bugs by a set of Travis CI

### 2020-07-13 ISSTA/introduction.tex
shrinking agian

@@
 perfectly.  In the real world,
-prediction may be possible but  impractical.  We write
+prediction may be possible but  impractical, or some details neither
+predictable nor interesting.  We write
 complex software systems because we cannot predict their
-behavior.  As a consequence, rather than
+behavior, and we usually want to abstract away from machine-level details.  As a consequence, rather than
 defining determinism in terms of prediction, we usually 
-say that a system is deterministic if, \emph{given a certain limited abstraction of state
+say that a system is deterministic if, \emph{given a certain limited
+  higher-level abstraction of state
 and inputs, observable behavior is repeatable}.  %This abstraction,
@@
 
+\begin{comment}
 Because the
@@
 higher level of abstraction.
+\end{comment}
 
@@
 run multiple times, wasting computing resources and
-delaying acceptance of code changes.  As an analogy, we note that  a canary in a coal mine is of
+delaying code changes.  Flaky tests are, for us, simply a special case of general test 
+nondeterminism, in particular the 
+\emph{horizontal} nondeterminism 
+described below.  Our focus is on detecting sources of 
+flaky-ness, \emph{before} they propagate.
+\begin{comment}
+  As an analogy, we note that  a canary in a coal mine is of
 little use if canaries frequently become ill for reasons unrelated to
@@
 canaries into the coal mine that canary-care itself becomes a serious burden.
+\end{comment}
 
-Flaky tests are, for us, simply a special case of general test
-nondeterminism, where the nondeterminism of test values is sufficient
-to cause the pass/fail result of the test to vary.  The definitions
-and techniques proposed in this paper all apply to flaky tests, as a
-special case of various kinds of \emph{horizontal} nondeterminism
-described below.  Our focus is on detecting sources of
-flaky-ness in libraries, \emph{before} they propagate.
 %unlike all other tools and
@@
 We also introduce necessary modifications to the widely used delta-debugging algorithm in order to better handle
-nondeterminism as a test property.
+nondeterminism as a test property.  Evaluating these approaches with
+realistic Python libraries, we show large improvements to both fault
+detection and test reduction.
 
@@
 
+\begin{comment}
 Our methods enable automated production of
@@
 additional specification or test design burden.
+\end{comment}

### 2020-07-15 ISSTA/concepts.tex
down to size

@@
 nondeterminism detection as an unusual kind of \emph{differential
-  testing} \cite{Differential,ICSEDiff}.  In differential testing, a
+  testing} \cite{Differential}.  In differential testing, a
 system is compared against a reference in order to ensure that
@@
 used for systems software such as compilers
-\cite{Differential,csmith} and file systems \cite{CFV08,AMAI}.  The major limitation of differential
+\cite{Differential,csmith} and file systems \cite{AMAI}.  The major limitation of differential
 testing is that multiple implementations of a system are
@@
 nondeterminism.  Attempting to login with the root account with an
-empty password appeared to fail, then, on a repeated try, succeeded
-\cite{applebug1,applebug2}.   Many library APIs are largely failure deterministic.  For instance, if
+empty password appeared to fail, then, on a repeated try, succeeded.   Many library APIs are largely failure deterministic.  For instance, if
 we exclude actual I/O errors, most POSIX file system calls either
@@
 Haskell, and ML), failure determinism can be automatically checked.
-The overhead should be even lower than for other vertical determinism, in that we can expect most failing
-operations to be fast.  Checking equivalence of full
+The overhead should be even lower than for other vertical determinism, in that most failing
+operations are fast.  Checking equivalence of full
 observable state, however, is still expensive, and requires defining
-the observable components of a state.  A more lightweight, but (as we
-will show) still powerful way of checking for failure nondeteterminism
-is to simply
-\emph{repeat the failing operation, and check if it still fails}.  This more restrictive notion still captures such issues as the Apple login bug.
+the observable components of a state, suggesting the shortcut of simply
+\emph{repeating the failing operation, and checking if it still fails}
+which still catches such issues as the Apple login bug.
 
@@
 action needs to be introduced that indicates the generation of a fresh
-process.  Address Space Layout Randomization (ASLR) \cite{ASLR}  is probably the most
+process.  Address Space Layout Randomization (ASLR), which scrambles the layout of memory of the process in which
+an executable runs in order to make it harder to exploit memory-safety
+vulnerabilities, is probably the most
 important source of nondeterminism that arises (only) from change in
-process.  ASLR scrambles the layout of memory of the process in which
-an executable runs in order to make it harder to exploit memory-safety
-vulnerabilities in code, but also makes tests more flaky
+process.  
 %as nondeterminism.
@@
 Delta-debugging as originally proposed uses a
-modified binary search, but we rely only on the general structure: \cite{CReduce,onetest}:  Given a
+modified binary search, but we rely only on the general structure \cite{onetest}:  Given a
 test case, $T_{\mathit{INIT}} : \mathit{test}$ and a predicate
@@
 respect to an arbitrary predicate $\mathit{PRED}$ (including failure,
-but also code coverage, etc. \cite{icst2014,stvrcausereduce}) that \emph{only holds with a
+but also code coverage, etc. \cite{stvrcausereduce}) that \emph{only holds with a
 certain probability} (the predicate itself is nondeterministic).
@@
 predicate still holds.   E.g., predicates, related to code-coverage can be very useful
-\cite{icst2014,stvrcausereduce}.
+\cite{stvrcausereduce}.
 \end{comment}

### 2020-07-15 ISSTA/conclusions.tex
start QRS shrink
down to size

@@
 Nondeterminism is even more pernicious in software testing,
-frustrating debugging efforts (Heisenbugs \cite{Heisenbug} and
-Mandelbugs \cite{GrottkeBugs,FaultTriggers} are widely loathed), and
+frustrating debugging efforts, and
 leading to the costly problem of flaky tests
@@
 practical approach to using automated test generation to detect and understand
-nondeterminism, especially in the library code that underlies most
-systems. In addition to traditional \emph{horizontal} nondeterminism, we introduce
+nondeterminism in Python library code.
+
+\begin{comment}
+In addition to traditional \emph{horizontal} nondeterminism, we introduce
 the concept of \emph{vertical} nondeterminism bugs, and show their
@@
 strong differential testing harness.
-\begin{comment}
+
 We implemented our approach in the
@@
 simplicity and utility of the approach on real-world examples.
-\end{comment}
 
@@
 an approach efficient and reliable.
+\end{comment}
 

### 2020-07-15 ISSTA/experiments.tex
down to size

@@
 of the problematic commands are obvious simply by inspection (e.g.,
-{\tt randomkey, srandmember}). It is also clear that 
-commands producing data with a limited lifetime (e.g., {\tt expire,
-  pexpire}) introduce timing-based nondeterminism.
-However, that a command such as {\tt restore} takes an expiration
-argument is not obvious to a non-expert, and it is very
-difficult to guess whether the {\tt pipe} mechanism, which allows a
-large sequence of commands to be queued up and executed
-at once, introduces nondeterminism.
+{\tt randomkey, srandmember, expire}). 
+However, issues such as whether the mechanism that allows a
+sequence of commands to be queued up and executed
+at once introduces nondeterminism are much more subtle.
 \begin{comment}
@@
 sources of nondeterminism and produce minimal tests is 3.2 
-hours.  Our answer to {\bf RQ1} is thus a strong affirmative.
+hours ( {\bf RQ1}).
 
@@
 nondeterminism, but rose to 98.32\% using the harness with 
-nondeterministic operations removed.
-
-The overhead for
+nondeterministic operations removed.  The overhead for
 determinism checks, with no delay between operations, is only
@@
 probability of nondeterminism.  Using 10 samples and 10 replications,
-on the other hand, actually improves the probability of
-nondeterministic behavior, and (in this case) even produces a smaller test (only 14 steps) in just
+on the other hand, improves the probability of
+nondeterministic behavior, and produces a test with only 14 steps in just
 over 10 minutes.
@@
 The {\tt datarray} module \cite{datarray} is a prototype
-implementation for numpy arrays with named axes to improve data
-management, developed by the Berkeley Institute for Data Science.  As part of its code, it provides a set of algorithms for
+implementation for numpy arrays with named axes, which also provides a set of algorithms for
 inference in
-Bayesian belief networks \cite{russell2016artificial}.  An earlier
-version of these algorithms produced nondeterministic (and in some
-cases incorrect) results due to dependence on the order of values in
-an iterator over a Python dictionary, on Python versions above 3.2,
-until 3.6 (see Section \ref{sec:pnondet}).
-
-\begin{comment}
-Figure \ref{hashbug} shows TSTL code for generating inputs to the {\tt
-  datarray} algorithms.  The {\tt flat\_sort} function is
-needed because we care about actual differences in probability values,
-not simply the order of list, tuple, or dictionary items.
-\end{comment}
+Bayesian belief networks.  An earlier
+version of these algorithms sometimes produced incorrect results due to dependence on the order of values in
+an iterator over a Python dictionary (see Section \ref{sec:pnondet}).  
 Running our TSTL
-harness to test {\tt datarray} consistently requires less than 10 seconds to produce a test exhibiting process-level nondeterminism in
-the {\tt calc\_marginals\_sumproduct} function (the only broken
-algorithm), answering {\bf RQ1} again in the affirmative.  Reducing this 60 step test to a minimal test of only 6 steps,
-showing an extremely simple input producing the issue, 
-required another 92 seconds ({\bf RQ3}) on average.  Interestingly, the way that {\tt
-  python-afl} is used in TSTL means that AFL stability is 100\% even
-for the original, broken, code,
-because the {\tt PYTHONHASHSEED} has already been chosen before the
-fork in AFL.  Removing the nondeterministic call, we can see that the cost of
-checking for process nondeterminism, with no delay between operations,
-is high, a mean 93\% slowdown ({\bf RQ2}.  This is due to the high cost of subprocess
-creation and communication. However, even with this essentially worst
-case behavior (where test runtime is very low compared to process overheads),
-we don't quite halve test executions.
+harness to test {\tt datarray} consistently requires less than 10 seconds to find process-level nondeterminism in
+the {\tt calc\_marginals\_sumproduct} function, answering {\bf RQ1}
+again in the affirmative.  Reducing this 60 step test to a 6-step test 
+required another 92 seconds ({\bf RQ3}) on average.  The cost of
+checking for process nondeterminism 
+is a mean 93\% slowdown ({\bf RQ2}).
 
@@
 file system changes that would not be safe or easily performed using
-real persistent storage.  Originally developed in 2006 at Google by
-Mike Bland, {\tt pyfakefs} is now used in over 2,000 Python tests,
-inside and outside Google \cite{pyfakefs}.
+real persistent storage. It is used in over 2,000 Python tests
+\cite{pyfakefs}.
 The TSTL harness for {\tt pyfakefs} has been used to detect (and
@@
 
-\subsubsection{Manually Inserted Fault}
+%\subsubsection{Manually Inserted Fault}
 
@@
 
-\subsubsection{Mutation Analysis}
+%\subsubsection{Mutation Analysis}
 
@@
 {\tt universalmutator} \cite{RegExpMut} to produce 2,350 mutants of
-the {\tt pyfakefs} core file system code.   %We restricted the generation
+the core file system code.   %We restricted the generation
 %to only mutate code covered during a 60 second run of the test

### 2020-07-15 ISSTA/introduction.tex
down to size

@@
 to debug a system
-face a serious challenge in reasoning about causality; this
-pernicious phenomenon has been sometimes referred to as the dreaded
-``Heisenbug'' \cite{Heisenbug}, though the more accurate usage is
-``Mandelbug'' \cite{GrottkeBugs,FaultTriggers}, since Heisenbugs, properly speaking, are bugs that disappear or
-alter behavior under instrumentation.
-Even more significantly, regression testing
+face a serious challenge.
+Regression testing
 effectiveness can be significantly reduced if tests cover code or,
 worse yet, fail only
-intermittently and unpredictably, as a result of environmental
-factors rather than bugs in changed code.  Such behavior is
+intermittently and unpredictably.  Such behavior is
 unfortunately all too common:  Gao et
 al. \cite{Gao:2015:MSU:2818754.2818764} observed coverage differences
-of up to 184 lines of code for the same test, and false positive rates
-as high as 96\%.  Nondeterminism is often problematic for
+of up to 184 LOC for the same test, and false positive rates
+as high as 96\%.  Nondeterminism is problematic for
 developers, who want to  assume that library code behaves in a  predictable fashion; nondeterminism is
-frequently vexing in debugging efforts; nondeterminism is \emph{often disastrous} for large-scale highly
+vexing in debugging; nondeterminism is \emph{often disastrous} for large-scale 
 automated testing.
@@
 \emph{regression tests}.  A regression test suite is a
-set of tests that can be run against softwarre every time it
+set of tests that can be run every time code
 is modified, to ensure that the modification has not broken the system.  \emph{Flaky tests} \cite{miccoflaky} are regression
 tests that fail in an intermittent, unreliable fashion.  The essence of a flaky
-test is that, for the same snapshot of test code and code under test
-(CUT), it sometimes fails and sometimes passes: the pass/fail result (disposition) of the
+test is that, for the same snapshot of test code and code under test, it sometimes fails and sometimes passes: the pass/fail result (disposition) of the
 test is not a deterministic property of the test code, code under
@@
 problems: first, a flaky test often wastes developer time and delays
-software changes by forcing the investigation of code-under-test that
-is not actually incorrect.  Second, failures in flaky tests are often ignored, and therefore serious software faults may
-be missed.  Finally, to mitigate these problems, flaky tests are often
+software changes by forcing the investigation of \emph{correct} code-under-test.
+Second, failures in flaky tests are often ignored, and therefore serious software faults missed.
+Finally, to mitigate the problem, flaky tests are often
 run multiple times, wasting computing resources and
-delaying code changes.  Flaky tests are, for us, simply a special case of general test 
-nondeterminism, in particular the 
+delaying code changes.  Flaky tests are, for us, simply a special case the 
 \emph{horizontal} nondeterminism 

### 2020-07-15 ISSTA/relatedwork.tex
shrink
down to size

@@
 
-At a broad level, the topic of uncertainty (and thus nondeterminism) in software engineering has been addressed by Garlan \cite{GarlanUncertain}, Elbaum and Rosenblum \cite{Unknowns,lu2015roundtable}, and Ziv and Richardson \cite{UncertaintyPrinciple}.
-There is a general agreement that as systems become more complex, more distributed, and more \emph{statistical}, these problems will only grow \cite{lu2015roundtable}.
-Gao et al.
-examined the question of how to make tests repeatable \cite{Gao:2015:MSU:2818754.2818764}, in the context of system-wide user interaction tests, focusing on systemic factors in contrast to our focus on  the library level.
+The general topic of nondeterminism in software engineering has long been considered important  \cite{Unknowns,UncertaintyPrinciple}. 
+Gao et al. examined the question of how to make tests repeatable \cite{Gao:2015:MSU:2818754.2818764}, in the context of system-wide user interaction tests, focusing on systemic factors.  
 Shi et.
-al \cite{DetermImp} examined what might be seen as the opposite, problem:  they detect code that assumes a deterministic implementation of a non-deterministic specification.
+al \cite{DetermImp} examined code that wrongly enforces a deterministic implementation.
 %E.g., they detect instances when  iteration through items of a set is explicitly not guaranteed to provide any particular order of items, but code depends on the order.
@@
 The problem of test nondeterminism is closely related to the issue of flaky tests \cite{miccoflaky, luo2014empirical,palomba2017does,listfieldtestanalysis}.
-How to handle flaky tests in practice (when they cannot be avoided) is a major issue in Google-scale continuous testing \cite{memon2017taming}, and, as Memon et al.
-describe, the problem of flaky tests influences the general design of Google test automation.
+How to handle flaky tests in practice is a major issue in Google-scale continuous testing \cite{memon2017taming}.
 Previous work on flaky tests has either focused on test inter-dependence \cite{LamZE2015}, or large-scale empirical examination \cite{luo2014empirical,palomba2017does}.  
@@
 iDFlakies \cite{idflakies} is a framework and dataset for flaky tests, but again focuses on whole-tests, and detecting flakiness by actually observing it.
-Their work shows the order-dependence accounts for a little more than 50\% of flaky test causes, and they classify the rest as due to ``concurrency, timeouts,
-network/IO, etc.''---the causes that our approach focuses on identifying.
 
@@
 Vertical/failure determinism is also a less-studied concept, and to our knowledge our formulation is novel.
-The kinds of errors that are exposed, however, are not new.
-In particular, we believe that many such faults are related to the propagation of error conditions, which has been studied using static analysis \cite{FileProp}.
+The kinds of errors that are exposed, however, are not new, e.g., faults related to the propagation of error conditions \cite{FileProp}.
 
 %TSTL's approach to test generation, and our instantiation of the formal definitions of nondeterminism are based on a formulation of unit tests using \emph{pools} of values \cite{AndrewsTR}, which provides a practical solution to the problem of defining the visible state to be compared when checking for nondeterminism.
-
-\begin{comment}
-Finally, Cotroneo et al.
-\cite{CompBugs,FaultTriggers}, and Grottke and Trivedi \cite{GrottkeBugs} have followed on early work on understanding bugs and how they manifest, including transient \cite{Transient} software faults.
-This work informs our attempt to identify sources of nondeterminism, and should provide other, context and project-specific, sources that could be introduced into our general framework.
-\end{comment}

### 2020-07-16 ISSTA/introduction.tex
small fixes

@@
 effectiveness can be significantly reduced if tests cover code or,
-worse yet, fail only
+worse yet, fail, only
 intermittently and unpredictably.  Such behavior is
@@
 run multiple times, wasting computing resources and
-delaying code changes.  Flaky tests are, for us, simply a special case the 
+delaying code changes.  Flaky tests are, for us, simply a special case
+of the 
 \emph{horizontal} nondeterminism 
